---
title: "Example workflow"
output: bookdown::html_document2
---


# installation

```{r load libraries}
#library(plant)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(patchwork)

devtools::load_all()
```

A basic run without 

```{r}
p0 <- Parameters("TF24", "TF24_Env")()

traits <- trait_matrix(c(0.07), c("lma"))
p1 <- expand_parameters(traits, p0)

traits6 <- trait_matrix(rep(c(0.07), 6), c("lma"))
p6 <- expand_parameters(traits6, p0, birth_rate_list = rep(1, 6))

env <- TF24_Environment(
                   soil_number_of_depths = 1, 
                   delta_z = 9999,
                   soil_moist_sat = 0.453,
                   K_sat = 440.628,
                   a_psi = 8.7,
                   n_psi = 4.8,
                   b_infil = 8)
ctrl <- fast_control()
```

```{r}
# existing - using wrappers
results <- run_scm(p1, env = TF24_Environment())
results$net_reproduction_ratios


# existing - using base functions
scm <- SCM("TF24", "TF24_Env")(p1, env, ctrl)
scm$run()
scm$net_reproduction_ratios


system.time(
  run_scm(p1, env, ctrl)
)

system.time({
  scm <- SCM("TF24", "TF24_Env")(p1, env, ctrl)
  scm$run()
})

env1 <- TF24_Environment()

system.time({
  scm <- SCM("TF24", "TF24_Env")(p1, env1, ctrl)
  scm$run()
})
```

To simplify
- unify `run_scm`, `run_scm_collect`, `build_schedule`
  - add collect into scm
  - integrate `build_schedule` into scm

- simplify handling of offspring calculations in SCM 
  - add as auxillary variable or ODE to species?

collect

```{r}
env <- FF16_Environment()
ctrl <- fast_control()

scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
res <- list(scm$state)

while (!scm$complete) {
  scm$run_next()
  res <- c(res, list(scm$state))
}
```

`scm$state` calls `get_state` function defined in [`get_state.h`](inst/include/plant/get_state.h)
  - move these into each class - i.e. scm, patch, node, species, environemt. 
  - as system progresses, archive a copy of 
    - patch OR
    - state vector

We can access all the states at all times at different levls, so why not just archive the patch in C++ and then retreive through patch history on R side?

```{r}
scm
scm$patch$ode_size

scm$patch$species[[1]]$ode_size
scm$patch$species[[1]]$nodes[[1]]$ode_names

scm$patch$species[[1]]$nodes[[1]]$ode_state

scm$patch$species$ode_size
```

Try simple C++ vector push back to store patch throught time

How does this differ to mutants?
- mutants need to store patch environment at all intermediate time points, whereas we only need to 

- also, we have set_state, get_state functions. Can input/output large vector of states
- can iterator return colnames for state?

```{r}
devtools::load_all()
scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
scm$collect <- TRUE
scm$run()
scm$patch
scm$history[[1]]$ode_state
scm$history[[4]]$ode_state
scm$history[[141]]$ode_state

all(scm$patch$ode_state == scm$history[[141]]$ode_state)
```

compare times, what is cost of collect

```{r}
scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
scm$collect <- TRUE
system.time(scm$run())

scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
scm$collect <- FALSE
system.time(scm$run())
```

Extract details

```{r}
# can access history using map
scm$history |> map_dbl(~ .x$ode_size)
scm$history |> map_dbl(~ .x$environment$ode_size)

```

Extarct state

```{r}
# old
s1 <- scm$state$species[[1]] |>
  t() |>
  as_tibble()

# new
scm$patch$species[[1]]$nodes[[1]]$ode_names
nm <- scm$patch$species[[1]]$nodes[[1]]$ode_names
s2 <- scm$patch$species[[1]]$nodes |> map_df(~ .x$ode_state |> setNames(nm))

s1 |> ggplot(aes(height, log_density)) +
  geom_line()

s2 |> ggplot(aes(height, log_density)) +
  geom_line()

all.equal(s1 |> slice_head( n = nrow(s2)), s2)
```

Functions

```{r}
get_state_species <- function(species, i = NA_real_, time = NA_real_, step = NA_real_) {
  # nm <- species$nodes[[1]]$ode_names
  # extracting odeNames is really slow
  nm <- c("height", "mortality", "fecundity", "area_heartwood", "mass_heartwood", "offspring_produced_survival_weighted", "log_density")

  # Convert vector to tibble
  df <- 
    species$ode_state |>
    matrix(ncol = length(nm), byrow = TRUE)
  colnames(df) <- nm
  
  df |> 
    dplyr::as_tibble() |>
    dplyr::mutate(
      step = step,
      time = time,
      species = i,
      node = seq_len(n())
    ) |>
    dplyr::select(step, time, species, node, dplyr::everything())
}

system.time(y <- scm$patch$species[[1]] |> get_state_species())


get_state_community <- function(patch, step_i = NA_real_) {
  patch$species |>
    purrr::imap(get_state_species) |>
    purrr::list_rbind() |>
    dplyr::mutate(
      time = patch$time,
      step = step_i
    )
}

system.time(
  y <- scm$patch |> get_state_community(step_i = 5)
)

get_state_history <- function(history) {
  history |> purrr::imap_dfr(get_state_community)
}

system.time(
  y <- scm$history |> get_state_history()
)
```

Many species

Neglible cost of collecting via either methods
```{r}
scm <- SCM("FF16", "FF16_Env")(p6, env, ctrl)
scm$collect <- FALSE
system.time(scm$run())

scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
scm$collect <- TRUE
system.time(scm$run())

system.time(
  run_scm_collect(p6, env, ctrl)
)

system.time(
  run_scm(p6, env, ctrl)
)

```


```{r}
# existing - using base functions


run_scm(p1, env, ctrl)
run_collect(p1, env, ctrl)

get_state

build_schedule(p1, env, ctrl)
-run_scm_error()

scm <- SCM("TF24", "TF24_Env")(p1, env, ctrl)
scm$collect <- FALSE
scm$refine <- TRUE
scm$run

scm$run_next

ParametersFF16,FF16_Env>()

Environment(x)
  FF16_Environment()

make_environemnt
  - FF16_make_environment()
test_environment
  - FF16_test_environment()

run_patch <- function() {

  p <- Parameters("FF16")
  p$max_patch_lifetime <- 40

  p1 <- expand_parameters("lma", 0.1)

  # create environment object
  env <- Environment("FF16")

  env$extrinsic_drivers$get()
  env$extrinsic_drivers$set_constant("ca", 50)
  
  ctrl <- Control()
  
  scm <- SCM("TF24", "TF24_Env")(p1, env, ctrl)
  scm$refine_node_schedule <- TRUE
  scm$collect_history <- TRUE
  scm$run
    - run
    - (if node schedule) check errors, add nodes
    - (if collect histoiry) collect
  data <- scm |> tidy_outputs
  scm$net_reproduction_ratios

}

scm <- patch_TF24(p1, env, ctrl)
       multi_patch_TF24(p1, env, ctrl)

scm$net_reproduction_ratios
system.time(y <- scm$patch$state)
system.time(yy <- scm$history |> map(~ .x$state))
all.equal(y, last(yy))
```


```{r}
env1 <- TF24_Environment()

env2 <- TF24_Environment()
# env3 <- env2
# env3$extrinsic_drivers <- env1$extrinsic_drivers

scm <- SCM("TF24", "TF24_Env")(p1, env2, ctrl)

```
