---
title: "Example workflow"
output: bookdown::html_document2
---

**authors**: DS Falster
**date**: 2022

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# installation

```{r load libraries}
#library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```

Activate logger

```{r}
plant_log_console()
```

Very minimalist 

A basic run without 

```{r}
p0 <- Parameters("FF16", "FF16_Env")()

traits <- trait_matrix(c(0.07), c("lma"))

p1 <- expand_parameters(traits, p0)

# existing - using wrappers
results <- run_scm(p1)
results$net_reproduction_ratios


# existing - using base functions
scm <- SCM("FF16","FF16_Env")(p1, 
        environment = FF16_Environment(light_availability_spline_rescale_usually = TRUE)
        control = fast_control() # Control()).
scm$run()
scm$net_reproduction_ratios

```

To simplify
- unify `run_scm`, `run_scm_collect`, `build_schedule`
  - add collect into scm
  - integrate `build_schedule` into scm

- simplify handling of offspring calculations in SCM 
  - add as auxillary variable or ODE to species?

collect

```{r}
env <- FF16_Environment(light_availability_spline_rescale_usually = TRUE)
ctrl <- fast_control()

scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
res <- list(scm$state)

while (!scm$complete) {
  scm$run_next()
  res <- c(res, list(scm$state))
}
```

`scm$state` calls `get_state` function defined in [`get_state.h`](inst/include/plant/get_state.h)
  - move these into each class - i.e. scm, patch, node, species, environemt. 
  - as system progresses, archive a copy of 
    - patch OR
    - state vector

We can access all the states at all times at different levls, so why not just archive the patch in C++ and then retreive through patch history on R side?

```{r}
scm
scm$patch$ode_size

scm$patch$species[[1]]$ode_size
scm$patch$species[[1]]$nodes[[1]]$ode_names

scm$patch$species[[1]]$nodes[[1]]$ode_state

scm$patch$species$ode_size
```

Try simple C++ vector push back to store patch throught time

How does this differ to mutants?
- mutants need to store patch environment at all intermediate time points, whereas we only need to 

- also, we have set_state, get_state functions. Can input/output large vector of states
- can iterator return colnames for state?

```{r}
devtools::load_all()
scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
scm$collect <- TRUE
scm$run()
scm$patch
scm$history[[1]]$ode_state
scm$history[[4]]$ode_state
scm$history[[141]]$ode_state

all(scm$patch$ode_state == scm$history[[141]]$ode_state)
```

compare times, what is cost of collect

```{r}
scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
scm$collect <- TRUE
system.time(scm$run())

scm <- SCM("FF16", "FF16_Env")(p1, env, ctrl)
scm$collect <- FALSE
system.time(scm$run())

```

