---
title: "AWRA_soil_water_model"
author: "Isaac Towers"
date: "2025-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
```

# Load observational data

First lets open up some observational data from Edaphic at the Corrie Downs site. It has precipitation and soil moisture observations.

```{r}
read_csv("~/Documents/GitHub/plant/vignettes/models/export-edap61.Falster.NSW.Corrie Downs weather-6.csv") %>%
  filter(Id != "Units") %>%
  setNames(as.character(.[1,])) %>%
  slice(-1) -> soil_moist_obs

# Extract layer names to identify depths. Layer 1 is 10cm deep, 2 20cm, etc.
str_extract_all(names(soil_moist_obs)[-1], "\\d+") %>%
  map(~.x[3]) %>%
  unlist() -> layernames

names(soil_moist_obs)[-1] <- layernames

# Turn timedate stamps into R readable dates
soil_moist_obs %>%
  rename(Prec_mm_per_15min = 17) %>%
  pivot_longer(cols = -c(Prec_mm_per_15min,Timestamp), names_to = "depth", values_to = "Soil_moist") %>%
  mutate(dom = lubridate::day(Timestamp)) %>%
  mutate(mod = lubridate::minute(Timestamp)) %>%
  mutate(hod = lubridate::hour(Timestamp)) %>% 
  mutate(week = lubridate::week(Timestamp)) %>% 
  mutate(moy = lubridate::month(Timestamp)) %>%
  # Precipitation is in units of mm/15mins
  mutate(Prec_mm_per_15min = as.numeric(Prec_mm_per_15min)) %>%
  # Convert those rates into units of mm/day (for each 15 step) %>% 
  mutate(Prec_mm_per_day = as.numeric(Prec_mm_per_15min*96)) %>% 
  # Calculating total rainfall per week
  group_by(week, depth) %>% 
  mutate(Prec_mm_per_week = sum(Prec_mm_per_15min)) %>%
  # mutate(Prec_mm_per_week = Prec_mm_per_15min*96*7) %>% 
  # m3 water per m3 soil - measured using electrical field
  mutate(Soil_moist = as.numeric(Soil_moist)) -> soil_moist_obs
```

```{r}
soil_moist_obs %>%
  ggplot(aes(x = lubridate::as_date(Timestamp), y = as.numeric(Soil_moist))) +
  geom_line(aes(colour = depth, group = depth)) -> a

soil_moist_obs %>%
  ggplot(aes(x = lubridate::as_date(Timestamp), y = as.numeric(Prec_mm_per_day))) +
  geom_line() -> b

a/b
```

# Prepare observational data

```{r}
# Make date-delisted timestep
soil_moist_obs%>%
  group_by(Timestamp) %>%
  nest() %>%
  ungroup() %>%
  mutate(timestep = seq(1:n()) - 1) %>%
  mutate(timestep = timestep / 96) %>%
  filter(timestep > 100) %>%
  unnest(data) -> soil_moist_obs
  

# Creating precipitation data table using the total rainfall per week data
soil_moist_obs  %>%
  group_by(week) %>%
  slice(1) %>%
  arrange(timestep) -> Prec_obs

# Spline function using the precipitation data
Prec_spline <- splinefun(x = Prec_obs$timestep, y = Prec_obs$Prec_mm_per_week)
```

# Number of soil layers

```{r}
soil_layers = length(unique(soil_moist_obs$depth))
```

# Frost, A. J., and Shokri, A., (2021) The Australian Landscape Water Balance model (AWRA-L v7). Technical Description of the Australian Water Resources Assessment Landscape model version 7. Bureau of Meteorology Technical Report

# Melton et al., (2017) Tiling soil textures for terrestrial ecosystem modelling via clustering analysis: a case study with CLASS-CTEM (version 2.1) depending on 

# Cosby et al. (1984) A Statistical Exploration of the Relationships of Soil Moisture Characteristics to the Physical Properties of Soils

## Functions

```{r}

#### Parameters functions

# Saturated hydraulic conductivity (mm day^-1) depends on sand content

K_sat_ <- function(sand_con){
  # Melton et al., (2017)
  7.0556e-6*exp(0.0352*sand_con-2.035) * 1000 * 60*60*24
}

# Slope (%)

Beta_ <- function(avg_Slope){
  avg_Slope/(2*pi)*100
}

# Unscaled reference precipitation controls infiltration-excess runoff (mm day^-1)

P_refmap_ <- function(sand_con, Beta){
  20*(2 + log(K_sat_(sand_con)/Beta))
}


# Soil saturation capacity (mm)

# Parameters from Cosby et al. (1984)
Sz_sat_ <- function(sand_con, dz){
  dz*((50.5 -0.142*sand_con)/100)
}

#### Rates functions

# Infiltration-excess runoff (mm day^-1)

Qh_ <- function(P, Qh_switch){
  Qh_switch * P_refmap*tanh(P/P_refmap)
}

# Drainage rate (mm day ^ -1)

Dz_ <- function(Sz, i){
  K_sat*(Sz/Sz_sat[i])^2
}
```

## Soil water model

```{r}
rates_psi_FS_2021 <- function(t, Sz, params) {
  with(as.list(c(Sz, params)), {
    # Sz values are in mm
    dSz_dt <- rep(NA_real_, nlayers)
    Dz <- rep(NA_real_, nlayers)
    Qh <- rep(NA_real_)
    # Calculate fluxes at boundaries
    # top boundary layer - depends on soil water content, rain, soil water evaporation
    P <- pmax(P_func_(t), 0) # pmax(x, 0) forces all negative values to zero
    Qh <- Qh_(P, Qh_switch)
    I = P - Qh
    Dz[1] <- Dz_(Sz[1], 1)    
    dSz_dt[1] = I - Dz[1]
    # Boundary below each node, except the last one
    # Eq. 3 + 11, Ireson et al. (2023)
    for (i in seq_len(nlayers-1)) {
      Dz[i + 1] <- Dz_(Sz[i + 1], i + 1)    
      dSz_dt[i + 1] = -Dz[i + 1] + Dz[i]
    }

    return(list(dSz_dt = dSz_dt, Qh = Qh, P = P, I = I))
  })
}
```

## Ode solver

```{r}
solve <- function(times, rates, state, method = "rk4", output = "theta") {
  
  deSolve::ode(
    y = state,
    times = times,
    func = rates,
    parms = params,
    method = method) %>% 
  as_tibble() %>%
  pivot_longer(-c(time, Qh, P, I), names_to = "depth", values_to = output)
}
```

## Parameters

```{r}
  # Soil info
  nlayers = soil_layers
  # Soil depth (mm)s
  soil_depth = 1.5e3
  # Different depths through sequence (mm)
  z = seq(soil_depth/nlayers, soil_depth, length.out= nlayers)
  # Distances between depths starting from 0 (mm)
  dz = diff(c(0,z))
  clay_con = 34 # Clay (%)
  sand_con = 10 # Sand (%)
  # Saturated conducitivity (mm day ^-1)
  K_sat = K_sat_(sand_con)
    
  # One value per soil layer
  Sz_sat = Sz_sat_(sand_con, dz)
  
  # Slope parameters (i.e. relating to the effect of slope on drainage)
  P_refscale = 2.637 # unitless
  avg_Slope = 10^-3 # radians
  Beta = Beta_(avg_Slope) # Slope (percent)
  P_refmap = P_refmap_(sand_con, Beta) # Unscaled reference precipitation (mm day^-1)
  
  # Set to 0 to disable runoff
  Qh_switch = 0

  P_ref = P_refmap*P_refscale # Scaled reference precipitation
  
  # Change which function is used to represent precipitation in the model
  
  # P_func_ <- function(t) {Prec_spline(t)} # Use spline function to represent precipitation
  # P_func_ <- function(t) {30*sin(t)+30} # Change precipitation function to a sine wave. First number is amplitude, second is the y-axis midpoint.
  P_func_ <- function(t) {t = 1} # Testing constant precipitation
  
params <- list(
  # Soil info
  nlayers = soil_layers,
  soil_depth = soil_depth,
  z = z,
  clay_con = clay_con, # Clay (%)
  sand_con = sand_con, # Sand (%)
  K_sat = K_sat,
  Sz_sat = Sz_sat,
  
  # Slope parameters
  P_refscale = P_refscale, # unitless
  avg_Slope = avg_Slope, # radians
  Beta = Beta, # Slope (percent)
  P_refmap = P_refmap, # Unscaled reference precipitation
  P_ref = P_ref, # Scaled reference precipitation
  Qh_switch = Qh_switch, # Runoff switch
  P_func_ = P_func_
)
```

```{r}
# generates timesteps per 15 minutes
timesteps <- seq(min(soil_moist_obs$timestep), max(soil_moist_obs$timestep), by = 1/96)

# generates timesteps per 10 days
timesteps <- seq(min(soil_moist_obs$timestep), max(soil_moist_obs$timestep), by = 10)

```

```{r}
# Setting initial state of all soil layers
init_state <- soil_moist_obs %>%
  slice(1:nlayers) %>%
  arrange(as.numeric(depth)) %>%
  pull(Soil_moist) %>%
  as.numeric()

Sz_solve_ode23 <- solve(times = timesteps, rates=rates_psi_FS_2021, state=init_state, method="ode23", output = c("Sz"))
```

```{r}
soil_moist_obs %>%
  ungroup() %>% 
  mutate(depth = as.numeric(depth)) %>%
  mutate(timestep = as.numeric(timestep) %>% round(digits = 4)) %>% # standardising decimal places for left join
  select(Soil_moist, Timestamp, depth, timestep) -> soil_moist_obs_rearranged

# Sz_solve %>%
#   mutate(timestep = as.numeric(time) %>% round(digits = 4)) %>%
#   mutate(depth = as.numeric(depth)) -> Sz_solve 

ggplot()+
  geom_line(data = soil_moist_obs_rearranged, aes(y = Soil_moist, x = timestep), col = "black") +
  geom_line(data = Sz_solve_ode23, aes(y = Sz, x = time), col = "red", linetype = 2) +
  ylim(0, 50) +
  # geom_line(data = Sz_solve_lsoda, aes(y = Sz, x = time), col = "blue", linetype = 2) +
  facet_wrap(~as.numeric(depth)) -> a

Sz_solve_ode23 %>%
  # left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = time)) +
  # ylim(0, 100) +
  geom_line(aes(y = as.numeric(P))) -> b


Sz_solve_ode23 %>%
  # left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = time)) +
  # ylim(0, 100) +
  geom_line(aes(y = as.numeric(Qh))) -> c


Sz_solve_ode23 %>%
  # left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = time)) +
  # ylim(0, 100) +
  geom_line(aes(y = as.numeric(I))) -> d


a/(b+c+d)

a+b
```

# cpp implementation

```{r}
devtools::load_all("~/Documents/GitHub/plant/")
```

```{r}
p0 <- scm_base_parameters("TF24")
p0$max_patch_lifetime <- (max(soil_moist_obs$timestep) - min(soil_moist_obs$timestep))/365
p1 <- expand_parameters(trait_matrix(0.0825, "lma"), p0)
```

```{r}
n_depths = 15
env <- Environment("TF24")
env$set_soil_number_of_depths(n_depths)
env$K_sat <- K_sat/1000*365
env$soil_moist_sat <- Sz_sat[1]/100
env$b_infil <- 0
inits <- soil_moist_obs %>% group_by(timestep) %>% nest() %>% ungroup() %>% slice(1)  %>% unnest(data) %>% arrange(as.numeric(depth)) %>% pull(Soil_moist)
env$set_soil_water_state(c(rep(inits/100)))
env$extrinsic_drivers_set_constant("rainfall", 0.365)

soil_moist_obs %>%
  group_by(timestep, Prec_mm_per_15min) %>%
  nest() -> soil_moist_obs_nested


# env$extrinsic_drivers_set_variable(driver_name = "rainfall", x = (soil_moist_obs_nested$timestep - min(soil_moist_obs$timestep))/365, y = soil_moist_obs_nested$Prec_mm_per_15min/1000*(4*24*365))
# 
# env$extrinsic_drivers_evaluate_range(driver_name = "rainfall", x = seq(0,0.2,length.out = 100))

```

```{r}
out <- run_scm_collect(p1, env)

out$env$soil_moist %>%
  mutate(soil_depth = out$env$soil_depth$soil_depth) -> Sz_solve_cpp

Sz_solve_cpp %>%
  mutate(soil_depth = soil_depth*10) -> Sz_solve_cpp 

Sz_solve_cpp %>%
  mutate(time = time*365 + min(soil_moist_obs$timestep)) -> Sz_solve_cpp


Sz_solve_cpp %>%
  mutate(depth = as.numeric(soil_depth)) -> Sz_solve_cpp 
```

```{r}
ggplot()+
  geom_line(data = soil_moist_obs_rearranged, aes(y = Soil_moist, x = timestep), col = "black") +
  geom_line(data = Sz_solve_cpp, aes(y = soil_moist*100, x = time), col = "red", linetype = 2) +
  ylim(0, 50) +
  # geom_line(data = Sz_solve_lsoda, aes(y = Sz, x = time), col = "blue", linetype = 2) +
  facet_wrap(~depth) -> a
```


```{r}
soiltype <- tibble(soil = c("sand","loamsand","sandyloam","loam",
                "siltyloam","sandyclayloam","clayloam",                "siltyclayloam","sandyclay","siltyclay",
                "clay","fullclay"),
       sand = c(92,82,58,43,17,58,32,10,52,6,22,0),
       clay = c(3,6,10,18,13,27,34,34,42,47,58,100))

rmse <- c()
bias <- c()
```

```{r}
for(i in 1:nrow(soiltype)){

    # Soil info
  nlayers = soil_layers
  # Soil depth (mm)s
  soil_depth = 1.5e3
  # Different depths through sequence (mm)
  z = seq(soil_depth/nlayers, soil_depth, length.out= nlayers)
  # Distances between depths starting from 0 (mm)
  dz = diff(c(0,z))
  clay_con = soiltype$clay[i] # Clay (%)
  sand_con = soiltype$sand[i] # Sand (%)
  # Saturated conducitivity (mm day ^-1)
  K_sat = K_sat_(sand_con)
    
  # One value per soil layer
  Sz_sat = Sz_sat_(sand_con, dz)
  
  # Slope parameters (i.e. relating to the effect of slope on drainage)
  P_refscale = 2.637 # unitless
  avg_Slope = 10^-5 # radians
  Beta = Beta_(avg_Slope) # Slope (percent)
  P_refmap = P_refmap_(clay_con, Beta) # Unscaled reference precipitation
  P_ref = P_refmap*P_refscale # Scaled reference precipitation

params <- list(
  # Soil info
  nlayers = soil_layers,
  soil_depth = soil_depth,
  z = z,
  clay_con = clay_con, # Clay (%)
  sand_con = sand_con, # Sand (%)
  K_sat = K_sat,
  Sz_sat = Sz_sat,
  
  # Slope parameters
  P_refscale = P_refscale, # unitless
  avg_Slope = avg_Slope, # radians
  Beta = Beta, # Slope (percent)
  P_refmap = P_refmap, # Unscaled reference precipitation
  P_ref = P_ref # Scaled reference precipitation
)


state <- soil_moist_obs %>%
  slice(1:nlayers) %>%
  arrange(as.numeric(depth)) %>%
  pull(Soil_moist) %>%
  as.numeric()
  

Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="ode45", output = "Sz")

soil_moist_obs %>% 
  ungroup() %>% 
  mutate(depth = as.numeric(depth)) %>%
  select(Soil_moist, Timestamp, depth, timestep) -> soil_moist_obs_rearranged


Sz_solve %>%
  mutate(timestep = as.numeric(time)) %>%
  mutate(depth = as.numeric(depth)) %>%
  left_join(soil_moist_obs_rearranged) %>%
  select(Sz, Soil_moist,timestep) -> res

rmse[i] <- Metrics::rmse(res$Sz, res$Soil_moist)
bias[i] <- Metrics::bias(res$Sz, res$Soil_moist)

Sz_solve %>%
  rename(timestep = time) %>%
  mutate(timestep = as.numeric(timestep)) %>%
  mutate(depth = as.numeric(depth)) %>%
  # mutate(date = soil_moist_obs_rearranged$date) %>%
  left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = timestep)) +
  geom_line(aes(y = Sz)) +
  geom_line(aes(y = Soil_moist), col = "red") +
  facet_wrap(~as.numeric(depth)) -> p

print(p)

}

rmse
bias

```


# Minimise rmse

```{r}
i = 1


## Parameters

  # Soil info
  nlayers = soil_layers
  soil_depth = 1.5e3
  z = seq(soil_depth/nlayers, soil_depth, length.out= nlayers)
  clay_con = soiltype[i,]$clay # Clay (%)
  sand_con = soiltype[i,]$sand # Sand (%)
  K_sat = K_sat_(clay_con, Kzsatscale_(z))
  Sz_sat = Sz_sat_(clay_con, sand_con, z)
  
  # Slope parameters
  P_refscale = 2.637 # unitless
  avg_Slope = 10^-5 # radians
  Beta = Beta_(avg_Slope) # Slope (percent)
  P_refmap = P_refmap_(clay_con, Beta) # Unscaled reference precipitation
  P_ref = P_refmap*P_refscale # Scaled reference precipitation
  
  S_awc = S_awc_(58,22)
  
  # Rain
  P = as.numeric(Prec_obs$Prec_mm_per_day)

params <- list(
  # Soil info
  nlayers = soil_layers,
  soil_depth = soil_depth,
  z = z,
  clay_con = clay_con, # Clay (%)
  sand_con = sand_con, # Sand (%)
  K_sat = K_sat,
  Sz_sat = Sz_sat,
  
  # Slope parameters
  P_refscale = P_refscale, # unitless
  avg_Slope = avg_Slope, # radians
  Beta = Beta, # Slope (percent)
  P_refmap = P_refmap, # Unscaled reference precipitation
  P_ref = P_ref, # Scaled reference precipitation
  
  S_awc = S_awc,
  
  # Rain
  P = P
)


state <- soil_moist_obs %>%
  slice(1:nlayers) %>%
  arrange(depth) %>%
  pull(Soil_moist)
  

Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="lsoda", output = "Sz")

soil_moist_obs %>% 
  ungroup() %>% 
  mutate(depth = as.numeric(depth)) %>%
  select(Soil_moist, Timestamp, depth, timestep) -> soil_moist_obs_rearranged

Sz_solve %>%
  mutate(timestep = as.numeric(time)) %>%
  mutate(depth = as.numeric(depth)) %>%
  left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = as.numeric(Sz))) +
  geom_line(aes(y = as.numeric(Soil_moist)), col = "red") +
  facet_wrap(~as.numeric(depth))

```

# Minimise bias

```{r}
i = 4

## Parameters

  # Soil info
  nlayers = soil_layers
  soil_depth = 1.5e3
  z = seq(soil_depth/nlayers, soil_depth, length.out= nlayers)
  clay_con = soiltype[i,]$clay # Clay (%)
  sand_con = soiltype[i,]$sand # Sand (%)
  K_sat = K_sat_(clay_con, Kzsatscale_(z))
  Sz_sat = Sz_sat_(clay_con, sand_con, z)
  
  # Slope parameters
  P_refscale = 2.637 # unitless
  avg_Slope = 10^-5 # radians
  Beta = Beta_(avg_Slope) # Slope (percent)
  P_refmap = P_refmap_(clay_con, Beta) # Unscaled reference precipitation
  P_ref = P_refmap*P_refscale # Scaled reference precipitation
  
  S_awc = S_awc_(58,22)
  
  # Rain
  P = as.numeric(Prec_obs$Prec_mm_per_day)

params <- list(
  # Soil info
  nlayers = soil_layers,
  soil_depth = soil_depth,
  z = z,
  clay_con = clay_con, # Clay (%)
  sand_con = sand_con, # Sand (%)
  K_sat = K_sat,
  Sz_sat = Sz_sat,
  
  # Slope parameters
  P_refscale = P_refscale, # unitless
  avg_Slope = avg_Slope, # radians
  Beta = Beta, # Slope (percent)
  P_refmap = P_refmap, # Unscaled reference precipitation
  P_ref = P_ref, # Scaled reference precipitation
  
  S_awc = S_awc,
  
  # Rain
  P = P
)


state <- soil_moist_obs %>%
  slice(1:nlayers) %>%
  arrange(depth) %>%
  pull(Soil_moist)
  

Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="lsoda", output = "Sz")

soil_moist_obs %>% 
  ungroup() %>% 
  mutate(depth = as.numeric(depth)) %>%
  select(Soil_moist, Timestamp, depth, timestep) -> soil_moist_obs_rearranged

Sz_solve %>%
  mutate(timestep = as.numeric(time)) %>%
  mutate(depth = as.numeric(depth)) %>%
  left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = as.numeric(Sz))) +
  geom_line(aes(y = as.numeric(Soil_moist)), col = "red") +
  facet_wrap(~as.numeric(depth))

```

## Delta bare soil example

```{r}
read_csv("~/Documents/export-edap61.Falster.NSW.Delta weather (extra soil)-2.csv") %>%
  filter(Id != "Units") %>%
  setNames(as.character(.[1,])) %>%
  slice(-1) -> soil_moist_obs

str_extract_all(names(soil_moist_obs)[-1], "\\d+") %>%
  map(~.x[2]) %>%
  unlist() -> layernames

names(soil_moist_obs)[-1] <- layernames

soil_moist_obs %>%
  rename(Prec_mm_per_15min = 5) %>%
  pivot_longer(cols = -c(Prec_mm_per_15min,Timestamp), names_to = "depth", values_to = "Soil_moist") %>%
  mutate(date = lubridate::as_date(Timestamp)) %>%
  filter(date > lubridate::as_date("2025-09-14")) %>%
  mutate(dom = lubridate::day(Timestamp)) %>%
  mutate(mod = lubridate::minute(Timestamp)) %>%
  mutate(hod = lubridate::hour(Timestamp)) %>%
  mutate(moy = lubridate::month(Timestamp)) %>%
  mutate(Prec_mm_per_15min = as.numeric(Prec_mm_per_15min)) %>%
  mutate(Prec_mm_per_day = Prec_mm_per_15min*96) %>%
  mutate(Soil_moist = as.numeric(Soil_moist)) -> soil_moist_obs
```

```{r}
soil_moist_obs %>%
  ggplot(aes(x = lubridate::as_date(Timestamp), y = as.numeric(Soil_moist))) +
  geom_line(aes(colour = depth, group = depth))
```

```{r}
soil_moist_obs %>%
  ggplot(aes(x = lubridate::as_date(Timestamp), y = as.numeric(Prec_mm_per_15min))) +
  geom_line(aes(colour = depth, group = depth))
```


```{r}
soil_moist_obs%>%
  group_by(Timestamp) %>%
  nest() %>%
  arrange(Timestamp) %>%
  ungroup() %>%
  mutate(timestep = seq(1:n()) - 1) %>%
  mutate(timestep = timestep / 96) %>%
  unnest(data) -> soil_moist_obs
  
soil_moist_obs  %>%
  group_by(Timestamp) %>%
  slice(1) %>%
  arrange(Timestamp)-> Prec_obs
```


```{r}
soil_layers = length(unique(soil_moist_obs$depth))
```

```{r}

```


```{r}
rates_psi_FS_2021 <- function(t, Sz, params) {
  with(as.list(c(Sz, params)), {
    dSz_dt <- rep(NA_real_, nlayers)
    Dz_ <- rep(NA_real_, nlayers)

    # Calculate fluxes at boundaries
    # top boundary layer - depends on soil water content and rain
    I = P[t+1] - Qh_(P[t+1])
    Dz_[1] <- Dz(Sz[1], 1)    
    dSz_dt[1] = I - Dz_[1] - E(Sz[1])
    # Boundary below each node, except the last one
    # Eq. 3 + 11, Ireson et al. (2023)
    for (i in seq_len(nlayers-1)) {
      Dz_[i + 1] <- Dz(Sz[i + 1], i + 1)    
      dSz_dt[i + 1] = -Dz_[i + 1] + Dz_[i]
    }

    return(list(dSz_dt))
  })
}
```

```{r}
Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="lsoda", output = "Sz")
```

