---
title: "AWRA_soil_water_model"
author: "Isaac Towers"
date: "2025-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
```

# Load observational data

First lets open up some observational data from Edaphic at the Corrie Downs site. It has precipitation and soil moisture observations.

```{r}
read_csv("~/Documents/export-edap61.Falster.NSW.Corrie Downs weather-6.csv") %>%
  filter(Id != "Units") %>%
  setNames(as.character(.[1,])) %>%
  slice(-1) -> soil_moist_obs

# Extract layer names to identify depths. Layer 1 is 10cm deep, 2 20cm, etc.
str_extract_all(names(soil_moist_obs)[-1], "\\d+") %>%
  map(~.x[3]) %>%
  unlist() -> layernames

names(soil_moist_obs)[-1] <- layernames

# Turn timedate stamps into R readable dates
soil_moist_obs %>%
  rename(Prec_mm_per_15min = 17) %>%
  pivot_longer(cols = -c(Prec_mm_per_15min,Timestamp), names_to = "depth", values_to = "Soil_moist") %>%
  mutate(dom = lubridate::day(Timestamp)) %>%
  mutate(mod = lubridate::minute(Timestamp)) %>%
  mutate(hod = lubridate::hour(Timestamp)) %>%
  mutate(moy = lubridate::month(Timestamp)) %>%
  # Precipitation is in units of mm/15mins
  mutate(Prec_mm_per_15min = as.numeric(Prec_mm_per_15min)) %>%
  # Convert those rates into units of mm/day (for each 15 step)
  mutate(Prec_mm_per_day = Prec_mm_per_15min*96) %>%
  mutate(Soil_moist = as.numeric(Soil_moist)) -> soil_moist_obs
```

```{r}
soil_moist_obs %>%
  ggplot(aes(x = lubridate::as_date(Timestamp), y = as.numeric(Soil_moist))) +
  geom_line(aes(colour = depth, group = depth)) -> a

soil_moist_obs %>%
  ggplot(aes(x = lubridate::as_date(Timestamp), y = as.numeric(Prec_mm_per_day))) +
  geom_line() -> b

a/b
```

# Prepare observational data

```{r}
# Make date-delisted timestep
soil_moist_obs%>%
  group_by(Timestamp) %>%
  nest() %>%
  ungroup() %>%
  mutate(timestep = seq(1:n()) - 1) %>%
  mutate(timestep = timestep / 96) %>%
  unnest(data) -> soil_moist_obs
  
# Make data frame for precipitation timesteps (only need one for timestep)
soil_moist_obs  %>%
  group_by(Timestamp) %>%
  slice(1) %>%
  arrange(Timestamp)-> Prec_obs
```

# Number of soil layers

```{r}
soil_layers = length(unique(soil_moist_obs$depth))
```

# Frost, A. J., and Shokri, A., (2021) The Australian Landscape Water Balance model (AWRA-L v7). Technical Description of the Australian Water Resources Assessment Landscape model version 7. Bureau of Meteorology Technical Report

## Functions

```{r}
# Saturated hydraulic conductivity (mm day^-1)

K_satpedo_ <- function(clay_con){
  303.84*exp(-0.144*clay_con)
}

K_sat_ <- function(clay_con, K_satscale){
  K_satscale*K_satpedo_(clay_con)
}

# Slope (%)

Beta_ <- function(avg_Slope){
  avg_Slope/(2*pi)*100
}

# Unscaled reference precipitation controls infiltration-excess runoff (mm day^-1)

P_refmap_ <- function(clay_con, Beta){
  20*(2 + log(K_satpedo_(clay_con)/Beta))
}

# Infiltration-excess runoff (mm day^-1)

Qh_ <- function(P){
  P - P_refmap*tanh(P/P_refmap)
}

# Soil water retention at -10kPa

theta_10_ <- function(clay_con){
  0.3064*(1-exp(-0.0406*clay_con)) + 0.1452
}

# Soil water retention at field capacity

theta_33_ <- function(clay_con){
  0.3543*(1-exp(-0.0385*clay_con)) + 0.083
}

# Soil water retention at wilting point

theta_1500_ <- function(theta_1500){
  0.4016*(1-exp(-0.0230*clay_con)) + 0.0027
}

# Clapp and Hornberger (1978)

theta_to_psi_ <- function(theta){
  psi_s * (theta/theta_s)^(-b)
} 

psi_to_theta_ <- function(psi){
  exp((log(psi) - log(psi_s))/-b + log(theta_s))
} 

# Empirical relationships (Cosby 1984)

b_ <- function(clay_con){
  3.10 + 0.157*clay_con
}
# Soil water potential at saturation  (mm^3/mm^-3), Cosby

psi_s_ <- function(sand_con){
  exp(1.54 + -0.0095*sand_con)
}

# Maximum soil water storage (mm^3/mm^-3), Cosby

theta_s_ <- function(sand_con){
  50.5 + -0.142*sand_con
}

# Available water capacity in soil (mm^3/mm^-3)

S_awc_ <- function(clay_con, sand_con){
  b <- b_(clay_con)
  psi_s <- psi_s_(sand_con)
  theta_s <- theta_s_(sand_con)
  
  psi_to_theta_(33)/100 - psi_to_theta_(1500)/100
}


# Maximum soil water storage (mm)

Sz_sat_ <- function(clay_con, sand_con, z){
  c(z[1], diff(z))*S_awc_(clay_con, sand_con)*Szmaxscale_(z)
}

# Drainage rate (mm day ^ -1)

Dz_ <- function(Sz, i){
  K_sat[i]*(Sz/Sz_sat[i])^2
}

# Soil evaporation rate (mm day ^ -1);; TO-DO

```

Frost and Shokri (2021) provide empirical equations for the relationship between clay content and the theta at -33kPa and -1500kPa. These values are then used to calculate soil water available content (soilawc). However, there is no way in this method to convert resulting soil moistures back into soil water potentials which we need to run the plant hydraulics module. So, we replace these empirical equations with the Clapp and Hornberger relationship between psi and theta (function: psi_) and estimate the parameters for this function from Cosby 1984. We can check that we get roughly equivalent results below. 

Works well for clay loam and clay but not so much for almost pure sand

```{r}
# Test that it works for clay loam

clay_con = 34 # Clay (%)
sand_con = 32 # Sand (%)

# Frost and Shokri emp equations for theta at 10,33,1500 based on clay content

# 0.4225188 (42.25%)
theta_10 = theta_10_(clay_con)
# 0.3993 (39.93%)
theta_33 = theta_33_(clay_con)
# 0.2985 (29.85%)
theta_1500 = theta_1500_(clay_con)

FS_2021_psi_theta <- tibble(psi = c(10,33,1500), theta = c(theta_10, theta_33, theta_1500)*100)

# Estimation of parameters for Clapp and Hornberger equation, used in psi_

b = b_(clay_con = clay_con)
psi_s = psi_s_(sand_con = sand_con)
theta_s = theta_s_(sand_con = sand_con)

tibble(theta = seq(1,theta_s, length.out = 100)) %>%
  mutate(psi = theta_to_psi_(theta)) %>%
  ggplot() +
  geom_line(aes(x = theta, y = psi)) +
  geom_point(data = FS_2021_psi_theta, aes(x = theta, y = psi)) +
  scale_y_log10() +
  theme_bw()
```


```{r}
# Test that it works for clay

clay_con = 58 # Clay (%)
sand_con = 22 # Sand (%)

# Frost and Shokri emp equations for theta at 10,33,1500 based on clay content

# 0.4225188 (42.25%)
theta_10 = theta_10_(clay_con)
# 0.3993 (39.93%)
theta_33 = theta_33_(clay_con)
# 0.2985 (29.85%)
theta_1500 = theta_1500_(clay_con)

FS_2021_psi_theta <- tibble(psi = c(10,33,1500), theta = c(theta_10, theta_33, theta_1500)*100)

# Estimation of parameters for Clapp and Hornberger equation, used in psi_

b = b_(clay_con = clay_con)
psi_s = psi_s_(sand_con = sand_con)
theta_s = theta_s_(sand_con = sand_con)

tibble(theta = seq(1,theta_s, length.out = 100)) %>%
  mutate(psi = theta_to_psi_(theta)) %>%
  ggplot() +
  geom_line(aes(x = theta, y = psi)) +
  geom_point(data = FS_2021_psi_theta, aes(x = theta, y = psi)) +
  scale_y_log10() +
  theme_bw()
```


```{r}
# Test that it works for sand

clay_con = 3 # Clay (%)
sand_con = 92 # Sand (%)

# Frost and Shokri emp equations for theta at 10,33,1500 based on clay content

# 0.4225188 (42.25%)
theta_10 = theta_10_(clay_con)
# 0.3993 (39.93%)
theta_33 = theta_33_(clay_con)
# 0.2985 (29.85%)
theta_1500 = theta_1500_(clay_con)

FS_2021_psi_theta <- tibble(psi = c(10,33,1500), theta = c(theta_10, theta_33, theta_1500)*100)

# Estimation of parameters for Clapp and Hornberger equation, used in psi_

b = b_(clay_con = clay_con)
psi_s = psi_s_(sand_con = sand_con)
theta_s = theta_s_(sand_con = sand_con)

tibble(theta = seq(1,theta_s, length.out = 100)) %>%
  mutate(psi = theta_to_psi_(theta)) %>%
  ggplot() +
  geom_line(aes(x = theta, y = psi)) +
  geom_point(data = FS_2021_psi_theta, aes(x = theta, y = psi)) +
  scale_y_log10() +
  theme_bw()
```

There are a variety of scaling parameters in the Frost and Shokri model which vary with depth. Unclear how this should vary if model has different depths. Could try and fit a spline to it? Spline for ksat looks good.

```{r}
#Spline for Szmaxscale

tibble(z = c(1e2, 1e3, 6e3),
       scale_factor = c(2.465, 1.637, 0.904)) -> data_for_spline

spline(x = data_for_spline$z,
       y = data_for_spline$scale_factor,
       method = "hyman") -> spline_values  

ggplot() +
  geom_point(data =data_for_spline, aes(x = z, y = scale_factor)) +
  geom_line(data = as.tibble(spline_values), aes(x = x, y = y))


#Make Szmaxscale spline
Szmaxscale_ <- splinefun(x = data_for_spline$z,
       y = data_for_spline$scale_factor,
       method = "hyman")
```


```{r}
#Spline for Ksat going negative because last two values are so low. Instead, just assume that this value must drop off extremely quickly

tibble(z = c(1e2, 1e3, 6e3),
       scale_factor = c(8.307, 0.016, 0.043)) -> data_for_spline

spline(x = data_for_spline$z,
       y = data_for_spline$scale_factor,
       method = "fmm") -> spline_values  

ggplot() +
  geom_point(data =data_for_spline, aes(x = z, y = scale_factor)) +
  geom_line(data = as.tibble(spline_values), aes(x = x, y = y))
```

```{r}
#Set the last two values to equal to make the spline instead

tibble(z = c(1e2, 1e3, 6e3),
       scale_factor = c(8.307, 0.016, 0.016)) -> data_for_spline

spline(x = data_for_spline$z,
       y = data_for_spline$scale_factor,
       method = "hyman") -> spline_values  

ggplot() +
  geom_point(data =data_for_spline, aes(x = z, y = scale_factor)) +
  geom_line(data = as.tibble(spline_values), aes(x = x, y = y))

#Make Kzsatscale spline
Kzsatscale_ <- splinefun(x = data_for_spline$z,
       y = data_for_spline$scale_factor,
       method = "hyman")
```

## Soil water model

```{r}
rates_psi_FS_2021 <- function(t, Sz, params) {
  with(as.list(c(Sz, params)), {
    dSz_dt <- rep(NA_real_, nlayers)
    Dz <- rep(NA_real_, nlayers)

    # Calculate fluxes at boundaries
    # top boundary layer - depends on soil water content, rain, soil water evaporation
    I = P[t+1] - Qh_(P[t+1])
    Dz[1] <- Dz_(Sz[1], 1)    
    dSz_dt[1] = I - Dz[1]
    # Boundary below each node, except the last one
    # Eq. 3 + 11, Ireson et al. (2023)
    for (i in seq_len(nlayers-1)) {
      Dz[i + 1] <- Dz_(Sz[i + 1], i + 1)    
      dSz_dt[i + 1] = -Dz[i + 1] + Dz[i]
    }

    return(list(dSz_dt))
  })
}
```

## Ode solver

```{r}
solve <- function(times, rates, state, method = "rk4", output = "theta") {
  deSolve::ode(
    y = state,
    times = times,
    func = rates,
    parms = params,
    method = method) %>% 
  as_tibble() %>% 
  pivot_longer(-time, names_to = "depth", values_to = output)
}
```

## Parameters

```{r}
  # Soil info
  nlayers = soil_layers
  soil_depth = 1.5e3
  z = seq(soil_depth/nlayers, soil_depth, length.out= nlayers)
  clay_con = 58 # Clay (%)
  sand_con = 22 # Sand (%)
  K_sat = K_sat_(clay_con, Kzsatscale_(z))
  Sz_sat = Sz_sat_(clay_con, sand_con, z)
  
  # Slope parameters
  P_refscale = 2.637 # unitless
  avg_Slope = 10^-5 # radians
  Beta = Beta_(avg_Slope) # Slope (percent)
  P_refmap = P_refmap_(clay_con, Beta) # Unscaled reference precipitation
  P_ref = P_refmap*P_refscale # Scaled reference precipitation
  
  S_awc = S_awc_(clay_con,sand_con)
  
  # Rain
  P = as.numeric(Prec_obs$Prec_mm_per_day)

params <- list(
  # Soil info
  nlayers = soil_layers,
  soil_depth = soil_depth,
  z = z,
  clay_con = clay_con, # Clay (%)
  sand_con = sand_con, # Sand (%)
  K_sat = K_sat,
  Sz_sat = Sz_sat,
  
  # Slope parameters
  P_refscale = P_refscale, # unitless
  avg_Slope = avg_Slope, # radians
  Beta = Beta, # Slope (percent)
  P_refmap = P_refmap, # Unscaled reference precipitation
  P_ref = P_ref, # Scaled reference precipitation
  
  S_awc = S_awc,
  
  # Rain
  P = P
)
```

```{r}
timesteps <- soil_moist_obs %>%
  group_by(Timestamp) %>%
  slice(1) %>%
  pull(timestep)
```



```{r}
state <- soil_moist_obs %>%
  slice(1:nlayers) %>%
  arrange(as.numeric(depth)) %>%
  pull(Soil_moist) %>%
  as.numeric()
  

Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="lsoda", output = "Sz")
```

```{r}
soil_moist_obs %>% 
  ungroup() %>% 
  mutate(depth = as.numeric(depth)) %>%
  select(Soil_moist, Timestamp, depth, timestep) -> soil_moist_obs_rearranged

Sz_solve %>%
  mutate(timestep = as.numeric(time)) %>%
  mutate(depth = as.numeric(depth)) %>%
  left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = as.numeric(Sz))) +
  geom_line(aes(y = as.numeric(Soil_moist)), col = "red") +
  facet_wrap(~as.numeric(depth))
```

```{r}
soiltype <- tibble(soil = c("sand","loamsand","sandyloam","loam",
                "siltyloam","sandyclayloam","clayloam",                "siltyclayloam","sandyclay","siltyclay",
                "clay"),
       sand = c(92,82,58,43,17,58,32,10,52,6,22),
       clay = c(3,6,10,18,13,27,34,34,42,47,58))

rmse <- c()
bias <- c()
```

```{r}
for(i in 1:nrow(soiltype)){


## Parameters

  # Soil info
  nlayers = soil_layers
  soil_depth = 1.5e3
  z = seq(soil_depth/nlayers, soil_depth, length.out= nlayers)
  clay_con = soiltype[i,]$clay # Clay (%)
  sand_con = soiltype[i,]$sand # Sand (%)
  K_sat = K_sat_(clay_con, Kzsatscale_(z))
  Sz_sat = Sz_sat_(clay_con, sand_con, z)
  
  # Slope parameters
  P_refscale = 2.637 # unitless
  avg_Slope = 10^-5 # radians
  Beta = Beta_(avg_Slope) # Slope (percent)
  P_refmap = P_refmap_(clay_con, Beta) # Unscaled reference precipitation
  P_ref = P_refmap*P_refscale # Scaled reference precipitation
  
  S_awc = S_awc_(58,22)
  
  # Rain
  P = as.numeric(Prec_obs$Prec_mm_per_day)

params <- list(
  # Soil info
  nlayers = soil_layers,
  soil_depth = soil_depth,
  z = z,
  clay_con = clay_con, # Clay (%)
  sand_con = sand_con, # Sand (%)
  K_sat = K_sat,
  Sz_sat = Sz_sat,
  
  # Slope parameters
  P_refscale = P_refscale, # unitless
  avg_Slope = avg_Slope, # radians
  Beta = Beta, # Slope (percent)
  P_refmap = P_refmap, # Unscaled reference precipitation
  P_ref = P_ref, # Scaled reference precipitation
  
  S_awc = S_awc,
  
  # Rain
  P = P
)


state <- soil_moist_obs %>%
  slice(1:nlayers) %>%
  arrange(as.numeric(depth)) %>%
  pull(Soil_moist) %>%
  as.numeric()
  

Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="lsoda", output = "Sz")

soil_moist_obs %>% 
  ungroup() %>% 
  mutate(depth = as.numeric(depth)) %>%
  select(Soil_moist, Timestamp, depth, timestep) -> soil_moist_obs_rearranged


Sz_solve %>%
  mutate(timestep = as.numeric(time)) %>%
  mutate(depth = as.numeric(depth)) %>%
  left_join(soil_moist_obs_rearranged) %>%
  select(Sz, Soil_moist,timestep) -> res

rmse[i] <- Metrics::rmse(res$Sz, res$Soil_moist)
bias[i] <- Metrics::bias(res$Sz, res$Soil_moist)

Sz_solve %>%
  rename(timestep = time) %>%
  mutate(timestep = as.numeric(timestep)) %>%
  mutate(depth = as.numeric(depth)) %>%
  # mutate(date = soil_moist_obs_rearranged$date) %>%
  left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = timestep)) +
  geom_line(aes(y = Sz)) +
  geom_line(aes(y = Soil_moist), col = "red") +
  facet_wrap(~as.numeric(depth)) -> p

print(p)

}

rmse
bias

```


# Minimise rmse

```{r}
i = 1


## Parameters

  # Soil info
  nlayers = soil_layers
  soil_depth = 1.5e3
  z = seq(soil_depth/nlayers, soil_depth, length.out= nlayers)
  clay_con = soiltype[i,]$clay # Clay (%)
  sand_con = soiltype[i,]$sand # Sand (%)
  K_sat = K_sat_(clay_con, Kzsatscale_(z))
  Sz_sat = Sz_sat_(clay_con, sand_con, z)
  
  # Slope parameters
  P_refscale = 2.637 # unitless
  avg_Slope = 10^-5 # radians
  Beta = Beta_(avg_Slope) # Slope (percent)
  P_refmap = P_refmap_(clay_con, Beta) # Unscaled reference precipitation
  P_ref = P_refmap*P_refscale # Scaled reference precipitation
  
  S_awc = S_awc_(58,22)
  
  # Rain
  P = as.numeric(Prec_obs$Prec_mm_per_day)

params <- list(
  # Soil info
  nlayers = soil_layers,
  soil_depth = soil_depth,
  z = z,
  clay_con = clay_con, # Clay (%)
  sand_con = sand_con, # Sand (%)
  K_sat = K_sat,
  Sz_sat = Sz_sat,
  
  # Slope parameters
  P_refscale = P_refscale, # unitless
  avg_Slope = avg_Slope, # radians
  Beta = Beta, # Slope (percent)
  P_refmap = P_refmap, # Unscaled reference precipitation
  P_ref = P_ref, # Scaled reference precipitation
  
  S_awc = S_awc,
  
  # Rain
  P = P
)


state <- soil_moist_obs %>%
  slice(1:nlayers) %>%
  arrange(depth) %>%
  pull(Soil_moist)
  

Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="lsoda", output = "Sz")

soil_moist_obs %>% 
  ungroup() %>% 
  mutate(depth = as.numeric(depth)) %>%
  select(Soil_moist, Timestamp, depth, timestep) -> soil_moist_obs_rearranged

Sz_solve %>%
  mutate(timestep = as.numeric(time)) %>%
  mutate(depth = as.numeric(depth)) %>%
  left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = as.numeric(Sz))) +
  geom_line(aes(y = as.numeric(Soil_moist)), col = "red") +
  facet_wrap(~as.numeric(depth))

```

# Minimise bias

```{r}
i = 4

## Parameters

  # Soil info
  nlayers = soil_layers
  soil_depth = 1.5e3
  z = seq(soil_depth/nlayers, soil_depth, length.out= nlayers)
  clay_con = soiltype[i,]$clay # Clay (%)
  sand_con = soiltype[i,]$sand # Sand (%)
  K_sat = K_sat_(clay_con, Kzsatscale_(z))
  Sz_sat = Sz_sat_(clay_con, sand_con, z)
  
  # Slope parameters
  P_refscale = 2.637 # unitless
  avg_Slope = 10^-5 # radians
  Beta = Beta_(avg_Slope) # Slope (percent)
  P_refmap = P_refmap_(clay_con, Beta) # Unscaled reference precipitation
  P_ref = P_refmap*P_refscale # Scaled reference precipitation
  
  S_awc = S_awc_(58,22)
  
  # Rain
  P = as.numeric(Prec_obs$Prec_mm_per_day)

params <- list(
  # Soil info
  nlayers = soil_layers,
  soil_depth = soil_depth,
  z = z,
  clay_con = clay_con, # Clay (%)
  sand_con = sand_con, # Sand (%)
  K_sat = K_sat,
  Sz_sat = Sz_sat,
  
  # Slope parameters
  P_refscale = P_refscale, # unitless
  avg_Slope = avg_Slope, # radians
  Beta = Beta, # Slope (percent)
  P_refmap = P_refmap, # Unscaled reference precipitation
  P_ref = P_ref, # Scaled reference precipitation
  
  S_awc = S_awc,
  
  # Rain
  P = P
)


state <- soil_moist_obs %>%
  slice(1:nlayers) %>%
  arrange(depth) %>%
  pull(Soil_moist)
  

Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="lsoda", output = "Sz")

soil_moist_obs %>% 
  ungroup() %>% 
  mutate(depth = as.numeric(depth)) %>%
  select(Soil_moist, Timestamp, depth, timestep) -> soil_moist_obs_rearranged

Sz_solve %>%
  mutate(timestep = as.numeric(time)) %>%
  mutate(depth = as.numeric(depth)) %>%
  left_join(soil_moist_obs_rearranged) %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = as.numeric(Sz))) +
  geom_line(aes(y = as.numeric(Soil_moist)), col = "red") +
  facet_wrap(~as.numeric(depth))

```

## Delta bare soil example

```{r}
read_csv("~/Documents/export-edap61.Falster.NSW.Delta weather (extra soil)-2.csv") %>%
  filter(Id != "Units") %>%
  setNames(as.character(.[1,])) %>%
  slice(-1) -> soil_moist_obs

str_extract_all(names(soil_moist_obs)[-1], "\\d+") %>%
  map(~.x[2]) %>%
  unlist() -> layernames

names(soil_moist_obs)[-1] <- layernames

soil_moist_obs %>%
  rename(Prec_mm_per_15min = 5) %>%
  pivot_longer(cols = -c(Prec_mm_per_15min,Timestamp), names_to = "depth", values_to = "Soil_moist") %>%
  mutate(date = lubridate::as_date(Timestamp)) %>%
  filter(date > lubridate::as_date("2025-09-14")) %>%
  mutate(dom = lubridate::day(Timestamp)) %>%
  mutate(mod = lubridate::minute(Timestamp)) %>%
  mutate(hod = lubridate::hour(Timestamp)) %>%
  mutate(moy = lubridate::month(Timestamp)) %>%
  mutate(Prec_mm_per_15min = as.numeric(Prec_mm_per_15min)) %>%
  mutate(Prec_mm_per_day = Prec_mm_per_15min*96) %>%
  mutate(Soil_moist = as.numeric(Soil_moist)) -> soil_moist_obs
```

```{r}
soil_moist_obs %>%
  ggplot(aes(x = lubridate::as_date(Timestamp), y = as.numeric(Soil_moist))) +
  geom_line(aes(colour = depth, group = depth))
```

```{r}
soil_moist_obs %>%
  ggplot(aes(x = lubridate::as_date(Timestamp), y = as.numeric(Prec_mm_per_15min))) +
  geom_line(aes(colour = depth, group = depth))
```


```{r}
soil_moist_obs%>%
  group_by(Timestamp) %>%
  nest() %>%
  arrange(Timestamp) %>%
  ungroup() %>%
  mutate(timestep = seq(1:n()) - 1) %>%
  mutate(timestep = timestep / 96) %>%
  unnest(data) -> soil_moist_obs
  
soil_moist_obs  %>%
  group_by(Timestamp) %>%
  slice(1) %>%
  arrange(Timestamp)-> Prec_obs
```


```{r}
soil_layers = length(unique(soil_moist_obs$depth))
```

```{r}

```


```{r}
rates_psi_FS_2021 <- function(t, Sz, params) {
  with(as.list(c(Sz, params)), {
    dSz_dt <- rep(NA_real_, nlayers)
    Dz_ <- rep(NA_real_, nlayers)

    # Calculate fluxes at boundaries
    # top boundary layer - depends on soil water content and rain
    I = P[t+1] - Qh_(P[t+1])
    Dz_[1] <- Dz(Sz[1], 1)    
    dSz_dt[1] = I - Dz_[1] - E(Sz[1])
    # Boundary below each node, except the last one
    # Eq. 3 + 11, Ireson et al. (2023)
    for (i in seq_len(nlayers-1)) {
      Dz_[i + 1] <- Dz(Sz[i + 1], i + 1)    
      dSz_dt[i + 1] = -Dz_[i + 1] + Dz_[i]
    }

    return(list(dSz_dt))
  })
}
```

```{r}
Sz_solve <- solve(times = timesteps, rates=rates_psi_FS_2021, state=state, method="lsoda", output = "Sz")
```

