---
title: "Testing root model"
format: html
editor_options:
  chunk_output_type: console
---

To run this vignette, make and rebuild plant using root_water_uptake branch. Then, run below.

This vignette exercises the TF24 strategy with the new multi-layer root water
uptake model. A single cohort of one species (`lma = 0.07`) is grown over a
patch lifetime in a soil column discretised into 15 layers, driven by a
seasonal (sinusoidal) rainfall series. `run_scm_collect` integrates the size-
structured population *and* the per-layer soil water balance simultaneously.

PERFORMANCE NOTE: runtime is dominated by the plant hydraulic solver
(`Leaf::find_root_collar_psi`). To profile, drop `max_patch_lifetime` to ~5
(≈40 s) rather than 75 (≈8 min); the per-step cost is unchanged, only the number
of steps differs. See **Implementation & performance notes** at the end of this
document for the model overview, a profile, and an optimisation roadmap.

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(plant)
# devtools::load_all("/Users/z3533545/Documents/GitHub/plant/")
```

```{r}
# Patch lifetime
max_patch_lifetime <-75
# Basic parameters
p0 <- scm_base_parameters("TF24", "TF24_Env")
p0$max_patch_lifetime <- max_patch_lifetime
traits <- trait_matrix(c(0.07), c("lma"))
p1 <- expand_parameters(traits, p0)

env <- Environment("TF24")

# fifteen soil layers
env$set_soil_number_of_depths(15)
# set soil moisture 0.2 (sat is 0.428)
env$set_soil_water_state(rep(c(0.2), times = 15))

# create rainfall vector across patch lifetime
x = seq(0,max_patch_lifetime,length.out = 1000)
y = 0.25*sin(2*pi*x) + 1
env$extrinsic_drivers_set_variable("rainfall", x=x, y=y)

# no change to ctrl at this point
ctrl <- scm_base_control()
# ctrl$ode_tol_rel <- 1e-4
# ctrl$ode_tol_abs <- 1e-4

system.time(results <- run_scm_collect(p1, env = env, ctrl = ctrl))
saveRDS(results, "results_high.RDS")
results <- readRDS("results_high.RDS")
```

Create diagnostic plots 

```{r}
source("vignettes/models/plot_diagnostics.R")
plot_diagnostics(results, x, y, filename = "diagnostic_plot5_high.png")
```

Plot the root distribution across soil depths (TODO: I think this is broken at the moment)

```{r}
# Replicate the C++ root mass distribution from tf24_strategy.cpp:
#   Q(z, height, eta_x) = (1 - (z/height)^eta_x)^2  for z <= height, else 0
#   rooting_depth       = min(height, 1.5)
#   root_mass_scale     = 83.26 * 0.5 * mass_root
#   mass_root_prop[a]   = root_mass_scale * (Q(prev_depth) - Q(curr_depth))

Q_fn <- function(z, height, eta_x = 0.2) {
  ifelse(z > height, 0.0, (1 - (z / height)^eta_x)^2)
}

Q_fn(0.0001,0.4)

s <- p1$strategies[[1]]

soil_depths <- results$env$soil_depth$soil_depth   # length-15 vector of cumulative depths

results %>%
  expand_state() %>%
  pluck("species") %>%
  select(species, time, node, step, density, patch_density, height, area_leaf) %>%
  mutate(
    mass_root       = s$a_r1 * area_leaf,
    rooting_depth   = pmin(height, 1.5),
    root_mass_scale = 83.26 * 0.5 * mass_root
  ) %>%
  # Expand to one row per soil layer WITHOUT deduplication.
  # tidyr::crossing() de-dupes its inputs, which merges nodes with identical
  # state values and corrupts the rooting_depth/layer_idx alignment.
  # mutate + unnest preserves every original row.
  mutate(layer_idx = list(seq_along(soil_depths))) %>%
  tidyr::unnest(layer_idx) %>%
  mutate(soil_depth = soil_depths[layer_idx],
         q_curr     = Q_fn(soil_depth, rooting_depth)) %>%
  # Mirror C++ exactly: prev_q starts at 1.0, then is the previous layer's q.
  # Using lag() within each plant avoids recomputing Q(0, height) which can
  # have floating-point inconsistency vs Q(prev_depth, height).
  group_by(species, step, node, time) %>%
  arrange(layer_idx, .by_group = TRUE) %>%
  mutate(
    q_prev          = lag(q_curr, default = 1.0),
    mass_root_layer = root_mass_scale * pmax(0.0, q_prev - q_curr)
  ) %>%
  ungroup() %>%
  # Community-level: integrate over size distribution per layer
  group_by(time, step, soil_depth) %>%
  summarise(
    mass_root_layer = sum(mass_root_layer * density * patch_density),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = time, y = mass_root_layer, colour = soil_depth, group = soil_depth)) +
  geom_line() +
  scale_colour_viridis_c(name = "Soil depth (m)") +
  scale_y_log10() +
  theme_bw() +
  theme(text = element_text(size = 16)) +
  ylab(expression(paste(Root~mass~per~layer, " (kg ", m^{-2}, ")"))) +
  xlab("Time (years)") -> root_mass_plot

root_mass_plot
```

# Implementation & performance notes

This section documents the multi-layer root water-uptake model added on the
`root_water_uptake` branch: what it does, where the runtime goes, and which
optimisations have been tried or are worth trying. (The corresponding R
prototype lives in `vignettes/models/root_water_uptake.Rmd`; per-function
scientific commentary lives in the C++ sources referenced below.)

## Model overview

The branch couples three pieces, integrated together by `run_scm_collect`:

- **Soil water (bucket model)** — `inst/include/plant/models/tf24_environment.h`.
  The soil column is discretised into `soil_number_of_depths` layers, each a
  water bucket with its own moisture state. `compute_rates` does a per-layer
  mass balance: infiltration (rainfall with saturation-excess runoff) →
  gravitational drainage (Clapp–Hornberger / Zeng–Decker unsaturated
  conductivity) → root uptake. Moisture↔potential uses a retention curve, with
  a `psi_soil` cache keyed on the moisture state.
- **Root distribution** — `src/tf24_strategy.cpp` (`net_mass_production_dt`).
  Total fine-root mass is spread over depth with the same `Q()` cumulative
  shape function used for the canopy, then passed to the leaf.
- **Plant hydraulics** — `src/leaf_model.cpp` (the core new science, after
  Potkay et al. 2021):
  - `set_physiology` builds a per-layer root resistance network — horizontal
    (intra-layer) plus cumulative vertical (inter-layer) resistances, scaled by
    root carbon and modulated by a root vulnerability curve.
  - `E_from_Soil_to_Root_Collar` sums water drawn from each layer for a given
    root-collar potential (including hydraulic redistribution into wetter
    layers).
  - `find_root_collar_psi` is the master solver, with **two nested levels**:
    (1) a continuity root-find (soil supply = stem transpiration), and (2) a
    golden-section search for the profit-maximising (assimilation − hydraulic
    cost) operating point — plus several early-exit short-circuits for plant
    shut-down conditions.

## Phase 1 — initial optimisation roadmap (#1–#5)

This phase is the original record: it took the new root-water model from its
first working version to roughly 4–6× faster through five ranked items.

### Where the runtime goes (initial profile)

Profiled with `max_patch_lifetime = 5` (≈40 s, a faithful proxy for the full
≈8-min run — only the number of ODE steps differs). Runtime is ~entirely in
`net_mass_production_dt → find_root_collar_psi`. The flat profile:

| Hot spot | Why |
|---|---|
| **Spline evaluation** (`tk::spline::operator()` via `Interpolator::eval`) — the single hottest leaf | `root_vuln_from_psi.eval` is called `(n+1)=6 × n_layers` times per `E_from_Soil_to_Root_Collar` call |
| **`E_from_Soil_to_Root_Collar`** | called inside **two bisection root-finds** (`find_root_psi`, ~16 iters each) |
| **`psi_stem_to_ci`** bisection | called inside every golden-section profit evaluation |

The cost is **algorithmic structure — nested root-finds × spline evals — not
memory allocation.** The per-call `mass_root_prop_` / `psi_soil_inverted_`
allocations and the vector copies in `set_physiology` are <0.1% of runtime;
preallocating them will not move the needle (profile-confirmed). This whole
optimum is recomputed for every individual at every ODE sub-step.

### Tried and REJECTED: super-linear root-finder (TOMS748)

Replacing the bisection in `util::uniroot` with a Brent-like super-linear
solver (TOMS748) is the obvious "biggest lever" — it normally needs ~5–8
evaluations where bisection needs ~16–30. It was implemented and **reverted**:
in this code path it produced **NaN soil potentials and ran *slower*** (the
coupled soil-water ODE destabilised). The hydraulic functions have kinks/clamps
and use splines with extrapolation disabled, so the interpolation steps probe
bad points; **bisection's robustness is load-bearing here.** A documented but
unused `util::uniroot_smooth` helper remains in `inst/include/plant/uniroot.h`
with this caveat, so the experiment is not silently repeated.

### Optimisation roadmap (ranked)

**DONE (#2 + #3): ~1.9× speed-up, bit-identical output** (MPL=5: ≈41 s → ≈21 s;
`all.equal` vs baseline TRUE). Both are numerically identical so the science is
unchanged. Implemented:

- **#3 O(1) spline index** — `tk::spline::operator()` now detects an
  equidistant x-grid in `set_points` and replaces the `std::lower_bound` binary
  search with direct arithmetic plus a ±1 correction that reproduces the
  binary-search index exactly (see `src/tk_spline.cpp`, `m_uniform`). This was
  the single hottest leaf in the profile; the same fix was reported externally
  in traitecoevo/plant#435 (~18% there).
- **#2 Memoised `transpiration()`** — a 1-entry cache on `(psi_stem,
  psi_upstream)` skips the repeated identical spline lookups within each profit
  evaluation; invalidated in `set_physiology` (see `src/leaf_model.cpp`).

**DONE (#1): ~1.25× speed-up, numerics changed (toward more accurate).** The
6-point (`n+1`) sample-mean of root conductivity in `E_from_Soil_to_Root_Collar`
is replaced by the *exact* interval mean `(1/(b−a))·∫ₐᵇ f_r dψ`, obtained from a
pre-integrated curve `root_vuln_integral_from_psi` (`G(m)=∫₀ᵐ f_r ds`, built in
`setup_root_vulnerability`) with **2 spline evals instead of 6** — the same
pre-integrated-curve trick as `setup_transpiration` for the xylem. The interval
is split at ψ=0 (ψ>0 ⇒ vulnerability 1).

This is **more accurate**: vs a fine-grained reference the new interval mean errs
~1e−3 where the old sample-mean erred up to ~5e−2. It therefore **changes the
trajectory** slightly — on a single-species MPL=10 run, `offspring_production`
shifts 4.712e−06 → 4.789e−06 (+1.6% rel, 7.6e−8 abs; toward the more accurate
value). Existing leaf/TF24 tests pass unchanged (their tolerances are absolute at
this magnitude). **Measured speed-up** (identical `-g -O2` build, old vs new,
median of 3 runs each on the same MPL=10 single-species run): **26.2 s → 21.0 s,
≈1.25× (~20%)**. The hot function gets the predicted ~3× fewer spline evals; the
whole-run gain is smaller because that function is one stage inside two nested
root-finds plus the golden-section search.

**DONE (#4): ~1.7× speed-up, numerics changed (tolerance-level shift).** The
golden-section search (GSS) over the feasible collar-potential interval in
`find_root_collar_psi` is replaced by **Brent's method** (golden-section +
parabolic interpolation — R's `Brent_fmin`, new `util::brent_fmin` in
`inst/include/plant/optimize.h`). On the smooth profit objective the parabolic
step converges super-linearly, while the golden-section fallback preserves GSS's
robustness. It is a **bounded** minimiser: it uses the same `[bound_a, bound_b]`
bracket and only ever evaluates *strictly interior* points, so it never probes
the clamped vulnerability-curve endpoints — this is why it is safe here where the
TOMS748 root-finder was not (it stays stable at the production tol `1e-3`; only
an absurdly tight tol `1e-9` over-resolves into the documented NaN instability).

**Measured** (clean `-g -O2` builds, MPL=10 single-species run): **24.7 s →
14.6 s, ≈1.70×**. The gain beats the ~17→~10 iteration estimate because each GSS
iteration also carries an inner `find_psi_stem_from_psi_root` root-find, so
fewer iterations pays double. **Numerics change at the tolerance level**:
`offspring_production` shifts 0.39954 → 0.39637 (−0.79% rel); `height_sum` /
`height_max` move <0.01%. Both optimisers stop at a ψ-tolerance of `1e-3`;
Brent locates the smooth optimum more precisely than GSS's bracket-midpoint at
the same tolerance. The shift is the same order as this system's *intrinsic*
chaotic build-to-build sensitivity (a stale-object incremental build of identical
source moved offspring +0.4% on its own — see build note below). Existing
leaf/TF24/environment tests pass unchanged.

**Build caveat (observed):** incremental `make compile` can leave stale `.o`
files after header/branch changes, yielding an inconsistent binary. Use
`make clean` (or `touch` the changed sources) before benchmarking.

No remaining items on the ranked roadmap.

### Post-optimisation profile (after #1–#4)

Re-profiled after all four items, on the same single-species MPL=10 run. macOS
`sample` was used (the `jointprof`/`pprof` path documented above is unavailable
here; `Rprof`/`profvis` can't see into the `.so` — it's a single Rcpp call).
The `-g` build keeps C++ symbols; a few small functions inline under `-O2`.
Self-time over ~31.6 k active samples (idle wait excluded):

| Function | Self % | Note |
|---|---|---|
| `Leaf::psi_stem_to_ci` | **31.4 %** | inner Cᵢ bisection — **not touched by the roadmap** |
| `tk::spline::operator()` | 22.4 % | spline eval (already O(1)-indexed by #3) |
| `Leaf::E_from_Soil_to_Root_Collar` | 10.5 % | already 6→2 spline evals by #1 |
| `Interpolator::eval` | 7.6 % | spline wrapper |
| `pow` (+ stub) | ~7 % | transcendentals in hydraulic / vulnerability curves |
| `Leaf::find_root_collar_psi` | 6.0 % | master solver (Brent'd by #4) |
| `Leaf::find_root_psi` | 1.7 % | |
| `std::map<string,int>::at` + `memcmp` | ~3 % | string-keyed name→index lookups in the hot path — **now resolved, see below** |

The picture has shifted: spline evaluation was *the* hottest leaf before and is
still ~30 % combined (`tk::spline` + `Interpolator::eval`), but it's no longer
dominant — work is now spread across the solver structure.

#### DONE: string-keyed lookups (~3 % → ~0.5 %)

The two string-lookup hot spots flagged above were both removed (numerically
bit-identical, `offspring_production` unchanged at 4.797128277e−06):

- **aux/state index cache** — `compute_rates` did ~10 `aux_index.at("…")` /
  `state_index.at("…")` (`std::map<string,int>`, i.e. red-black-tree string
  compares) per derivs evaluation per individual. The keys are now resolved to
  ints once in `refresh_indices()` and cached as members (`src/tf24_strategy.cpp`).
- **extrinsic-driver memo** — the 6 time-varying drivers (PPFD, atm_vpd, ca,
  leaf_temp, atm_o2_kpa, atm_kpa) were each an `unordered_map<string,…>::at`
  (plus a by-value key copy) per individual per derivs evaluation. The `get_*`
  accessors now memoise per driver keyed on `time`, so repeated same-time reads
  across individuals collapse to one lookup per distinct time
  (`inst/include/plant/models/tf24_environment.h`); `evaluate` also takes the
  name by `const&` now (`inst/include/plant/extrinsic_drivers.h`).

Re-profiled afterwards: `std::map<…>::at` 1.7 %→0.3 %, `memcmp` 1.6 %→0.2 %,
`ExtrinsicDrivers::evaluate` dropped off entirely (the residual is the expected
one-lookup-per-distinct-time cache misses). Measured ~2 % + ~3 % wall-time
(controlled same-session A/B, 10 reps each); the rest of the profile is
unchanged in shape.

**Remaining lever (a possible roadmap #5):**

- **`psi_stem_to_ci` (~32 %)** is the inner Cᵢ bisection that #4's Brent did
  *not* touch (Brent went on the *outer* GSS profit-max). Giving it the same
  bounded-Brent treatment, or memoising it the way #2 did for `transpiration()`,
  is the obvious biggest remaining target.

#### DONE (#5): `E_from_Soil` integral-spline cache — and why `psi_stem_to_ci` was a red herring

The roadmap-#5 premise above **does not reproduce on the current binary.**
Re-profiling HEAD (macOS `sample`) on *both* the 15-layer root-water scenario at
the top of this document (MPL=5) and the `bench_tf24.R` default-`Environment` run
(MPL=10), `psi_stem_to_ci` is only **~2–3 %** — its symbol appears in 127 / 4012
stacks, and since self-time ⊆ inclusive that bounds it at **≤3.2 %**, not 31 %.
(The 31.4 % figure was measured at `b6ec6045`; it no longer holds.) The actual
hot path in *both* scenarios is **`E_from_Soil_to_Root_Collar` (~22–24 %) plus
the spline evals it drives** (`tk::spline` + `Interpolator::eval`, ~30 %
combined). `psi_stem_to_ci` is also a poor target on its own terms: it can't be
sped **bit-identically** (its bisection *result* is load-bearing for FP
reproducibility — the same reason TOMS748 is rejected above — and Brent probes
distinct points each call, so there is nothing to memoise).

So #5 instead caches the **cumulative root-vulnerability integral** lookups in the
`E_from_Soil` layer loop. That loop did **two `root_vuln_integral_from_psi.eval`
per soil layer**; the two arguments are always `{−ψ_soil[i], −P_x_r, 0}`. `−P_x_r`
is constant across layers (hoisted to **one eval per call**) and `−ψ_soil[i]` is
constant for the whole solve (precomputed once per `find_root_collar_psi` into the
new member `root_vuln_integral_soil_`). A memoised accessor keyed on
exact-`double` equality returns the cached values, engaging only when `psi_soil`
aliases `psi_soil_inverted_`, so the R-exposed by-value method keeps the unchanged
direct-eval fallback. **~30 spline evals/call → ~1.**

**Bit-identical** (this one really is, unlike #1/#4): `identical()` on the full
MPL=5 `run_scm_collect` trajectory vs HEAD; `offspring_production` unchanged at
4.797128277e−06 on the bench run; existing leaf/TF24 tests pass. Because the
change is a value cache rather than new index logic, it was also checked with a
standalone cached-vs-uncached C++ diff over a 27 113-point `P_x_r` sweep (exact
layer knots, `nextafter` neighbours, negative interior, and the ψ>0 /
`hi_neg→0` extrapolation tail the sim never exercises) — **0 mismatches,
max|ΔE| = 0**.

**Measured speed-up** (clean `make compile` builds, `pkgload::load_all(recompile
= FALSE)`, median of 5–7 timed runs):

| Scenario | Build | Median | Speed-up |
|---|---|---|---|
| 15-layer root-water, MPL=5 | pre-#5 → #5 | 8.84 s → 6.72 s | **≈1.32× (−24 %)** |
| bench default-`Environment`, MPL=10 | pre-#5 → #5 | 8.67 s → 7.82 s | **≈1.11× (−10 %)** |

The gain is larger in the 15-layer scenario because the root-vuln integral spline
is a bigger share of total spline time there (it scales with soil-layer count ×
collar-solve evals); the default-`Environment` run has fewer layers, so the
xylem/transpiration splines — which #5 doesn't touch — dominate the spline budget.
(Commit `Cache soil-side root-vuln integral lookups in E_from_Soil`.)

## Phase 2 — reprofile & reassessment (2026-06-18)

After Phase 1 the model was re-profiled from scratch on the 15-layer root-water
scenario at the top of this document (MPL = 5), to check whether the documented
hot-spot ranking still held and whether any cheap levers remained. It produced
one significant correction and one rejected experiment — but no new speed-up.

### Corrected profile (15-layer root-water, MPL = 5)

macOS `sample`, 50 326 active samples; self-time computed as
`count − Σ(child counts)` and cross-checked by a leaf-frame sum (both agree):

| Function | Self % | Note |
|---|---|---|
| `Leaf::E_from_Soil_to_Root_Collar` | **39.0 %** | lean arithmetic over 15 layers × nested root-finds (spline evals counted separately) |
| `Leaf::psi_stem_to_ci` | **35.7 %** | inner Cᵢ bisection (`leaf_model.cpp:914`); almost all self-time is the inlined `util::uniroot` + `assim_colimited` |
| `tk::spline::operator()` | 8.1 % | already O(1)-indexed (#3) |
| `pow` | 8.0 % | **fractional-exponent only** (see below) |
| `Leaf::find_root_collar_psi` | 7.0 % | master solver (Brent'd by #4) |
| `Interpolator::eval` | 3.2 % | spline wrapper |

### Correction: `psi_stem_to_ci` was mis-ranked in Phase 1

The Phase-1 #5 note claimed `psi_stem_to_ci` was "only ~2–3 %". **That was a
measurement error.** It counted the number of distinct stack *contexts* the
symbol appears in (~127–147 tree nodes) and treated that as an inclusive-time
bound. But a deep bisection loop appears in *few* stack contexts while each
carries *many* samples — the correct metric is sample weight, which here is
18 079 samples ≈ **35.7 %**. So `psi_stem_to_ci` is in fact **co-dominant with
`E_from_Soil_to_Root_Collar`, not negligible.**

It still cannot be sped **bit-identically**, for the reasons given in Phase 1:
its bisection *result* is load-bearing for FP reproducibility (same reason
TOMS748 is rejected), and there is no closed form for `A(cᵢ) = gc·(ca − cᵢ)`.

### Tried and REJECTED: `pow(x, 2) → x*x`

The 8 % `pow` self-time looked like an easy win, so the integer `pow(z, 2)` in
`assim_colimited` / `electron_transport` was replaced with `z*z`. **Rejected:**

- **No speed-up** — MPL = 5 median 6.83 s (baseline) vs 7.12 s (candidate); if
  anything slightly slower, i.e. within noise.
- **Not bit-identical** — `offspring_production` 1.64825e−06 → 1.65255e−06
  (+0.26 %); `pow(x, 2.0) ≠ x*x` on Apple libm, and the ODE amplifies the
  difference.

Profiling the `pow` *callers* explains why it was a wash: the 8 % is **entirely
fractional-exponent** pows, not the integer one — `find_root_collar_psi` (xylem
vulnerability `exp(-pow(ψ/b, c))` and hydraulic cost `pow(1 − vuln, beta2)`),
and `compute_competition` / `net_mass_production_dt` (the
`Q() = (1 − (z/h)^eta)²` canopy/root shape function). The integer `pow(z, 2)` in
`assim_colimited` was a non-bottleneck. Reverted.

### Tried and REJECTED: spline the xylem hydraulic cost

`hydraulic_cost_TF(ψ) = g1·(1 − vuln(ψ))^beta2` (where `vuln = exp(-pow(ψ/b, c))`)
is a pure function of ψ — `b, c, g1, beta2` are all constant for the run — so it
looked like the best remaining lever: pre-build it once as a spline
(`setup_hydraulic_cost`, same domain as the transpiration supply curve) and
replace the per-profit-evaluation `exp + pow + pow` with one spline eval. It was
implemented (member `hydraulic_cost_from_psi`, built in both constructors) and
**reverted**:

- **Slower, not faster** — a tightly interleaved A/B (swap pre-built `.so`,
  fresh R per run, 6 alternating pairs to cancel thermal drift) gave baseline
  median **6.91 s** vs candidate **7.26 s** — the candidate was ~5 % *slower* in
  5 / 6 pairs.
- **Also numerics-changing** — at the supply-curve resolution (100 pts)
  `offspring_production` shifted +0.56 %; a 10× finer grid cut that to −0.062 %
  (within build-to-build noise) but did **not** recover the speed.

Why it loses: `Leaf` is constructed once per *strategy* (not per call), so the
build cost is irrelevant — the regression is the **eval**. On this Apple-ARM
machine the `Interpolator::eval → tk::spline::operator()` path (wrapper call +
bounds/extrapolate check + index arithmetic + cubic eval) is *more* expensive
than the inlined `exp + pow + pow` it replaced. Spline eval is already ~11 % of
runtime; adding one more per profit evaluation costs more than the cheap,
well-optimised hardware transcendentals saved. **The spline-everything strategy
has hit diminishing returns against fast hardware transcendentals.**

### Remaining levers (all numerics-changing)

The two dominant blocks (≈75 % combined) are the **nested root-find structure**;
their algorithmic wins are already taken (Brent on the outer GSS, integral-spline
caches, transpiration memo, O(1) splines), and neither can be sped
bit-identically. With the hydraulic-cost spline now rejected, the only untried
levers left both shift numerics at the tolerance level and are speculative:

1. **Spline / cache `Q()`'s `pow(z/h, eta)`** in `compute_competition` (~4.5 % of
   the pow time; canopy-side, independent of hydraulics). Caveat: by the same
   logic as the rejected hydraulic-cost spline, a spline eval may not beat the
   `pow` it replaces — would need the same interleaved A/B to confirm.
2. **Loosen the Cᵢ-bisection tolerance** (currently `1e-7`, nested inside an outer
   `1e-3` solve — over-resolved). Cheapest potential win on the 35.7 % block, but
   explicitly changes results.

**Assessment:** the code is structurally well-optimised and has reached the point
where micro-optimising the transcendentals no longer pays (spline eval ≈ the cost
of the math it would replace). Absent appetite for tolerance-level numerics
shifts, Phase 2 stops here.

### Code-consistency follow-up: Brent for the single-layer optimisers (OFF the hot path)

A unification cleanup, **not** a production speed-up. The two R-exposed
single-layer profit optimisers — `optimise_psi_stem_Sperry` /
`optimise_psi_stem_TF` (`src/leaf_model.cpp`) — were the last hand-rolled
golden-section loops in the tree (they shared a loose file-global `gr`). They
were first factored into a shared `util::golden_section_max` in
`inst/include/plant/optimize.h`, then switched to **`util::brent_fmin`** — the
same bounded minimiser #4 put on the *outer* collar solve — so the codebase now
does 1-D profit maximisation through one routine everywhere.

**This does not touch `run_scm_collect`.** These two functions have **no caller
on the production path**: `TF24_Strategy::refresh_indices` calls
`find_root_collar_psi` and the `optimise_psi_stem_TF` line is commented out
(`src/tf24_strategy.cpp:381`). Their only callers are the RcppR6 bindings /
`tests/testthat/test-leaf.r`. So the SCM trajectory is **bit-identical** and the
whole-run profile is unchanged — profiling the scenario at the top of this
document would only measure build noise. The change is therefore profiled in
**isolation**, by calling `optimise_psi_stem_TF` directly.

**Measured** (`make compile` builds, `pkgload::load_all(recompile = FALSE)`;
46-point `psi_soil` sweep × 200 reps, median of 7; same Leaf params as
`test-leaf.r`):

| `GSS_tol_abs` | GSS | Brent | Speed-up |
|---|---|---|---|
| `1e-3` (production default) | 0.143 s | 0.063 s | **≈2.27×** |
| `1e-8` (test setting) | 0.327 s | 0.127 s | **≈2.57×** |

**Numerics change at the tolerance level** (same character as #4 — Brent locates
the smooth optimum more precisely than GSS's bracket-midpoint at the same tol).
At the production tol `1e-3`: max |Δ `opt_psi_stem`| ≈ 3.4e−4, max rel
|Δ `profit`| ≈ 5.7e−5; profit *at* the optimum is essentially unchanged because
the objective is flat near the max. Note the argmax still differs ~3.8e−4 even
at tol `1e-8` (while rel |Δ profit| < 6e−7): the inner Cᵢ bisection
(`ci_abs_tol`) caps how precisely the outer argmax is defined, so tightening the
outer tol below ~1e−4 does not sharpen it. Existing leaf tests pass 151/151
(their assertions are qualitative).

### Cumulative progress (baseline → end of Phase 1)

The per-item figures above were each measured against the build immediately
before them, so they don't compound cleanly. Here is the **end-to-end** result,
both binaries built clean from the same toolchain (`make clean && make compile`,
`-g -O2`) and loaded with `pkgload::load_all(".", recompile = FALSE)`, on the
**single-species MPL=10** SCM run (`birth_rate = 20`, default `Environment`,
median of 5 timed runs after a warm-up):

| Build | Commit | Median run | offspring_production |
|---|---|---|---|
| Baseline (roadmap, pre-optimisation) | `ff75ad22` | **39.78 s** | 4.712317419e−06 |
| HEAD (#1–#4 done) | `b6ec6045` | **9.05 s** | 4.797128277e−06 |

- **≈4.40× faster** (39.8 s → 9.0 s, −77 % wall-time; min-of-5 ratio 4.38×).
  This slightly beats the product of the individual claims (1.9 × 1.25 × 1.7 ≈
  4.0×) because removing the hottest spline evals also shrinks every inner
  root-find the later items still call.
- **#5 adds a further ≈1.32×** on top (15-layer, MPL = 5; see the #5 entry in
  Phase 1), so the full Phase-1 stack is ≈5–6× over baseline.
- **Numerics:** offspring_production shifts **+1.80 %** (+8.5e−08), the net of
  the two accuracy-improving items (#1 interval-mean, #4 Brent) — #2/#3/#5 are
  bit-identical. Direction is *toward* the fine-grained reference; magnitude is
  within the system's intrinsic build-to-build sensitivity. Existing
  leaf/TF24/environment tests pass unchanged.

(Reproduce with `scripts/bench_tf24.R`; absolute seconds are machine-specific
but the ratio is stable.)

## Analytical hydraulic transform (issue #468)

[Issue #468](https://github.com/traitecoevo/plant/issues/468) noted that the
Weibull hydraulic transform has a closed form involving the incomplete gamma
function, and asked (1) whether it matches the current numerical implementation,
(2) how large the spline interpolation error is, and (3) whether a monotone
interpolator (PCHIP) would help. This section answers (1) and (2) and resolves
(3); the runtime question was already settled in Phase 2 (the spline path is the
right choice for forward simulation — see *Tried and REJECTED: spline the xylem
hydraulic cost*).

### The closed form

The supply-side transpiration integrates the xylem vulnerability curve
`proportion_of_conductivity(m) = exp(-(m/b)^c)` (where `m = -psi` is the
potential magnitude in MPa), giving the hydraulic transform

$$F(m)=\int_0^m \exp\!\big[-(s/b)^c\big]\,ds .$$

Substituting $u=(s/b)^c$ yields a **lower** incomplete gamma:

$$F(m)=\frac{b}{c}\,\gamma\!\left(\tfrac1c,\,(m/b)^c\right)
      =\frac{b}{c}\,\Gamma\!\left(\tfrac1c\right)\,
       P\!\left(\tfrac1c,\,(m/b)^c\right),$$

i.e. in R, `F = (b/c) * gamma(1/c) * pgamma((m/b)^c, 1/c)`. This is identical to
the issue's $P_{50}$-parameterised, *upper*-incomplete-gamma form: the code sets
`b = p_50 / (ln 2)^(1/c)`, so `ln2·(−ψ/p_50)^c = (m/b)^c`, and lower-vs-upper
differ only by the additive constant $(b/c)\Gamma(1/c)$, which cancels in
$F(\psi_2)-F(\psi_1)$.

### Verification and error (reproduce with `scripts/validate_gamma_transform.R`)

The script builds the real C++ `Leaf` and compares the closed form against both
numerical paths — `transpiration()` (the production `tk::spline` of $F$) and
`transpiration_full_integration()` (QAG adaptive quadrature) — plus the inverse
spline `psi_from_transpiration`. The table shows the trapezoid-knot build
(*before*) and the analytic-knot build now in the code (*after*, see
*Implemented* below); QAG is build-independent. Over a 2001-point sweep:

| Parameter set (`ncontrol`) | QAG (rel) | spline $F$ before (abs) | spline $F$ after (abs) | inverse before | inverse after |
|---|---|---|---|---|---|
| TF24 default `p_50=1.85` (100) | 1.4e−6 | 2.2e−4 | 1.5e−4 | 2.0e−2 MPa | 2.7e−4 MPa |
| test-leaf `c=2.04,b=3` (100)   | 1.7e−10 | 9.7e−5 | 3.2e−6 | 8.1e−4 MPa | 3.2e−4 MPa |
| TF24 default (1000)            | 1.4e−6 | 1.8e−6 | 9.5e−7 | 1.8e−4 MPa | 2.4e−6 MPa |
| TF24 default (30)              | 1.3e−6 | 2.7e−3 | 1.8e−3 | 2.1e−1 MPa | 3.4e−3 MPa |

**(1) The closed form is correct.** QAG quadrature — an independent numerical
method — agrees with it to ~1e−6 relative (its own tolerance floor), and to
~1e−10 for the steeper test curve. The analytical transform can stand as a
validation reference.

**(2) The spline error has *two* parts; only one is quadrature bias.** The build
splits into (a) knot error — the trapezoid quadrature bias baked into the
construction values — and (b) cubic interpolation error *between* knots. Seeding
the knots analytically (below) drives (a) to machine epsilon (knot error
2.2e−4 → 3e−15) and isolates (b). The earlier read of this section — "99.7%
trapezoid, 0.3% cubic" — was **wrong**: it subtracted the max knot error from the
max total error, but the two maxima occur at *different* potentials. For the
production curve ($c=1.09$) the residual after exact seeding is **1.46e−4, all of
it in the first inter-knot cell** ($\psi\approx0.04$, before the first knot at
0.105). There $F$ has unbounded higher derivatives at the origin
($F'''\sim m^{c-2}$ diverges for $c<2$), so a single cubic cell cannot track it —
this is genuine interpolation error, not quadrature. The trapezoid bias, large at
*high* $\psi$, dominated the *before* total and masked it. For the steeper
test-leaf curve ($c=2.04$, no origin singularity) seeding does help in full:
9.7e−5 → 3.2e−6 (~30×).

### Implemented

`setup_transpiration` and `setup_root_vulnerability` now seed their cumulative-
integral knots from the gamma closed form (`boost::math::tgamma_lower`) instead of
a running trapezoid sum — same knots, same `tk::spline`, same O(1) hot-path eval.
Measured effect (table above):

- **Inverse transform (used in simulation): ~60–75× more accurate.** `p_50=1.85`,
  `ncontrol=100`: 2.0e−2 → 2.7e−4 MPa. The inverse's error *was* bias-dominated,
  so removing the bias is almost the whole story — this is the real payoff.
- **Forward $F$: modest for the production curve (~1.5×, 2.2e−4 → 1.5e−4)** because
  its error is the near-origin interpolation term (b), not bias; large (~30×) for
  steep curves with no origin singularity.
- **Trajectory shift** (MPL=5 self-comparison, `run_scm_collect`): not
  bit-identical — by design. Hydraulics-driven states move ~0.2–0.3%
  (`transpiration`/`E_up_`/`opt_psi_stem`/`height` rel ≤3e−3), `offspring_production`
  ~1e−3; the shift is smooth (large relative diffs appear only at near-zero-density
  cohort boundaries and sign crossings), with identical cohort/step structure. The
  move is toward the more accurate integral.

### Consequences

- **(3) PCHIP would not help.** The production residual is the first-cell origin
  singularity in $F$ ($F'''$ diverges for $c<2$) — a *lower*-order monotone
  interpolator would do worse there, not better, and monotonicity/overshoot was
  never the issue ($F$, $F^{-1}$ are smooth and the cubic does not overshoot). To
  shave the residual further the lever is a non-uniform knot grid (denser near
  $\psi=0$), not a different interpolator. Out of scope.
- **Calibration / AD (issue step 5)** remains the closed form's other natural
  home: it is a smooth expression in `(p_50, c, K_max)` with no spline-build
  graph to differentiate through. Out of scope here.

## Build note: use `make`, not `load_all`

Build the package with **`make compile`** (`pkgbuild::compile_dll(debug=FALSE)`,
`-O3 -march=native`). `devtools::load_all()` recompiles at `-g -O2`, which is
both **slower at runtime** and produces **slightly different floating-point
results** (FMA/reassociation differ from `-O3 -march=native`); under the
adaptive ODE solver those tiny differences amplify, so trajectories from a
`load_all` build will not match a `make` build bit-for-bit even when the source
is identical. For benchmarking/profiling, load the `make`-built `.so` without
recompiling, e.g. `pkgload::load_all(".", recompile = FALSE)`.