---
title: "Testing root model"
format: html
editor_options:
  chunk_output_type: console
---

To run this vignette, make and rebuild plant using root_water_uptake branch. Then, run below.

This vignette exercises the TF24 strategy with the new multi-layer root water
uptake model. A single cohort of one species (`lma = 0.07`) is grown over a
patch lifetime in a soil column discretised into 15 layers, driven by a
seasonal (sinusoidal) rainfall series. `run_scm_collect` integrates the size-
structured population *and* the per-layer soil water balance simultaneously.

PERFORMANCE NOTE: runtime is dominated by the plant hydraulic solver
(`Leaf::find_root_collar_psi`). To profile, drop `max_patch_lifetime` to ~5
(≈40 s) rather than 75 (≈8 min); the per-step cost is unchanged, only the number
of steps differs. See **Implementation & performance notes** at the end of this
document for the model overview, a profile, and an optimisation roadmap.

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(plant)
# devtools::load_all("/Users/z3533545/Documents/GitHub/plant/")
```

```{r}
# Patch lifetime
max_patch_lifetime <-75
# Basic parameters
p0 <- scm_base_parameters("TF24", "TF24_Env")
p0$max_patch_lifetime <- max_patch_lifetime
traits <- trait_matrix(c(0.07), c("lma"))
p1 <- expand_parameters(traits, p0)

env <- Environment("TF24")

# fifteen soil layers
env$set_soil_number_of_depths(15)
# set soil moisture 0.2 (sat is 0.428)
env$set_soil_water_state(rep(c(0.2), times = 15))

# create rainfall vector across patch lifetime
x = seq(0,max_patch_lifetime,length.out = 1000)
y = 0.25*sin(2*pi*x) + 1
env$extrinsic_drivers_set_variable("rainfall", x=x, y=y)

# no change to ctrl at this point
ctrl <- scm_base_control()
# ctrl$ode_tol_rel <- 1e-4
# ctrl$ode_tol_abs <- 1e-4

system.time(results <- run_scm_collect(p1, env = env, ctrl = ctrl))
saveRDS(results, "results_high.RDS")
results <- readRDS("results_high.RDS")
```

Create diagnostic plots 

```{r}
source("vignettes/models/plot_diagnostics.R")
plot_diagnostics(results, x, y, filename = "diagnostic_plot5_high.png")
```

Plot the root distribution across soil depths (TODO: I think this is broken at the moment)

```{r}
# Replicate the C++ root mass distribution from tf24_strategy.cpp:
#   Q(z, height, eta_x) = (1 - (z/height)^eta_x)^2  for z <= height, else 0
#   rooting_depth       = min(height, 1.5)
#   root_mass_scale     = 83.26 * 0.5 * mass_root
#   mass_root_prop[a]   = root_mass_scale * (Q(prev_depth) - Q(curr_depth))

Q_fn <- function(z, height, eta_x = 0.2) {
  ifelse(z > height, 0.0, (1 - (z / height)^eta_x)^2)
}

Q_fn(0.0001,0.4)

s <- p1$strategies[[1]]

soil_depths <- results$env$soil_depth$soil_depth   # length-15 vector of cumulative depths

results %>%
  expand_state() %>%
  pluck("species") %>%
  select(species, time, node, step, density, patch_density, height, area_leaf) %>%
  mutate(
    mass_root       = s$a_r1 * area_leaf,
    rooting_depth   = pmin(height, 1.5),
    root_mass_scale = 83.26 * 0.5 * mass_root
  ) %>%
  # Expand to one row per soil layer WITHOUT deduplication.
  # tidyr::crossing() de-dupes its inputs, which merges nodes with identical
  # state values and corrupts the rooting_depth/layer_idx alignment.
  # mutate + unnest preserves every original row.
  mutate(layer_idx = list(seq_along(soil_depths))) %>%
  tidyr::unnest(layer_idx) %>%
  mutate(soil_depth = soil_depths[layer_idx],
         q_curr     = Q_fn(soil_depth, rooting_depth)) %>%
  # Mirror C++ exactly: prev_q starts at 1.0, then is the previous layer's q.
  # Using lag() within each plant avoids recomputing Q(0, height) which can
  # have floating-point inconsistency vs Q(prev_depth, height).
  group_by(species, step, node, time) %>%
  arrange(layer_idx, .by_group = TRUE) %>%
  mutate(
    q_prev          = lag(q_curr, default = 1.0),
    mass_root_layer = root_mass_scale * pmax(0.0, q_prev - q_curr)
  ) %>%
  ungroup() %>%
  # Community-level: integrate over size distribution per layer
  group_by(time, step, soil_depth) %>%
  summarise(
    mass_root_layer = sum(mass_root_layer * density * patch_density),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = time, y = mass_root_layer, colour = soil_depth, group = soil_depth)) +
  geom_line() +
  scale_colour_viridis_c(name = "Soil depth (m)") +
  scale_y_log10() +
  theme_bw() +
  theme(text = element_text(size = 16)) +
  ylab(expression(paste(Root~mass~per~layer, " (kg ", m^{-2}, ")"))) +
  xlab("Time (years)") -> root_mass_plot

root_mass_plot
```

# Implementation & performance notes

This section documents the multi-layer root water-uptake model added on the
`root_water_uptake` branch: what it does, where the runtime goes, and which
optimisations have been tried or are worth trying. (The corresponding R
prototype lives in `vignettes/models/root_water_uptake.Rmd`; per-function
scientific commentary lives in the C++ sources referenced below.)

## Model overview

The branch couples three pieces, integrated together by `run_scm_collect`:

- **Soil water (bucket model)** — `inst/include/plant/models/tf24_environment.h`.
  The soil column is discretised into `soil_number_of_depths` layers, each a
  water bucket with its own moisture state. `compute_rates` does a per-layer
  mass balance: infiltration (rainfall with saturation-excess runoff) →
  gravitational drainage (Clapp–Hornberger / Zeng–Decker unsaturated
  conductivity) → root uptake. Moisture↔potential uses a retention curve, with
  a `psi_soil` cache keyed on the moisture state.
- **Root distribution** — `src/tf24_strategy.cpp` (`net_mass_production_dt`).
  Total fine-root mass is spread over depth with the same `Q()` cumulative
  shape function used for the canopy, then passed to the leaf.
- **Plant hydraulics** — `src/leaf_model.cpp` (the core new science, after
  Potkay et al. 2021):
  - `set_physiology` builds a per-layer root resistance network — horizontal
    (intra-layer) plus cumulative vertical (inter-layer) resistances, scaled by
    root carbon and modulated by a root vulnerability curve.
  - `E_from_Soil_to_Root_Collar` sums water drawn from each layer for a given
    root-collar potential (including hydraulic redistribution into wetter
    layers).
  - `find_root_collar_psi` is the master solver, with **two nested levels**:
    (1) a continuity root-find (soil supply = stem transpiration), and (2) a
    golden-section search for the profit-maximising (assimilation − hydraulic
    cost) operating point — plus several early-exit short-circuits for plant
    shut-down conditions.

## Where the runtime goes (profile)

Profiled with `max_patch_lifetime = 5` (≈40 s, a faithful proxy for the full
≈8-min run — only the number of ODE steps differs). Runtime is ~entirely in
`net_mass_production_dt → find_root_collar_psi`. The flat profile:

| Hot spot | Why |
|---|---|
| **Spline evaluation** (`tk::spline::operator()` via `Interpolator::eval`) — the single hottest leaf | `root_vuln_from_psi.eval` is called `(n+1)=6 × n_layers` times per `E_from_Soil_to_Root_Collar` call |
| **`E_from_Soil_to_Root_Collar`** | called inside **two bisection root-finds** (`find_root_psi`, ~16 iters each) |
| **`psi_stem_to_ci`** bisection | called inside every golden-section profit evaluation |

The cost is **algorithmic structure — nested root-finds × spline evals — not
memory allocation.** The per-call `mass_root_prop_` / `psi_soil_inverted_`
allocations and the vector copies in `set_physiology` are <0.1% of runtime;
preallocating them will not move the needle (profile-confirmed). This whole
optimum is recomputed for every individual at every ODE sub-step.

## Tried and REJECTED: super-linear root-finder (TOMS748)

Replacing the bisection in `util::uniroot` with a Brent-like super-linear
solver (TOMS748) is the obvious "biggest lever" — it normally needs ~5–8
evaluations where bisection needs ~16–30. It was implemented and **reverted**:
in this code path it produced **NaN soil potentials and ran *slower*** (the
coupled soil-water ODE destabilised). The hydraulic functions have kinks/clamps
and use splines with extrapolation disabled, so the interpolation steps probe
bad points; **bisection's robustness is load-bearing here.** A documented but
unused `util::uniroot_smooth` helper remains in `inst/include/plant/uniroot.h`
with this caveat, so the experiment is not silently repeated.

## Optimisation roadmap (ranked)

**DONE (#2 + #3): ~1.9× speed-up, bit-identical output** (MPL=5: ≈41 s → ≈21 s;
`all.equal` vs baseline TRUE). Both are numerically identical so the science is
unchanged. Implemented:

- **#3 O(1) spline index** — `tk::spline::operator()` now detects an
  equidistant x-grid in `set_points` and replaces the `std::lower_bound` binary
  search with direct arithmetic plus a ±1 correction that reproduces the
  binary-search index exactly (see `src/tk_spline.cpp`, `m_uniform`). This was
  the single hottest leaf in the profile; the same fix was reported externally
  in traitecoevo/plant#435 (~18% there).
- **#2 Memoised `transpiration()`** — a 1-entry cache on `(psi_stem,
  psi_upstream)` skips the repeated identical spline lookups within each profit
  evaluation; invalidated in `set_physiology` (see `src/leaf_model.cpp`).

**DONE (#1): ~1.25× speed-up, numerics changed (toward more accurate).** The
6-point (`n+1`) sample-mean of root conductivity in `E_from_Soil_to_Root_Collar`
is replaced by the *exact* interval mean `(1/(b−a))·∫ₐᵇ f_r dψ`, obtained from a
pre-integrated curve `root_vuln_integral_from_psi` (`G(m)=∫₀ᵐ f_r ds`, built in
`setup_root_vulnerability`) with **2 spline evals instead of 6** — the same
pre-integrated-curve trick as `setup_transpiration` for the xylem. The interval
is split at ψ=0 (ψ>0 ⇒ vulnerability 1).

This is **more accurate**: vs a fine-grained reference the new interval mean errs
~1e−3 where the old sample-mean erred up to ~5e−2. It therefore **changes the
trajectory** slightly — on a single-species MPL=10 run, `offspring_production`
shifts 4.712e−06 → 4.789e−06 (+1.6% rel, 7.6e−8 abs; toward the more accurate
value). Existing leaf/TF24 tests pass unchanged (their tolerances are absolute at
this magnitude). **Measured speed-up** (identical `-g -O2` build, old vs new,
median of 3 runs each on the same MPL=10 single-species run): **26.2 s → 21.0 s,
≈1.25× (~20%)**. The hot function gets the predicted ~3× fewer spline evals; the
whole-run gain is smaller because that function is one stage inside two nested
root-finds plus the golden-section search.

Remaining (these **change numerics** — validate against the science before
adoption: capture a baseline trajectory, re-run, compare):

4. **Brent for the golden-section profit max** (~17 → ~10 iters). Moderate
   risk; the profit objective is smooth near the optimum.

## Build note: use `make`, not `load_all`

Build the package with **`make compile`** (`pkgbuild::compile_dll(debug=FALSE)`,
`-O3 -march=native`). `devtools::load_all()` recompiles at `-g -O2`, which is
both **slower at runtime** and produces **slightly different floating-point
results** (FMA/reassociation differ from `-O3 -march=native`); under the
adaptive ODE solver those tiny differences amplify, so trajectories from a
`load_all` build will not match a `make` build bit-for-bit even when the source
is identical. For benchmarking/profiling, load the `make`-built `.so` without
recompiling, e.g. `pkgload::load_all(".", recompile = FALSE)`.