---
title: "Testing root model"
format: html
editor_options:
  chunk_output_type: console
---

To run this vignette, make and rebuild plant using root_water_uptake branch. Then, run below.

This vignette exercises the TF24 strategy with the new multi-layer root water
uptake model. A single cohort of one species (`lma = 0.07`) is grown over a
patch lifetime in a soil column discretised into 15 layers, driven by a
seasonal (sinusoidal) rainfall series. `run_scm_collect` integrates the size-
structured population *and* the per-layer soil water balance simultaneously.

PERFORMANCE NOTE: runtime is dominated by the plant hydraulic solver
(`Leaf::find_root_collar_psi`). To profile, drop `max_patch_lifetime` to ~5
(≈40 s) rather than 75 (≈8 min); the per-step cost is unchanged, only the number
of steps differs. See **Implementation & performance notes** at the end of this
document for the model overview, a profile, and an optimisation roadmap.

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(plant)
# devtools::load_all(".")

```

```{r}
# Patch lifetime
max_patch_lifetime <- 70
# Basic parameters
p0 <- scm_base_parameters("TF24", "TF24_Env")
p0$max_patch_lifetime <- max_patch_lifetime
traits <- trait_matrix(c(0.07), c("lma"))
p1 <- expand_parameters(traits, p0)

# Force finer node-introduction spacing so the *collected* output is sampled at
# <=0.25 yr everywhere. run_scm_collect saves state once per node introduction
# (NOT per ODE step), and the default schedule coarsens to ~0.5-1 yr late, which
# under-samples the 1-yr seasonal rainfall signal -> jagged aggregate trajectories.
# The node schedule is the lever for output smoothness here, NOT ode_* controls.
# See traitecoevo/plant#474. We keep the dense early default spacing and add a
# regular <=0.25 yr grid on top.
fine_times <- sort(unique(c(
  node_schedule_times_default(max_patch_lifetime),
  seq(0, max_patch_lifetime, by = 0.125)
)))
# p1$node_schedule_times <- list(fine_times)

env <- Environment("TF24")

# fifteen soil layers
env$set_soil_number_of_depths(15)
# set soil moisture 0.2 (sat is 0.428)
env$set_soil_water_state(rep(c(0.2), times = 15))

# create rainfall vector across patch lifetime
x = seq(0,max_patch_lifetime,length.out = 1000)
# NOTE: mean rainfall 1.0 keeps this a healthy-cohort speed test. Lower means
# (e.g. 0.5) dry the column and drive the cohort towards extinction; those
# drought regimes are explored in the run_rainfall_regime sweep below. They now
# run without aborting: the soil model floors moisture at a residual theta_r so
# the retention curve psi_from_soil_moist stays finite (was traitecoevo/plant#485,
# "set_physiology received non-finite psi_soil").
y = 0.25*sin(2*pi*x) + 1.0
env$extrinsic_drivers_set_variable("rainfall", x=x, y=y)

ctrl <- control()
# NOTE: ode_step_size_max only bounds the *internal* adaptive step between node
# introductions; it does NOT add collected output points and does NOT smooth the
# jaggedness (that's the node schedule above). Left commented out to avoid a
# needlessly slow run. See traitecoevo/plant#474.
# ctrl$ode_step_size_max <- 1e-03

# ctrl$ode_tol_rel <- 1e-5
# ctrl$ode_tol_abs <- 1e-5

system.time(results <- run_scm(p1, env = env, ctrl = ctrl, collect = TRUE))
saveRDS(results, "results_high.RDS")
results <- readRDS("results_high.RDS")

source("R/plot_hydraulic_diagnostics.R")
plot_diagnostics(results, x, y, filename = "diagnostic_plot5_low.pdf")
```

```{r}
run_rainfall_regime <- function(mean_rainfall = mean_rainfall, amplitude_rainfall = amplitude_rainfall){
max_patch_lifetime <-20
# Basic parameters
p0 <- scm_base_parameters("TF24", "TF24_Env")
p0$max_patch_lifetime <- max_patch_lifetime
traits <- trait_matrix(c(0.07), c("lma"))
p1 <- expand_parameters(traits, p0)

env <- Environment("TF24")

# fifteen soil layers
env$set_soil_number_of_depths(15)
# set soil moisture 0.2 (sat is 0.428)
env$set_soil_water_state(rep(c(0.2), times = 15))

# create rainfall vector across patch lifetime
x = seq(0,max_patch_lifetime,length.out = 1000)
y = amplitude_rainfall*sin(2*pi*x) + mean_rainfall
env$extrinsic_drivers_set_variable("rainfall", x=x, y=y)

# no change to ctrl at this point
ctrl <- scm_base_control()

# ctrl$ode_tol_rel <- 1e-5
# ctrl$ode_tol_abs <- 1e-5
print(paste0("mean_rainfall = ", mean_rainfall, "amplitude_rainfall = ", amplitude_rainfall))
system.time(results1 <- run_scm_collect(p1, env = env, ctrl = ctrl))
results1
}

tibble(mean_rainfall = rep(seq(0.5,2,length.out = 10))) %>%
expand_grid(amplitude_rainfall = c(0,0.5)) %>%
mutate(amplitude_rainfall = amplitude_rainfall * mean_rainfall) -> env_drivers

future::plan("multisession", workers = 2)


result_list <- furrr::future_map2(
  env_drivers$mean_rainfall,
  env_drivers$amplitude_rainfall,
  ~ tryCatch(
    run_rainfall_regime(.x, .y),
    error = function(e) {
      structure(
        list(
          error = conditionMessage(e),
          mean_rainfall = .x,
          amplitude_rainfall = .y
        ),
        class = "run_rainfall_error"
      )
    }
  )
)
```

```{r}
find_total_community <- function(results, mean_rainfall, amplitude_rainfall){
  results %>%
    expand_state() %>%
    pluck("species") %>%
    integrate_over_size_distribution() %>%
    mutate(
      mean_rainfall = mean_rainfall,
      amplitude_rainfall = amplitude_rainfall
    )
}

combined_results <- pmap_df(
  list(
    result = result_list,
    mean_rainfall = rep(seq(0.5, 2, length.out = 10), each = 2),
    amplitude_rainfall = rep(c(0, 0.25), 10)
  ),
  ~ {
    if (inherits(..1, "run_rainfall_error")) {
      tibble()
    } else {
      find_total_community(..1, ..2, ..3)
    }
  }
)

combined_results %>%
ggplot(aes(x = time, y = area_leaf)) +
geom_line(aes(colour = as.factor(amplitude_rainfall), group = as.factor(amplitude_rainfall))) +
facet_wrap(~mean_rainfall)
```

Plot the root distribution across soil depths (TODO: I think this is broken at the moment)

```{r}
# Replicate the C++ root mass distribution from tf24_strategy.cpp:
#   Q(z, height, eta_x) = (1 - (z/height)^eta_x)^2  for z <= height, else 0
#   rooting_depth       = min(height, 1.5)
#   root_mass_scale     = 83.26 * 0.5 * mass_root
#   mass_root_prop[a]   = root_mass_scale * (Q(prev_depth) - Q(curr_depth))

Q_fn <- function(z, height, eta_x = 0.2) {
  ifelse(z > height, 0.0, (1 - (z / height)^eta_x)^2)
}

Q_fn(0.0001,0.4)

s <- p1$strategies[[1]]

soil_depths <- results$env$soil_depth$soil_depth   # length-15 vector of cumulative depths

results %>%
  expand_state() %>%
  pluck("species") %>%
  select(species, time, node, step, density, patch_density, height, area_leaf) %>%
  mutate(
    mass_root       = s$a_r1 * area_leaf,
    rooting_depth   = pmin(height, 1.5),
    root_mass_scale = 83.26 * 0.5 * mass_root
  ) %>%
  # Expand to one row per soil layer WITHOUT deduplication.
  # tidyr::crossing() de-dupes its inputs, which merges nodes with identical
  # state values and corrupts the rooting_depth/layer_idx alignment.
  # mutate + unnest preserves every original row.
  mutate(layer_idx = list(seq_along(soil_depths))) %>%
  tidyr::unnest(layer_idx) %>%
  mutate(soil_depth = soil_depths[layer_idx],
         q_curr     = Q_fn(soil_depth, rooting_depth)) %>%
  # Mirror C++ exactly: prev_q starts at 1.0, then is the previous layer's q.
  # Using lag() within each plant avoids recomputing Q(0, height) which can
  # have floating-point inconsistency vs Q(prev_depth, height).
  group_by(species, step, node, time) %>%
  arrange(layer_idx, .by_group = TRUE) %>%
  mutate(
    q_prev          = lag(q_curr, default = 1.0),
    mass_root_layer = root_mass_scale * pmax(0.0, q_prev - q_curr)
  ) %>%
  ungroup() %>%
  # Community-level: integrate over size distribution per layer
  group_by(time, step, soil_depth) %>%
  summarise(
    density_layer = sum(density),
    patch_density_layer = sum(patch_density),
    mass_root_layer = sum(mass_root_layer),
    mass_root_int_layer = sum(mass_root_layer * density * patch_density),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(density_layer, patch_density_layer, mass_root_layer, mass_root_int_layer)) %>%
  ggplot(aes(x = time, y = value, colour = soil_depth, group = soil_depth)) +
  geom_line() +
  facet_wrap(~name, scales = "free") +
  scale_colour_viridis_c(name = "Soil depth (m)") +
  theme_bw() +
  theme(text = element_text(size = 16)) +
  ylab(expression(paste(Root~mass~per~layer, " (kg ", m^{-2}, ")"))) +
  xlab("Time (years)") -> root_mass_plot

root_mass_plot
```

```{r}

results %>%
  expand_state() %>%
  pluck("species") %>%
  select(species, time, node, step, density, patch_density, height, area_leaf) %>%
  mutate(
    mass_root       = s$a_r1 * area_leaf,
    rooting_depth   = pmin(height, 1.5),
    root_mass_scale = 83.26 * 0.5 * mass_root
  ) %>%
  # Expand to one row per soil layer WITHOUT deduplication.
  # tidyr::crossing() de-dupes its inputs, which merges nodes with identical
  # state values and corrupts the rooting_depth/layer_idx alignment.
  # mutate + unnest preserves every original row.
  mutate(layer_idx = list(seq_along(soil_depths))) %>%
  tidyr::unnest(layer_idx) %>%
  mutate(soil_depth = soil_depths[layer_idx],
         q_curr     = Q_fn(soil_depth, rooting_depth)) %>%
  # Mirror C++ exactly: prev_q starts at 1.0, then is the previous layer's q.
  # Using lag() within each plant avoids recomputing Q(0, height) which can
  # have floating-point inconsistency vs Q(prev_depth, height).
  group_by(species, step, node, time) %>%
  arrange(layer_idx, .by_group = TRUE) %>%
  mutate(
    q_prev          = lag(q_curr, default = 1.0),
    mass_root_layer = root_mass_scale * pmax(0.0, q_prev - q_curr)
  ) %>%
  ungroup() %>%
  # Community-level: integrate over size distribution per layer
  group_by(time, step) %>%
  summarise(mass_root = sum(mass_root),
            area_leaf = sum(area_leaf)) %>%
            ungroup() %>%
            pivot_longer(cols = c(mass_root, area_leaf)) %>%
            ggplot(aes(x = time, y = value)) +
            geom_line() +
            facet_wrap(~name, scales= "free")
```


# Implementation & performance notes

This section documents the multi-layer root water-uptake model added on the
`root_water_uptake` branch: what it does, where the runtime goes, and which
optimisations have been tried or are worth trying. (The corresponding R
prototype lives in `vignettes/models/root_water_uptake.Rmd`; per-function
scientific commentary lives in the C++ sources referenced below.)

## Model overview

The branch couples three pieces, integrated together by `run_scm_collect`:

- **Soil water (bucket model)** — `inst/include/plant/models/tf24_environment.h`.
  The soil column is discretised into `soil_number_of_depths` layers, each a
  water bucket with its own moisture state. `compute_rates` does a per-layer
  mass balance: infiltration (rainfall with saturation-excess runoff) →
  gravitational drainage (Clapp–Hornberger / Zeng–Decker unsaturated
  conductivity) → root uptake. Moisture↔potential uses a retention curve, with
  a `psi_soil` cache keyed on the moisture state.
- **Root distribution** — `src/tf24_strategy.cpp` (`net_mass_production_dt`).
  Total fine-root mass is spread over depth with the same `Q()` cumulative
  shape function used for the canopy, then passed to the leaf.
- **Plant hydraulics** — `src/leaf_model.cpp` (the core new science, after
  Potkay et al. 2021):
  - `set_physiology` builds a per-layer root resistance network — horizontal
    (intra-layer) plus cumulative vertical (inter-layer) resistances, scaled by
    root carbon and modulated by a root vulnerability curve.
  - `E_from_Soil_to_Root_Collar` sums water drawn from each layer for a given
    root-collar potential (including hydraulic redistribution into wetter
    layers).
  - `find_root_collar_psi` is the master solver, with **two nested levels**:
    (1) a continuity root-find (soil supply = stem transpiration), and (2) a
    golden-section search for the profit-maximising (assimilation − hydraulic
    cost) operating point — plus several early-exit short-circuits for plant
    shut-down conditions.

## Phase 1 — initial optimisation roadmap (#1–#5)

This phase is the original record: it took the new root-water model from its
first working version to roughly 4–6× faster through five ranked items.

### Where the runtime goes (initial profile)

Profiled with `max_patch_lifetime = 5` (≈40 s, a faithful proxy for the full
≈8-min run — only the number of ODE steps differs). Runtime is ~entirely in
`net_mass_production_dt → find_root_collar_psi`. The flat profile:

| Hot spot | Why |
|---|---|
| **Spline evaluation** (`tk::spline::operator()` via `Interpolator::eval`) — the single hottest leaf | `root_vuln_from_psi.eval` is called `(n+1)=6 × n_layers` times per `E_from_Soil_to_Root_Collar` call |
| **`E_from_Soil_to_Root_Collar`** | called inside **two bisection root-finds** (`find_root_psi`, ~16 iters each) |
| **`psi_stem_to_ci`** bisection | called inside every golden-section profit evaluation |

The cost is **algorithmic structure — nested root-finds × spline evals — not
memory allocation.** The per-call `mass_root_prop_` / `psi_soil_inverted_`
allocations and the vector copies in `set_physiology` are <0.1% of runtime;
preallocating them will not move the needle (profile-confirmed). This whole
optimum is recomputed for every individual at every ODE sub-step.

### Tried and REJECTED: super-linear root-finder (TOMS748)

Replacing the bisection in `util::uniroot` with a Brent-like super-linear
solver (TOMS748) is the obvious "biggest lever" — it normally needs ~5–8
evaluations where bisection needs ~16–30. It was implemented and **reverted**:
in this code path it produced **NaN soil potentials and ran *slower*** (the
coupled soil-water ODE destabilised). The hydraulic functions have kinks/clamps
and use splines with extrapolation disabled, so the interpolation steps probe
bad points; **bisection's robustness is load-bearing here.** A documented but
unused `util::uniroot_smooth` helper remains in `inst/include/plant/uniroot.h`
with this caveat, so the experiment is not silently repeated.

### Optimisation roadmap (ranked)

**DONE (#2 + #3): ~1.9× speed-up, bit-identical output** (MPL=5: ≈41 s → ≈21 s;
`all.equal` vs baseline TRUE). Both are numerically identical so the science is
unchanged. Implemented:

- **#3 O(1) spline index** — `tk::spline::operator()` now detects an
  equidistant x-grid in `set_points` and replaces the `std::lower_bound` binary
  search with direct arithmetic plus a ±1 correction that reproduces the
  binary-search index exactly (see `src/tk_spline.cpp`, `m_uniform`). This was
  the single hottest leaf in the profile; the same fix was reported externally
  in traitecoevo/plant#435 (~18% there).
- **#2 Memoised `transpiration()`** — a 1-entry cache on `(psi_stem,
  psi_upstream)` skips the repeated identical spline lookups within each profit
  evaluation; invalidated in `set_physiology` (see `src/leaf_model.cpp`).

**DONE (#1): ~1.25× speed-up, numerics changed (toward more accurate).** The
6-point (`n+1`) sample-mean of root conductivity in `E_from_Soil_to_Root_Collar`
is replaced by the *exact* interval mean `(1/(b−a))·∫ₐᵇ f_r dψ`, obtained from a
pre-integrated curve `root_vuln_integral_from_psi` (`G(m)=∫₀ᵐ f_r ds`, built in
`setup_root_vulnerability`) with **2 spline evals instead of 6** — the same
pre-integrated-curve trick as `setup_transpiration` for the xylem. The interval
is split at ψ=0 (ψ>0 ⇒ vulnerability 1).

This is **more accurate**: vs a fine-grained reference the new interval mean errs
~1e−3 where the old sample-mean erred up to ~5e−2. It therefore **changes the
trajectory** slightly — on a single-species MPL=10 run, `offspring_production`
shifts 4.712e−06 → 4.789e−06 (+1.6% rel, 7.6e−8 abs; toward the more accurate
value). Existing leaf/TF24 tests pass unchanged (their tolerances are absolute at
this magnitude). **Measured speed-up** (identical `-g -O2` build, old vs new,
median of 3 runs each on the same MPL=10 single-species run): **26.2 s → 21.0 s,
≈1.25× (~20%)**. The hot function gets the predicted ~3× fewer spline evals; the
whole-run gain is smaller because that function is one stage inside two nested
root-finds plus the golden-section search.

**DONE (#4): ~1.7× speed-up, numerics changed (tolerance-level shift).** The
golden-section search (GSS) over the feasible collar-potential interval in
`find_root_collar_psi` is replaced by **Brent's method** (golden-section +
parabolic interpolation — R's `Brent_fmin`, new `util::brent_fmin` in
`inst/include/plant/optimize.h`). On the smooth profit objective the parabolic
step converges super-linearly, while the golden-section fallback preserves GSS's
robustness. It is a **bounded** minimiser: it uses the same `[bound_a, bound_b]`
bracket and only ever evaluates *strictly interior* points, so it never probes
the clamped vulnerability-curve endpoints — this is why it is safe here where the
TOMS748 root-finder was not (it stays stable at the production tol `1e-3`; only
an absurdly tight tol `1e-9` over-resolves into the documented NaN instability).

> **LATER REVERTED for `find_root_collar_psi` (gradient smoothness).** Brent's
> parabolic step makes its argmax a *non-smooth* function of the inputs. Because
> this solver's argmax feeds the demographic growth-rate gradient (which needs
> the operating point to vary smoothly with plant state), the **outer** collar
> optimisation was switched **back to golden-section search** — now via the
> shared `util::golden_section_max` in `optimize.h` (the efficient one-eval-per-
> iteration reuse variant), alongside `util::brent_fmin`. The #4 ~1.7× figure no
> longer applies to this solver; GSS is accepted as the slower-but-smoother
> choice here. **Brent is retained** for the off-path single-layer
> `optimise_psi_stem_*` optimisers, whose argmax feeds no gradient and where it
> measured ~2.3–2.6× faster. Output shifts vs the Brent build at the
> tolerance level (offspring 1.64927e−06 → 1.63974e−06 on the 15-layer MPL=5
> scenario) — same character as #4, the smooth optimum located slightly
> differently.

**Measured** (clean `-g -O2` builds, MPL=10 single-species run): **24.7 s →
14.6 s, ≈1.70×**. The gain beats the ~17→~10 iteration estimate because each GSS
iteration also carries an inner `find_psi_stem_from_psi_root` root-find, so
fewer iterations pays double. **Numerics change at the tolerance level**:
`offspring_production` shifts 0.39954 → 0.39637 (−0.79% rel); `height_sum` /
`height_max` move <0.01%. Both optimisers stop at a ψ-tolerance of `1e-3`;
Brent locates the smooth optimum more precisely than GSS's bracket-midpoint at
the same tolerance. The shift is the same order as this system's *intrinsic*
chaotic build-to-build sensitivity (a stale-object incremental build of identical
source moved offspring +0.4% on its own — see build note below). Existing
leaf/TF24/environment tests pass unchanged.

**Build caveat (observed):** incremental `make compile` can leave stale `.o`
files after header/branch changes, yielding an inconsistent binary. Use
`make clean` (or `touch` the changed sources) before benchmarking.

No remaining items on the ranked roadmap.

### Post-optimisation profile (after #1–#4)

Re-profiled after all four items, on the same single-species MPL=10 run. macOS
`sample` was used (the `jointprof`/`pprof` path documented above is unavailable
here; `Rprof`/`profvis` can't see into the `.so` — it's a single Rcpp call).
The `-g` build keeps C++ symbols; a few small functions inline under `-O2`.
Self-time over ~31.6 k active samples (idle wait excluded):

| Function | Self % | Note |
|---|---|---|
| `Leaf::psi_stem_to_ci` | **31.4 %** | inner Cᵢ bisection — **not touched by the roadmap** |
| `tk::spline::operator()` | 22.4 % | spline eval (already O(1)-indexed by #3) |
| `Leaf::E_from_Soil_to_Root_Collar` | 10.5 % | already 6→2 spline evals by #1 |
| `Interpolator::eval` | 7.6 % | spline wrapper |
| `pow` (+ stub) | ~7 % | transcendentals in hydraulic / vulnerability curves |
| `Leaf::find_root_collar_psi` | 6.0 % | master solver (Brent'd by #4) |
| `Leaf::find_root_psi` | 1.7 % | |
| `std::map<string,int>::at` + `memcmp` | ~3 % | string-keyed name→index lookups in the hot path — **now resolved, see below** |

The picture has shifted: spline evaluation was *the* hottest leaf before and is
still ~30 % combined (`tk::spline` + `Interpolator::eval`), but it's no longer
dominant — work is now spread across the solver structure.

#### DONE: string-keyed lookups (~3 % → ~0.5 %)

The two string-lookup hot spots flagged above were both removed (numerically
bit-identical, `offspring_production` unchanged at 4.797128277e−06):

- **aux/state index cache** — `compute_rates` did ~10 `aux_index.at("…")` /
  `state_index.at("…")` (`std::map<string,int>`, i.e. red-black-tree string
  compares) per derivs evaluation per individual. The keys are now resolved to
  ints once in `refresh_indices()` and cached as members (`src/tf24_strategy.cpp`).
- **extrinsic-driver memo** — the 6 time-varying drivers (PPFD, atm_vpd, ca,
  leaf_temp, atm_o2_kpa, atm_kpa) were each an `unordered_map<string,…>::at`
  (plus a by-value key copy) per individual per derivs evaluation. The `get_*`
  accessors now memoise per driver keyed on `time`, so repeated same-time reads
  across individuals collapse to one lookup per distinct time
  (`inst/include/plant/models/tf24_environment.h`); `evaluate` also takes the
  name by `const&` now (`inst/include/plant/extrinsic_drivers.h`).

Re-profiled afterwards: `std::map<…>::at` 1.7 %→0.3 %, `memcmp` 1.6 %→0.2 %,
`ExtrinsicDrivers::evaluate` dropped off entirely (the residual is the expected
one-lookup-per-distinct-time cache misses). Measured ~2 % + ~3 % wall-time
(controlled same-session A/B, 10 reps each); the rest of the profile is
unchanged in shape.

**Remaining lever (a possible roadmap #5):**

- **`psi_stem_to_ci` (~32 %)** is the inner Cᵢ bisection that #4's Brent did
  *not* touch (Brent went on the *outer* GSS profit-max). Giving it the same
  bounded-Brent treatment, or memoising it the way #2 did for `transpiration()`,
  is the obvious biggest remaining target.

#### DONE (#5): `E_from_Soil` integral-spline cache — and why `psi_stem_to_ci` was a red herring

The roadmap-#5 premise above **does not reproduce on the current binary.**
Re-profiling HEAD (macOS `sample`) on *both* the 15-layer root-water scenario at
the top of this document (MPL=5) and the `bench_tf24.R` default-`Environment` run
(MPL=10), `psi_stem_to_ci` is only **~2–3 %** — its symbol appears in 127 / 4012
stacks, and since self-time ⊆ inclusive that bounds it at **≤3.2 %**, not 31 %.
(The 31.4 % figure was measured at `b6ec6045`; it no longer holds.) The actual
hot path in *both* scenarios is **`E_from_Soil_to_Root_Collar` (~22–24 %) plus
the spline evals it drives** (`tk::spline` + `Interpolator::eval`, ~30 %
combined). `psi_stem_to_ci` is also a poor target on its own terms: it can't be
sped **bit-identically** (its bisection *result* is load-bearing for FP
reproducibility — the same reason TOMS748 is rejected above — and Brent probes
distinct points each call, so there is nothing to memoise).

So #5 instead caches the **cumulative root-vulnerability integral** lookups in the
`E_from_Soil` layer loop. That loop did **two `root_vuln_integral_from_psi.eval`
per soil layer**; the two arguments are always `{−ψ_soil[i], −P_x_r, 0}`. `−P_x_r`
is constant across layers (hoisted to **one eval per call**) and `−ψ_soil[i]` is
constant for the whole solve (precomputed once per `find_root_collar_psi` into the
new member `root_vuln_integral_soil_`). A memoised accessor keyed on
exact-`double` equality returns the cached values, engaging only when `psi_soil`
aliases `psi_soil_inverted_`, so the R-exposed by-value method keeps the unchanged
direct-eval fallback. **~30 spline evals/call → ~1.**

**Bit-identical** (this one really is, unlike #1/#4): `identical()` on the full
MPL=5 `run_scm_collect` trajectory vs HEAD; `offspring_production` unchanged at
4.797128277e−06 on the bench run; existing leaf/TF24 tests pass. Because the
change is a value cache rather than new index logic, it was also checked with a
standalone cached-vs-uncached C++ diff over a 27 113-point `P_x_r` sweep (exact
layer knots, `nextafter` neighbours, negative interior, and the ψ>0 /
`hi_neg→0` extrapolation tail the sim never exercises) — **0 mismatches,
max|ΔE| = 0**.

**Measured speed-up** (clean `make compile` builds, `pkgload::load_all(recompile
= FALSE)`, median of 5–7 timed runs):

| Scenario | Build | Median | Speed-up |
|---|---|---|---|
| 15-layer root-water, MPL=5 | pre-#5 → #5 | 8.84 s → 6.72 s | **≈1.32× (−24 %)** |
| bench default-`Environment`, MPL=10 | pre-#5 → #5 | 8.67 s → 7.82 s | **≈1.11× (−10 %)** |

The gain is larger in the 15-layer scenario because the root-vuln integral spline
is a bigger share of total spline time there (it scales with soil-layer count ×
collar-solve evals); the default-`Environment` run has fewer layers, so the
xylem/transpiration splines — which #5 doesn't touch — dominate the spline budget.
(Commit `Cache soil-side root-vuln integral lookups in E_from_Soil`.)

## Phase 2 — reprofile & reassessment (2026-06-18)

After Phase 1 the model was re-profiled from scratch on the 15-layer root-water
scenario at the top of this document (MPL = 5), to check whether the documented
hot-spot ranking still held and whether any cheap levers remained. It produced
one significant correction and one rejected experiment — but no new speed-up.

### Corrected profile (15-layer root-water, MPL = 5)

macOS `sample`, 50 326 active samples; self-time computed as
`count − Σ(child counts)` and cross-checked by a leaf-frame sum (both agree):

| Function | Self % | Note |
|---|---|---|
| `Leaf::E_from_Soil_to_Root_Collar` | **39.0 %** | lean arithmetic over 15 layers × nested root-finds (spline evals counted separately) |
| `Leaf::psi_stem_to_ci` | **35.7 %** | inner Cᵢ bisection (`leaf_model.cpp:914`); almost all self-time is the inlined `util::uniroot` + `assim_colimited` |
| `tk::spline::operator()` | 8.1 % | already O(1)-indexed (#3) |
| `pow` | 8.0 % | **fractional-exponent only** (see below) |
| `Leaf::find_root_collar_psi` | 7.0 % | master solver (Brent'd by #4) |
| `Interpolator::eval` | 3.2 % | spline wrapper |

### Correction: `psi_stem_to_ci` was mis-ranked in Phase 1

The Phase-1 #5 note claimed `psi_stem_to_ci` was "only ~2–3 %". **That was a
measurement error.** It counted the number of distinct stack *contexts* the
symbol appears in (~127–147 tree nodes) and treated that as an inclusive-time
bound. But a deep bisection loop appears in *few* stack contexts while each
carries *many* samples — the correct metric is sample weight, which here is
18 079 samples ≈ **35.7 %**. So `psi_stem_to_ci` is in fact **co-dominant with
`E_from_Soil_to_Root_Collar`, not negligible.**

It still cannot be sped **bit-identically**, for the reasons given in Phase 1:
its bisection *result* is load-bearing for FP reproducibility (same reason
TOMS748 is rejected), and there is no closed form for `A(cᵢ) = gc·(ca − cᵢ)`.

### Tried and REJECTED: `pow(x, 2) → x*x`

The 8 % `pow` self-time looked like an easy win, so the integer `pow(z, 2)` in
`assim_colimited` / `electron_transport` was replaced with `z*z`. **Rejected:**

- **No speed-up** — MPL = 5 median 6.83 s (baseline) vs 7.12 s (candidate); if
  anything slightly slower, i.e. within noise.
- **Not bit-identical** — `offspring_production` 1.64825e−06 → 1.65255e−06
  (+0.26 %); `pow(x, 2.0) ≠ x*x` on Apple libm, and the ODE amplifies the
  difference.

Profiling the `pow` *callers* explains why it was a wash: the 8 % is **entirely
fractional-exponent** pows, not the integer one — `find_root_collar_psi` (xylem
vulnerability `exp(-pow(ψ/b, c))` and hydraulic cost `pow(1 − vuln, beta2)`),
and `compute_competition` / `net_mass_production_dt` (the
`Q() = (1 − (z/h)^eta)²` canopy/root shape function). The integer `pow(z, 2)` in
`assim_colimited` was a non-bottleneck. Reverted.

### Tried and REJECTED: spline the xylem hydraulic cost

`hydraulic_cost_TF(ψ) = g1·(1 − vuln(ψ))^beta2` (where `vuln = exp(-pow(ψ/b, c))`)
is a pure function of ψ — `b, c, g1, beta2` are all constant for the run — so it
looked like the best remaining lever: pre-build it once as a spline
(`setup_hydraulic_cost`, same domain as the transpiration supply curve) and
replace the per-profit-evaluation `exp + pow + pow` with one spline eval. It was
implemented (member `hydraulic_cost_from_psi`, built in both constructors) and
**reverted**:

- **Slower, not faster** — a tightly interleaved A/B (swap pre-built `.so`,
  fresh R per run, 6 alternating pairs to cancel thermal drift) gave baseline
  median **6.91 s** vs candidate **7.26 s** — the candidate was ~5 % *slower* in
  5 / 6 pairs.
- **Also numerics-changing** — at the supply-curve resolution (100 pts)
  `offspring_production` shifted +0.56 %; a 10× finer grid cut that to −0.062 %
  (within build-to-build noise) but did **not** recover the speed.

Why it loses: `Leaf` is constructed once per *strategy* (not per call), so the
build cost is irrelevant — the regression is the **eval**. On this Apple-ARM
machine the `Interpolator::eval → tk::spline::operator()` path (wrapper call +
bounds/extrapolate check + index arithmetic + cubic eval) is *more* expensive
than the inlined `exp + pow + pow` it replaced. Spline eval is already ~11 % of
runtime; adding one more per profit evaluation costs more than the cheap,
well-optimised hardware transcendentals saved. **The spline-everything strategy
has hit diminishing returns against fast hardware transcendentals.**

### Remaining levers (all numerics-changing)

The two dominant blocks (≈75 % combined) are the **nested root-find structure**;
their algorithmic wins are already taken (Brent on the outer GSS, integral-spline
caches, transpiration memo, O(1) splines), and neither can be sped
bit-identically. With the hydraulic-cost spline now rejected, the only untried
levers left both shift numerics at the tolerance level and are speculative:

1. **Spline / cache `Q()`'s `pow(z/h, eta)`** in `compute_competition` (~4.5 % of
   the pow time; canopy-side, independent of hydraulics). Caveat: by the same
   logic as the rejected hydraulic-cost spline, a spline eval may not beat the
   `pow` it replaces — would need the same interleaved A/B to confirm.
2. **Loosen the Cᵢ-bisection tolerance** (currently `1e-7`, nested inside an outer
   `1e-3` solve — over-resolved). Cheapest potential win on the 35.7 % block, but
   explicitly changes results.

**Assessment:** the code is structurally well-optimised and has reached the point
where micro-optimising the transcendentals no longer pays (spline eval ≈ the cost
of the math it would replace). Absent appetite for tolerance-level numerics
shifts, Phase 2 stops here.

### Code-consistency follow-up: Brent for the single-layer optimisers (OFF the hot path)

A unification cleanup, **not** a production speed-up. The two R-exposed
single-layer profit optimisers — `optimise_psi_stem_Sperry` /
`optimise_psi_stem_TF` (`src/leaf_model.cpp`) — were the last hand-rolled
golden-section loops in the tree (they shared a loose file-global `gr`). They
were first factored into a shared `util::golden_section_max` in
`inst/include/plant/optimize.h`, then switched to **`util::brent_fmin`** — the
same bounded minimiser #4 put on the *outer* collar solve — so the codebase now
does 1-D profit maximisation through one routine everywhere.

> **Update:** `util::golden_section_max` was later restored to `optimize.h` (the
> efficient reuse variant) when `find_root_collar_psi` reverted to GSS for
> gradient smoothness (see the LATER REVERTED note under #4). So the codebase now
> offers *both* `brent_fmin` and `golden_section_max`: the production collar solve
> uses GSS; these off-path single-layer optimisers keep `brent_fmin` for its
> speed, since their argmax feeds no gradient.

**This does not touch `run_scm_collect`.** These two functions have **no caller
on the production path**: `TF24_Strategy::refresh_indices` calls
`find_root_collar_psi` and the `optimise_psi_stem_TF` line is commented out
(`src/tf24_strategy.cpp:381`). Their only callers are the RcppR6 bindings /
`tests/testthat/test-leaf.r`. So the SCM trajectory is **bit-identical** and the
whole-run profile is unchanged — profiling the scenario at the top of this
document would only measure build noise. The change is therefore profiled in
**isolation**, by calling `optimise_psi_stem_TF` directly.

**Measured** (`make compile` builds, `pkgload::load_all(recompile = FALSE)`;
46-point `psi_soil` sweep × 200 reps, median of 7; same Leaf params as
`test-leaf.r`):

| `GSS_tol_abs` | GSS | Brent | Speed-up |
|---|---|---|---|
| `1e-3` (production default) | 0.143 s | 0.063 s | **≈2.27×** |
| `1e-8` (test setting) | 0.327 s | 0.127 s | **≈2.57×** |

**Numerics change at the tolerance level** (same character as #4 — Brent locates
the smooth optimum more precisely than GSS's bracket-midpoint at the same tol).
At the production tol `1e-3`: max |Δ `opt_psi_stem`| ≈ 3.4e−4, max rel
|Δ `profit`| ≈ 5.7e−5; profit *at* the optimum is essentially unchanged because
the objective is flat near the max. Note the argmax still differs ~3.8e−4 even
at tol `1e-8` (while rel |Δ profit| < 6e−7): the inner Cᵢ bisection
(`ci_abs_tol`) caps how precisely the outer argmax is defined, so tightening the
outer tol below ~1e−4 does not sharpen it. Existing leaf tests pass 151/151
(their assertions are qualitative).

### Cumulative progress (baseline → end of Phase 1)

The per-item figures above were each measured against the build immediately
before them, so they don't compound cleanly. Here is the **end-to-end** result,
both binaries built clean from the same toolchain (`make clean && make compile`,
`-g -O2`) and loaded with `pkgload::load_all(".", recompile = FALSE)`, on the
**single-species MPL=10** SCM run (`birth_rate = 20`, default `Environment`,
median of 5 timed runs after a warm-up):

| Build | Commit | Median run | offspring_production |
|---|---|---|---|
| Baseline (roadmap, pre-optimisation) | `ff75ad22` | **39.78 s** | 4.712317419e−06 |
| HEAD (#1–#4 done) | `b6ec6045` | **9.05 s** | 4.797128277e−06 |

- **≈4.40× faster** (39.8 s → 9.0 s, −77 % wall-time; min-of-5 ratio 4.38×).
  This slightly beats the product of the individual claims (1.9 × 1.25 × 1.7 ≈
  4.0×) because removing the hottest spline evals also shrinks every inner
  root-find the later items still call.
- **#5 adds a further ≈1.32×** on top (15-layer, MPL = 5; see the #5 entry in
  Phase 1), so the full Phase-1 stack is ≈5–6× over baseline.
- **Numerics:** offspring_production shifts **+1.80 %** (+8.5e−08), the net of
  the two accuracy-improving items (#1 interval-mean, #4 Brent) — #2/#3/#5 are
  bit-identical. Direction is *toward* the fine-grained reference; magnitude is
  within the system's intrinsic build-to-build sensitivity. Existing
  leaf/TF24/environment tests pass unchanged.

(Reproduce with `scripts/bench_tf24.R`; absolute seconds are machine-specific
but the ratio is stable.)

## Phase 3 — structural code review (2026-06-18)

A clarity/maintainability pass over `src/leaf_model.cpp` and
`src/tf24_strategy.cpp`, tracked in
[`code_review_leaf_tf24.qmd`](code_review_leaf_tf24.qmd). **Numbering there is a
separate namespace from the performance roadmap above** (its #1–#11 are review
findings, not the Phase-1 perf items).

**Speed outcome: no gain — and none expected.** Phase 2 already concluded the
hot path is structurally optimised; this round was deliberately clarity-first,
not a performance pass. The changes that landed (review #2 reusable root-mass
buffer, #5 shared spline builder, #6 shut-down-state helper, #8 de-shadowed
locals, #9/#10 named constants, #11 naming) are all **bit-identical and
runtime-neutral by construction** — they dedupe code or rename, they do not touch
the arithmetic or the number of evals.

**One intentional output change (review #7).** The psi sign-convention cleanup
made `root_collar_psi_` (the `opt_root_psi` aux) sign-consistent across all
`find_root_collar_psi` branches. A per-column A/B confirms every *driver* (height,
profit, transpiration, `E_up_`, soil consumption, `offspring_production`) is
bit-identical; the **only** change is the `opt_root_psi` diagnostic, which becomes
a pure sign flip. So a full-trajectory digest will differ from pre-#7 builds *in
that aux alone* — by design, not a science change.

### Tried: cache the temperature/O₂ photosynthesis params (review #1)

The one item with a speed *hypothesis*: `set_physiology` recomputed ~8 Arrhenius
transcendentals (`vcmax_, jmax_, gamma_, ko_, kc_, R_d_, km_`) every call though
they depend only on `(leaf_temp_, atm_o2_kpa_)`. Cached them on that key
(`electron_transport_` excluded — it also depends on per-call `PPFD_`).

A/B on the 15-layer MPL=5 scenario (median of 5, `pkgload::load_all(recompile =
FALSE)`):

| Build | Median | offspring_production | trajectory digest |
|---|---|---|---|
| baseline (`4541acbf`) | **6.98 s** | 1.64927177539e−06 | 164392.960996 |
| + temp-param cache | **7.07 s** | 1.64927177539e−06 | 164392.960996 |

**Bit-identical, but a speed wash** (within run-to-run noise). This confirms the
Phase 2 profile: `set_physiology` is not on the hot path — those transcendentals
run once per `net_mass_production_dt`, while the nested collar solver it precedes
does hundreds of `E_from_Soil`/`psi_stem_to_ci` evals per call. **Kept anyway**
as a clarity change (it removes a genuine redundant recompute and is exact), but
recorded here so the negative speed result is not silently re-litigated.

## Phase 4 — develop merge (2026-06-19)

`develop` was merged into `root_water_uptake` (merge `2f9d5337`), bringing in
**#471 "Optimise FF16/TF24 hot paths (~2.7-3.5× FF16, ~1.3× TF24)"** plus the
SCM/patch/control refactors (#408, #459, #462, #463). The merge also renamed the
R API this vignette used: `scm_base_control()` → `control()` and
`run_scm_collect()` → `run_scm(..., collect = TRUE)`.

### Speed: negligible gain on this branch (≈1.03×)

A/B on the canonical `scripts/bench_tf24.R` benchmark (single-species MPL=10,
default `Environment`, `birth_rate = 20`, `lma = 0.0825`). Both sides built clean
(`make clean && make compile`, `-O2 -g`), loaded with
`pkgload::load_all(recompile = FALSE)`, **identical control values pinned
explicitly** (the merge's `scm_base_control` → `control` rename otherwise risks a
silent default drift; the documented fast defaults are the same on both sides),
median of 10 timed reps after a warm-up:

| Build | Commit | median | min | offspring_production |
|---|---|---|---|---|
| before develop | `5cf6694e` | 20.06 s | 20.02 s | 4.758172414e-06 |
| after develop  | `2f9d5337` | 19.45 s | 18.44 s | 4.730799193e-06 |

**≈1.03× (median, ~3 %) — essentially within run-to-run/thermal noise**, NOT the
~1.3× that #471 reports for TF24 in isolation (the post-merge build was timed
first/cooler each round, so even the 3 % likely overstates it). Two compounding
reasons:

1. **This branch's runtime is dominated by the *new* multi-layer root-water
   hydraulic solver** — `find_root_collar_psi` and its nested root-finds in
   `leaf_model.cpp` (Phase 2: `E_from_Soil_to_Root_Collar` ≈39 %, `psi_stem_to_ci`
   ≈36 % self-time). That code exists only on this branch; #471 on `develop` never
   saw it. #471's wins land on FF16-canopy / general TF24 paths that are a small
   share of per-step cost here.
2. **The branch already independently carries the Phase-1 hot-path optimisations**
   (#1-#5: O(1) spline index, transpiration memo, integral-spline caches), which
   overlap with the spline/transcendental work in #471 — so there is little left
   for #471 to win on the parts it *does* touch.

Net: the develop merge is **speed-neutral for the root-water model**; its value
here is API/refactor alignment, not throughput.

### Numerics: −0.58 % offspring shift (tolerance-level)

`offspring_production` moves 4.758172414e-06 → 4.730799193e-06 (−0.58 %), a
tolerance-level shift from #471's edits to `leaf_model.cpp` / `tk_spline.cpp` /
`interpolator.h`. Same character as the #1/#4 shifts — smooth, not structural.

### Unrelated finding: drought NaN in the vignette scenario (issue #485)

While setting up the benchmark, the 15-layer scenario at the top of this document
was found to abort with `set_physiology received non-finite psi_soil` whenever
the rainfall mean was 0.5 (`y = 0.25*sin(2*pi*x) + 0.5`). This is **not** caused
by the merge — pre-merge `5cf6694e` and post-merge `2f9d5337` fail identically,
and the current build runs the scenario fine at the original mean of 1.0
(offspring 0.39 at MPL=10). Under sustained low rainfall a soil layer dries to ~0
moisture, where `psi_from_soil_moist` (a `pow` with negative exponent) goes
non-finite; the bucket model has no residual-moisture floor. Filed as
traitecoevo/plant#485; the vignette rainfall has been reverted to mean 1.0.

## Phase 5 — reprofile on current HEAD (2026-06-19)

A fresh baseline + profile on the post-merge tip (`3611fca5`), to refresh the
optimisation targets after Phases 1–4. It produced an updated hot-spot ranking,
one explained ranking shift (not a regression), and one **rejected** lever — but
no new speed-up, confirming Phases 2–3's "structurally optimised" verdict.

### New baseline (15-layer root-water, MPL = 5)

Clean `make clean && make compile` (`-O2 -g`), loaded with
`pkgload::load_all(".", recompile = FALSE)`, median of 7 timed reps on the
canonical 15-layer scenario at the top of this document:

| Metric | Value |
|---|---|
| Median run | **≈11.3 s** (min 11.26, single-warm-session range 11.26–11.74) |
| `offspring_production` | **1.651178375e-06** |

### Refreshed profile

macOS `sample` (35 s, the scenario looped), ≈28.8 k C++-active samples; self-time
via `count − Σ(child counts)`, cross-checked by leaf-frame sum:

| Function | Self % | Note |
|---|---|---|
| `Leaf::psi_stem_to_ci` | **44.1 %** | inner Cᵢ bisection (`leaf_model.cpp:1005`, `util::uniroot` → `assim_colimited`) |
| `Leaf::E_from_Soil_to_Root_Collar` | **30.8 %** | lean arithmetic over 15 layers × nested root-finds |
| `tk::spline::operator()` | 8.4 % | already O(1)-indexed (#3) |
| `pow` | 6.5 % | fractional-exponent only (xylem/root vuln, `Q()` shape) |
| `Leaf::find_root_collar_psi` | 3.4 % | master solver (now GSS — see below) |
| `Leaf::set_leaf_states_rates_from_psi_stem` | 1.9 % | profit-eval wrapper around `psi_stem_to_ci` |
| `Interpolator::eval` | 1.3 % | spline wrapper |
| `exp` | 0.9 % | |
| `Leaf::find_root_psi` | 0.5 % | |

### Ranking flip vs Phase 2 is the GSS revert — NOT a regression

Two things changed against the Phase-2 profile: `psi_stem_to_ci` overtook
`E_from_Soil_to_Root_Collar` as the single leader (44 % vs Phase-2's co-dominant
36 %/39 %), and wall-time on this scenario rose from the ~7 s of Phases 1–3 to
~11.3 s (≈1.6×). Both are the **documented `find_root_collar_psi` Brent→GSS
revert** (commit `3129ebb5` "Use golden_section_max for the collar solver", the
*LATER REVERTED* note under roadmap #4), which landed **after** the Phase-3
baseline `4541acbf`. GSS does ~17 outer profit iterations where Brent did ~10,
and each iteration carries an inner `find_psi_stem` root-find that drives
`psi_stem_to_ci` — so reverting to GSS both **raises wall-time** and **inflates
`psi_stem_to_ci`'s share**. The trajectory is otherwise unchanged:
`offspring_production` 1.651178375e-06 here vs 1.64927e-06 at `4541acbf` (+0.1 %),
confirming this is the same scenario at a slower-but-smoother solver, not a
different/regressed one. The ~7 s figures in Phases 1–3 belong to the Brent build
and **no longer apply** — this ≈11.3 s is the accepted cost of the smooth collar
argmax the demographic gradient needs.

### Tried and REJECTED: loosen the Cᵢ-bisection tolerance

Phase 2's remaining-lever #2 — the inner Cᵢ bisection runs to `1e-7`
(`leaf_model.cpp:1005`) nested inside the outer `1e-3` collar solve, so it looked
over-resolved, and it is now the dominant 44 % block. **Rejected:** loosening it
even one decade destabilises the coupled soil-water ODE.

- `1e-7 → 1e-5`: aborts `set_physiology received non-finite psi_soil at layer=4`.
- `1e-7 → 1e-6`: aborts `… non-finite psi_soil at layer=13`.

The looser inner solve perturbs the trajectory just enough to drive a soil layer
to ~0 moisture, where `psi_from_soil_moist` (a `pow` with negative exponent) goes
non-finite — the same drought singularity as issue #485, here tripped at the
otherwise-stable mean-1.0 rainfall purely by the tolerance change. So the `1e-7`
Cᵢ tolerance is **load-bearing for coupled-ODE stability**, exactly the lesson
of the rejected TOMS748 root-finder (Phase 1): bisection precision on this leaf
path is not slack to be reclaimed. **Lever closed.**

### Assessment

The other Phase-2 lever (spline `Q()`'s `pow` in `compute_competition`) is
confirmed negligible here — `compute_competition` is ~0.1 % self. With the
Cᵢ-tolerance lever now rejected, **no safe remaining lever exists**: ~75 % of
runtime is the nested root-find structure, every bit-identical win is already
taken (O(1) spline index #3, transpiration memo #2, integral-spline caches #1/#5,
gamma-seeded knots #468, Brent on the off-path single-layer optimisers), and the
one numerics-changing lever left is unsafe. Phase 5 delivers a refreshed baseline
and profile and closes the last flagged target; it finds no new speed-up,
consistent with the Phase 2–3 conclusion that the hot path is structurally
optimised.

## Phase 6 — Brent on `psi_stem_to_ci` only (2026-06-19)

Phases 1–5 concluded "no safe remaining lever". **That was too pessimistic for
`psi_stem_to_ci`** — it conflated two different things the earlier rejections
lumped together:

- **Loosening the Cᵢ tolerance** (Phase 2 lever, Phase 5 rejected) changes the
  *answer*: a looser solve returns a different, less-accurate Cᵢ, which perturbs
  the coupled ODE into the #485 drought singularity. Unsafe.
- **Swapping the solver** (TOMS748 for bisection) at the *same* `1e-7` tolerance
  returns the *same root in fewer evaluations*. It does not loosen anything.

The Phase-1 TOMS748 rejection was a **global** swap across *both* nested
root-finds, and it failed on the hydraulic `find_root_psi` path — whose target is
genuinely non-smooth (vulnerability-curve clamps, splines with extrapolation
disabled, near-flat regions). But the `psi_stem_to_ci` target is different:

$$f(c_i) = A_\text{colim}(c_i)\,\kappa_1 - g_c\,(c_a - c_i)\,\kappa_2$$

is **smooth and strictly monotone** over $(\Gamma^*, c_a]$. There is *no*
singularity at the lower bracket: near $\Gamma^*$ both Rubisco- and
electron-limited rates vanish *linearly*, so the colimitation discriminant
$\sim(c_i-\Gamma^*)^2$ and $\sqrt{\cdot}$ is linear (finite derivative). Its only
sharp feature is the colimitation elbow ($c_i\approx4.3$), far below the
operating root ($c_i\approx12$). This is a textbook case for a superlinear
*bracketing* solver. Measured eval counts: bisection needs **~29**
`assim_colimited` evaluations to reach `1e-7` over the 35.7 Pa bracket; TOMS748
reaches the same root in **~9**.

**Change.** `util::uniroot → util::uniroot_smooth` (boost TOMS748) in
`Leaf::psi_stem_to_ci` *only*; `find_root_psi` keeps bisection. One guard was
required: TOMS748 validates its bracket and *throws* on non-finite endpoints,
whereas `bisect` silently returned NaN. So a `psi_stem = NA` call (e.g.
`profit_psi_stem_TF(NA, ·)`) now needs an explicit
`if (!std::isfinite(gc)) return ci_ = NA_REAL;` to preserve the NA-in→NA-out
contract (a unit test in `test-leaf.r` caught this).

**Result (clean sequential A/B, both `make compile` `-g -O2`, deterministic,
15-layer MPL = 5, mean-1.0 rainfall):**

| build | median time | offspring |
|---|---|---|
| bisection (baseline) | 19.08 s | 1.647796754e-06 |
| TOMS748 (`psi_stem_to_ci`) | **10.75 s** | 1.649221246e-06 |

**≈1.78× faster, +0.086 % offspring shift.** The speed-up *exceeds* the naive
44%-self-time Amdahl bound (1.43×) because that 44% was psi_stem_to_ci
*self*-time only: cutting ~29→~9 evals also removes ~20 `assim_colimited` /
`sqrt` / `pow` calls per solve whose cost was attributed to *those* symbols, so
the inclusive saving is larger.

**Numerics — safe, unlike the rejected levers.** It is *not* bit-identical (no
solver swap can be — bisection returns an interval midpoint, TOMS748 the
interpolated root), but it converges to the *same* `1e-7` tolerance, so the shift
is tolerance-level and smooth (same cohort/step structure). Critically, the
**drought stability gate passes**: the mean-0.5 regime completes with finite
output (offspring 2.11e-10), **no NaN** — the failure mode that killed the
tolerance-loosening lever does not appear, because the answer is the same, just
reached faster. Full test suite: 1912 pass / 0 fail.

The Cᵢ tolerance is still load-bearing (Phase 5 stands — don't *loosen* it); the
lever Phases 1–5 missed is converging to it *faster*. This is the first
post-Phase-1 hot-path speed-up.

## Phase 7 — inline `Interpolator::eval` into the header (2026-06-19)

A portable, **bit-identical** code-move on top of Phase 6 (commit `3593929c`,
"Inline `Interpolator::eval` into header"). The leaf root-finders call the spline
once per iteration (`Leaf::transpiration` / `E_from_Soil_to_Root_Collar` →
`.eval()`), but `Interpolator::eval` (and `check_active`) were defined
out-of-line in `src/interpolator.cpp`, so the cross-TU call barred the
already-header-inline `tk_spline(u)` from folding into those hot loops. Moving
`eval`/`check_active` into `inst/include/plant/interpolator.h` as `inline` (the
cold `util::stop` error paths still reach the out-of-line symbol) lets them
inline. This was the win an LTO investigation (#470) was chasing — LTO is a
no-op via `UseLTO` on a stock R toolchain and gave only ~3 % even when forced
with `-flto`; the code-move captures it portably.

### Cumulative timing (15-layer root-water, MPL = 5)

Same-session A/B, clean `make clean && make compile` (`-g -O2`) on each side,
`pkgload::load_all(".", recompile = FALSE)`, median of 7 timed reps. The "before"
side is the commit's parent `3f8e17f3` — a docs-only change, so its binary is the
**Phase 6 TOMS748 build** (the C++ diff vs HEAD is *only* `interpolator.h` /
`interpolator.cpp`):

| Build | Commit | Median | offspring_production | Speed-up |
|---|---|---|---|---|
| Phase 6 (TOMS748 `psi_stem_to_ci`) | `3f8e17f3` | 5.225 s | 1.651052959e-06 | — |
| + `Interpolator::eval` inline | `3593929c` (HEAD) | 5.161 s | 1.651052959e-06 | **≈1.01× (~1.2 %)** |

**Bit-identical** (offspring unchanged to 10 sig-figs; min-of-7 5.195 s → 5.115 s,
~1.6 %). The gain is smaller here than the ~5.5 % the commit measures on
`bench_tf24.R` (default `Environment`, MPL=10): on this hydraulics-dominated
scenario the spline path is only ~8–10 % of runtime (Phase 5 profile:
`tk::spline` 8.4 % + `Interpolator::eval` 1.3 %), whereas the default-`Environment`
bench is dominated by the xylem/transpiration splines the inline most helps. The
change removes a call barrier around the spline eval; it does not touch the
nested-root-find structure that is ~75 % of cost here.

**Note on absolute seconds — different session, not a regression of the ~11 s
Phase 5/6 figures.** This A/B ran the same scenario+build at ~5.2 s where Phases
5–6 measured ~11 s (a ~2× cross-session/machine-state difference). Only the
**same-session ratio** is quoted, exactly as the build note below warns: absolute
seconds are machine-specific, comparisons are valid only within one session built
both-sides the same way.

## Phase 8 — Brent on `find_root_psi` too (2026-06-19)

Phase 6 swapped the *inner* `psi_stem_to_ci` solve to TOMS748 and explicitly
kept bisection on the hydraulic `find_root_psi` path, on the Phase-1 belief that
*that* target is "genuinely non-smooth (vulnerability-curve clamps, splines with
extrapolation disabled, near-flat regions)". A refreshed self-time profile then
made `Leaf::E_from_Soil_to_Root_Collar` **the** dominant hot-spot (~46 %, up from
30.8 %; `psi_stem_to_ci` fell to ~14 %). `E_from_Soil` is per-layer arithmetic
over the 15 soil layers, evaluated inside `find_root_psi` (bisection, ~16 iters)
which is itself inside the golden-section collar optimisation — **~270
`E_from_Soil` calls per collar solve**. Cutting `find_root_psi`'s iteration count
attacks that 46 % directly.

The Phase-1 rejection turns out to have been the same over-broad call it was for
`psi_stem_to_ci`: it conflated the two nested root-finds in *one* global swap. So
the `find_root_psi` target was characterised on its own (densely sampled in R,
both `find_root_crit` modes, six soil scenarios):

- **Smooth and strictly monotone** over the bracket `[-psi_crit,
  wettest_soil_layer]` in the normal operating regime: a clean single
  sign-change, and derivatives **continuous across every `x == psi_soil[i]`
  layer crossing** — the per-layer branch switches in `E_from_Soil` are bit-level
  kinks (relative slope jump ~1e-6), not real corners.
- **Eval counts** (to the existing `1e-4` tol): bisection **15–16** vs TOMS748
  **6–8** `E_from_Soil` evals per solve — a ~2× cut.
- The genuine non-smoothness *is* real but lives **in `E_from_Soil` itself**: for
  soil drier than the root vulnerability spline domain (~`root_psi_crit`) the
  spline extrapolates to a *negative* conductivity and `E_from_Soil` `util::stop`s
  — identically under *both* solvers. Crucially it does **not** arise on the
  brackets `find_root_collar_psi` actually hands `find_root_psi`: the crit=1 lower
  endpoint is exactly the `E_column(-psi_crit) < 0` shut-down test, and crit=0 is
  only reached when soil is wetter than `psi_crit`. Both brackets are therefore
  guaranteed opposite-sign and finite.

**Change.** `util::uniroot → util::uniroot_smooth` in `Leaf::find_root_psi` for
*both* `find_root_crit` modes, at the *same* `1e-4` tolerance (same root, fewer
evals — a method swap, not a tolerance loosening). No NaN-bracket guard was
needed (unlike `psi_stem_to_ci` Phase 6): the brackets are non-degenerate by the
early-exits above, and `E_from_Soil` `util::stop`s rather than returning a silent
NaN, so the existing `try/catch` already surfaces any genuine failure.

**Result (clean sequential A/B, both `make compile` `-g -O2`, deterministic,
15-layer MPL = 5, mean-1.0 rainfall, median of timed reps):**

| Build | Median | offspring_production | Speed-up |
|---|---|---|---|
| bisection `find_root_psi` (Phase 6/7) | 5.037 s | 1.6510529588e-06 | — |
| TOMS748 `find_root_psi` (this) | **4.148 s** | 1.64950903698e-06 | **≈1.21× (~18 %)** |

**Numerics — safe.** Not bit-identical (no solver swap can be), but converges to
the same `1e-4` tolerance, so the shift is tolerance-level and smooth:
**−0.094 %** offspring. The **drought-stability gate passes**: the mean-0.5
regime completes with finite output (offspring 2.068e-10 vs the bisection
baseline's 2.114e-10, a −2.1 % shift on a near-knife-edge value), **no NaN** —
the failure mode that killed the tolerance-loosening lever (#485) does not
appear, because the answer is the same, just reached faster. Full test suite:
**1943 pass / 0 fail** (incl. a new method-independent `find_root_psi` contract
block in `test-leaf.r`).

This is the second post-Phase-1 hot-path speed-up, and it confirms that the
remaining lever on `E_from_Soil` is its **per-layer arithmetic** (lever 2), not
the root-finder — both nested solvers are now superlinear.

## Phase 9 — `E_from_Soil` per-layer arithmetic (2026-06-19)

With both nested solvers superlinear (Phases 6, 8), `E_from_Soil_to_Root_Collar`
is the lone dominant hot-spot: a fresh `sample` profile (idle threads excluded)
puts it at **~31 % of compute self-time**, ahead of `tk::spline` (11.6 %), the
`psi_stem_to_ci` TOMS748 brackets (9.9 %) and `pow` (7.4 %). It is called ~33×
per collar solve, each looping all 15 soil layers.

**Diagnosis — divisions, not guards.** The line-level self-time pointed at the
`std::isfinite` guard lines, but the disassembly showed `isfinite` is **inlined
as integer bit-ops** (the `R_finite` libcall stubs come from `set_physiology`,
not the loop). The guard lines are hot because each *consumes a freshly-divided
value*, and `fdiv` (~10–15-cycle latency on Apple Silicon) smears its stall onto
the dependent instruction. The loop did **4 divisions per layer**
(`f_r_average`, `r_R_H`, `/r_R`, `/area_leaf_`).

**Change 1 — halve the divisions (the speed win, commit `dfc9e723`).** 4 → 2
div/layer: fold the call-constant `1/area_leaf_` into a multiply, and merge
`f_r_average = integral/span` + `r_R_H = r_R_H_min/f_r_average` into
`r_R_H = r_R_H_min·span/integral` (one division; since `span>0`, the
`f_r_average>0` guard is exactly `integral>0`).

**Change 2 — precompute grav head + prune drought-era guards (cleanup, commit
`8074c536`).** Cache `gravity_head·z_soil_mid_[i]` in `grav_head_z_` (fixed
across the solve, used 3×/layer), and remove 6 per-layer defensive guards left
over from the #485 drought-NaN debugging (now fixed at source by the soil
residual-moisture floor). Two guards are **kept** because the post-loop
`isfinite(E_up_)` net cannot replace them: the equal-branch `f_ri ≤ 0` check
(`root_vuln_from_psi` extrapolates *negative* beyond its domain → negative-but-
*finite* `r_R`) and the post-loop aggregate `isfinite(E_up_)` itself. The
removed ones are provably finite on the valid path (`psi_soil` validated in
`set_physiology`; `P_src_min ≤ P_src_max` by construction; the integral comes
from a monotone-increasing spline so `integral > 0`, giving `r_R > 0`) and any
stray NaN/Inf still reaches the post-loop net.

**Result (clean sequential A/B, both `make compile` `-g -O2`, 15-layer MPL = 5,
mean-1.0 rainfall):**

| Build | Median | offspring_production | E_from_Soil self-time |
|---|---|---|---|
| Phase 8 (bisection→Brent both solvers) | 4.41 s | 1.64950903698e-06 | 30.8 % |
| + division reduction | 4.08 s | 1.64884081609e-06 | — |
| + grav head + guard cleanup | **4.08 s** | 1.64840041329e-06 | **25.2 %** |

**≈1.08× (~7.4 %)**, E_from_Soil self-time **30.8 % → 25.2 %**. Two honest
findings: (1) the **guard removal added ~0.1 %** — confirming the hot guard lines
were division-latency smear, so it is a maintainability change, not a speed one;
the division reduction is essentially the entire gain. (2) **Not bit-identical**
(tolerance-level, like #468 / Phases 6, 8): folding `/area → ·inv_area` and
`/f_avg → ·span/integral` shifts the last ULP, and the precomputed grav head is
rounded before the subtraction where the old inline product was FMA-contracted
into it. Offspring shift **−0.067 %**, smooth. **Drought gate (mean-0.5)** finite
(2.069e-10), **no NaN**. Full suite **1943 pass / 0 fail**.

## Analytical hydraulic transform (issue #468)

[Issue #468](https://github.com/traitecoevo/plant/issues/468) noted that the
Weibull hydraulic transform has a closed form involving the incomplete gamma
function, and asked (1) whether it matches the current numerical implementation,
(2) how large the spline interpolation error is, and (3) whether a monotone
interpolator (PCHIP) would help. This section answers (1) and (2) and resolves
(3); the runtime question was already settled in Phase 2 (the spline path is the
right choice for forward simulation — see *Tried and REJECTED: spline the xylem
hydraulic cost*).

### The closed form

The supply-side transpiration integrates the xylem vulnerability curve
`proportion_of_conductivity(m) = exp(-(m/b)^c)` (where `m = -psi` is the
potential magnitude in MPa), giving the hydraulic transform

$$F(m)=\int_0^m \exp\!\big[-(s/b)^c\big]\,ds .$$

Substituting $u=(s/b)^c$ yields a **lower** incomplete gamma:

$$F(m)=\frac{b}{c}\,\gamma\!\left(\tfrac1c,\,(m/b)^c\right)
      =\frac{b}{c}\,\Gamma\!\left(\tfrac1c\right)\,
       P\!\left(\tfrac1c,\,(m/b)^c\right),$$

i.e. in R, `F = (b/c) * gamma(1/c) * pgamma((m/b)^c, 1/c)`. This is identical to
the issue's $P_{50}$-parameterised, *upper*-incomplete-gamma form: the code sets
`b = p_50 / (ln 2)^(1/c)`, so `ln2·(−ψ/p_50)^c = (m/b)^c`, and lower-vs-upper
differ only by the additive constant $(b/c)\Gamma(1/c)$, which cancels in
$F(\psi_2)-F(\psi_1)$.

### Verification and error (reproduce with `scripts/validate_gamma_transform.R`)

The script builds the real C++ `Leaf` and compares the closed form against both
numerical paths — `transpiration()` (the production `tk::spline` of $F$) and
`transpiration_full_integration()` (QAG adaptive quadrature) — plus the inverse
spline `psi_from_transpiration`. The table shows the trapezoid-knot build
(*before*) and the analytic-knot build now in the code (*after*, see
*Implemented* below); QAG is build-independent. Over a 2001-point sweep:

| Parameter set (`ncontrol`) | QAG (rel) | spline $F$ before (abs) | spline $F$ after (abs) | inverse before | inverse after |
|---|---|---|---|---|---|
| TF24 default `p_50=1.85` (100) | 1.4e−6 | 2.2e−4 | 1.5e−4 | 2.0e−2 MPa | 2.7e−4 MPa |
| test-leaf `c=2.04,b=3` (100)   | 1.7e−10 | 9.7e−5 | 3.2e−6 | 8.1e−4 MPa | 3.2e−4 MPa |
| TF24 default (1000)            | 1.4e−6 | 1.8e−6 | 9.5e−7 | 1.8e−4 MPa | 2.4e−6 MPa |
| TF24 default (30)              | 1.3e−6 | 2.7e−3 | 1.8e−3 | 2.1e−1 MPa | 3.4e−3 MPa |

**(1) The closed form is correct.** QAG quadrature — an independent numerical
method — agrees with it to ~1e−6 relative (its own tolerance floor), and to
~1e−10 for the steeper test curve. The analytical transform can stand as a
validation reference.

**(2) The spline error has *two* parts; only one is quadrature bias.** The build
splits into (a) knot error — the trapezoid quadrature bias baked into the
construction values — and (b) cubic interpolation error *between* knots. Seeding
the knots analytically (below) drives (a) to machine epsilon (knot error
2.2e−4 → 3e−15) and isolates (b). The earlier read of this section — "99.7%
trapezoid, 0.3% cubic" — was **wrong**: it subtracted the max knot error from the
max total error, but the two maxima occur at *different* potentials. For the
production curve ($c=1.09$) the residual after exact seeding is **1.46e−4, all of
it in the first inter-knot cell** ($\psi\approx0.04$, before the first knot at
0.105). There $F$ has unbounded higher derivatives at the origin
($F'''\sim m^{c-2}$ diverges for $c<2$), so a single cubic cell cannot track it —
this is genuine interpolation error, not quadrature. The trapezoid bias, large at
*high* $\psi$, dominated the *before* total and masked it. For the steeper
test-leaf curve ($c=2.04$, no origin singularity) seeding does help in full:
9.7e−5 → 3.2e−6 (~30×).

### Implemented

`setup_transpiration` and `setup_root_vulnerability` now seed their cumulative-
integral knots from the gamma closed form (`boost::math::tgamma_lower`) instead of
a running trapezoid sum — same knots, same `tk::spline`, same O(1) hot-path eval.
Measured effect (table above):

- **Inverse transform (used in simulation): ~60–75× more accurate.** `p_50=1.85`,
  `ncontrol=100`: 2.0e−2 → 2.7e−4 MPa. The inverse's error *was* bias-dominated,
  so removing the bias is almost the whole story — this is the real payoff.
- **Forward $F$: modest for the production curve (~1.5×, 2.2e−4 → 1.5e−4)** because
  its error is the near-origin interpolation term (b), not bias; large (~30×) for
  steep curves with no origin singularity.
- **Trajectory shift** (MPL=5 self-comparison, `run_scm_collect`): not
  bit-identical — by design. Hydraulics-driven states move ~0.2–0.3%
  (`transpiration`/`E_up_`/`opt_psi_stem`/`height` rel ≤3e−3), `offspring_production`
  ~1e−3; the shift is smooth (large relative diffs appear only at near-zero-density
  cohort boundaries and sign crossings), with identical cohort/step structure. The
  move is toward the more accurate integral.

### Consequences

- **(3) PCHIP would not help.** The production residual is the first-cell origin
  singularity in $F$ ($F'''$ diverges for $c<2$) — a *lower*-order monotone
  interpolator would do worse there, not better, and monotonicity/overshoot was
  never the issue ($F$, $F^{-1}$ are smooth and the cubic does not overshoot). To
  shave the residual further the lever is a non-uniform knot grid (denser near
  $\psi=0$), not a different interpolator. Out of scope.
- **Calibration / AD (issue step 5)** remains the closed form's other natural
  home: it is a smooth expression in `(p_50, c, K_max)` with no spline-build
  graph to differentiate through. Out of scope here.

## Logistic vulnerability curve (issue #467)

[Issue #467](https://github.com/traitecoevo/plant/issues/467) asks whether a
**logistic** vulnerability curve could replace the Weibull — the argument being
that it is "slightly less accurate biologically but substantially easier
computationally": an elementary integral (no incomplete gamma), an explicit
inverse, and smooth, AD-friendly derivatives. This section evaluates that for
*plant* specifically. Reproduce with `scripts/compare_logistic_weibull.R` (pure
R — both curves are closed form, no build needed).

### The two curves

With magnitude $m=-\psi\ge0$ (matching the `+ve psi` argument of
`proportion_of_conductivity`),

$$\underbrace{f(m)=\exp\!\big[-(m/b)^c\big]}_{\text{Weibull (current)}}
  \qquad
  \underbrace{g(m)=\frac{1}{1+\exp[a\,(m-m_{50})]}}_{\text{logistic (proposed)}}.$$

The logistic forward transform $F(m)=\int_0^m g\,ds$ **is** elementary,

$$F(m)=m-\tfrac1a\ln\!\big(1+e^{a(m-m_{50})}\big)
        +\tfrac1a\ln\!\big(1+e^{-a\,m_{50}}\big),$$

so it needs no incomplete gamma. **But the issue's "explicit inverse" claim does
not hold**: $m$ appears both linearly and inside the log, so $F^{-1}$ is
transcendental (and ill-conditioned near saturation, where $F$ flattens). $F$ and
$F'$ are cheap, so the inverse would still be a Newton iterate or a spline-seed —
exactly the machinery the Weibull path already has (issue #468). The logistic
removes the incomplete gamma at *build* time and one fractional `pow` from the
hot `proportion_of_conductivity`; it does **not** remove the need to invert.

### How well does a logistic match plant's Weibull?

Fitting a logistic to each production curve two ways — slope-matched at the 50%
point ($m_{50}=p_{50}$, $a=2c\ln2/p_{50}$) and least-squares over the working
domain $[0,\psi_\text{crit}]$ — and measuring both the conductivity deviation and
the deviation of the model-relevant transform $F$ (relative to the full integral
$F(\psi_\text{crit})$):

| Curve | fit | $a$ | $m_{50}$ | cond max-dev | cond RMS | **$F$ rel-dev** |
|---|---|---|---|---|---|---|
| Stem ($c=1.09$) | slope-match | 0.817 | 1.85 | 0.181 | 0.056 | **0.135** |
| Stem ($c=1.09$) | best-fit LS | 0.782 | 2.10 | 0.162 | 0.045 | **0.042** |
| Root ($c=2.68$) | slope-match | 1.093 | 3.40 | 0.042 | 0.022 | **0.024** |
| Root ($c=2.68$) | best-fit LS | 1.191 | 3.43 | 0.028 | 0.015 | **0.012** |

**The fit quality is entirely controlled by the Weibull shape $c$.** The root
curve ($c=2.68$) is itself sigmoidal — its inflection sits away from the origin —
so a logistic tracks it to ~1–2% in $F$. The production *stem* curve
($c\approx1.09$) is not: for $c<2$ the Weibull is concave-down from the origin
(heavy conductivity loss at low stress) with a fat tail, a shape a symmetric
sigmoid cannot reproduce. The best a logistic manages is a **4% bias in the
transpiration integral** (13% if you also insist on preserving $p_{50}$ and the
50%-point slope) — and it buys that only by displacing $m_{50}$ to 2.10, i.e. by
**abandoning the interpretable $P_{50}$** that was the logistic's main selling
point. The tails diverge in the obvious way: at $m=5$ MPa the slope-matched
logistic gives half the Weibull conductivity (0.071 vs 0.129).

### Recommendation: not worth adopting for plant's core

- **Accuracy.** For the production parameterisation the logistic biases the
  transpiration integral by 4–13% — one to two orders of magnitude larger than
  the spline/quadrature errors quantified for #468 (≤1.5e−4). This is a model
  change, not a numerical refinement, and not bit-identical by construction.
- **Performance.** The win is small and already largely banked elsewhere. The
  forward and inverse transforms are O(1) spline lookups (#468), and *Tried and
  REJECTED: spline the xylem hydraulic cost* showed a spline eval already costs
  ~the transcendental it replaces on Apple-ARM. The logistic's only hot-path
  saving is one fractional `pow` inside `proportion_of_conductivity`/
  `hydraulic_cost_TF` — a slice of the ~8% `pow` budget that is *shared* with the
  unrelated $Q(z,h)$ canopy-shape `pow`. Net: a few percent at most, against a
  4–13% biology shift.
- **Where it does belong.** The issue's real motivation — smooth, special-
  function-free, AD-friendly curves for Bayesian calibration and gradient-based
  fitting — is genuine, and there the logistic is attractive. But that is an
  *external fitting/inference* concern, the same niche flagged for the closed-form
  Weibull transform (#468 *Consequences*), not the SCM forward-simulation core
  this branch optimises.

**Verdict.** Keep the Weibull for the stem. A logistic is a defensible surrogate
only for steep curves ($c\gtrsim2$, e.g. the root), and even there the payoff is
marginal given the transforms are already splined. Revisit only if *plant* grows
an AD/calibration path where special-function-free curves matter more than the
near-origin fit.

## Build note: build once with `make`, load with `recompile = FALSE`

Build the package with **`make compile`** (`pkgbuild::compile_dll(debug = FALSE)`),
then load the built binary without recompiling:
`pkgload::load_all(".", recompile = FALSE)`.

**On the flags (correcting an earlier claim here that `make` is
`-O3 -march=native`).** It is not. Because the package sets `CXX_STD = CXX20`,
C++ sources compile with R's **`CXX20FLAGS`** — here `-g -O2` — and *not* with
`CXXFLAGS`, so a `~/.R/Makevars` `CXXFLAGS = -O3 -march=native` line does **not**
apply to this package. `make compile` and a plain `devtools::load_all()`
therefore use the *same* `-g -O2`; with identical flags and compiler the two
builds are deterministic and match bit-for-bit.

The reason to prefer `make compile` + `recompile = FALSE` is thus **not**
different optimisation — it is **iteration speed and consistency**: you build the
`.so` once and benchmark the exact same binary across A/B runs, instead of
`load_all()` recompiling on every load (and risking accidental flag drift). For a
valid speed/numerics comparison, always build *both* sides the same way.