---
title: "Initialising a patch"
---

<!--
STAGING COPY. This guide is written for the Overstorey site
(https://github.com/traitecoevo/overstorey) but is kept in the plant repo for
now because it documents the `init-from-nodes` branch (reviving PR #304), which
is not yet merged. When that branch lands, move this file to
`overstorey/guides/initialising-a-patch.qmd`, add it to the Guides sidebar in
`_quarto.yml`, and delete the `overstorey-staging/` directory here.
-->

::: {.eyebrow}
guide
:::

By default a `plant` patch starts empty and grows from seed: nodes (cohorts) are
introduced on a schedule and integrated forward as characteristics of the
[size-structured PDE](../theory/size-structured-pde.qmd). Sometimes you instead
want a patch to *start* with plants already in it — either to **resume** a run
you stopped earlier, or to **seed** a patch with an arbitrary initial size
distribution and watch it evolve.

This guide covers both. The mechanism is the same in each case: an initial state
is attached to the `Parameters` object and applied when the patch is reset, so an
ordinary `run_scm(p)` picks it up with no special arguments.

```{r}
#| label: setup
library(plant)
library(dplyr)
library(ggplot2)

p0 <- scm_base_parameters("FF16")
p  <- expand_parameters(trait_matrix(0.08, "lma"), p0, birth_rate_list = 1.0)
```

## Exporting and resuming a run

Run a patch, collecting a snapshot of the whole population after each step, then
export the state part-way through:

```{r}
#| label: run-and-export
scm <- SCM("FF16", "FF16_Env")(p, Environment("FF16"), Control())
scm$collect <- TRUE
scm$run()

# export the patch about half-way through the run
k <- floor(length(scm$history) / 2)
state <- export_patch_state(scm, step = k)
```

`export_patch_state()` captures everything needed to continue the run: every
node's ODE state, the per-node *birth bookkeeping* (introduction time, patch-age
density and survival probability at birth — quantities that are not part of the
ODE state but feed the growth/mortality rates and the lifetime-fitness
integrals), the patch age, and the portion of the introduction schedule that has
not happened yet.

`set_initial_state()` writes that onto a `Parameters` object, and `run_scm()`
resumes from it:

```{r}
#| label: resume
p_resumed <- set_initial_state(p, state)
resumed   <- run_scm(p_resumed, collect = TRUE)
```

The resumed run reproduces the original. The seed is loaded bit-for-bit; only the
adaptive ODE stepper, whose internal step size is not part of the checkpoint,
diverges slightly going forward — so the recovered fitness matches to a few parts
in $10^5$ rather than exactly:

```{r}
#| label: compare-fitness
c(original = scm$net_reproduction_ratios,
  resumed  = resumed$net_reproduction_ratios)
```

Because the state lives on the `Parameters` object, the resumed run is fully
self-describing: you could `saveRDS(p_resumed)` and reproduce the continuation in
a fresh session.

## Seeding an arbitrary size distribution

To start a patch from scratch with a chosen distribution of plants, build an
initial state from node heights and densities with `make_initial_state()`. The
remaining state (mortality, fecundity, accumulated reproduction, heartwood) starts
at zero, every node is introduced at patch age 0, and recruitment continues
normally for $t > 0$ (the seeded distribution stands in for the $t = 0$ recruit).

```{r}
#| label: seed-distribution
heights <- seq(0.5, 8, length.out = 12)
state0  <- make_initial_state(p, heights = heights, densities = rep(0.2, 12))

p_seeded <- set_initial_state(p, state0)
seeded   <- run_scm(p_seeded, collect = TRUE)
```

The patch begins with the twelve plants we specified and then develops under
competition and ongoing recruitment:

```{r}
#| label: plot-trajectories
#| fig-alt: "Height of each node over patch age, starting from twelve seeded plants."
seeded$species |>
  ggplot(aes(time, height, group = node)) +
  geom_line(alpha = 0.5) +
  labs(x = "Patch age (years)", y = "Height (m)") +
  theme_classic()
```

The twelve tall lines starting at $t = 0$ are the seeded plants; the lines that
appear later are recruits arriving on the introduction schedule.

## Plausible initial conditions

Very large plants at very high density are not a state the model can integrate:
they self-shade so heavily that net production goes negative, mortality blows up,
and node densities race to non-finite values within a few steps. `plant` checks
for this when the patch is seeded and refuses rather than producing garbage:

```{r}
#| label: guard
#| error: true
bad <- make_initial_state(p, heights = seq(8, 12, length.out = 20),
                          densities = rep(50, 20))
run_scm(set_initial_state(p, bad))
```

If you hit this, reduce the sizes and/or densities of the initial plants — a
distribution the solver itself could have produced (for example, one obtained via
`export_patch_state()`) is always safe to re-import.
