---
title: "Initialising a patch"
---

<!--
STAGING COPY. This guide is written for the Overstorey site
(https://github.com/traitecoevo/overstorey) but is kept in the plant repo for
now because it documents the `init-from-nodes` branch (reviving PR #304), which
is not yet merged. When that branch lands, move this file to
`overstorey/guides/initialising-a-patch.qmd`, add it to the Guides sidebar in
`_quarto.yml`, and delete the `overstorey-staging/` directory here. All figures
are generated by the code chunks below (the site uses `freeze`), so nothing
static needs to move with it.
-->

::: {.eyebrow}
guide
:::

By default a `plant` patch starts empty and grows from seed: nodes (cohorts) are
introduced on a schedule and integrated forward as characteristics of the
[size-structured PDE](../theory/size-structured-pde.qmd). Sometimes you instead
want a patch to *start* with plants already in it — either to **resume** a run
you stopped earlier, or to **seed** a patch with an arbitrary initial size
distribution and watch it evolve.

This guide covers both. The mechanism is the same in each case: an initial state
is attached to the `Parameters` object and applied when the patch is reset, so an
ordinary `run_scm(p)` picks it up with no special arguments.

```{r}
#| label: setup
library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)

p0 <- scm_base_parameters("FF16")
p  <- expand_parameters(trait_matrix(0.0825, "lma"), p0, birth_rate_list = 20)
```

## Exporting and resuming a run

Run a patch, collecting a snapshot of the whole population after each step, then
export the state part-way through:

```{r}
#| label: run-and-export
scm <- SCM("FF16", "FF16_Env")(p, Environment("FF16"), Control())
scm$collect <- TRUE
scm$run()

# export the patch about half-way through the run
k <- floor(length(scm$history) / 2)
state <- export_patch_state(scm, step = k)
```

`export_patch_state()` captures everything needed to continue the run: every
node's ODE state, the per-node *birth bookkeeping* (introduction time, patch-age
density and survival probability at birth — quantities that are not part of the
ODE state but feed the growth/mortality rates and the lifetime-fitness
integrals), the patch age, and the portion of the introduction schedule that has
not happened yet.

`set_initial_state()` writes that onto a `Parameters` object, and `run_scm()`
resumes from it:

```{r}
#| label: resume
p_resumed <- set_initial_state(p, state)
resumed   <- run_scm(p_resumed, collect = TRUE)
```

The resumed run reproduces the original. The seed is loaded bit-for-bit; only the
adaptive ODE stepper, whose internal step size is not part of the checkpoint,
diverges slightly going forward — so the recovered fitness matches to a few parts
in $10^5$ rather than exactly:

```{r}
#| label: compare-fitness
c(original = scm$net_reproduction_ratios,
  resumed  = resumed$net_reproduction_ratios)
```

Because the state lives on the `Parameters` object, the resumed run is fully
self-describing: you could `saveRDS(p_resumed)` and reproduce the continuation in
a fresh session.

## Seeding an arbitrary size distribution

To start a patch from scratch with a chosen distribution of plants, build an
initial state from node heights and densities with `make_initial_state()`. The
remaining state (mortality, fecundity, accumulated reproduction, heartwood) starts
at zero, every node is introduced at patch age 0, and recruitment continues
normally for $t > 0$ (the seeded distribution stands in for the $t = 0$ recruit).

As an example, take a log-normal distribution of heights, normalised to a target
total density — the same shape PR #304 used to motivate this feature:

```{r}
#| label: lognormal-distribution
# a log-normal density over heights, scaled to a target total density and
# expressed as node heights + densities
ln_dist <- function(mean_height = 6, log_sd = 0.5, density = 1,
                    min_h = 0.5, max_h = 10, n = 40) {
  h <- seq(min_h, max_h, length.out = n)
  d <- dlnorm(h, log(mean_height), log_sd)
  list(heights = h, densities = d / trapezium(h, d) * density)
}

dist <- ln_dist(mean_height = 6, density = 1)

tibble(height = dist$heights, density = dist$densities) |>
  ggplot(aes(height, density)) +
  geom_area(fill = "grey80", colour = "black") +
  labs(x = "Height (m)", y = "Density (/m / m2)",
       title = "Seeded initial size distribution (t = 0)") +
  coord_cartesian(expand = FALSE) +
  theme_classic()
```

Seed a patch with it and run as usual. `run_scm(collect = TRUE)` returns the
standard tidy output, so the seeded run drops straight into the usual plotting
and aggregation helpers:

```{r}
#| label: seed-and-run
state0 <- make_initial_state(p, heights = dist$heights,
                             densities = dist$densities)
seeded <- run_scm(set_initial_state(p, state0), collect = TRUE)
```

```{r}
#| label: plot-size-distribution
#| fig-alt: "Node heights over patch age from a seeded log-normal distribution."
plot_size_distribution(seeded$species |> drop_na()) +
  coord_cartesian(expand = FALSE) +
  labs(title = "Patch evolution from a seeded distribution")
```

The fan of lines starting at $t = 0$ is the seeded population growing and being
shaded out; the lines that appear later are recruits arriving on the schedule.

Aggregate properties come from the same tidy table via `FF16_expand_state()` and
`integrate_over_size_distribution()`:

```{r}
#| label: aggregate-dynamics
#| fig-alt: "Leaf area index and stem density over patch age."
totals <- seeded |> FF16_expand_state() |> _$species |>
  integrate_over_size_distribution()

totals |>
  select(time, area_leaf, density) |>
  rename(`leaf area index` = area_leaf, `individuals (/m2)` = density) |>
  pivot_longer(-time) |>
  ggplot(aes(time, value)) +
  geom_line() +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Patch age (years)", y = NULL) +
  coord_cartesian(expand = FALSE) +
  theme_classic()
```

Seeding the *same* distribution shape at different total densities shows how the
starting density shapes the early dynamics before the patch converges:

```{r}
#| label: density-panel
#| fig-alt: "Size-distribution trajectories for three decreasing initial densities."
densities <- c(1.0, 0.2, 0.05)

panel <- lapply(densities, function(dn) {
  d <- ln_dist(mean_height = 6, density = dn)
  r <- run_scm(set_initial_state(p, make_initial_state(p, d$heights, d$densities)),
               collect = TRUE)
  r$species |> drop_na() |>
    mutate(scenario = sprintf("initial density = %.2f /m2", dn))
}) |> bind_rows()

plot_size_distribution(panel) +
  facet_wrap(~scenario) +
  coord_cartesian(expand = FALSE) +
  labs(title = "Same distribution shape, decreasing initial density")
```

## Plausible initial conditions

Very large plants at very high density are not a state the model can integrate:
they self-shade so heavily that net production goes negative, mortality blows up,
and node densities race to non-finite values within a few steps. `plant` checks
for this when the patch is seeded and refuses rather than producing garbage:

```{r}
#| label: guard
#| error: true
bad <- make_initial_state(p, heights = seq(8, 12, length.out = 20),
                          densities = rep(50, 20))
run_scm(set_initial_state(p, bad))
```

If you hit this, reduce the sizes and/or densities of the initial plants — a
distribution the solver itself could have produced (for example, one obtained via
`export_patch_state()`) is always safe to re-import.
