---
title: "TF24: hydraulics, and a fast acclimating variant"
---

<!--
STAGING COPY. This guide is written for the Overstorey site
(https://github.com/traitecoevo/overstorey) but is kept in the plant repo for
now because it documents the `feature/tf24f-acclimation` branch (TF24f, issue
#525), which is not yet merged. When that branch lands, move this file to
`overstorey/guides/tf24-hydraulics-and-acclimation.qmd`, add it to the Guides
sidebar in `_quarto.yml`, and delete the `overstorey-staging/` directory here.
All figures are generated by the code chunks below (the site uses `freeze`), so
nothing static needs to move with it.
-->

::: {.eyebrow}
guide
:::

`TF24` is `plant`'s hydraulically-explicit strategy. Where the simpler `FF16`
strategy reads a single light value and applies a fixed light–response curve,
`TF24` resolves, for every plant at every step, the **leaf water potential** the
plant should operate at — trading carbon gain (open stomata, more
photosynthesis) against hydraulic risk (more negative water potential, closer to
loss of conductivity). Concretely it maximises a carbon *profit* over the
root-collar water potential, which itself requires solving the soil → root →
stem water-transport problem.

That optimisation is the model's accuracy *and* its cost: each rate evaluation
runs a golden-section search over collar potential, and inside every step of that
search a root-find for the water-transport continuity. Under the adaptive ODE
solver this happens many times per plant per patch step.

This guide explains the idea behind **TF24f** ("f" for fast / forecasting), a
variant that keeps the same biology but stops re-solving the optimisation from
scratch every step.

```{r}
#| label: setup
library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)

base <- function(type) {
  scm_base_parameters(type) |>
    add_strategies(trait_matrix(0.0825, "lma"), birth_rate = 20)
}
```

## The idea: let the optimum chase a moving target

The optimal operating point changes only slowly as a plant grows, its light
environment shifts and the water environment changes. So instead of re-optimising it from scratch, `TF24f` carries
the optimal root-collar potential as an **extra ODE state** that *relaxes toward*
its optimum — gradient ascent on the carbon profit:

$$\frac{\mathrm{d}\psi}{\mathrm{d}t} = k \,\frac{\partial\,\text{profit}}{\partial\psi}.$$

The fixed point of this ODE is exactly the optimum (where the profit gradient is
zero), so $\psi$ tracks the moving optimum *without ever solving for it*. Each
step then needs only a pair of perturbed evaluations either side of the current
$\psi$ for a centred finite-difference gradient — no golden-section search. The gain $k$ sets how
tightly the state tracks: this is both a performance lever and, at finite $k$, a
genuine model of **acclimation** with a finite adjustment lag (issues #405/#406).

`TF24f` is implemented as a thin C++ subclass of `TF24` (it inherits all of
TF24's biology and reuses its environment and parameters), adding just the
tracked state and the gradient-ascent rate. From R it behaves like any other
strategy, with two extra knobs: `k_acclim` (the gain $k$) and `psi_fd_step` (the
finite-difference step for the gradient).

## Running TF24 and TF24f side by side

```{r}
#| label: run-both
tf24  <- run_scm(base("TF24"),  collect = TRUE, refine_schedule = FALSE)

p_f <- base("TF24f")
s <- p_f$strategies[[1]]
s$k_acclim   <- 2     # gain; low values track best (see below)
s$psi_fd_step <- 0.05
p_f$strategies[[1]] <- s
tf24f <- run_scm(p_f, collect = TRUE, refine_schedule = FALSE)
```

The two patches develop near-identically — the size-structure trajectories track
each other closely:

```{r}
#| label: size-distributions
#| fig-alt: "Node height trajectories for TF24 and TF24f, which closely overlap."
bind_rows(
  tf24$species  |> drop_na() |> mutate(model = "TF24 (optimised)"),
  tf24f$species |> drop_na() |> mutate(model = "TF24f (tracked)")
) |>
  plot_size_distribution() +
  facet_wrap(~model) +
  coord_cartesian(expand = FALSE) +
  labs(title = "Same biology, two ways of resolving the leaf optimum")
```

`TF24f` adds one diagnostic column, `opt_root_psi_state` — the tracked collar
potential — which we can compare against TF24's per-step optimum (`opt_root_psi`,
reported as a signed potential). They sit close throughout. The visible gap is
the **acclimation lag** early in life, when the optimum drops quickly as the
seedling grows and the finite-$k$ state cannot descend quite as fast; once growth
slows the two converge to within a fraction of a millivolt:

```{r}
#| label: tracking
#| fig-alt: "Tracked collar potential closely follows TF24's per-step optimum."
node1 <- function(d) d |> filter(node == 1) |> drop_na()
bind_rows(
  node1(tf24$species)  |> transmute(time, psi = -opt_root_psi,       model = "TF24 optimum"),
  node1(tf24f$species) |> transmute(time, psi = opt_root_psi_state, model = "TF24f tracked")
) |>
  ggplot(aes(time, psi, colour = model)) +
  geom_line() +
  labs(x = "Patch age (years)", y = "Root-collar potential (MPa)",
       colour = NULL, title = "The tracked state follows the moving optimum") +
  theme_classic()
```

## Choosing the gain `k`

It is tempting to think a very large `k` (track the optimum instantly) should
recover `TF24` exactly. It does **not**. Two things get worse as `k` grows:

- **Stiffness.** A fast-relaxing extra state forces the adaptive solver into ever
  smaller steps; beyond a moderate `k` the run becomes effectively unintegrable.
- **Gradient noise.** The profit gradient here is a finite difference, and the
  underlying leaf solve has its own internal tolerances, so a large `k` amplifies
  that numerical noise into the dynamics.

The result is a *low* optimal gain. On the standard single-species patch above,
`k ≈ 2` gives the best agreement with `TF24` — median height difference a few
millimetres, ~93% of the population within 1% — while running `run_scm` in
roughly **half the time** of `TF24`. Larger `k` is both slower and less accurate.

The gradient is a *centred* finite difference, so its fixed point sits at the
true optimum to $O(h^2)$ rather than the $O(h)$ of a one-sided difference — at
`psi_fd_step = 0.05` the steady-state offset from `TF24` is well under a
millivolt. (An earlier one-sided difference left a constant `psi_fd_step`/2 ≈
0.025 MPa offset that did not shrink as the patch aged.) The remaining mismatch
is a floor set by the finite-difference gradient, not by the tracking idea
itself; a smooth/analytic gradient (automatic differentiation) is the route to
tightening it further.

## A note on initialisation

A subtlety worth flagging: a newly recruited seedling must *start* at its optimal
operating point. If the tracked state instead starts at zero, shaded understory
recruits spend their first steps at a non-physical operating point and can
**escape light suppression** — growing into the canopy where `TF24` would keep
them suppressed. `TF24f` avoids this by seeding the tracked potential at its true
optimum (one full solve) when each plant is born, so gradient ascent starts at
the optimum rather than racing toward it.

## Status

`TF24f` currently tracks the root-collar potential only, under the mean-light
canopy model, and matches `TF24` well (not exactly) at roughly twice the speed.
Tightening the match with a smooth gradient, and extending to the deep-crown
canopy model, are the next steps.
