---
title: "TF24: carbohydrate storage, buffered growth and mortality"
---

<!--
STAGING COPY. This guide is written for the Overstorey site
(https://github.com/traitecoevo/overstorey) but is kept in the plant repo for
now because it documents the `feature/tf24-nsc-517` branch (NSC storage, issue
#517, PR #554), which is not yet merged. When that branch lands, move this file
to `overstorey/guides/tf24-nsc-storage.qmd`, add it to the Guides sidebar in
`_quarto.yml`, and delete the `overstorey-staging/` directory here. All figures
are generated by the code chunks below (the site uses `freeze`), so nothing
static needs to move with it.
-->

::: {.eyebrow}
guide
:::

By default `TF24` maps carbon supply to demography *instantaneously*: the height
growth rate and the growth-dependent mortality rate both read the current net
mass production. That coupling is brittle. Under a deep productivity trough net
production goes sharply negative, and the growth-dependent mortality term
`a_dG1 · exp(-a_dG2 · productivity)` runs away to enormous values — in the SCM it
reached ~10³² and overflowed the coupled size-density equations (issue #550).
Biologically it is also wrong: real trees do not die the instant a bad season
starts. They draw down stored **non-structural carbohydrate (NSC)** and die
gradually only once reserves are exhausted (Fiona Robinson's thesis; Stefaniak et
al. 2026).

This guide describes the NSC **storage pool** added to `TF24` (issue #517). A new
state variable stores carbon; growth and mortality read that *buffered* reserve
rather than instantaneous production. The governing principle, following the
biology, is simple:

> a plant should not grow unless it has ample carbon in storage.

```{r}
#| label: setup
#| message: false
library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## The carbon flow

Each plant carries a storage pool $S$ (kg of NSC) as an extra ODE state. Net mass
production $P$ (the usual TF24 quantity: assimilation, less respiration and
turnover) charges the pool; growth and reproduction are funded from it, gated on
how full it is; the balance stays in reserve. In symbols:

$$
\frac{\mathrm{d}S}{\mathrm{d}t} = P - U,
\qquad
U = P^{+}\,G(r),
$$

where $U$ is the carbon actually committed to growth **and** reproduction (split
between them by TF24's usual allocation fractions), $P^{+}$ is a smooth
positive part of net production, and $G(r)$ is a **reserve gate** on the relative
fullness of the pool.

### Capacity and relative reserves

Storage capacity scales with sapwood mass — NSC is held largely in living
sapwood parenchyma:

$$
S_{\max} = a_{\mathrm{st1}}\, m_{s},
\qquad
r = \frac{S}{S_{\max}} \in [0, 1],
$$

with $m_s$ the sapwood mass and $a_{\mathrm{st1}}$ the storable fraction of it
(default `0.10`). $r$ is the plant's relative reserve level, and it is the single
signal that both the growth gate and the mortality term read.

### The reserve gate on growth

Growth proceeds at the **production rate**, scaled by a smooth logistic gate that
is $\approx 0$ when reserves are low and $\approx 1$ when they are ample:

$$
G(r) = \frac{1}{1 + \exp\!\big(-(r - a_{\mathrm{st2}})/w\big)},
\qquad
P^{+} = \tfrac{1}{2}\Big(P + \sqrt{P^{2} + \varepsilon^{2}}\Big).
$$

$a_{\mathrm{st2}}$ is the reserve fraction at which growth is half-on (default
`0.10`), $w$ a fixed width, and $P^{+}$ a differentiable stand-in for
$\max(P, 0)$ that removes the old hard `net > 0` growth cutoff. Carbon not spent
on growth (because the gate is partly closed, or because $P > U$) stays in $S$ and
refills the pool.

A deliberate modelling choice sits here. Growth is gated *at the production rate*,
not metered out of the pool as a discharge $a_{\mathrm{st2}}\,S$ (the literal
"utilisation" of Stefaniak et al. 2026). With capacity tied to sapwood mass, a
seedling's pool is tiny relative to its high specific productivity — a full pool
can fund only a few percent of its production rate — so a discharge-rate limit
would choke establishment and collapse reproduction under light competition.
Gating the *rate* instead preserves healthy growth while still enforcing "no
growth without reserves".

### Reserves-based mortality

The growth-dependent mortality term now reads relative reserves instead of
instantaneous productivity:

$$
d(r) = d_I + a_{\mathrm{dG1}}\,\exp(-a_{\mathrm{dG2}}\, r).
$$

Because $r \in [0,1]$, this is **bounded** in
$[\,a_{\mathrm{dG1}} e^{-a_{\mathrm{dG2}}},\; a_{\mathrm{dG1}}\,]$: near-zero
excess mortality at full reserves, rising smoothly to a finite maximum
$a_{\mathrm{dG1}}$ as the pool empties. This is the structural fix for the #550
blow-up — the runaway is impossible by construction — and it makes death
*gradual* as reserves deplete rather than instantaneous when a season turns bad.

A newborn seedling is seeded with reserves at a fraction $a_{\mathrm{st3}}$ of its
capacity (default `0.8`), so it can establish rather than dying immediately with
an empty pool.

```{r}
#| label: mortality-bounded
#| fig-alt: "Storage-dependent mortality is bounded between ~0 and a_dG1 as reserves vary 0 to 1."
p <- TF24_Strategy()$pars
tibble(r = seq(0, 1, length.out = 200)) |>
  mutate(mortality = p$d_I + p$a_dG1 * exp(-p$a_dG2 * r)) |>
  ggplot(aes(r, mortality)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = p$d_I + p$a_dG1, linetype = "dashed") +
  annotate("text", x = 0.5, y = p$d_I + p$a_dG1, vjust = -0.5,
           label = "bounded maximum = d_I + a_dG1") +
  labs(x = "Relative reserves  r = S / S_max",
       y = "Instantaneous mortality rate (/yr)",
       title = "Reserves-based mortality is bounded",
       subtitle = "Empty reserves give a finite ~5.5/yr, not the old ~1e32 spike") +
  theme_classic()
```

## Behaviour: growth pauses in drought and resumes on recovery

The clearest way to see the buffer at work is to drive a single plant through a
drought and back. The helper below reproduces the C++ storage capacity
($S_{\max} = a_{\mathrm{st1}}\, m_s$) so we can report the relative reserve $r$
alongside height and mortality.

```{r}
#| label: capacity-helper
storage_capacity <- function(s, height) {
  p <- s$pars
  area_leaf    <- (height / p$a_l1)^(1 / p$a_l2)
  eta_c        <- 1 - 2 / (1 + p$eta) + 1 / (1 + 2 * p$eta)
  mass_sapwood <- (area_leaf * p$theta) * height * eta_c * p$rho
  p$a_st1 * mass_sapwood
}

fixed_env <- function(theta) {
  e <- Environment("TF24")
  e$set_soil_number_of_depths(5)
  e$set_soil_water_state(rep(theta, 5))
  e$set_fixed_environment(1.0, 40)   # full light to 40 m
  e
}
```

We integrate one plant with a simple fixed step: three years in a well-watered
soil, then a sharp drought (soil moisture dropped to near the residual floor) from
year 3 to 4.5, then recovery.

```{r}
#| label: drought-probe
#| fig-alt: "Height growth pauses during drought while reserves deplete and mortality rises to a bounded level; growth resumes after reserves refill."
s   <- TF24_Strategy()
wet <- fixed_env(0.30)
dry <- fixed_env(0.08)

pl <- TF24_Individual(s)
h0 <- pl$state("height")
pl$set_state("storage", s$pars$a_st3 * storage_capacity(s, h0))

dt   <- 0.02
tmax <- 6
rows <- vector("list", round(tmax / dt) + 1)
for (i in seq_along(rows)) {
  t <- (i - 1) * dt
  env <- if (t >= 3 && t < 4.5) dry else wet
  pl$compute_rates(env)
  h <- pl$state("height")
  rows[[i]] <- tibble(
    time      = t,
    height    = h,
    dheight   = pl$rate("height"),
    reserves  = pmax(pl$state("storage"), 0) / storage_capacity(s, h),
    mortality = pl$rate("mortality"),
    soil_water_state = env$get_soil_water_state()[1]
  )
  pl$set_state("height",  h + dt * pl$rate("height"))
  pl$set_state("storage", pl$state("storage") + dt * pl$rate("storage"))
}
traj <- bind_rows(rows)

traj |>
  pivot_longer(c(dheight, reserves, mortality,soil_water_state)) |>
  mutate(name = recode(name,
    dheight   = "Height growth (m/yr)",
    reserves  = "Relative reserves r",
    mortality = "Mortality rate (/yr)",
    soil_water_state = "Soil moisture (m3/m3)")) |>
  ggplot(aes(time, value)) +
  annotate("rect", xmin = 3, xmax = 4.5, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "brown") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(x = "Time (years)", y = NULL,
       title = "Storage buffers the drought (shaded)",
       subtitle = "Growth pauses and reserves draw down; mortality rises to a bounded rate; growth resumes once reserves refill") +
  theme_classic()

ggsave("harsh_soil_moist.png")

```

Three things happen when the drought hits (shaded band) and none of them are the
old failure mode:

- **Growth pauses.** As reserves fall below the gate threshold, the growth rate
  drops smoothly to ~0 — the plant stops building tissue rather than growing down
  its stores.
- **Reserves deplete, mortality rises — but stays bounded.** The mortality rate
  climbs toward its finite maximum ($d_I + a_{\mathrm{dG1}} \approx 5.5$/yr), so
  death is gradual over the drought rather than an instantaneous overflow.
- **Recovery is buffered, not instantaneous.** When water returns, carbon first
  *refills* the pool; growth resumes only once reserves have recovered, a short
  lag after the environment improves.

## The growth threshold `a_st2`

`a_st2` is the reserve fraction below which growth throttles — the knob that sets
how strictly "you need carbon to grow". Because a healthy plant self-regulates to
reserves around $r \approx 0.45$, thresholds up to ~0.2 leave healthy growth
essentially untouched and bite only as reserves draw down in drought; the default
is `0.10`. Lower values behave more like the original instantaneous model (faster
maturation, higher lifetime reproduction); higher values enforce a stronger
storage-first strategy at the cost of slower growth. It is exposed as an ordinary
strategy parameter:

```{r}
#| label: params
s <- TF24_Strategy()
s$pars[c("a_st1", "a_st2", "a_st3")]
```

| parameter | meaning | default |
|---|---|---|
| `a_st1` | storage capacity per unit sapwood mass (kg NSC / kg) | 0.10 |
| `a_st2` | reserve fraction at which growth is half-on | 0.10 |
| `a_st3` | initial storage at birth, as a fraction of capacity | 0.80 |

## Status

The storage pool makes the #550 drought regime complete robustly (across patch
lifetimes 20–70 yr and drought amplitudes 0.30–0.40) with bounded, finite
densities, and reproduces the qualitative growth-and-mortality buffering above.
The parameters are **defensible prototype defaults, not calibrated**; fitting
$a_{\mathrm{st1}}$, $a_{\mathrm{st2}}$, $a_{\mathrm{st3}}$ and the
reserve→mortality shape to Fiona Robinson's thesis and Stefaniak et al. (2026) is
the next step. `TF24f` (the fast acclimating variant) inherits the storage pool
automatically.

Reaching this regime also required fixing an independent soil-water singularity
(issue #549), where a layer at the residual moisture floor produced a non-finite
soil water potential; that is documented in PR #554.

## Smoother soil moisture variation

```{r}
#| label: drought-probe
#| fig-alt: "Height growth pauses during drought while reserves deplete and mortality rises to a bounded level; growth resumes after reserves refill."
s   <- TF24_Strategy()
wet <- fixed_env(0.30)
dry <- fixed_env(0.08)

pl <- TF24_Individual(s)
h0 <- pl$state("height")
pl$set_state("storage", s$pars$a_st3 * storage_capacity(s, h0))

dt   <- 0.02
tmax <- 6
rows <- vector("list", round(tmax / dt) + 1)
for (i in seq_along(rows)) {
  t <- (i - 1) * dt
  env <- if (t >= 3 && t < 4.5) fixed_env(0.3 + 0.22*(16/9)*(t - 3)*(t - 4.5)) else wet
  pl$compute_rates(env)
  h <- pl$state("height")
  rows[[i]] <- tibble(
    time      = t,
    height    = h,
    dheight   = pl$rate("height"),
    reserves  = pmax(pl$state("storage"), 0) / storage_capacity(s, h),
    mortality = pl$rate("mortality"),
    soil_water_state = env$get_soil_water_state()[1]
  )
  pl$set_state("height",  h + dt * pl$rate("height"))
  pl$set_state("storage", pl$state("storage") + dt * pl$rate("storage"))
}
traj <- bind_rows(rows)

traj |>
  pivot_longer(c(dheight, reserves, mortality,soil_water_state)) |>
  mutate(name = recode(name,
    dheight   = "Height growth (m/yr)",
    reserves  = "Relative reserves r",
    mortality = "Mortality rate (/yr)",
    soil_water_state = "Soil moisture (m3/m3)")) |>
  ggplot(aes(time, value)) +
  annotate("rect", xmin = 3, xmax = 4.5, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "brown") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(x = "Time (years)", y = NULL,
       title = "Storage buffers the drought (shaded)",
       subtitle = "Growth pauses and reserves draw down; mortality rises to a bounded rate; growth resumes once reserves refill") +
  theme_classic()

  ggsave("smooth_soil_moist.png")
```
