---
title: "Canopy shading methods"
output: bookdown::html_document2

---

**authors**: DS Falster
**date**: 2026

# Background

## Light interception, and why it matters

Light is the currency of competition in a forest. A plant's carbon gain depends
on the light its leaves intercept, and that light has been filtered twice: by
the crowns of taller neighbours above it, and by the plant's own leaves higher
in its crown (self-shading). `plant` captures this through a single shared
object -- the patch **light profile** $E(z)$, the fraction of open-sky light
reaching height $z$. Every plant both *reads* this profile (to photosynthesise)
and *contributes* to it (the shade its leaves cast lowers $E$ below them).

Getting the light profile right is not a detail. It sets each plant's growth
rate, which sets the size structure of the stand, which sets the competitive
hierarchy, which ultimately decides which strategies persist and which traits
selection favours. Errors here propagate all the way up to the ecological and
evolutionary predictions the model exists to make.

## Realism versus speed

A fully realistic treatment would resolve the light field through each crown and
integrate photosynthesis over it. That is expensive -- especially when the
leaf-level photosynthesis model is itself costly (as in `TF24`, whose coupled
stomatal/hydraulic optimisation is far heavier than a simple light-response
curve). The classic response is to *simplify*: treat the crown as a single
point, or the canopy as a few discrete layers -- trading fidelity for speed.
`plant` lets you choose where to sit on that spectrum. The job of this vignette
is to ask **which simplifications are safe, and for what purpose** -- because, as
we will see, a model can be perfectly adequate for one question and badly wrong
for another.

## The models

The `FF16` strategy offers six shading models, selected with
`control$shading_model`. Four of them share the *same* per-plant competition --
the smooth Yokozawa leaf-area profile -- and differ only in how a plant's own
assimilation is computed (and, for PPA, how the light profile is assembled):

* **`"deep-crown"`** (the default) -- the original `plant` behaviour. Leaf area
  is distributed continuously through the crown following the Yokozawa (1995)
  foliage profile, and a plant's gross assimilation is computed by *integrating*
  photosynthesis over crown depth against the light profile. The patch light
  profile is a smooth function of height.

* **`"crown-centre"`** -- the light profile is built exactly as for `"deep-crown"`,
  but a plant's assimilation is approximated by a *single evaluation* of the
  light at its crown centre rather than an integral over depth. This skips the
  per-plant quadrature, so it is faster, and removes within-crown self-shading.

* **`"mean-light"`** -- partway between deep-crown and crown-centre. Rather than
  integrating the *photosynthetic rate* over depth, it integrates the *light*
  over depth (weighted by leaf-area density) to a single leaf-area-weighted mean
  light, then takes one photosynthesis evaluation of that mean. It therefore
  still feels the within-crown change in light environment -- and how that environment
  shifts as the canopy develops -- but collapses the expensive photosynthesis
  calculation to a *single* call per plant rather than one per crown depth. For
  FF16 that saving is modest (its leaf photosynthesis is a cheap saturating
  function), but it becomes important for models with an expensive leaf
  submodel: the `TF24` strategy's coupled stomatal/hydraulic optimisation is far
  more costly to evaluate, so integrating the light and optimising once is the
  natural choice there. The cost is bias: because photosynthesis saturates with
  light, evaluating the rate at the mean light overestimates the depth-averaged
  rate (Jensen's inequality). So `"mean-light"` always gives at least as much
  assimilation as `"deep-crown"`, the two agreeing only when the within-crown
  light is uniform.

* **`"ppa"`** -- the perfect-plasticity approximation, the scheme that underlies
  large models such as FATES. The patch light profile is discretised into
  discrete canopy *layers*: light steps down by a fixed amount of optical depth
  (`control$ppa_layer_optical_depth`) per layer.
  Assimilation is evaluated at the crown centre as in `"crown-centre"`, but now
  reads this stepped profile. Because a hard step is not differentiable (and
  makes the ODE solver fail), the layer boundaries are smoothed with a
  C^1^-continuous ramp whose width is set by `control$ppa_layer_smoothing`.
  This gives two variants of the *one* `"ppa"` model, exactly paralleling the
  Flat Top pair below: **PPA (hard step)** (`ppa_layer_smoothing = 0`) is the
  literal field discretisation (e.g. Strigul 2008, FATES) -- discontinuous, and
  it does not run in `plant`'s adaptive solver -- while **PPA (smoothed)**
  (`ppa_layer_smoothing > 0`, default `0.3`) is the runnable version that every
  PPA result reported here uses.

Two further models instead change the *competition* itself -- the shade a plant
casts -- collapsing its leaf area towards a thin crown-centre layer rather than
spreading it through the crown. This discrete-layer view is close to how many
large vegetation models (e.g. LPJ-GUESS) represent canopy competition, so it is
worth taking seriously rather than dismissing:

* **`"flat-top-box"`** -- the shade cast is a hard *step*: full below the crown
  centre, none above. Summed over cohorts this makes the patch light profile
  discontinuous, and in `plant` the stand then cannot be solved at all (see
  [mutant fitness](#mutant-fitness)).
* **`"flat-top-soft-box"`** -- the same box competition with the step smoothed to
  a continuous drop, so the light environment can be built. It runs, but the
  mis-shaped shade biases its predictions.

The model is resolved once when a strategy is prepared, so there is no per-call
cost to having the choice available. The `TF24` strategy supports `"deep-crown"`,
`"mean-light"` (its default) and `"crown-centre"`, but not the PPA or box
variants.

## How we evaluate a shading model

We put the models through a ladder of increasingly demanding tests of biological
adequacy, noting their computational cost and numerical behaviour along the way.
Each rung asks more of the light profile than the last:

1. **Stand dynamics** -- does a stand develop sensibly? We compare the size
   distribution, leaf area index, and vertical light profile as a patch grows
   and self-thins.
2. **Demographic equilibrium** -- does the population settle to a sensible,
   self-replacing seed rain?
3. **Invasion (mutant) fitness** -- can the model support evolutionary analysis?
   This requires the light a plant experiences to vary *continuously* with its
   size, so the fitness landscape is smooth.
4. **Evolutionary endpoint** -- given a smooth landscape, *where does evolution
   settle*, and do the methods agree on that singular strategy? This is the most
   demanding rung. It compounds every earlier difference, because the endpoint
   depends on the equilibrium stand each model builds -- so two models can pass
   every rung below and still predict different optima.

The recurring finding is that the cheap models are often fine for stand dynamics
and acceptable for equilibrium, but invasion fitness is unforgiving: it exposes
which simplifications break -- PPA's stepped profile, and the discrete-layer
("box") competition that large vegetation models rely on -- and shows that the
key requirement is **continuity of the shade each plant casts as a function of
height**. The final rung sharpens the warning: even a model that clears the
fitness-landscape bar can still settle evolution in the wrong place.

```{r, message = FALSE}
library(plant)
library(dplyr)
library(ggplot2)
library(patchwork)
```

# Simulating a patch under each model

We grow a single `FF16` species in a patch, and run the same patch under each of
the four smooth-competition models (the box variants are revisited later, once
we reach the question they bear on). Everything except `control$shading_model`
is held fixed, so any differences are due to the shading model alone.

```{r}
params <- scm_base_parameters("FF16")
patch  <- expand_parameters(trait_matrix(0.0825, "lma"), params)

models <- c("deep-crown", "mean-light", "crown-centre", "ppa")

run_model <- function(model) {
  ctrl <- Control()
  ctrl$shading_model <- model
  run_scm(patch, ctrl = ctrl, collect = TRUE)
}

results <- lapply(models, run_model)
names(results) <- models
```

Each run returns the full collected state (`run_scm(collect = TRUE)`): a tidy
`species` table (one row per node per step) and the light environment
(`env$light_availability`). We tag each run's `species` table with its shading
model and bind them together so we can use `plant`'s tidy helpers across all
four.

```{r}
species_all <- dplyr::bind_rows(
  lapply(models, function(m) dplyr::mutate(results[[m]]$species, method = m))
)
species_all$method <- factor(species_all$method, levels = models)

# Light-extinction coefficient, used throughout to convert canopy openness to
# cumulative leaf area: E = exp(-k_I * LAI), so LAI_above = -log(E) / k_I.
k_I <- results[[1]]$p$strategies[[1]]$k_I
```

# Stand dynamics

The first and least demanding test is whether a stand develops sensibly. We look
at three views of the same patch as it grows and self-thins: the size
distribution, the leaf area index, and the vertical light profile.

## Size distribution

The size distribution is the set of node height trajectories: each line is a
cohort introduced at a given patch age and followed as it grows, with line
opacity scaled by the cohort's density (faint lines are sparse cohorts).
`plant`'s `plot_size_distribution()` builds this directly from the tidy
`species` table; faceting by `method` compares the shading models.

```{r sizedist, fig.width = 10, fig.height = 3.4}
p_size <- plot_size_distribution(species_all) +
  facet_wrap(~method) +
  theme(legend.position = "bottom")
p_size
```

The discrete layering in `"ppa"` produces visibly different growth trajectories:
cohorts grow at near-constant light within a layer, then accelerate as they
emerge into a brighter layer.

## Leaf area index

Leaf area index (LAI) -- the total one-sided leaf area per unit ground -- is the
cumulative leaf area above the ground, which is already encoded in the recorded
light profile: with Beer's law $E = \exp(-k_I\,\mathrm{LAI})$, the patch LAI is
just $-\log E(0) / k_I$, where $E(0)$ is canopy openness at ground level. This
reads straight off `env$light_availability` and needs no integration over the
size distribution. (LAI is real leaf area, so the smooth recorded profile gives
the true value for all four models -- the PPA layering only affects the light
plants *experience*, not the leaf area present.)

```{r}
lai_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    results[[m]]$env$light_availability |>
      dplyr::group_by(time) |>
      dplyr::slice_min(height, n = 1, with_ties = FALSE) |>
      dplyr::ungroup() |>
      dplyr::transmute(method = m, time, lai = -log(light_availability) / k_I)
  })
)
lai_all$method <- factor(lai_all$method, levels = models)
```

```{r lai, fig.width = 7, fig.height = 4}
p_lai <- ggplot(lai_all, aes(time, lai, colour = method)) +
  geom_line(linewidth = 1) +
  labs(x = "Patch age (years)", y = "Leaf area index", colour = "Shading model") +
  theme_classic()
p_lai
```

`"deep-crown"`, `"mean-light"` and `"crown-centre"` track each other closely and
settle to a steady LAI: under a near-uniform light profile, all three ways of
summarising within-crown light give similar carbon gain (with
`"mean-light"` and `"crown-centre"` running slightly richer, since they do not
fully resolve self-shading). The `"ppa"` model instead shows large, sustained
oscillations -- successive waves of recruitment and self-thinning -- because the
discrete light layers create sharp thresholds for cohort success (visible as the
pulses of emerging cohorts in the size-distribution panel above).

## Vertical light profile

We can also look *within* the canopy at a few patch ages, at the vertical
profile of cumulative leaf area above each height -- $\mathrm{LAI}_{above}(z) =
-\log E(z) / k_I$, where $E$ is canopy openness. This is the quantity that sets
how much each individual is shaded.

The recorded `env$light_availability` is the smooth underlying profile shared by
all models. The light that plants actually *experience* under PPA is the stepped
version, so for the `"ppa"` panel we apply the same layering transform used
internally (see `FF16_Environment::step_light`).

```{r}
d <- Control()$ppa_layer_optical_depth   # layer thickness (optical depth)
w <- Control()$ppa_layer_smoothing       # boundary smoothing fraction

# C1-smoothed staircase: flat over the lower (1 - w) of each layer, cubic ramp
# over the top w. Mirrors FF16_Environment::smooth_floor() / step_light().
smooth_floor <- function(u, w) {
  n <- floor(u)
  f <- u - n
  t <- pmax(0, f - (1 - w)) / w
  n + ifelse(f <= 1 - w, 0, t^2 * (3 - 2 * t))
}
ppa_step_light <- function(E, d, w) {
  ifelse(E >= 1 | E <= 0, E,
         exp(-d * smooth_floor(-log(E) / d, w)))
}

target_ages <- c(5, 10, 20, 60)

profile_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    light <- results[[m]]$env$light_availability
    if (m == "ppa") {
      light$light_availability <- ppa_step_light(light$light_availability, d, w)
    }
    # nearest recorded step to each target age
    steps <- results[[m]]$steps
    keep <- sapply(target_ages, function(a) steps$step[which.min(abs(steps$time - a))])
    light |>
      dplyr::filter(step %in% keep) |>
      dplyr::mutate(method = m,
                    age = target_ages[match(step, keep)],
                    lai_above = -log(light_availability) / k_I)
  })
)
profile_all$method <- factor(profile_all$method, levels = models)
```

```{r vprofile, fig.width = 10, fig.height = 3.6}
p_profile <- ggplot(profile_all,
                    aes(lai_above, height, colour = factor(age), group = age)) +
  geom_path() +
  facet_wrap(~method) +
  labs(x = expression(LAI[above]~(m^2/m^2)), y = "Height (m)",
       colour = "Patch age\n(years)") +
  theme_classic()
p_profile
```

`"deep-crown"`, `"mean-light"` and `"crown-centre"` share an identical smooth
profile (they build the light environment the same way and differ only in how
assimilation is evaluated); `"ppa"` shows the characteristic staircase, with
cumulative leaf area held constant within a layer and jumping at each layer
boundary.

# Computational cost

Speed is the whole reason the simpler models exist. A single patch is cheap, but
`plant` is usually run thousands of times over -- sweeping traits, fitting to
data, or computing the fitness landscapes below -- so the per-run cost compounds,
and shaving the most expensive inner calculation is worthwhile. This is the
payoff side of the realism-versus-speed trade.

How much you save depends on two competing effects, and on a feature that is
relatively unusual in vegetation models: `plant` integrates the patch with an
**adaptive** ODE solver, which chooses its own step sizes (large where the
dynamics are smooth, small only where they are not) to hit a requested accuracy
cheaply. So the cost of a model is set both by the work *per step* and by the
*number of steps* the solver decides to take:

* **Per step**, `"deep-crown"` is the most expensive -- it performs a
  Gauss-Kronrod quadrature over crown depth for every individual -- while
  `"crown-centre"` and `"ppa"` replace that with a single light evaluation.
* **Number of steps**: a model whose light profile is smooth lets the solver
  stride; PPA's near-stepped profile is much stiffer, forcing many more, smaller
  steps.

Which effect dominates depends on how the system is stepped, so we report both
the adaptive cost (what you actually pay) and the per-step cost on a fixed
schedule.

```{r speed-helper}
# Median wall-clock over a few repeats. `times = NULL` uses the adaptive solver;
# supplying a fixed schedule pins every model to the same steps.
time_model <- function(model, times = NULL, reps = 5) {
  p <- patch
  use_fixed <- !is.null(times)
  if (use_fixed) p$ode_times <- times
  ctrl <- Control()
  ctrl$shading_model <- model
  run1 <- function() run_scm(p, ctrl = ctrl, use_ode_times = use_fixed)
  run1()                                   # warm up
  median(replicate(reps, system.time(run1())[["elapsed"]]))
}

speed_table <- function(timings) {
  data.frame(
    model = names(timings),
    `median time (s)` = round(unname(timings), 4),
    `relative to deep-crown` = round(unname(timings / timings[["deep-crown"]]), 2),
    check.names = FALSE,
    row.names = NULL
  )
}
```

## Adaptive solver (real-world cost)

This is how `plant` runs by default, and the cost you actually pay. The adaptive
solver is part of what makes `plant` efficient: it spends steps only where the
dynamics demand them. But it also means a model's cost is not just its per-step
work -- a model that makes the dynamics harder to integrate pays for it in extra
steps. (Most large vegetation models instead march on a *fixed* timestep, where
this consideration simply does not arise; we return to that below.)

```{r speed-adaptive}
timings_adaptive <- vapply(models, time_model, numeric(1))
knitr::kable(speed_table(timings_adaptive))
```

`"crown-centre"` is fastest -- it shares deep-crown's smooth light profile (so the
same number of solver steps) but skips the per-plant crown-depth integral.
`"mean-light"` costs about the same as `"deep-crown"`: it also integrates over
crown depth (just the light rather than the rate), so the per-step work and the
step count are both similar. `"ppa"` is *slower* overall despite the cheaper
per-plant assimilation, because its near-stepped profile forces the adaptive
solver to take many more steps; finer layers (smaller `ppa_layer_optical_depth`)
or sharper boundaries (smaller `ppa_layer_smoothing`) make this worse.

## Fixed ODE schedule (per-step cost)

If instead every model is pinned to the *same* fixed schedule
(`run_scm(..., use_ode_times = TRUE)`), the step-count effect is removed and only
the per-step cost remains. Now PPA's advantage shows: like `"crown-centre"`, it
skips the crown-depth quadrature, so both are markedly faster than
`"deep-crown"`.

```{r speed-fixed}
times <- seq(0, params$max_patch_lifetime, length.out = 4000)
timings_fixed <- vapply(models, time_model, numeric(1), times = times)
knitr::kable(speed_table(timings_fixed))
```

So PPA is not inherently slow -- on a fixed schedule it is *faster* than
`"deep-crown"`. Its apparent cost under the adaptive solver comes entirely from
the extra steps the stiff profile demands. This is also why discrete-layer
schemes are unproblematic in the large models that use them: those models march
on a fixed timestep, so a stepped, stiff light profile costs nothing extra and
the difficulty never surfaces. It is `plant`'s adaptive solver that makes the
stiffness visible -- and, as the next section shows, a *hard* step visible as
outright failure. `"deep-crown"` nonetheless remains the default here: it is the
most faithful representation of within-crown light capture and carries no speed
penalty relative to earlier versions of `plant`.

# Why PPA needs smoothing

A literal PPA discretisation is a *hard* step function: as a plant grows and its
crown centre crosses a layer boundary, the light it experiences jumps
discontinuously, and so does its growth rate. This is what `ppa_layer_smoothing`
exists to tame, and it is worth being explicit about *why* -- because the
problem is specifically one of **adaptive ODE stepping**, not of the model
itself.

## Adaptive stepping cannot integrate a discontinuity

`plant` solves the patch dynamics with an adaptive, error-controlled solver
(an embedded Runge-Kutta scheme). At each step it forms two estimates of the
state at different orders, and uses their difference as an estimate of the local
truncation error. If that error exceeds the requested tolerance, the step is
*rejected* and retried with a smaller size; the method assumes the solution is
smooth, so halving the step should roughly quarter the error.

Across a discontinuity that assumption breaks. No matter how small the step, a
step that straddles the jump carries an O(jump) error that never falls below
tolerance, so the solver rejects and shrinks the step indefinitely until it
gives up. With a hard step (`ppa_layer_smoothing = 0`) this is exactly what
happens:

```{r ppa-hard}
ctrl_hard <- Control()
ctrl_hard$shading_model <- "ppa"
ctrl_hard$ppa_layer_smoothing <- 0      # hard step, no smoothing

tryCatch(
  run_scm(patch, ctrl = ctrl_hard),
  error = function(e) conditionMessage(e)
)
```

A *fixed* schedule (`run_scm(..., use_ode_times = TRUE)`) sidesteps the
error-control loop -- it simply takes the prescribed steps without rejecting any
-- but then nothing protects against a large explicit step overshooting the jump,
which sends growth to non-finite values unless the schedule is very fine. Either
way, a hard step is not something the solver can handle robustly.

Switching the rounding direction (light at the top vs the bottom of a layer)
does not help: the difficulty is the discontinuity itself, not where the step
sits.

## The smoothed profile

`ppa_layer_smoothing` replaces each hard boundary with a C^1^-continuous cubic
ramp over the top fraction of the layer (see `FF16_Environment::smooth_floor`).
The light profile -- and hence each plant's growth rate -- is now continuous with
a continuous first derivative, so the adaptive error estimate behaves as the
solver assumes and steps are accepted normally. Even a small amount of smoothing
is enough to recover a well-behaved integration, and with the default the
adaptive and fixed-step solutions agree.

Two parameters control the discretisation:

* `ppa_layer_optical_depth` -- the optical-depth thickness of one layer
  (default `0.5`, i.e. one unit of leaf area index per layer at the default
  extinction coefficient `k_I = 0.5`). Smaller values give finer layering.
* `ppa_layer_smoothing` -- the fraction of each layer over which the boundary is
  smoothed (default `0.3`). Values approaching `0` recover a hard step (and the
  instability above); `1` removes the flat region and approaches the smooth
  `"deep-crown"`-style profile. There is a tension here: smaller values are
  closer to a true PPA step but stiffer (slower, per the speed table above),
  while larger values integrate faster but blur the layer structure.

# Demographic equilibrium

The next test asks whether the population closes its own loop sensibly, and it
also supplies the resident we need for the fitness analysis that follows. Rather
than seed a stand at an arbitrary rate, we put it at its **demographic
equilibrium**: the seed rain at which the population exactly replaces itself
(net reproductive ratio R0 = 1). A stand seeded above or below this is over- or
under-crowded, and would cast an unrepresentative light environment.

The equilibrium is a fixed point of the patch's own seed production: put a seed
rain in, run the patch, read the seed rain out (`offspring_production`), and feed
it back. It converges in a handful of iterations.

```{r equil}
p_fit <- scm_base_parameters("FF16")
p_fit$max_patch_lifetime <- 50
lma_resident <- 0.0825

# Iterate seed-rain-out -> seed-rain-in until it stops moving (R0 = 1). Re-run
# this for any model / trait / max_patch_lifetime to re-derive the table below.
find_equilibrium_birth_rate <- function(model, n_iter = 6, birth_rate = 1) {
  ctrl <- Control()
  ctrl$shading_model <- model
  for (i in seq_len(n_iter)) {
    resident <- expand_parameters(trait_matrix(lma_resident, "lma"), p_fit,
                                  birth_rate_list = birth_rate)
    birth_rate <- run_scm(resident, ctrl = ctrl)$offspring_production
  }
  birth_rate
}
find_equilibrium_birth_rate("deep-crown")   # converges to ~1.66
```

Crucially, the equilibrium **depends on the shading model**: one that lets
plants capture more light supports a denser, more productive stand. Running the
same iteration for each model gives equilibria spanning an order of magnitude --
PPA's brighter layered light supports more than ten times the seed rain of
deep-crown. We hardcode them so the build does not re-run the slower models:

```{r equil-table}
equilibrium_birth_rate <- c(
  "deep-crown"        = 1.662,
  "mean-light"     = 2.178,
  "crown-centre"          = 2.185,
  "ppa"               = 18.835,
  "flat-top-soft-box" = 3.101
)
knitr::kable(data.frame(
  model = names(equilibrium_birth_rate),
  `equilibrium seed rain` = unname(equilibrium_birth_rate),
  check.names = FALSE
))
```

That an order-of-magnitude difference in predicted stand density follows from
the shading assumption alone is the first sign that these models, so alike in
their stand-development *trajectories*, are not interchangeable once the
population has to sustain itself.

# Mutant fitness

Stand dynamics and equilibrium both tolerated the cheap models reasonably well.
Invasion fitness is far more demanding. The fitness of a rare mutant is its
lifetime reproductive output as it grows through the light environment cast by an
established resident -- so it depends not on the light at one size, but on how
the light a plant experiences *changes as it grows*. A fitness landscape (fitness
versus a mutant trait) is the basis for evolutionary analysis, and is only useful
if it is smooth: kinks and jumps leave optima, selection gradients and
evolutionary endpoints ill-defined.

We compute invasion fitness for mutants spanning a range of leaf mass per area
(LMA), at fine trait resolution to expose any roughness, and run **two**
comparisons that differ only in the resident's seed rain:

1. **Same seed rain** -- every model uses the same resident seed input, so any
   difference is due purely to the shading model.
2. **Each at its own equilibrium** -- each resident uses its own equilibrium seed
   rain (from the table above), so each casts a realistic, self-consistent stand.

The resident is run once with `save_RK45_cache = TRUE` to cache its light
environment; mutants are introduced into that fixed landscape with
`run_mutant()`, and fitness is `log(R0)`.

```{r fitness, fig.width = 10, fig.height = 3.6}
# p_fit, lma_resident and equilibrium_birth_rate come from the section above.
mutant_lma <- seq(0.07, 0.11, length.out = 60)

fitness_landscape <- function(model, birth_rate) {
  ctrl <- Control()
  ctrl$shading_model <- model
  ctrl$save_RK45_cache <- TRUE                       # required by run_mutant()
  resident <- expand_parameters(trait_matrix(lma_resident, "lma"), p_fit,
                                birth_rate_list = birth_rate)
  scm <- run_scm(resident, ctrl = ctrl)              # caches the light environment
  mutants <- expand_parameters(trait_matrix(mutant_lma, "lma"), resident,
                               birth_rate_list = rep(1, length(mutant_lma)))
  scm$run_mutant(mutants)
  r0 <- scm$net_reproduction_ratios                  # resident first, then mutants
  data.frame(method = model, lma = mutant_lma,
             fitness = log(tail(r0, length(mutant_lma))))
}

# (1) all models at a common seed rain (the deep-crown equilibrium), and
# (2) each model at its own equilibrium seed rain.
common_rain <- equilibrium_birth_rate[["deep-crown"]]
fit_same  <- dplyr::bind_rows(lapply(models, fitness_landscape, birth_rate = common_rain))
fit_equil <- dplyr::bind_rows(lapply(models, function(m)
  fitness_landscape(m, equilibrium_birth_rate[[m]])))

fit_two <- dplyr::bind_rows(
  dplyr::mutate(fit_same,  seeding = "same seed rain"),
  dplyr::mutate(fit_equil, seeding = "each at its own equilibrium")
)
fit_two$method  <- factor(fit_two$method, levels = models)
fit_two$seeding <- factor(fit_two$seeding,
                          levels = c("same seed rain", "each at its own equilibrium"))

ggplot(fit_two, aes(lma, fitness, colour = seeding)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~method, nrow = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey70") +
  labs(x = "Mutant LMA", y = "Invasion fitness, log(R0)", colour = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

Two things stand out. First, **`"deep-crown"`, `"mean-light"` and
`"crown-centre"` yield smooth landscapes under both seedings, while `"ppa"` is
jagged** -- and at its own (much denser) equilibrium, wildly so, with a
non-monotonic dip where small trait changes flip which layer the mutant ends in.
For the deep-crown panel the two lines coincide, because "same seed rain" *is*
its own equilibrium -- a useful check.

Second, **the seeding assumption matters**: because the models reach different
equilibrium stands (the table above), their fitness landscapes diverge more when
each sits at its own equilibrium than when all are seeded alike. So a cheap
shading model can mislead in two compounding ways -- a different fitness gradient
in a given stand, *and* a different equilibrium stand to begin with.

The deeper lesson is about *continuity of the light profile*, not the cost of
the assimilation calculation. Invasion fitness needs the shading a plant
experiences to vary continuously with its size; deep-crown, mean-light and
crown-centre all build a continuous (smooth) light profile and differ only in how
they summarise assimilation, so all three give smooth fitness. PPA breaks this:
its light profile is a discretised step function of height, so as a mutant grows
its light -- and hence its fitness -- jumps. The layer smoothing
(`ppa_layer_smoothing`) that makes PPA's *dynamics* solvable softens but does not
remove this: a mutant still crosses layer boundaries as it grows, so PPA remains
the wrong tool for evolutionary analysis even when it is fine for a single stand
simulation.

## Getting the competition profile right

It is worth being clear about *why* crown-centre survives this test. It approximates
a plant's own assimilation crudely (a single crown-centre evaluation), but it
still computes the shade a plant *casts* from the full smooth Yokozawa crown.
The continuity that fitness needs comes from that competition profile, not from
the assimilation calculation.

This is exactly what the box models change, and -- as noted earlier -- the way
they change it mirrors the discrete-layer competition of large vegetation
models. They keep crown-centre's crown-centre assimilation but collapse the shade a
plant casts towards that same thin layer:

* `"flat-top-box"` -- a hard step: full shade below the crown centre, none above.
* `"flat-top-soft-box"` -- the same idea but smoothed into a continuous drop, so
  the light environment can still be built.

So the question is not whether these are crude -- they are the workhorse
representation in many large models -- but whether that crudeness is safe for the
evolutionary question we are now asking.

```{r compbox, fig.width = 7, fig.height = 3.6}
comp_profile <- function(model, height = 10) {
  s <- FF16_Strategy()
  s$control$shading_model <- model
  ind <- FF16_Individual(s)
  ind$set_state("height", height)
  z <- seq(0, height, length.out = 400)
  data.frame(model = model, z = z,
             competition = sapply(z, function(zz) ind$compute_competition(zz)))
}
comp <- dplyr::bind_rows(
  comp_profile("crown-centre"),
  comp_profile("flat-top-box"),
  comp_profile("flat-top-soft-box")
)

ggplot(comp, aes(competition, z, colour = model)) +
  geom_path() +
  labs(x = "Shade cast (competition) by a 10 m plant", y = "Height (m)",
       colour = "Model") +
  theme_classic()
```

The correct profile (crown-centre) tapers continuously to zero through the crown;
the box profile is a step, and the soft box a continuous-but-wrong drop
concentrated near the crown centre.

**The hard step does not even run.** Summed over a population of discrete
cohorts, a step competition makes the patch light profile a discontinuous,
piecewise-constant function of height, which the adaptive light-environment
spline cannot represent -- the model fails before a stand can develop:

```{r box-fail}
ctrl_box <- Control()
ctrl_box$shading_model <- "flat-top-box"
tryCatch(
  run_scm(patch, ctrl = ctrl_box),
  error = function(e) conditionMessage(e)
)
```

**The soft box runs, but gives the wrong answer.** Because its competition
profile is continuous, the light environment builds and a fitness landscape can
be computed -- but the shade is mis-distributed through the canopy, so the
landscape is biased relative to the deep-crown reference. The *correct* crown-centre
tracks deep-crown; the soft box does not.

```{r softbox-fitness, fig.width = 7, fig.height = 4}
box_models <- c("deep-crown", "crown-centre", "flat-top-soft-box")
fit_compare <- dplyr::bind_rows(lapply(box_models, function(m)
  fitness_landscape(m, equilibrium_birth_rate[[m]])))  # each at its own equilibrium
fit_compare$method <- factor(fit_compare$method, levels = box_models)

ggplot(fit_compare, aes(lma, fitness, colour = method)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(x = "Mutant LMA", y = "Invasion fitness, log(R0)", colour = "Model") +
  theme_classic()
```

So getting the competition profile right is not a refinement but a prerequisite.
The result generalises the PPA story: a plant model can tolerate a crude
*assimilation* approximation (crown-centre, mean-light) and still behave well for
both dynamics and fitness, but the shade each plant *casts* as a function of
height must be both continuous (or the light environment is ill-defined, as with
the hard box) and the right shape (or evolutionary predictions are biased, as
with the soft box).

None of this means the large models are wrong to use discrete layers: for
stand-level simulation on a fixed timestep, the representation is fast, robust,
and entirely adequate. The caution is specific to the question `plant` is built
for. Resolving how traits evolve needs invasion fitness, invasion fitness needs
continuous responses to size, and that is exactly what a layered competition
profile does not provide.

This is worth flagging as trait-based DGVMs (FATES and the like) are
increasingly pushed towards eco-evolutionary questions -- many strategies
competing, and selection on their traits. For strategies that differ *markedly*
the discretisation is harmless: the fitness differences between them dwarf the
layer-boundary noise, so coexistence or exclusion of distinct types is captured
fine. The trouble is confined to the opposite, and most delicate, case --
discriminating mutants that differ only slightly. Two near-identical strategies
can fall on opposite sides of a layer boundary and receive incomparable
fitnesses, so the selection gradient between them -- the very quantity that
decides which way a trait evolves -- becomes erratic. A model can give
sensible-looking community simulations, and resolve clearly distinct strategies,
yet still be unable to say which of two similar strategies wins.

# The evolutionary attractor

The invasion-fitness section asked whether each model gives a *smooth* fitness
landscape -- essentially a yes/no question about whether evolutionary analysis is
possible at all. The smooth-competition models passed. But a usable landscape is
not the point in itself: the reason we want one is to ask **where evolution
settles**, and whether the methods that agree a landscape is smooth also agree on
its endpoint. They do not -- and the way they part company is the subtlest lesson
in this vignette.

## Finding the singular strategy

The optimum of an invasion-fitness landscape is not a fixed target. As the
resident trait shifts, the resident *stand* it builds shifts too, and with it the
light environment every mutant grows through -- so the landscape itself moves. The
evolutionary endpoint is the trait that is its own best reply: the **singular
strategy** $s^\*$, where the selection gradient vanishes. We locate it by
root-solving that gradient.

1. **Equilibrium resident.** For a candidate resident LMA, find the
   demographic-equilibrium seed rain (R0 = 1) -- the same fixed-point iteration as
   the equilibrium section, but re-solved for whatever resident we are testing, so
   the stand is always self-consistent.
2. **Selection gradient.** Introduce mutants just either side of the resident; the
   gradient is the slope of mutant $\log(R_0)$ at mutant = resident (a central
   finite difference of the same `run_mutant()` fitness used above).
3. **Root-find.** Move the resident along the gradient and repeat. The attractor
   is where the gradient crosses zero. A *negative slope* at the crossing means
   the gradient pushes the resident back toward $s^\*$ from either side, so
   evolution actually arrives there -- the strategy is **convergence stable**. A
   *negative fitness curvature* at $s^\*$ means no nearby mutant can invade -- it is
   **uninvadable** (an ESS). Both together make it a **CSS**: a continuously
   stable endpoint, where evolution both converges and stays. (A positive
   curvature would instead be a *branching* point, where the population splits.)

Mechanically this is just a one-dimensional root-find on the selection gradient --
far cheaper than mapping the whole landscape, since each step needs only the local
slope, not the full curve. We lift the machinery straight from the
invasion-fitness section: `equilibrium_birth_rate_at()` is the equilibrium
iteration re-parameterised by resident LMA, and `sel_gradient()` /
`fitness_curvature()` are finite-difference first and second derivatives of the
cached `run_mutant()` fitness.

```{r attractor-machinery}
# Demographic-equilibrium seed rain for a resident of given LMA (cf.
# find_equilibrium_birth_rate above, now a function of the resident trait).
equilibrium_birth_rate_at <- function(model, lma, n_iter = 6, birth_rate = 1) {
  ctrl <- Control(); ctrl$shading_model <- model
  for (i in seq_len(n_iter)) {
    resident <- expand_parameters(trait_matrix(lma, "lma"), p_fit,
                                  birth_rate_list = birth_rate)
    birth_rate <- run_scm(resident, ctrl = ctrl)$offspring_production
  }
  birth_rate
}

# Selection gradient d log(R0)/d(LMA) at mutant = resident, with the resident at
# its own demographic equilibrium. Central finite difference.
sel_gradient <- function(model, lma, delta = 0.0015, n_iter = 6) {
  ctrl <- Control(); ctrl$shading_model <- model; ctrl$save_RK45_cache <- TRUE
  br <- equilibrium_birth_rate_at(model, lma, n_iter)
  resident <- expand_parameters(trait_matrix(lma, "lma"), p_fit,
                                birth_rate_list = br)
  scm <- run_scm(resident, ctrl = ctrl)
  mutants <- expand_parameters(trait_matrix(c(lma - delta, lma + delta), "lma"),
                               resident, birth_rate_list = rep(1, 2))
  scm$run_mutant(mutants)
  f <- log(tail(scm$net_reproduction_ratios, 2))
  (f[2] - f[1]) / (2 * delta)
}

# Curvature d2 log(R0)/d(LMA)^2 at s* (uninvadability test): < 0 => CSS.
fitness_curvature <- function(model, s, delta = 0.0015, n_iter = 6) {
  ctrl <- Control(); ctrl$shading_model <- model; ctrl$save_RK45_cache <- TRUE
  br <- equilibrium_birth_rate_at(model, s, n_iter)
  resident <- expand_parameters(trait_matrix(s, "lma"), p_fit,
                                birth_rate_list = br)
  scm <- run_scm(resident, ctrl = ctrl)
  mutants <- expand_parameters(trait_matrix(c(s - delta, s, s + delta), "lma"),
                               resident, birth_rate_list = rep(1, 3))
  scm$run_mutant(mutants)
  f <- log(tail(scm$net_reproduction_ratios, 3))
  (f[3] - 2 * f[2] + f[1]) / delta^2
}
```

As a live check, we root-find the default `"deep-crown"` attractor over the LMA
interval where its gradient changes sign (read off the grid below). A single
`uniroot()` converges in a handful of equilibrium solves:

```{r attractor-live}
deep_crown_star <- uniroot(function(x) sel_gradient("deep-crown", x),
                           interval = c(0.100, 0.1025), tol = 1e-5)$root
deep_crown_star                                   # ~0.1007
```

Mapping the gradient over a grid and root-solving + curvature-testing all five
models is the expensive part -- a few minutes, since every grid point re-solves an
equilibrium and PPA is slow. So, exactly as with the equilibrium table, we
hardcode the precomputed results and plot from them. They were produced by
`examples/DunFalster2026/make_figure_attractor.R`, which writes out the gradient
grid and the singular-strategy table reproduced here.

```{r attractor-data}
attractor_gradients <- read.csv(strip.white = TRUE, blank.lines.skip = TRUE,
  text = "method,lma,gradient
deep-crown,0.0650,437.8
deep-crown,0.0675,410.9
deep-crown,0.0700,382.6
deep-crown,0.0725,353.4
deep-crown,0.0750,323.2
deep-crown,0.0775,292.2
deep-crown,0.0800,260.7
deep-crown,0.0825,229.0
deep-crown,0.0850,197.1
deep-crown,0.0875,165.2
deep-crown,0.0900,133.4
deep-crown,0.0925,101.8
deep-crown,0.0950,70.5
deep-crown,0.0975,39.5
deep-crown,0.1000,8.9
deep-crown,0.1025,-21.2
deep-crown,0.1050,-50.8
deep-crown,0.1075,-79.8
deep-crown,0.1100,-108.2
deep-crown,0.1125,-136.0
deep-crown,0.1150,-163.2
deep-crown,0.1175,-189.8
deep-crown,0.1200,-215.8
deep-crown,0.1225,-241.3
deep-crown,0.1250,-266.1
mean-light,0.0650,496.5
mean-light,0.0675,466.6
mean-light,0.0700,434.0
mean-light,0.0725,399.1
mean-light,0.0750,362.5
mean-light,0.0775,324.5
mean-light,0.0800,285.4
mean-light,0.0825,245.6
mean-light,0.0850,205.1
mean-light,0.0875,164.3
mean-light,0.0900,123.4
mean-light,0.0925,82.5
mean-light,0.0950,41.8
mean-light,0.0975,1.3
mean-light,0.1000,-38.9
mean-light,0.1025,-78.6
mean-light,0.1050,-117.9
mean-light,0.1075,-156.7
mean-light,0.1100,-195.1
mean-light,0.1125,-233.0
mean-light,0.1150,-270.2
mean-light,0.1175,-306.9
mean-light,0.1200,-343.2
mean-light,0.1225,-379.0
mean-light,0.1250,-414.2
crown-centre,0.0650,571.2
crown-centre,0.0675,531.3
crown-centre,0.0700,487.5
crown-centre,0.0725,440.0
crown-centre,0.0750,389.4
crown-centre,0.0775,336.6
crown-centre,0.0800,281.9
crown-centre,0.0825,226.1
crown-centre,0.0850,169.5
crown-centre,0.0875,112.4
crown-centre,0.0900,54.8
crown-centre,0.0925,-2.5
crown-centre,0.0950,-59.2
crown-centre,0.0975,-115.4
crown-centre,0.1000,-171.1
crown-centre,0.1025,-225.7
crown-centre,0.1050,-279.6
crown-centre,0.1075,-332.4
crown-centre,0.1100,-384.2
crown-centre,0.1125,-434.7
crown-centre,0.1150,-484.2
crown-centre,0.1175,-532.7
crown-centre,0.1200,-579.9
crown-centre,0.1225,-626.2
crown-centre,0.1250,-671.0
flat-top-soft-box,0.0650,717.9
flat-top-soft-box,0.0675,662.2
flat-top-soft-box,0.0700,600.0
flat-top-soft-box,0.0725,532.7
flat-top-soft-box,0.0750,461.6
flat-top-soft-box,0.0775,387.6
flat-top-soft-box,0.0800,312.1
flat-top-soft-box,0.0825,235.5
flat-top-soft-box,0.0850,158.5
flat-top-soft-box,0.0875,81.0
flat-top-soft-box,0.0900,3.7
flat-top-soft-box,0.0925,-73.5
flat-top-soft-box,0.0950,-150.6
flat-top-soft-box,0.0975,-227.7
flat-top-soft-box,0.1000,-304.6
flat-top-soft-box,0.1025,-381.6
flat-top-soft-box,0.1050,-458.8
flat-top-soft-box,0.1075,-536.6
flat-top-soft-box,0.1100,-615.1
flat-top-soft-box,0.1125,-694.2
flat-top-soft-box,0.1150,-774.1
flat-top-soft-box,0.1175,-854.8
flat-top-soft-box,0.1200,-936.9
flat-top-soft-box,0.1225,-1020.0
flat-top-soft-box,0.1250,-1104.0
ppa,0.0650,3461.1
ppa,0.0675,2945.3
ppa,0.0700,3120.4
ppa,0.0725,2647.5
ppa,0.0750,2729.6
ppa,0.0775,2445.4
ppa,0.0800,2650.1
ppa,0.0825,1093.1
ppa,0.0850,271.6
ppa,0.0875,-310.5
ppa,0.0900,-2335.3
ppa,0.0925,-2840.7
ppa,0.0950,-3062.7
ppa,0.0975,-2961.9
ppa,0.1000,-3913.1
ppa,0.1025,-3968.6
ppa,0.1050,-4075.2
ppa,0.1075,-2792.7
ppa,0.1100,-2631.3
ppa,0.1125,-2557.8
ppa,0.1150,-4407.0
ppa,0.1175,-86.9
ppa,0.1200,-4012.5
ppa,0.1225,-461.9
ppa,0.1250,-167.4")

singular_strategies <- read.csv(strip.white = TRUE, blank.lines.skip = TRUE,
  text = "method,lma_star,curvature,class,robust
deep-crown,0.1007,-12477,CSS,TRUE
mean-light,0.0976,-17176,CSS,TRUE
crown-centre,0.0924,-20663,CSS,TRUE
flat-top-soft-box,0.0901,-30828,CSS,TRUE
ppa,0.0859,-250149,CSS,FALSE")

model_order <- c("deep-crown", "mean-light", "crown-centre",
                 "flat-top-soft-box", "ppa")
attractor_gradients$method <- factor(attractor_gradients$method, levels = model_order)
singular_strategies$method <- factor(singular_strategies$method, levels = model_order)
```

```{r attractor-table}
knitr::kable(data.frame(
  model                   = as.character(singular_strategies$method),
  `attracted LMA`         = round(singular_strategies$lma_star, 4),
  `equilibrium seed rain` = unname(equilibrium_birth_rate[as.character(singular_strategies$method)]),
  `fitness curvature`     = singular_strategies$curvature,
  classification          = ifelse(singular_strategies$robust,
                                   singular_strategies$class,
                                   paste0(singular_strategies$class, " (stepped)")),
  check.names = FALSE, row.names = NULL))
```

```{r attractor-fig, fig.width = 8, fig.height = 7.5}
shape_vals <- c(`TRUE` = 19, `FALSE` = 1)   # filled = robust; open = stepped (PPA)

p_grad <- ggplot(attractor_gradients, aes(lma, gradient, colour = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.9) +
  geom_point(data = singular_strategies, aes(lma_star, 0, shape = robust),
             size = 3) +
  scale_shape_manual(values = shape_vals, guide = "none") +
  coord_cartesian(ylim = c(-400, 400)) +     # PPA's spikes reach ~4000; clip them
  labs(x = "Resident LMA", y = "Selection gradient,  d log(R0) / d(LMA)",
       colour = "Shading model",
       subtitle = "Zero crossing with negative slope = attractor (PPA clipped: |gradient| up to ~4000)") +
  theme_classic()

p_star <- ggplot(singular_strategies, aes(method, lma_star, colour = method)) +
  geom_point(aes(shape = robust), size = 3.5, show.legend = FALSE) +
  geom_text(aes(label = formatC(lma_star, format = "f", digits = 4)),
            vjust = -1.2, size = 3, show.legend = FALSE) +
  scale_shape_manual(values = shape_vals, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.15))) +
  labs(x = NULL, y = "Attracted LMA  (singular strategy)",
       subtitle = "Where evolution settles: the methods disagree (~15% spread)") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))

(p_grad / p_star) + plot_layout(heights = c(1.2, 1))
```

## The methods disagree on where evolution settles

Three things stand out.

**Every method has a genuine endpoint -- but they are different endpoints.** All
five gradients cross zero exactly once, each with a negative slope (convergence
stable) and a negative fitness curvature (uninvadable), so each is a CSS: a real
evolutionary stopping point, not a branching point. Yet the attracted LMA marches
steadily across the methods -- deep-crown `0.1007`, mean-light `0.0976`,
crown-centre `0.0924`, flat-top-soft-box `0.0901`, ppa `0.0859` -- a roughly 15%
spread.

**That disagreement is invisible at the rungs below.** This is the key new
lesson. deep-crown, mean-light and crown-centre agreed on stand dynamics and gave
fitness landscapes of the same *shape* (the previous section) -- yet they settle
on visibly different LMA. The mechanism ties straight back to demographic
equilibrium: each method supports a different equilibrium *density* (seed rains
`1.66 / 2.18 / 2.18 / 3.10 / 18.8` for the table above), so the realised light
environment a mutant grows through differs, which shifts the selection gradient
and moves its zero. A shading model can therefore be perfectly "safe" for the
*shape* of the fitness landscape and still relocate the predicted optimum --
because it gets the stand density wrong. Smoothness is necessary for evolutionary
analysis; it is not sufficient for the answer to be right.

**PPA reaches an endpoint, but a fragile one.** PPA's gradient does cross zero (at
the lowest LMA of all), but the crossing is buried in the staircase its discrete
layers impose: the gradient is enormous (clipped in the figure -- `|gradient|` up
to ~4000), takes a discontinuous jump just above the attractor, is erratic in the
high-LMA tail, and its curvature is roughly an order of magnitude steeper than the
smooth models'. So while a singular strategy formally exists, its location is set
by where the layer boundaries happen to fall -- nudge `ppa_layer_optical_depth`
and it moves. This is the same verdict as the invasion-fitness section, sharpened:
PPA is the wrong tool for evolutionary questions, and the attractor it reports
should not be trusted.
