---
title: "Canopy shading methods"
output: bookdown::html_document2

---

**authors**: DS Falster
**date**: 2026

# Background

To grow a plant, `plant` needs to know how much light each individual
intercepts. That depends on two things: how a plant's own leaf area is
distributed through its crown (which sets the light it captures for
photosynthesis), and how that leaf area shades everything below it (which sets
the patch-level light profile). Both can be modelled at different levels of
realism, trading detail for speed.

The `FF16` strategy offers four shading models, selected with
`control$shading_model`:

* **`"deep-crown"`** (the default) -- the original `plant` behaviour. Leaf area
  is distributed continuously through the crown following the Yokozawa (1995)
  foliage profile, and a plant's gross assimilation is computed by *integrating*
  photosynthesis over crown depth against the light profile. The patch light
  profile is a smooth function of height.

* **`"average-light"`** -- partway between deep-crown and flat-top. Rather than
  integrating the *photosynthetic rate* over depth, it integrates the *light*
  over depth (weighted by leaf-area density) to a single leaf-area-weighted mean
  light, then takes one photosynthesis evaluation of that mean. It therefore
  still feels the within-crown light environment -- and how that environment
  shifts as the canopy develops -- but collapses the expensive photosynthesis
  calculation to a *single* call per plant rather than one per crown depth. For
  FF16 that saving is modest (its leaf photosynthesis is a cheap saturating
  function), but it becomes important for models with an expensive leaf
  submodel: the `TF24` strategy's coupled stomatal/hydraulic optimisation is far
  more costly to evaluate, so integrating the light and optimising once is the
  natural choice there. The cost is bias: because photosynthesis saturates with
  light, evaluating the rate at the mean light overestimates the depth-averaged
  rate (Jensen's inequality). So `"average-light"` always gives at least as much
  assimilation as `"deep-crown"`, the two agreeing only when the within-crown
  light is uniform.

* **`"flat-top"`** -- the light profile is built exactly as for `"deep-crown"`,
  but a plant's assimilation is approximated by a *single evaluation* of the
  light at its crown centre rather than an integral over depth. This skips the
  per-plant quadrature, so it is faster, and removes within-crown self-shading.

* **`"ppa"`** -- a perfect-plasticity-style approximation. The patch light
  profile is discretised into discrete canopy *layers*: light steps down by a
  fixed amount of optical depth (`control$ppa_layer_optical_depth`) per layer.
  Assimilation is evaluated at the crown centre as in `"flat-top"`, but now
  reads this stepped profile. Because a hard step is not differentiable (and
  makes the ODE solver fail), the layer boundaries are smoothed with a
  C^1^-continuous ramp whose width is set by `control$ppa_layer_smoothing`.

All four share the *same* per-plant competition contribution (the smooth
Yokozawa leaf-area profile); they differ only in how assimilation is evaluated
and, for PPA, how the light profile is assembled. The model is resolved once
when a strategy is prepared, so there is no per-call cost to having the choice
available.

```{r, message = FALSE}
library(plant)
library(dplyr)
library(ggplot2)
library(patchwork)
```

# Simulating a patch under each model

We grow a single `FF16` species in a patch, and run the same patch under each
of the four shading models. Everything except `control$shading_model` is held
fixed, so any differences are due to the shading model alone.

```{r}
params <- scm_base_parameters("FF16")
patch  <- expand_parameters(trait_matrix(0.0825, "lma"), params)

models <- c("deep-crown", "average-light", "flat-top", "ppa")

run_model <- function(model) {
  ctrl <- Control()
  ctrl$shading_model <- model
  run_scm(patch, ctrl = ctrl, collect = TRUE)
}

results <- lapply(models, run_model)
names(results) <- models
```

Each run returns the full collected state (`run_scm(collect = TRUE)`): a tidy
`species` table (one row per node per step) and the light environment
(`env$light_availability`). We tag each run's `species` table with its shading
model and bind them together so we can use `plant`'s tidy helpers across all
four.

```{r}
species_all <- dplyr::bind_rows(
  lapply(models, function(m) dplyr::mutate(results[[m]]$species, method = m))
)
species_all$method <- factor(species_all$method, levels = models)

# Light-extinction coefficient, used throughout to convert canopy openness to
# cumulative leaf area: E = exp(-k_I * LAI), so LAI_above = -log(E) / k_I.
k_I <- results[[1]]$p$strategies[[1]]$k_I
```

# Size-distribution dynamics

The size distribution is the set of node height trajectories: each line is a
cohort introduced at a given patch age, growing (and being shaded) over time,
shaded by node density. `plant`'s `plot_size_distribution()` builds this
directly from the tidy `species` table; faceting by `method` compares the
shading models.

```{r sizedist, fig.width = 10, fig.height = 3.4}
p_size <- plot_size_distribution(species_all) +
  facet_wrap(~method) +
  theme(legend.position = "bottom")
p_size
```

The discrete layering in `"ppa"` produces visibly different growth trajectories:
cohorts grow at near-constant light within a layer, then accelerate as they
emerge into a brighter layer.

# Leaf area index dynamics

Leaf area index (LAI) -- the total one-sided leaf area per unit ground -- is the
cumulative leaf area above the ground, which is already encoded in the recorded
light profile: with Beer's law $E = \exp(-k_I\,\mathrm{LAI})$, the patch LAI is
just $-\log E(0) / k_I$, where $E(0)$ is canopy openness at ground level. This
reads straight off `env$light_availability` and needs no integration over the
size distribution. (LAI is real leaf area, so the smooth recorded profile gives
the true value for all four models -- the PPA layering only affects the light
plants *experience*, not the leaf area present.)

```{r}
lai_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    results[[m]]$env$light_availability |>
      dplyr::group_by(time) |>
      dplyr::slice_min(height, n = 1, with_ties = FALSE) |>
      dplyr::ungroup() |>
      dplyr::transmute(method = m, time, lai = -log(light_availability) / k_I)
  })
)
lai_all$method <- factor(lai_all$method, levels = models)
```

```{r lai, fig.width = 7, fig.height = 4}
p_lai <- ggplot(lai_all, aes(time, lai, colour = method)) +
  geom_line(linewidth = 1) +
  labs(x = "Patch age (years)", y = "Leaf area index", colour = "Shading model") +
  theme_classic()
p_lai
```

`"deep-crown"`, `"average-light"` and `"flat-top"` track each other closely and
settle to a steady LAI: under a near-uniform light profile, all three ways of
summarising within-crown light give similar carbon gain (with
`"average-light"` and `"flat-top"` running slightly richer, since they do not
fully resolve self-shading). The `"ppa"` model instead shows large, sustained
oscillations -- successive waves of recruitment and self-thinning -- because the
discrete light layers create sharp thresholds for cohort success (visible as the
pulses of emerging cohorts in the size-distribution panel above).

# Vertical light profile

We can also look *within* the canopy at a few patch ages, at the vertical
profile of cumulative leaf area above each height -- $\mathrm{LAI}_{above}(z) =
-\log E(z) / k_I$, where $E$ is canopy openness. This is the quantity that sets
how much each individual is shaded.

The recorded `env$light_availability` is the smooth underlying profile shared by
all models. The light that plants actually *experience* under PPA is the stepped
version, so for the `"ppa"` panel we apply the same layering transform used
internally (see `FF16_Environment::step_light`).

```{r}
d <- Control()$ppa_layer_optical_depth   # layer thickness (optical depth)
w <- Control()$ppa_layer_smoothing       # boundary smoothing fraction

# C1-smoothed staircase: flat over the lower (1 - w) of each layer, cubic ramp
# over the top w. Mirrors FF16_Environment::smooth_floor() / step_light().
smooth_floor <- function(u, w) {
  n <- floor(u)
  f <- u - n
  t <- pmax(0, f - (1 - w)) / w
  n + ifelse(f <= 1 - w, 0, t^2 * (3 - 2 * t))
}
ppa_step_light <- function(E, d, w) {
  ifelse(E >= 1 | E <= 0, E,
         exp(-d * smooth_floor(-log(E) / d, w)))
}

target_ages <- c(5, 10, 20, 60)

profile_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    light <- results[[m]]$env$light_availability
    if (m == "ppa") {
      light$light_availability <- ppa_step_light(light$light_availability, d, w)
    }
    # nearest recorded step to each target age
    steps <- results[[m]]$steps
    keep <- sapply(target_ages, function(a) steps$step[which.min(abs(steps$time - a))])
    light |>
      dplyr::filter(step %in% keep) |>
      dplyr::mutate(method = m,
                    age = target_ages[match(step, keep)],
                    lai_above = -log(light_availability) / k_I)
  })
)
profile_all$method <- factor(profile_all$method, levels = models)
```

```{r vprofile, fig.width = 10, fig.height = 3.6}
p_profile <- ggplot(profile_all,
                    aes(lai_above, height, colour = factor(age), group = age)) +
  geom_path() +
  facet_wrap(~method) +
  labs(x = expression(LAI[above]~(m^2/m^2)), y = "Height (m)",
       colour = "Patch age\n(years)") +
  theme_classic()
p_profile
```

`"deep-crown"`, `"average-light"` and `"flat-top"` share an identical smooth
profile (they build the light environment the same way and differ only in how
assimilation is evaluated); `"ppa"` shows the characteristic staircase, with
cumulative leaf area held constant within a layer and jumping at each layer
boundary.

# Comparison panel

```{r panel, fig.width = 9, fig.height = 10}
(p_size / p_lai  / p_profile) + patchwork::plot_annotation(tag_levels = "A")
```

# Speed comparison

Two effects compete. Per ODE step, `"deep-crown"` is the most expensive because
it performs a Gauss-Kronrod quadrature over crown depth for every individual,
whereas `"flat-top"` and `"ppa"` replace this with a single light evaluation. But
the *number* of steps also matters: PPA's layered light profile is much stiffer,
so the adaptive solver takes many more, smaller steps. Which effect dominates
depends on how the system is stepped.

```{r speed-helper}
# Median wall-clock over a few repeats. `times = NULL` uses the adaptive solver;
# supplying a fixed schedule pins every model to the same steps.
time_model <- function(model, times = NULL, reps = 5) {
  p <- patch
  use_fixed <- !is.null(times)
  if (use_fixed) p$ode_times <- times
  ctrl <- Control()
  ctrl$shading_model <- model
  run1 <- function() run_scm(p, ctrl = ctrl, use_ode_times = use_fixed)
  run1()                                   # warm up
  median(replicate(reps, system.time(run1())[["elapsed"]]))
}

speed_table <- function(timings) {
  data.frame(
    model = names(timings),
    `median time (s)` = round(unname(timings), 4),
    `relative to deep-crown` = round(unname(timings / timings[["deep-crown"]]), 2),
    check.names = FALSE,
    row.names = NULL
  )
}
```

## Adaptive solver (real-world cost)

This is how the models run by default, and is the cost you actually pay.

```{r speed-adaptive}
timings_adaptive <- vapply(models, time_model, numeric(1))
knitr::kable(speed_table(timings_adaptive))
```

`"flat-top"` is fastest -- it shares deep-crown's smooth light profile (so the
same number of solver steps) but skips the per-plant crown-depth integral.
`"average-light"` costs about the same as `"deep-crown"`: it also integrates over
crown depth (just the light rather than the rate), so the per-step work and the
step count are both similar. `"ppa"` is *slower* overall despite the cheaper
per-plant assimilation, because its near-stepped profile forces the adaptive
solver to take many more steps; finer layers (smaller `ppa_layer_optical_depth`)
or sharper boundaries (smaller `ppa_layer_smoothing`) make this worse.

## Fixed ODE schedule (per-step cost)

If instead every model is pinned to the *same* fixed schedule
(`run_scm(..., use_ode_times = TRUE)`), the step-count effect is removed and only
the per-step cost remains. Now PPA's advantage shows: like `"flat-top"`, it
skips the crown-depth quadrature, so both are markedly faster than
`"deep-crown"`.

```{r speed-fixed}
times <- seq(0, params$max_patch_lifetime, length.out = 4000)
timings_fixed <- vapply(models, time_model, numeric(1), times = times)
knitr::kable(speed_table(timings_fixed))
```

So PPA is not inherently slow -- on a fixed schedule it is *faster* than
`"deep-crown"`. Its apparent cost under the adaptive solver comes entirely from
the extra steps the stiff profile demands. Where a fixed schedule is acceptable
(or where the layering is coarse and lightly smoothed), PPA can be the cheaper
choice. `"deep-crown"` nonetheless remains the default: it is the most faithful
representation of within-crown light capture and carries no speed penalty
relative to earlier versions of `plant`.

# Why PPA needs smoothing

A literal PPA discretisation is a *hard* step function: as a plant grows and its
crown centre crosses a layer boundary, the light it experiences jumps
discontinuously, and so does its growth rate. This is what `ppa_layer_smoothing`
exists to tame, and it is worth being explicit about *why* -- because the
problem is specifically one of **adaptive ODE stepping**, not of the model
itself.

## Adaptive stepping cannot integrate a discontinuity

`plant` solves the patch dynamics with an adaptive, error-controlled solver
(an embedded Runge-Kutta scheme). At each step it forms two estimates of the
state at different orders, and uses their difference as an estimate of the local
truncation error. If that error exceeds the requested tolerance, the step is
*rejected* and retried with a smaller size; the method assumes the solution is
smooth, so halving the step should roughly quarter the error.

Across a discontinuity that assumption breaks. No matter how small the step, a
step that straddles the jump carries an O(jump) error that never falls below
tolerance, so the solver rejects and shrinks the step indefinitely until it
gives up. With a hard step (`ppa_layer_smoothing = 0`) this is exactly what
happens:

```{r ppa-hard}
ctrl_hard <- Control()
ctrl_hard$shading_model <- "ppa"
ctrl_hard$ppa_layer_smoothing <- 0      # hard step, no smoothing

tryCatch(
  run_scm(patch, ctrl = ctrl_hard),
  error = function(e) conditionMessage(e)
)
```

A *fixed* schedule (`run_scm(..., use_ode_times = TRUE)`) sidesteps the
error-control loop -- it simply takes the prescribed steps without rejecting any
-- but then nothing protects against a large explicit step overshooting the jump,
which sends growth to non-finite values unless the schedule is very fine. Either
way, a hard step is not something the solver can handle robustly.

Switching the rounding direction (light at the top vs the bottom of a layer)
does not help: the difficulty is the discontinuity itself, not where the step
sits.

## The smoothed profile

`ppa_layer_smoothing` replaces each hard boundary with a C^1^-continuous cubic
ramp over the top fraction of the layer (see `FF16_Environment::smooth_floor`).
The light profile -- and hence each plant's growth rate -- is now continuous with
a continuous first derivative, so the adaptive error estimate behaves as the
solver assumes and steps are accepted normally. Even a small amount of smoothing
is enough to recover a well-behaved integration, and with the default the
adaptive and fixed-step solutions agree.

Two parameters control the discretisation:

* `ppa_layer_optical_depth` -- the optical-depth thickness of one layer
  (default `0.5`, i.e. one unit of leaf area index per layer at the default
  extinction coefficient `k_I = 0.5`). Smaller values give finer layering.
* `ppa_layer_smoothing` -- the fraction of each layer over which the boundary is
  smoothed (default `0.3`). Values approaching `0` recover a hard step (and the
  instability above); `1` removes the flat region and approaches the smooth
  `"deep-crown"`-style profile. There is a tension here: smaller values are
  closer to a true PPA step but stiffer (slower, per the speed table above),
  while larger values integrate faster but blur the layer structure.

# Mutant fitness

So far the models have looked broadly interchangeable for *stand dynamics*: the
size distribution and LAI trajectories differ in detail but not in character,
and the simpler, cheaper models reproduce the overall behaviour reasonably well.
Invasion fitness is a far more demanding test. The fitness of a rare mutant is
its lifetime reproductive output while it grows through the light environment
cast by an established resident -- so it depends not just on the light at one
size, but on how the light a plant experiences *changes smoothly as it grows*. A
fitness landscape (fitness versus a mutant trait) is the basis for evolutionary
analysis, and it is only useful if it is smooth: kinks and jumps make optima,
gradients and evolutionary endpoints ill-defined.

We compute the invasion fitness of mutants spanning a range of leaf mass per
area (LMA) against a single resident, at fine trait resolution to expose any
roughness. The resident is run once with `save_RK45_cache = TRUE` so its
competitive light environment is cached; mutants are then introduced into that
fixed landscape with `run_mutant()`, and fitness is `log(R0)`, the log net
reproductive output.

```{r fitness, fig.width = 10, fig.height = 3.4}
p_fit <- scm_base_parameters("FF16")
p_fit$max_patch_lifetime <- 50
lma_resident <- 0.0825
mutant_lma   <- seq(0.07, 0.11, length.out = 60)

fitness_landscape <- function(model) {
  ctrl <- Control()
  ctrl$shading_model <- model
  ctrl$save_RK45_cache <- TRUE                       # required by run_mutant()
  resident <- expand_parameters(trait_matrix(lma_resident, "lma"), p_fit,
                                birth_rate_list = 1)
  scm <- run_scm(resident, ctrl = ctrl)              # caches the light environment
  mutants <- expand_parameters(trait_matrix(mutant_lma, "lma"), resident,
                               birth_rate_list = rep(1, length(mutant_lma)))
  scm$run_mutant(mutants)
  r0 <- scm$net_reproduction_ratios                  # resident first, then mutants
  data.frame(method = model, lma = mutant_lma,
             fitness = log(tail(r0, length(mutant_lma))))
}

fit_all <- dplyr::bind_rows(lapply(models, fitness_landscape))
fit_all$method <- factor(fit_all$method, levels = models)

ggplot(fit_all, aes(lma, fitness)) +
  geom_line() +
  geom_point(size = 0.5) +
  facet_wrap(~method, nrow = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(x = "Mutant LMA", y = "Invasion fitness, log(R0)") +
  theme_classic()
```

`"deep-crown"`, `"average-light"` and `"flat-top"` all yield smooth fitness
landscapes, and the two cheap approximations stay close to the deep-crown
reference -- so for fitness, as for stand dynamics, they are perfectly usable.
`"ppa"` is the exception: its landscape is jagged, with a sharp discontinuity
where a small change in the mutant trait flips which canopy layer the mutant
ends up in.

The lesson is about *continuity of the light profile*, not the cost of the
assimilation calculation. Invasion fitness needs the shading a plant experiences
to vary continuously with its size; deep-crown, average-light and flat-top all
build a continuous (smooth) light profile and differ only in how they summarise
assimilation, so all three give smooth fitness. PPA breaks this: its light
profile is a discretised step function of height, so as a mutant grows its light
-- and hence its fitness -- jumps. The layer smoothing (`ppa_layer_smoothing`)
that makes PPA's *dynamics* solvable softens but does not remove this: a mutant
still crosses layer boundaries as it grows, so PPA remains the wrong tool for
evolutionary analysis even when it is fine for a single stand simulation.

## Getting the competition profile right

It is worth being clear about *why* flat-top survives this test. It approximates
a plant's own assimilation crudely (a single crown-centre evaluation), but it
still computes the shade a plant *casts* from the full smooth Yokozawa crown.
The continuity that fitness needs comes from that competition profile, not from
the assimilation calculation.

To show this is what matters, two further variants keep flat-top's crown-centre
assimilation but get the *competition* wrong, collapsing each plant's leaf area
towards the thin crown-centre layer instead of spreading it through the crown:

* `"flat-top-box"` -- a hard step: full shade below the crown centre, none above.
* `"flat-top-soft-box"` -- the same idea but smoothed into a continuous drop, so
  the light environment can still be built.

```{r compbox, fig.width = 7, fig.height = 3.6}
comp_profile <- function(model, height = 10) {
  s <- FF16_Strategy()
  s$control$shading_model <- model
  ind <- FF16_Individual(s)
  ind$set_state("height", height)
  z <- seq(0, height, length.out = 400)
  data.frame(model = model, z = z,
             competition = sapply(z, function(zz) ind$compute_competition(zz)))
}
comp <- dplyr::bind_rows(
  comp_profile("flat-top"),
  comp_profile("flat-top-box"),
  comp_profile("flat-top-soft-box")
)

ggplot(comp, aes(competition, z, colour = model)) +
  geom_path() +
  labs(x = "Shade cast (competition) by a 10 m plant", y = "Height (m)",
       colour = "Model") +
  theme_classic()
```

The correct profile (flat-top) tapers continuously to zero through the crown;
the box profile is a step, and the soft box a continuous-but-wrong drop
concentrated near the crown centre.

**The hard step does not even run.** Summed over a population of discrete
cohorts, a step competition makes the patch light profile a discontinuous,
piecewise-constant function of height, which the adaptive light-environment
spline cannot represent -- the model fails before a stand can develop:

```{r box-fail}
ctrl_box <- Control()
ctrl_box$shading_model <- "flat-top-box"
tryCatch(
  run_scm(patch, ctrl = ctrl_box),
  error = function(e) conditionMessage(e)
)
```

**The soft box runs, but gives the wrong answer.** Because its competition
profile is continuous, the light environment builds and a fitness landscape can
be computed -- but the shade is mis-distributed through the canopy, so the
landscape is biased relative to the deep-crown reference. The *correct* flat-top
tracks deep-crown; the soft box does not.

```{r softbox-fitness, fig.width = 7, fig.height = 4}
fit_compare <- dplyr::bind_rows(
  fitness_landscape("deep-crown"),
  fitness_landscape("flat-top"),
  fitness_landscape("flat-top-soft-box")
)
fit_compare$method <- factor(
  fit_compare$method,
  levels = c("deep-crown", "flat-top", "flat-top-soft-box")
)

ggplot(fit_compare, aes(lma, fitness, colour = method)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(x = "Mutant LMA", y = "Invasion fitness, log(R0)", colour = "Model") +
  theme_classic()
```

So getting the competition profile right is not a refinement but a prerequisite.
The result generalises the PPA story: a plant model can tolerate a crude
*assimilation* approximation (flat-top, average-light) and still behave well for
both dynamics and fitness, but the shade each plant *casts* as a function of
height must be both continuous (or the light environment is ill-defined, as with
the hard box) and the right shape (or evolutionary predictions are biased, as
with the soft box).
