---
title: "Canopy shading methods"
output: bookdown::html_document2

---

**authors**: DS Falster
**date**: 2026

# Background

To grow a plant, `plant` needs to know how much light each individual
intercepts. That depends on two things: how a plant's own leaf area is
distributed through its crown (which sets the light it captures for
photosynthesis), and how that leaf area shades everything below it (which sets
the patch-level light profile). Both can be modelled at different levels of
realism, trading detail for speed.

The `FF16` strategy offers three shading models, selected with
`control$shading_model`:

* **`"deep-crown"`** (the default) -- the original `plant` behaviour. Leaf area
  is distributed continuously through the crown following the Yokozawa (1995)
  foliage profile, and a plant's gross assimilation is computed by *integrating*
  photosynthesis over crown depth against the light profile. The patch light
  profile is a smooth function of height.

* **`"flat-top"`** -- the light profile is built exactly as for `"deep-crown"`,
  but a plant's assimilation is approximated by a *single evaluation* of the
  light at its crown centre rather than an integral over depth. This skips the
  per-plant quadrature, so it is faster, and removes within-crown self-shading.

* **`"ppa"`** -- a perfect-plasticity-style approximation. The patch light
  profile is discretised into discrete canopy *layers*: light steps down by a
  fixed amount of optical depth (`control$ppa_layer_optical_depth`) per layer.
  Assimilation is evaluated at the crown centre as in `"flat-top"`, but now
  reads this stepped profile. Because a hard step is not differentiable (and
  makes the ODE solver fail), the layer boundaries are smoothed with a
  C^1^-continuous ramp whose width is set by `control$ppa_layer_smoothing`.

All three share the *same* per-plant competition contribution (the smooth
Yokozawa leaf-area profile); they differ only in how assimilation is evaluated
and, for PPA, how the light profile is assembled. The model is resolved once
when a strategy is prepared, so there is no per-call cost to having the choice
available.

```{r, message = FALSE}
library(plant)
library(dplyr)
library(ggplot2)
library(patchwork)
```

# Simulating a patch under each model

We grow a single `FF16` species in a patch, and run the same patch under each
of the three shading models. Everything except `control$shading_model` is held
fixed, so any differences are due to the shading model alone.

```{r}
params <- scm_base_parameters("FF16")
patch  <- expand_parameters(trait_matrix(0.0825, "lma"), params)

models <- c("deep-crown", "flat-top", "ppa")

run_model <- function(model) {
  ctrl <- Control()
  ctrl$shading_model <- model
  run_scm(patch, ctrl = ctrl, collect = TRUE)
}

results <- lapply(models, run_model)
names(results) <- models
```

Each run returns the full collected state (`run_scm(collect = TRUE)`): a tidy
`species` table (one row per node per step) and the light environment
(`env$light_availability`). We tag each run's `species` table with its shading
model and bind them together so we can use `plant`'s tidy helpers across all
three.

```{r}
species_all <- dplyr::bind_rows(
  lapply(models, function(m) dplyr::mutate(results[[m]]$species, method = m))
)
species_all$method <- factor(species_all$method, levels = models)

# Light-extinction coefficient, used throughout to convert canopy openness to
# cumulative leaf area: E = exp(-k_I * LAI), so LAI_above = -log(E) / k_I.
k_I <- results[[1]]$p$strategies[[1]]$k_I
```

# Size-distribution dynamics

The size distribution is the set of node height trajectories: each line is a
cohort introduced at a given patch age, growing (and being shaded) over time,
shaded by node density. `plant`'s `plot_size_distribution()` builds this
directly from the tidy `species` table; faceting by `method` compares the three
shading models.

```{r sizedist, fig.width = 9, fig.height = 3.8}
p_size <- plot_size_distribution(species_all) +
  facet_wrap(~method) +
  theme(legend.position = "bottom")
p_size
```

The discrete layering in `"ppa"` produces visibly different growth trajectories:
cohorts grow at near-constant light within a layer, then accelerate as they
emerge into a brighter layer.

# Leaf area index dynamics

Leaf area index (LAI) -- the total one-sided leaf area per unit ground -- is the
cumulative leaf area above the ground, which is already encoded in the recorded
light profile: with Beer's law $E = \exp(-k_I\,\mathrm{LAI})$, the patch LAI is
just $-\log E(0) / k_I$, where $E(0)$ is canopy openness at ground level. This
reads straight off `env$light_availability` and needs no integration over the
size distribution. (LAI is real leaf area, so the smooth recorded profile gives
the true value for all three models -- the PPA layering only affects the light
plants *experience*, not the leaf area present.)

```{r}
lai_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    results[[m]]$env$light_availability |>
      dplyr::group_by(time) |>
      dplyr::slice_min(height, n = 1, with_ties = FALSE) |>
      dplyr::ungroup() |>
      dplyr::transmute(method = m, time, lai = -log(light_availability) / k_I)
  })
)
lai_all$method <- factor(lai_all$method, levels = models)
```

```{r lai, fig.width = 7, fig.height = 4}
p_lai <- ggplot(lai_all, aes(time, lai, colour = method)) +
  geom_line(linewidth = 1) +
  labs(x = "Patch age (years)", y = "Leaf area index", colour = "Shading model") +
  theme_classic()
p_lai
```

`"deep-crown"` and `"flat-top"` track each other closely and settle to a steady
LAI: under a near-uniform light profile, integrating photosynthesis over the
crown and evaluating it once at the crown centre give similar carbon gain. The
`"ppa"` model instead shows large, sustained oscillations -- successive waves of
recruitment and self-thinning -- because the discrete light layers create sharp
thresholds for cohort success (visible as the pulses of emerging cohorts in the
size-distribution panel above).

# Vertical light profile

We can also look *within* the canopy at a few patch ages, at the vertical
profile of cumulative leaf area above each height -- $\mathrm{LAI}_{above}(z) =
-\log E(z) / k_I$, where $E$ is canopy openness. This is the quantity that sets
how much each individual is shaded.

The recorded `env$light_availability` is the smooth underlying profile shared by
all models. The light that plants actually *experience* under PPA is the stepped
version, so for the `"ppa"` panel we apply the same layering transform used
internally (see `FF16_Environment::step_light`).

```{r}
d <- Control()$ppa_layer_optical_depth   # layer thickness (optical depth)
w <- Control()$ppa_layer_smoothing       # boundary smoothing fraction

# C1-smoothed staircase: flat over the lower (1 - w) of each layer, cubic ramp
# over the top w. Mirrors FF16_Environment::smooth_floor() / step_light().
smooth_floor <- function(u, w) {
  n <- floor(u)
  f <- u - n
  t <- pmax(0, f - (1 - w)) / w
  n + ifelse(f <= 1 - w, 0, t^2 * (3 - 2 * t))
}
ppa_step_light <- function(E, d, w) {
  ifelse(E >= 1 | E <= 0, E,
         exp(-d * smooth_floor(-log(E) / d, w)))
}

target_ages <- c(5, 10, 20, 60)

profile_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    light <- results[[m]]$env$light_availability
    if (m == "ppa") {
      light$light_availability <- ppa_step_light(light$light_availability, d, w)
    }
    # nearest recorded step to each target age
    steps <- results[[m]]$steps
    keep <- sapply(target_ages, function(a) steps$step[which.min(abs(steps$time - a))])
    light |>
      dplyr::filter(step %in% keep) |>
      dplyr::mutate(method = m,
                    age = target_ages[match(step, keep)],
                    lai_above = -log(light_availability) / k_I)
  })
)
profile_all$method <- factor(profile_all$method, levels = models)
```

```{r vprofile, fig.width = 9, fig.height = 4}
p_profile <- ggplot(profile_all,
                    aes(lai_above, height, colour = factor(age), group = age)) +
  geom_path() +
  facet_wrap(~method) +
  labs(x = expression(LAI[above]~(m^2/m^2)), y = "Height (m)",
       colour = "Patch age\n(years)") +
  theme_classic()
p_profile
```

`"deep-crown"` and `"flat-top"` share a smooth profile; `"ppa"` shows the
characteristic staircase, with cumulative leaf area held constant within a layer
and jumping at each layer boundary.

# Comparison panel

```{r panel, fig.width = 9, fig.height = 10}
(p_lai / p_size / p_profile) + patchwork::plot_annotation(tag_levels = "A")
```

# Speed comparison

Two effects compete. Per ODE step, `"deep-crown"` is the most expensive because
it performs a Gauss-Kronrod quadrature over crown depth for every individual,
whereas `"flat-top"` and `"ppa"` replace this with a single light evaluation. But
the *number* of steps also matters: PPA's layered light profile is much stiffer,
so the adaptive solver takes many more, smaller steps. Which effect dominates
depends on how the system is stepped.

```{r speed-helper}
# Median wall-clock over a few repeats. `times = NULL` uses the adaptive solver;
# supplying a fixed schedule pins every model to the same steps.
time_model <- function(model, times = NULL, reps = 5) {
  p <- patch
  use_fixed <- !is.null(times)
  if (use_fixed) p$ode_times <- times
  ctrl <- Control()
  ctrl$shading_model <- model
  run1 <- function() run_scm(p, ctrl = ctrl, use_ode_times = use_fixed)
  run1()                                   # warm up
  median(replicate(reps, system.time(run1())[["elapsed"]]))
}

speed_table <- function(timings) {
  data.frame(
    model = names(timings),
    `median time (s)` = round(unname(timings), 4),
    `relative to deep-crown` = round(unname(timings / timings[["deep-crown"]]), 2),
    check.names = FALSE,
    row.names = NULL
  )
}
```

## Adaptive solver (real-world cost)

This is how the models run by default, and is the cost you actually pay.

```{r speed-adaptive}
timings_adaptive <- vapply(models, time_model, numeric(1))
knitr::kable(speed_table(timings_adaptive))
```

`"flat-top"` is fastest -- it shares deep-crown's smooth light profile (so the
same number of solver steps) but skips the per-plant crown-depth integral.
`"ppa"` is *slower* overall despite the cheaper per-plant assimilation, because
its near-stepped profile forces the adaptive solver to take many more steps;
finer layers (smaller `ppa_layer_optical_depth`) or sharper boundaries (smaller
`ppa_layer_smoothing`) make this worse.

## Fixed ODE schedule (per-step cost)

If instead every model is pinned to the *same* fixed schedule
(`run_scm(..., use_ode_times = TRUE)`), the step-count effect is removed and only
the per-step cost remains. Now PPA's advantage shows: like `"flat-top"`, it
skips the crown-depth quadrature, so both are markedly faster than
`"deep-crown"`.

```{r speed-fixed}
times <- seq(0, params$max_patch_lifetime, length.out = 4000)
timings_fixed <- vapply(models, time_model, numeric(1), times = times)
knitr::kable(speed_table(timings_fixed))
```

So PPA is not inherently slow -- on a fixed schedule it is *faster* than
`"deep-crown"`. Its apparent cost under the adaptive solver comes entirely from
the extra steps the stiff profile demands. Where a fixed schedule is acceptable
(or where the layering is coarse and lightly smoothed), PPA can be the cheaper
choice. `"deep-crown"` nonetheless remains the default: it is the most faithful
representation of within-crown light capture and carries no speed penalty
relative to earlier versions of `plant`.

# Why PPA needs smoothing

A literal PPA discretisation is a *hard* step function: as a plant grows and its
crown centre crosses a layer boundary, the light it experiences jumps
discontinuously, and so does its growth rate. This is what `ppa_layer_smoothing`
exists to tame, and it is worth being explicit about *why* -- because the
problem is specifically one of **adaptive ODE stepping**, not of the model
itself.

## Adaptive stepping cannot integrate a discontinuity

`plant` solves the patch dynamics with an adaptive, error-controlled solver
(an embedded Runge-Kutta scheme). At each step it forms two estimates of the
state at different orders, and uses their difference as an estimate of the local
truncation error. If that error exceeds the requested tolerance, the step is
*rejected* and retried with a smaller size; the method assumes the solution is
smooth, so halving the step should roughly quarter the error.

Across a discontinuity that assumption breaks. No matter how small the step, a
step that straddles the jump carries an O(jump) error that never falls below
tolerance, so the solver rejects and shrinks the step indefinitely until it
gives up. With a hard step (`ppa_layer_smoothing = 0`) this is exactly what
happens:

```{r ppa-hard}
ctrl_hard <- Control()
ctrl_hard$shading_model <- "ppa"
ctrl_hard$ppa_layer_smoothing <- 0      # hard step, no smoothing

tryCatch(
  run_scm(patch, ctrl = ctrl_hard),
  error = function(e) conditionMessage(e)
)
```

A *fixed* schedule (`run_scm(..., use_ode_times = TRUE)`) sidesteps the
error-control loop -- it simply takes the prescribed steps without rejecting any
-- but then nothing protects against a large explicit step overshooting the jump,
which sends growth to non-finite values unless the schedule is very fine. Either
way, a hard step is not something the solver can handle robustly.

Switching the rounding direction (light at the top vs the bottom of a layer)
does not help: the difficulty is the discontinuity itself, not where the step
sits.

## The smoothed profile

`ppa_layer_smoothing` replaces each hard boundary with a C^1^-continuous cubic
ramp over the top fraction of the layer (see `FF16_Environment::smooth_floor`).
The light profile -- and hence each plant's growth rate -- is now continuous with
a continuous first derivative, so the adaptive error estimate behaves as the
solver assumes and steps are accepted normally. Even a small amount of smoothing
is enough to recover a well-behaved integration, and with the default the
adaptive and fixed-step solutions agree.

Two parameters control the discretisation:

* `ppa_layer_optical_depth` -- the optical-depth thickness of one layer
  (default `0.5`, i.e. one unit of leaf area index per layer at the default
  extinction coefficient `k_I = 0.5`). Smaller values give finer layering.
* `ppa_layer_smoothing` -- the fraction of each layer over which the boundary is
  smoothed (default `0.3`). Values approaching `0` recover a hard step (and the
  instability above); `1` removes the flat region and approaches the smooth
  `"deep-crown"`-style profile. There is a tension here: smaller values are
  closer to a true PPA step but stiffer (slower, per the speed table above),
  while larger values integrate faster but blur the layer structure.
