---
title: "Canopy shading methods"
output: bookdown::html_document2

---

**authors**: DS Falster
**date**: 2026

# Background

To grow a plant, `plant` needs to know how much light each individual
intercepts. That depends on two things: how a plant's own leaf area is
distributed through its crown (which sets the light it captures for
photosynthesis), and how that leaf area shades everything below it (which sets
the patch-level light profile). Both can be modelled at different levels of
realism, trading detail for speed.

The `FF16` strategy offers three shading models, selected with
`control$shading_model`:

* **`"deep-crown"`** (the default) -- the original `plant` behaviour. Leaf area
  is distributed continuously through the crown following the Yokozawa (1995)
  foliage profile, and a plant's gross assimilation is computed by *integrating*
  photosynthesis over crown depth against the light profile. The patch light
  profile is a smooth function of height.

* **`"flat-top"`** -- the light profile is built exactly as for `"deep-crown"`,
  but a plant's assimilation is approximated by a *single evaluation* of the
  light at its crown centre rather than an integral over depth. This skips the
  per-plant quadrature, so it is faster, and removes within-crown self-shading.

* **`"ppa"`** -- a perfect-plasticity-style approximation. The patch light
  profile is discretised into discrete canopy *layers*: light steps down by a
  fixed amount of optical depth (`control$ppa_layer_optical_depth`) per layer.
  Assimilation is evaluated at the crown centre as in `"flat-top"`, but now
  reads this stepped profile. Because a hard step is not differentiable (and
  makes the ODE solver fail), the layer boundaries are smoothed with a
  C^1^-continuous ramp whose width is set by `control$ppa_layer_smoothing`.

All three share the *same* per-plant competition contribution (the smooth
Yokozawa leaf-area profile); they differ only in how assimilation is evaluated
and, for PPA, how the light profile is assembled. The model is resolved once
when a strategy is prepared, so there is no per-call cost to having the choice
available.

```{r, message = FALSE}
library(plant)
library(dplyr)
library(ggplot2)
library(patchwork)
```

# Simulating a patch under each model

We grow a single `FF16` species in a patch, and run the same patch under each
of the three shading models. Everything except `control$shading_model` is held
fixed, so any differences are due to the shading model alone.

```{r}
params <- scm_base_parameters("FF16")
patch  <- expand_parameters(trait_matrix(0.0825, "lma"), params)

models <- c("deep-crown", "flat-top", "ppa")

run_model <- function(model) {
  ctrl <- Control()
  ctrl$shading_model <- model
  run_scm(patch, ctrl = ctrl, collect = TRUE)
}

results <- lapply(models, run_model)
names(results) <- models
```

Each run returns the full collected state (`run_scm(collect = TRUE)`): a tidy
`species` table (one row per node per step) and the light environment
(`env$light_availability`). We tag each run's `species` table with its shading
model and bind them together so we can use `plant`'s tidy helpers across all
three.

```{r}
species_all <- dplyr::bind_rows(
  lapply(models, function(m) dplyr::mutate(results[[m]]$species, method = m))
)
species_all$method <- factor(species_all$method, levels = models)

# Light-extinction coefficient, used throughout to convert canopy openness to
# cumulative leaf area: E = exp(-k_I * LAI), so LAI_above = -log(E) / k_I.
k_I <- results[[1]]$p$strategies[[1]]$k_I
```

# Size-distribution dynamics

The size distribution is the set of node height trajectories: each line is a
cohort introduced at a given patch age, growing (and being shaded) over time,
shaded by node density. `plant`'s `plot_size_distribution()` builds this
directly from the tidy `species` table; faceting by `method` compares the three
shading models.

```{r sizedist, fig.width = 9, fig.height = 3.8}
p_size <- plot_size_distribution(species_all) +
  facet_wrap(~method) +
  theme(legend.position = "bottom")
p_size
```

The discrete layering in `"ppa"` produces visibly different growth trajectories:
cohorts grow at near-constant light within a layer, then accelerate as they
emerge into a brighter layer.

# Leaf area index dynamics

Leaf area index (LAI) -- the total one-sided leaf area per unit ground -- is the
cumulative leaf area above the ground, which is already encoded in the recorded
light profile: with Beer's law $E = \exp(-k_I\,\mathrm{LAI})$, the patch LAI is
just $-\log E(0) / k_I$, where $E(0)$ is canopy openness at ground level. This
reads straight off `env$light_availability` and needs no integration over the
size distribution. (LAI is real leaf area, so the smooth recorded profile gives
the true value for all three models -- the PPA layering only affects the light
plants *experience*, not the leaf area present.)

```{r}
lai_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    results[[m]]$env$light_availability |>
      dplyr::group_by(time) |>
      dplyr::slice_min(height, n = 1, with_ties = FALSE) |>
      dplyr::ungroup() |>
      dplyr::transmute(method = m, time, lai = -log(light_availability) / k_I)
  })
)
lai_all$method <- factor(lai_all$method, levels = models)
```

```{r lai, fig.width = 7, fig.height = 4}
p_lai <- ggplot(lai_all, aes(time, lai, colour = method)) +
  geom_line(linewidth = 1) +
  labs(x = "Patch age (years)", y = "Leaf area index", colour = "Shading model") +
  theme_classic()
p_lai
```

`"deep-crown"` and `"flat-top"` track each other closely and settle to a steady
LAI: under a near-uniform light profile, integrating photosynthesis over the
crown and evaluating it once at the crown centre give similar carbon gain. The
`"ppa"` model instead shows large, sustained oscillations -- successive waves of
recruitment and self-thinning -- because the discrete light layers create sharp
thresholds for cohort success (visible as the pulses of emerging cohorts in the
size-distribution panel above).

# Vertical light profile

We can also look *within* the canopy at a few patch ages, at the vertical
profile of cumulative leaf area above each height -- $\mathrm{LAI}_{above}(z) =
-\log E(z) / k_I$, where $E$ is canopy openness. This is the quantity that sets
how much each individual is shaded.

The recorded `env$light_availability` is the smooth underlying profile shared by
all models. The light that plants actually *experience* under PPA is the stepped
version, so for the `"ppa"` panel we apply the same layering transform used
internally (see `FF16_Environment::step_light`).

```{r}
d <- Control()$ppa_layer_optical_depth   # layer thickness (optical depth)
w <- Control()$ppa_layer_smoothing       # boundary smoothing fraction

# C1-smoothed staircase: flat over the lower (1 - w) of each layer, cubic ramp
# over the top w. Mirrors FF16_Environment::smooth_floor() / step_light().
smooth_floor <- function(u, w) {
  n <- floor(u)
  f <- u - n
  t <- pmax(0, f - (1 - w)) / w
  n + ifelse(f <= 1 - w, 0, t^2 * (3 - 2 * t))
}
ppa_step_light <- function(E, d, w) {
  ifelse(E >= 1 | E <= 0, E,
         exp(-d * smooth_floor(-log(E) / d, w)))
}

target_ages <- c(5, 20, 60)

profile_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    light <- results[[m]]$env$light_availability
    if (m == "ppa") {
      light$light_availability <- ppa_step_light(light$light_availability, d, w)
    }
    # nearest recorded step to each target age
    steps <- results[[m]]$steps
    keep <- sapply(target_ages, function(a) steps$step[which.min(abs(steps$time - a))])
    light |>
      dplyr::filter(step %in% keep) |>
      dplyr::mutate(method = m,
                    age = target_ages[match(step, keep)],
                    lai_above = -log(light_availability) / k_I)
  })
)
profile_all$method <- factor(profile_all$method, levels = models)
```

```{r vprofile, fig.width = 9, fig.height = 4}
p_profile <- ggplot(profile_all,
                    aes(lai_above, height, colour = factor(age), group = age)) +
  geom_path() +
  facet_wrap(~method) +
  labs(x = expression(LAI[above]~(m^2/m^2)), y = "Height (m)",
       colour = "Patch age\n(years)") +
  theme_classic()
p_profile
```

`"deep-crown"` and `"flat-top"` share a smooth profile; `"ppa"` shows the
characteristic staircase, with cumulative leaf area held constant within a layer
and jumping at each layer boundary.

# Comparison panel

```{r panel, fig.width = 9, fig.height = 10}
(p_lai / p_size / p_profile) + patchwork::plot_annotation(tag_levels = "A")
```

# Speed comparison

Two effects compete here. Per ODE step, `"deep-crown"` is the most expensive
because it performs a Gauss-Kronrod quadrature over crown depth for every
individual, whereas `"flat-top"` and `"ppa"` replace this with a single light
evaluation. But the *number* of steps also matters: PPA's layered light profile
is much stiffer, so the adaptive solver takes many more, smaller steps. The
timings below are for the normal adaptive solver, i.e. real end-to-end cost.

```{r speed}
time_model <- function(model, reps = 5) {
  ctrl <- Control()
  ctrl$shading_model <- model
  run_scm(patch, ctrl = ctrl)            # warm up
  t <- replicate(reps, system.time(run_scm(patch, ctrl = ctrl))[["elapsed"]])
  median(t)
}

timings <- vapply(models, time_model, numeric(1))

speed <- data.frame(
  model = models,
  `median time (s)` = round(timings, 4),
  `relative to deep-crown` = round(timings / timings[["deep-crown"]], 2),
  check.names = FALSE,
  row.names = NULL
)
knitr::kable(speed)
```

`"flat-top"` is the fastest -- it shares deep-crown's smooth light profile (so
the same number of solver steps) but skips the per-plant crown-depth integral.
`"ppa"` is *slower* overall despite the cheaper per-plant assimilation, because
its near-stepped profile forces the adaptive solver to take many more steps;
finer layers (smaller `ppa_layer_optical_depth`) or sharper boundaries (smaller
`ppa_layer_smoothing`) make this worse. `"deep-crown"` remains the default: it is
the most faithful representation of within-crown light capture and carries no
speed penalty relative to earlier versions of `plant`.

# Notes on the PPA model

The PPA light profile is a step function, which is not differentiable. A literal
hard step makes a plant's light jump as it grows across a layer boundary, which
both defeats the adaptive ODE solver's error control and can send growth to
non-finite values. The `control$ppa_layer_smoothing` parameter therefore
replaces each hard boundary with a C^1^-continuous cubic ramp over the top
fraction of each layer. With the default (`0.3`) the profile is smooth enough to
integrate with the normal adaptive solver, and adaptive and fixed-step solutions
agree. Two parameters control the discretisation:

* `ppa_layer_optical_depth` -- the optical-depth thickness of one layer
  (default `0.5`, i.e. one unit of leaf area index per layer at the default
  extinction coefficient `k_I = 0.5`). Smaller values give finer layering.
* `ppa_layer_smoothing` -- the fraction of each layer over which the boundary is
  smoothed (default `0.3`). Values approaching `0` recover a hard step (and its
  numerical instability); `1` removes the flat region and approaches the smooth
  `"deep-crown"`-style profile.
