---
title: "Patch-level emergent properties"
output: bookdown::html_document2
---

**authors**: DS Falster, RG FitzJohn
**date**: 2016

# Background

The aim here is to use the plant package to investigate dynamics
within a patch of competing individuals, focusing on emergent patch-level
properties, rather than properties of individuals within the patch.

We'll work with a single species at birth rate close to demographic equilibrium:


```{r}
library(plant)

p0 <- scm_base_parameters("FF16")
p <- expand_parameters(trait_matrix(0.0825, "lma"), p0, birth_rate_list = 17.31)
patch <- run_scm(p, refine_schedule = TRUE)$parameters
result <- run_scm(patch, collect = TRUE)

```

# Node densities

There was a bit of hassle in [Patches vignette](https://traitecoevo.github.io/plant/articles/patch.html)
in plotting all trajectories along with a couple of lines corresponding to focal
years.  We're going to do the same thing here:

```{r}
closest <- function(t, time) {
  which.min(abs(time - t))
}
last <- function(x) {
  x[[length(x)]]
}

times <- c(5, 10, 20, 40, dplyr::last(result$time))
i <- vapply(times, closest, integer(1), result$time)
blues <- c("#DEEBF7", "#C6DBEF", "#9ECAE1", "#6BAED6",
           "#4292C6", "#2171B5", "#08519C", "#08306B")
cols <- colorRampPalette(blues[-(1:2)])(length(i))

height      <- t(result$species[[1]]["height", , ])
log_density <- t(result$species[[1]]["log_density", , ])
```

As with height in the [Patches vignette](https://traitecoevo.github.io/plant/articles/patch.html),
node density is a matrix of time and node identity.  However, to show the
profiles for a given time slice using `matplot` it is convenient to transpose 
this matrix (`matplot` works column-by-column so we want each column to
to be the state of the system at a given time):

```{r}
density <- exp(log_density)
matplot(height, density, type="l", lty=1,
        col=util_colour_set_opacity("black", 0.15),
        xlab="Height (m)", ylab="Density (1 / m / m2)", las=1,
        log="y")
```

Note that the densities here can be *extremely* low, and yet
individuals within the nodes continue to grow in size; this is the "[atto-fox problem](http://en.wikipedia.org/wiki/Lotka-Volterra_equation)",
though here we drop down as low as `r formatC(signif(min(density, na.rm=TRUE) / 1e-24, 1))`
yocto individuals / m / metre squared (a yocto-x being 1 millionth of an
atto-x).  *Anyway*....

The trajectories are easier to understand if a few are
highlighted.

```{r}
xlim <- c(0, max(height, na.rm=TRUE) * 1.05)
matplot(height, density, type="l", lty=1,
        col=util_colour_set_opacity("black", 0.15),
        xlab="Height (m)", ylab="Density (1 / m / m2)", las=1,
        log="y", xlim=xlim)
matlines(height[, i], density[, i], col=cols, lty=1, type="l")
points(height[1, i], density[1, i], pch=19, col=cols)
text(height[1, i] + strwidth("x"), density[1, i],
     paste0(round(times), c(" years", rep("", length(times) - 1))),
     adj=c(0, 0))
```

Early on there is a low density "dip" caused by self thinning (5
years).  As the stand develops that dip broadens and deepens and
moves forward in height (these very low density nodes are still
traveling the characteristic equations of the SCM).  Later on (20
years) additional an second wave of recruitment gives a second
pulse of high density (around 4 m tall), which can be seen
traveling along in the 40 year profile to about 13 m.  When the
patch is very old the stand approaches a stable age distribution,
though a very narrow window of "missing" heights exists just below
the top of the canopy.

# Leaf area

It's also possible see where the leaf area in the patch is coming
from; a profile of leaf area with respect to height.  Again, this
requires reconstructing the patches, and using an unexported
function from `plant` to put this into a matrix:

```{r}
patches <- lapply(seq_along(result$time), scm_patch, result)
leaf_area <- lapply(patches, function(x) x$species[[1]]$competition_effects)
leaf_area <- plant:::pad_list_to_array(leaf_area)

matplot(height, leaf_area, type="l", lty=1, col="lightgrey",
        xlim=xlim, xlab="Height (m)",
        ylab="Leaf area density (m2 / m2 / m)", las=1)
matlines(height[, i], leaf_area[, i], col=cols, lty=1, type="l")
points(height[1, i], leaf_area[1, i], pch=19, col=cols)
text(height[1, i] + strwidth("x"), leaf_area[1, i],
     paste0(round(times), c(" years", rep("", length(times) - 1))),
     adj=c(0, 0))
```
# Growth rate

Finally, we can see where height growth rate is concentrated in the
population.  This differs from the profile in the [Individuals vignette](https://traitecoevo.github.io/plant/articles/individuals.html)
because it is reduced depending on the light environment, but that
environment is the one constructed by the individual itself.

```{r}
ode_size <- patches[[1]]$ode_size

growth_rate <- lapply(patches, function(x)
                      matrix(x$species[[1]]$ode_rates, ode_size)[1, ])
growth_rate <- plant:::pad_list_to_array(growth_rate)

matplot(height, growth_rate, type="l", lty=1, col="lightgrey",
        xlim=xlim, xlab="Height (m)",
        ylab="Height growth rate (m / year)", las=1)
matlines(height[, i], growth_rate[, i], col=cols, lty=1, type="l")
points(height[1, i], growth_rate[1, i], pch=19, col=cols)
text(height[1, i] + strwidth("x"), growth_rate[1, i],
     paste0(round(times), c(" years", rep("", length(times) - 1))),
     adj=c(0, 0))
```
# Integrating over a metapopulation

The above plots show relationships with patches of a given age
What about the average relationship across the entire metapopulation?
To get that, we  average (integrate) over the distribution over patch
(for formula see demography vignette). To achieve this we first need
the patch-level relationships to a series of fixed heights

```{r}
hh <- seq_log_range(range(height, na.rm=TRUE), 500)
```

We'll use a spline interpolation, on log-log-scaled data, clamped
so that for x values outside the observed range are set to zero.

```{r}
##' Spline interpolation in log-x space
##' @title Spline interpolation in log-x space
##' @param x,y Vectors giving coordinates of points to be
##' interpolated.  The x points should be naturally on a log scale,
##' and for \code{splinefun_loglog} both x and y should be on a log
##' scale.
##' @param ... Additional parameters passed to
##' @rdname splinefun_log
splinefun_loglog <- function(x, y, ...) {
        f <- splinefun(log(x), log(y), ...)
        function(x) {
                exp(f(log(x)))
        }
}

##' Clamp a function to particular values when outside of a given domain (r)
##'
##' Things like names on input and return vectors are not dealt with
##' very well, and would differ if the function was better
##' constructed.  Values falling outwide the domain are not evaluated
##' (which is useful if these would cause crashes, warnings, etc).
##'
##' @title Clamp function to domain
##' @param f A function that takes \code{x} as a first argument.
##' @param r Range of values (vector of length 2)
##' @param value (Single) value to use when out of domain.
##' @return A new function
##' @export
clamp_domain <- function(f, r, value = NA_real_) {
        f <- match.fun(f)
        if (length(r) != 2L) {
                stop("Expected length two range")
        }
        if (any(is.na(r)) || r[[2]] < r[[1]]) {
                stop("Values for range must be finite and not decreasing")
        }
        if (length(value) != 1L) {
                stop("value must be length 1")
        }
        function(x, ...) {
                ret <- rep_len(value, length(x))
                i <- x >= r[[1]] & x <= r[[2]]
                if (any(i)) {
                        ret[i] <- f(x[i])
                }
                ret
        }
}

f <- function(height, density, hout) {
  r <- range(height, na.rm=TRUE)
  clamp_domain(splinefun_loglog(height, density), r, 0)(hout)
}
```

Now interpolate the height-density relationship in each patch to
the series of  specified heights

```{r}
xx <- lapply(seq_along(result$time),
             function(i) f(height[, i], density[, i], hh))
n_hh <- plant:::pad_list_to_array(xx)
```

For each of these heights, we can now integrate across patches
(i.e. across rows), weighting by patch abundance

```{r}
trapezium <- plant:::trapezium
n_av <- apply(n_hh, 1,
              function(x) trapezium(result$time, x * result$patch_density))
```

Add this average to the plot (red line):

```{r}
xlim <- c(0, max(height, na.rm=TRUE) * 1.05)
matplot(height, density, type="l", lty=1,
        col=util_colour_set_opacity("black", 0.15),
        xlab="Height (m)", ylab="Density (1 / m / m2)", las=1,
        log="y", xlim=xlim)
matlines(height[, i], density[, i], col=cols, lty=1, type="l")
points(height[1, i], density[1, i], pch=19, col=cols)
text(height[1, i] + strwidth("x"), density[1, i],
     paste0(round(times), c(" years", rep("", length(times) - 1))),
     adj=c(0, 0))
points(hh, n_av, col="red", type='l')
```
