---
title: "Example worklfow"
author: 
- Daniel Falster
date: 2016
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{patch}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# installation

```{r load libraries}
devtools::load_all()

# library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```

Activate logger

```{r}
plant_log_console()
```

Setting some base parameters

```{r}
base_parameters <- function() {
  p0 <- scm_base_parameters("FF16", "FF16_Env")

  p0$strategy_default$lma <- 0.07
  p0$strategy_default$hmat <- 15
  p0$strategy_default$rho <- 700
  p0$strategy_default$a_l1 <- 2.17
  p0$strategy_default$a_l2 <- 0.5
  
  p0
}
```


```{r}
run_mypatch <- function(
  traits =  trait_matrix(c(0.07,0.24), c("lma")),
  B_lf1 = 1,
  seed_rain = 100,
  p0 = base_parameters(),
  find_equilbrium = FALSE
  ) {
  
  hyper_par_fn = make_FF16_hyperpar(B_lf1 = B_lf1, latitude = 28.182)
  p1 <- expand_parameters(traits, p0, hyper_par_fn, mutant = FALSE)
  p1$seed_rain <- seed_rain
  
  if(find_equilbrium)
    result <- equilibrium_seed_rain(p1)
  else
    result <- build_schedule(p1)
  
  # gather outputs at each time step
  run_scm_collect(result)
}
```


Run patch

```{r}
results <- 
  run_mypatch(seed_rain = 180)
```

Tidy into tidy state
```{r}
results_tidy <- 
  results %>% 
  tidy_patch()

results_tidy
```

## Plot size distirbuion

```{r}
results_tidy$species %>%
  drop_na() %>%
plot_size_distribution_patch()
```


## Calculate totals by species for each step

```{r}
data_species_tot <- 
  results_tidy$species %>% patch_species_total()

data_species_tot
```

total leaf area over time

```{r}
data_species_tot %>% 
  ggplot(aes(time, individuals)) +
  geom_line()
```

## Expand state

```{r}
results_tidy_expand <- results_tidy %>% expand_state()

results_tidy_expand$species
```

total leaf area over time

```{r}
data_species_tot <- 
  results_tidy_expand$species %>% patch_species_total()

data_species_tot %>% 
  ggplot(aes(time, area_leaf, colour=species)) +
  geom_line()
```


## total leaf area over time

```{r}
data_species_tot %>% 
  ggplot(aes(time, area_leaf, colour=species)) +
  geom_line()
```

## biomass components over time

Mass in kg/m2
```{r}
v <- c("mass_heartwood", "mass_sapwood", "mass_bark", "mass_leaf")

data_long <- 
  data_species_tot %>% 
  select(time,species, one_of(v)) %>%
  pivot_longer(cols=starts_with("mass"), names_to = "tissue")

data_long$tissue <- factor(data_long$tissue, levels = v)

species_names <- tibble(species = unique(data_long$species),
                      species_name = c("Species_1","Species_2"))

data_long %>%
  left_join(species_names) %>%
  ggplot(aes(time, value, fill=tissue)) +
  geom_area() +
  labs(x = "Patch age (yr)", y = "Above ground mass (kg/m2)") +
  theme_classic() + 
  xlim(c(0,50)) +
  facet_wrap(~species_name)
```


## Classic Self thinnning plot

For this we need average size over time. Get average size by dividing total by the sum of individuals per unit area

Note this plot has some noise in it. Suggests integration in previous step may need more resolution, or a better method. 

```{r}

data_species_tot <- 
  results_tidy_expand$species %>% patch_species_total()

p1 <- results_tidy_expand$species %>% plot_size_distribution_patch() 
p2 <- 
  data_species_tot %>%
  filter(species==1) %>%
  mutate(area_stem_av = area_stem / individuals) %>% 
  ggplot(aes(area_stem_av, individuals, colour=species)) +
  geom_line() +
  scale_x_log10() +
  scale_y_log10() + 
  theme_classic()

p1 + p2

results_tidy_expand$species %>% filter(height > 1, time> 20, time<50) %>% 
  patch_species_total() %>%
  mutate(area_stem_av = area_stem / individuals) %>% 
  ggplot(aes(area_stem_av, individuals, colour=species)) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() + 
  theme_classic()
```

Note, get cleaner results if use first cohorts

```{r}
results_tidy_expand$species %>% filter(cohort < 150) %>% 
  patch_species_total() %>%
  mutate(
    area_stem_av = area_stem / individuals,
    # A = pi*D^2/4 -> D = 
    diam_stem_av = sqrt(area_stem_av)
    ) %>% 
  ggplot(aes(area_stem_av, individuals, colour=species)) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() +
  theme_classic()
```


Note bottom density fluctuates over time, this makes sense if germination is responding to leaf area above

Possibility this is error in calculating competition -- i.e. total leaf area?

How is integral achieved in plant?

```{r}
results_tidy_expand$species %>% 
  drop_na() %>%
  group_by(step, species, time) %>%
  slice_tail() %>% 
  ungroup() %>%
  ggplot(aes(time, density)) + geom_line()
```

# Work in progress why is there noise in the self thinning relationship

```{r}
results_tidy_expand$species %>% 
  patch_species_total() %>%
  mutate(
    area_stem_av = area_stem / individuals,
    diam_av = 
    ) %>% 
  ggplot(aes(area_stem_av, individuals, colour=species)) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() +
  theme_classic()

  
results_tidy_expand$species %>% filter(height > 1, time> 20, time<50) %>% 
  patch_species_total() %>%
  mutate(area_stem_av = area_stem / individuals) %>% 
  ggplot(aes(area_stem_av, individuals, colour=species)) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() + 
  theme_classic()

results_tidy_expand$species %>% filter(height > 0, time> 0, time<150) %>% 
  patch_species_total() %>%
  ggplot(aes(time, individuals, colour=species)) +
  geom_line() + 
  theme_classic()

results_tidy_expand$species %>% filter(step%in% c(138, 139, 140, 141)) %>% 
  select(step, height, density) %>%
  arrange(step,height) %>%
  group_by(step) %>%
  mutate(
    dx = lead(height) -height,
    y = 0.5*(density + lead(density)),
    xy = dx*y
  ) %>% ungroup() -> tmp

tmp %>% filter(step==138) %>% pull(xy) %>% sum(na.rm=TRUE)

data_species_tot%>% filter(species==1, step==138)

ggplot(tmp %>% filter(height > 0.5), aes(height, y, group=step, col=as.character(step))) + geom_point() + geom_line() + scale_x_log10()

  
results_tidy_expand$species %>% filter(height > 0, time> 30, time<31) %>% View() 
  
results_tidy_expand$species %>%
ggplot(aes(height, density, colour=time, group=time)) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() + 
  theme_classic() + 
  facet_wrap(~species)


results_tidy_expand$species %>% filter(species==1) %>% plot_size_distribution_patch()   


```

## Two species

```{r}
results <- run_mypatch(traits =  trait_matrix(c(0.07, 0.24), c("lma")))

results_tidy <- results %>%
  tidy_patch()

```


## Interpolatign to specific time points

```{r}
times <- c(1, 5, 10, 1000)

tidy_species_data <- results_tidy$species

tidy_species_new <- interpolate_to_times(tidy_species_data, times)

```

Gives predictions at new time points. 

```{r}
tidy_species_new
```

To see how good these are, plot together with old points

```{r}
data_combined <- 
  tidy_species_data %>%
  bind_rows(tidy_species_new) %>%
  arrange(species, cohort, time) %>% 
  filter(cohort %in% seq(1, 101, by=20))

data_combined_long <- 
  data_combined  %>% 
  select(cohort, time, step, density, height, species) %>% 
  pivot_longer(cols = c("density", "height"))

data_combined_long %>% 
  ggplot(aes(time, value, group=cohort,colour=cohort)) +
  geom_line() +
  geom_point(data = data_combined_long %>% filter(is.na(step)), col=2) +
  scale_y_log10() +
  xlim(c(0, 20)) +
  facet_grid(name~species, scales="free") +
  theme_classic()
```

## Interpolate to heights

Similar to above, can interpolate across cohorts at given time to sepcific heights

```{r}
heights <- c(1, 5, 10)

tidy_species_new <- interpolate_to_heights(tidy_species_data, heights)
```

Gives predictions at new time points. 

```{r}
tidy_species_new %>% drop_na()
```

To see how good these are, plot together with old points

```{r}
data_combined <- 
  tidy_species_data %>%
  bind_rows(tidy_species_new) %>%
  arrange(species, time, height) %>% 
  filter(step %in% c(100, 150))

data_combined_long <- 
  data_combined  %>% 
  select(cohort, time, step, height, density, mortality, area_heartwood, species) %>% 
  pivot_longer(cols = c("density", "mortality", "area_heartwood"))

data_combined_long %>% 
    filter(!is.na(cohort)) %>% 
  ggplot() +
  geom_line(aes(height, value, group=step, colour=step)) +
  geom_point(data = data_combined_long %>% filter(is.na(cohort)), aes(x=height, y=value), col="red")+
  facet_grid(name~species, scales="free") +
  theme_classic()
```


## Sampling from population

(not implemented)

```{r}
sample_size_distribution(tidy_species_data, patch_area)
```

