---
title: "Adaptive vs fixed-step (daily Euler) integration"
output: bookdown::html_document2
---

**authors**: DS Falster
**date**: 2026

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
```

`plant` casts the whole patch as a system of ordinary differential equations and,
by default, integrates it with an **adaptive, error-controlled Cash–Karp
Runge–Kutta (RK45)** stepper (see the [ODE stepping](ode_solver.html) article).
The solver chooses each step size to keep the local error within a tolerance:
tiny steps early on, when newly introduced cohorts are diverging rapidly, and
much coarser steps later, once the stand settles.

Many widely used **dynamic global vegetation models (DGVMs)** instead march the
state forward on a *fixed* grid — most commonly a **daily** time step — using
plain forward (explicit) Euler. `plant` can now be run this way too, via the
`fixed_time_step` control. This article shows how to enable it and uses a
**work–precision diagram** to make the central trade-off concrete: at *matched
accuracy*, adaptive stepping does far less work.

```{r load libraries}
library(plant)
library(dplyr)
library(ggplot2)
```

# Enabling fixed-step forward Euler

The integration mode is selected by a single control field, `fixed_time_step`
(units: **years**):

- `fixed_time_step = 0` (the default) → adaptive RK45.
- `fixed_time_step > 0` → forward Euler on a uniform grid of that spacing.

A *daily* step is therefore `fixed_time_step = 1 / 365`.

```{r}
p0 <- scm_base_parameters("FF16")
p0$max_patch_lifetime <- 60

p <- expand_parameters(trait_matrix(0.0825, "lma"), p0, birth_rate_list = 1.0)
env <- Environment("FF16")

ctrl_adaptive <- Control()                       # fixed_time_step = 0
ctrl_daily    <- Control(fixed_time_step = 1 / 365)

out_adaptive <- run_scm(p, env, ctrl_adaptive)
out_daily    <- run_scm(p, env, ctrl_daily)
```

Forward Euler does **one** derivative evaluation per step (`y <- y + h * dy/dt`),
whereas each RK45 step evaluates the derivatives six times. The two runs report
slightly different fitness, and use a very different number of steps:

```{r}
tibble::tibble(
  mode    = c("adaptive RK45", "daily Euler"),
  R0      = c(out_adaptive$net_reproduction_ratios,
              out_daily$net_reproduction_ratios),
  n_steps = c(length(out_adaptive$ode_times),
              length(out_daily$ode_times))
)
```

> **Caution — don't read this as "Euler is faster".** A *single* daily step is
> cheaper than a single adaptive step (1 derivative evaluation versus 6), so a
> naive timing race makes Euler look quick. But that comparison holds the *step
> count* fixed, not the *accuracy*. Daily Euler here is visibly less accurate
> than the adaptive run. The honest question is: *how much work does each method
> need to reach the same accuracy?* That is what the work–precision diagram
> below answers — and there adaptive wins comfortably.

# A work–precision diagram

We sweep the adaptive solver over a range of error tolerances, and forward Euler
over a range of step sizes, and for each run record:

- **work**: total derivative evaluations — `6 × n_steps` for RK45, `1 × n_steps`
  for Euler (this is the dominant cost of the inner loop);
- **error**: relative error in the net reproduction ratio $R_0$ against a
  tight-tolerance adaptive **reference**.

```{r}
## Tight-tolerance adaptive reference.
ctrl_ref <- control_accurate()
ctrl_ref$ode_tol_rel <- 1e-8
ctrl_ref$ode_tol_abs <- 1e-8
R0_ref <- run_scm(p, env, ctrl_ref)$net_reproduction_ratios

measure <- function(ctrl, mode) {
  stages <- if (mode == "adaptive") 6 else 1   # derivative evals per step
  scm <- run_scm(p, env, ctrl)
  n   <- length(scm$ode_times)
  tibble::tibble(
    mode  = mode,
    work  = n * stages,
    error = abs(scm$net_reproduction_ratios - R0_ref) / R0_ref
  )
}

adaptive_sweep <- purrr::map_dfr(c(1e-2, 1e-3, 1e-4, 1e-5, 1e-6), function(tol) {
  ctrl <- Control()
  ctrl$ode_tol_rel <- tol
  ctrl$ode_tol_abs <- tol
  measure(ctrl, "adaptive")
})

euler_sweep <- purrr::map_dfr(c(2, 1, 0.5, 0.25, 0.125, 0.0625), function(dt) {
  measure(Control(fixed_time_step = dt), "euler")
})

work_precision <- bind_rows(adaptive_sweep, euler_sweep)
```

```{r work-precision, fig.cap = "Work-precision diagram. Lower and to the left is better (less work, smaller error). The adaptive RK45 frontier sits well below the fixed-step Euler frontier: for any target accuracy it reaches it with fewer derivative evaluations."}
ggplot(work_precision, aes(work, error, colour = mode)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Derivative evaluations (work)",
       y = expression("Relative error in " * R[0]),
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

# Why adaptive stepping dominates

The two curves tell a consistent story:

- **Forward Euler is first-order.** Halving the step size roughly halves the
  error, so the Euler curve descends slowly — buying an extra digit of accuracy
  costs roughly a *tenfold* increase in work.
- **Adaptive RK45 is high-order *and* adaptive.** It spends small steps only
  where the dynamics demand them (the first weeks, while cohorts diverge) and
  coarsens dramatically once the stand stabilises. It reaches an error floor —
  set by the node-introduction schedule rather than the ODE tolerance — at a
  fraction of the evaluations Euler would need to match it.

To get daily-Euler accuracy *down to* the adaptive frontier you would need a
sub-daily step everywhere, including the long, smooth tail of the simulation
where the adaptive solver is happily taking steps of years. That is wasted work,
and it is exactly the inefficiency adaptive stepping removes.

The practical conclusion: fixed-step Euler is valuable for **operational-realism
comparisons** — running `plant` on the same footing as a daily-step DGVM — but it
is not the way to get an accurate or cheap solution. For that, keep the default
adaptive solver.

# A run that completes is not a run that converged

Fixed-step integration removes the solver's safety net: forward Euler will march
across *anything*, including a discontinuous right-hand side that the adaptive
solver rightly refuses. When that happens the integrator still returns numbers —
but they are an artefact of where the step grid happens to fall, not an
approximation to a solution that gets better as you refine.

`plant`'s experimental **PPA hard-step** shading variant
(`ppa_layer_smoothing = 0`, the literal field discretisation) is exactly such a
case: its light profile is discontinuous in height, so the adaptive solver fails
with *"Cannot achieve the desired accuracy"*. Under a fixed step it runs. The
test of whether that result *means* anything is convergence: does it settle as
the step shrinks? We compare it with the smoothed variant
(`ppa_layer_smoothing = 0.3`), which is a well-posed, continuous problem.

```{r}
seed_output <- function(steps_per_year, smoothing) {
  ctrl <- Control(fixed_time_step = 1 / steps_per_year)
  ctrl$shading_model      <- "ppa"
  ctrl$ppa_layer_smoothing <- smoothing
  run_scm(p, env, ctrl)$offspring_production
}

convergence <- purrr::map_dfr(c(91, 182, 365, 730, 1460), function(spy) {
  tibble::tibble(
    steps_per_year = spy,
    `hard step (s = 0)`   = seed_output(spy, 0),
    `smoothed (s = 0.3)`  = seed_output(spy, 0.3))
}) |>
  tidyr::pivot_longer(-steps_per_year, names_to = "variant", values_to = "seed")
```

```{r hard-step-convergence, fig.width = 7, fig.height = 3.2, fig.cap = "Resident seed output as the fixed step shrinks. The smoothed variant converges; the hard step does not — it wanders by an order of magnitude with no limit. Note the independent y-scales: the hard step's values are ~1000x smaller, and erratic. A result that depends on the step size is a property of the grid, not of the model."}
ggplot(convergence, aes(steps_per_year, seed, colour = variant)) +
  geom_line() + geom_point(size = 2) +
  facet_wrap(~variant, scales = "free_y") +
  scale_x_log10() +
  labs(x = "Steps per year (1 / fixed_time_step)", y = "Resident seed output",
       colour = NULL) +
  theme_bw() + theme(legend.position = "none")
```

The smoothed variant marches monotonically toward a limit — refining the step
buys accuracy, as it should. The hard step does not converge to anything: each
step size gives a different answer, so no step size gives *the* answer. This is
why such variants must not be read as fitness or as a solution — the fixed-step
solver makes them *executable*, not *correct*.

# Caveats and scope

- **Mutant fitness / resident replay is not supported under `fixed_time_step`.**
  The mutant pathway replays a resident run by reusing the six cached RK
  sub-step environments per step; forward Euler has a single stage and no such
  cache. Combining `fixed_time_step > 0` with `save_RK45_cache` (or with a pinned
  `use_ode_times` replay) raises a clear error rather than returning a wrong
  fitness.
- **Discontinuous/"hard-step" environment variants** only *run to completion*
  under a fixed grid and do not converge as the step shrinks (see the section
  above) — any apparent result is a grid artefact, not a solution. Do not
  interpret such runs as fitness or as a converged answer.
```
