---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE}
rm(list=ls())

library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(deSolve)
library(lhs)
library(furrr)
```

```{r}
devtools::load_all(".")
# devtools::install(".")

# library(plant)
```


```{r warning=FALSE}
source("R/solar_model.R")
# source("water_bucket_submodel/R/reference_soil.R")
source("tests/reference_leaf_updated.R")
source("examples_vignettes/R/testing_integration.R")
```

```{r}
#vpsat in kPa
calc_vpsat <- function(air_temp_c){
  
    A = 17.27;
    T_star = 273.0; #K
    T_dash = 36.0;  #K
    es_T_star = 0.611; #kPa
  
  
  air_temp_K <- air_temp_c + 273.15
  
  esat = es_T_star * exp(A * (air_temp_K - T_star) / (air_temp_K - T_dash))
  esat
}
```


```{r}
p_50_2_K_s <- function(p50){
  1.638999*(p50/2)^(-1.38)
}

calc_vul_b <- function(p_50, c){
  num <- p_50
  den <- (-log(1-50/100))^(1/c)
  num/den
}


calc_psi_crit <- function(b,c) {
  b*(log(1/0.05))^(1/c)
}


```


```{r}
calc_temp <- function(instant_of_day, day_length, mean_temp, temp_diff, sun_rise, sun_down){

# Parton and Logan (1981) A model for dirunal variation in soil and
#        air temperature. Agricultural Meteorology, 23, 205--216.


a_t = 1.86;
b_t = 2.2;     #nighttime coeffcient
c_t = -0.17;   #lag of the min temp from the time of runrise

min_temp = mean_temp - temp_diff/2
max_temp = mean_temp + temp_diff/2

sun_rise = sun_rise
sun_down = sun_down

m = instant_of_day - sun_rise + c_t

if(instant_of_day >= sun_rise & instant_of_day <= sun_down){
  min_temp + (max_temp - min_temp) * sin((pi * m) / (day_length + 2.0 * a_t))
} else {
    NA
  }
}

```

```{r}
LHS <- randomLHS(5000, 11)

params <- tibble(mean_temp = 20, temp_diff = qunif(LHS[,1], min = 5, max = 20), latitude = qunif(LHS[,2], min = 0, max = 60), VP = qunif(LHS[,3], 0.5, 2),
                   vcmax = qunif(LHS[,4], 30, 100), p_50 = qunif(LHS[,5], min = 1, max = 10), huber_value = qunif(LHS[,6], min = 0.000157/2, max =  0.000157*2),
                   day = qunif(LHS[,7], min = 0, 180), h = qunif(LHS[,8], min = 5, max = 80), E = qunif(LHS[,9], min = 0, max = 1), psi_soil = qunif(LHS[,10], 0, 10), ca = qunif(LHS[,11], 30, 45))
```


```{r}
# future::plan("multisession", workers = 8)
```

```{r}
params %>%
  head(50) %>%
  pmap(expand_input_data, focal_points = ) %>%
  bind_rows() %>%
  ungroup() -> expanded_input_data
```

```{r}
expanded_input_data %>% 
 mutate(leaf = pmap(., make_leaf)) %>%
  mutate(leaf_states_rate = map(leaf, leaf_states_rates_from_leaf)) %>%
  unnest(leaf_states_rate) %>% 
  nest(outputs = c(leaf, instant_through_day_hr, instant_through_day_dec, dec_day_time, solar_angle, PAR, temp, VPD_hr, profit, GSS_count, count, method, transpiration, psi_stem, g_c)) -> expanded_output
```

```{r}
expanded_output %>%
  slice(30) %>%
  mutate(param_set = seq(1:n())) %>%
  mutate(integrated_daily_profit = map_dbl(outputs, ~plant:::trapezium(.x$instant_through_day_hr, .x$profit))) %>%
  mutate(integrated_daily_transpiration = map_dbl(outputs, ~plant:::trapezium(.x$instant_through_day_hr, .x$transpiration))) %>%
  expand_grid(focal_points = c(1,2,3,50)) %>%
  mutate(point_estimate_integrals = pmap(., calc_mult_point)) %>%
  mutate(point_estimate_integrated_daily_profit = map_dbl(point_estimate_integrals, ~sum(.x$profit_integrated))) %>%
  mutate(point_estimate_integrated_daily_transpiraiton = map_dbl(point_estimate_integrals, ~sum(.x$transpiration_integrated))) -> expanded_output_integration
```

```{r}
expanded_output_integration %>%
  filter(psi_soil < psi_crit) %>%
  group_by(param_set) %>%
  sample_n(1) %>%
  ungroup() %>% 
  sample_n(4) -> data_for_plot
  

data_for_plot %>%
  unnest(outputs) %>%
  ggplot(aes(x = instant_through_day_hr, y = profit)) +
  geom_line() + 
  scale_x_continuous(breaks = seq(0,24,4), labels = seq(0,24,4), limits = c(0,24)) +
  facet_wrap(~param_set, scales = "free_x") +
  theme_classic() +
  theme(strip.background = element_blank(),
  strip.text.x = element_blank(),
  text = element_text(size=16),
  axis.title = element_text(size=18)
  ) +
  ylab(expression(paste("Profit (",mu,"mol ", m^{-2}~s^{-1},")"))) +
  xlab("Time (hr)") -> a

  
data_for_plot %>%
  unnest(outputs) %>%
  ggplot() +
  geom_line(aes(x = instant_through_day_hr, y = transpiration)) + 
  scale_x_continuous(breaks = seq(0,24,4), labels = seq(0,24,4), limits = c(0,24)) +
  facet_wrap(~param_set, scales = "free_x") +
  theme_classic() +
  theme(strip.background = element_blank(),
  strip.text.x = element_blank(),
  text = element_text(size=16),
  axis.title = element_text(size=18)
  ) +
  ylab(expression(paste("Evapotranspiration (kg ", m^{-2}~s^{-1},")"))) +
  xlab("Time (hr)")-> b

cowplot::plot_grid(a,b, labels = c("a)", "b)"))
  

```


```{r}
expanded_output_integration %>%
  mutate(run = map(point_estimate_integrals, ~.x %>%
                     mutate(run = seq(1:n())))) %>% 
  unnest(run) %>% 
  nest(run, point_estimate_integrals, cols) %>% 
  unnest(data) %>% 
  unnest(cols) %>% 
  select(-point_estimate_integrals) %>%
  mutate(coords= pmap(., make_polygon_coordinates)) %>%
  group_by(param_set) %>%
  nest(params = -param_set) %>%
  ungroup() -> data_for_plot
  
data_for_plot %>%
  # slice(30) %>%
  {map(.$params, visualise_integrals, variable = "profit")} -> a

data_for_plot %>%
  # slice(30) %>%
  {map(.$params, visualise_integrals, variable = "transpiration")} -> b

a[[1]] +
  scale_x_continuous(breaks = seq(0,24,4), labels = seq(0,24,4), limits = c(0,24)) +
  theme_classic() +
  theme(strip.background = element_blank(),
  strip.text.x = element_blank(),
  text = element_text(size=16),
  axis.title = element_text(size=18)
  ) +
  ylab(expression(paste("Profit (",mu,"mol ", m^{-2}~s^{-1},")"))) +
  xlab("Time (hr)") -> a

b[[1]] +
  scale_x_continuous(breaks = seq(0,24,4), labels = seq(0,24,4), limits = c(0,24)) +
  theme_classic() +
  theme(strip.background = element_blank(),
  strip.text.x = element_blank(),
  text = element_text(size=16),
  axis.title = element_text(size=18)
  ) +
  ylab(expression(paste("Evapotranspiration (kg ", m^{-2}~s^{-1},")"))) +
  xlab("Time (hr)") -> b
  

cowplot::plot_grid(a,b, labels = c("a)", "b)"))

```


```{r}
#add example points and also increase the number of them

expanded_output %>% 
  ggplot(aes(x =integrated_daily_profit,  y= one_point_profit)) +
  geom_point() +
  geom_abline(slope = 1, intercept =0) +
  theme_classic() +
  xlab("Integrated daily profit") +
  ylab("One-point esimate daily profit") +
  geom_text(data = expanded_output %>%
               filter(abs(integrated_daily_profit - 13.89234) == min(abs(integrated_daily_profit - 13.89234)) |
                        abs(integrated_daily_profit - 46.13512) == min(abs(integrated_daily_profit - 46.13512)) | 
                        abs(integrated_daily_profit - 89.16447) == min(abs(integrated_daily_profit - 89.16447))),
             aes(x = integrated_daily_profit, y = one_point_profit, label = c("B", "D", "C")), nudge_y = c(-2, -2, -2)) +
  geom_point(data = expanded_output %>%
               filter(abs(integrated_daily_profit - 13.89234) == min(abs(integrated_daily_profit - 13.89234)) |
                        abs(integrated_daily_profit - 46.13512) == min(abs(integrated_daily_profit - 46.13512)) | 
                        abs(integrated_daily_profit - 89.16447) == min(abs(integrated_daily_profit - 89.16447))),
             aes(x = integrated_daily_profit, y = one_point_profit), size = 3, shape = 21) +
    geom_point(data = expanded_output %>%
               filter(abs(integrated_daily_profit - 13.89234) == min(abs(integrated_daily_profit - 13.89234)) |
                        abs(integrated_daily_profit - 46.13512) == min(abs(integrated_daily_profit - 46.13512)) | 
                        abs(integrated_daily_profit - 89.16447) == min(abs(integrated_daily_profit - 89.16447))),
             aes(x = integrated_daily_profit, y = one_point_profit), colour = "maroon")-> a

expanded_output %>%
  filter(abs(integrated_daily_profit - 13.89234) == min(abs(integrated_daily_profit - 13.89234))) %>%
  unnest(outputs) %>%
  ggplot(aes(x = instant_through_day_hr, y= profit)) +
  geom_line() +
  geom_polygon(data = . %>%
  filter(abs(instant_through_day_hr - 12) == min(abs(instant_through_day_hr - 12))) %>%
    sample_n(1) %>%
  tibble(x = c(sun_rise, sun_rise, 12, sun_down, sun_down), y = c(0, min(profit), min(profit), min(profit), 0)),
  aes(x = x, y = y), fill = NA, colour = "black", linetype = "dashed") +
  geom_point(data = . %>%
  filter(abs(instant_through_day_hr - 12) == min(abs(instant_through_day_hr - 12))) %>%sample_n(1),
  aes(x = instant_through_day_hr, y = profit), colour = "black") +
  xlab("Time (hrs)") +
  ylab(expression(Profit~(umol~mol^{-1}~m^{-2}~s^{-1}))) +
  theme_classic() -> b

expanded_output %>%
    filter(abs(integrated_daily_profit - 46.13512) == min(abs(integrated_daily_profit - 46.13512))) %>%
    unnest(outputs) %>%
  ggplot(aes(x = instant_through_day_hr, y= profit)) +
  geom_line() +
  geom_polygon(data = . %>%
  filter(abs(instant_through_day_hr - 12) == min(abs(instant_through_day_hr - 12))) %>%
  tibble(x = c(sun_rise, sun_rise, 12, sun_down, sun_down), y = c(0, min(profit), min(profit), min(profit), 0)),
  aes(x = x, y = y), fill = NA, colour = "black", linetype = "dashed") +
  geom_point(data = . %>%
  filter(abs(instant_through_day_hr - 12) == min(abs(instant_through_day_hr - 12))),
  aes(x = instant_through_day_hr, y = profit), colour = "black") +
  xlab("Time (hrs)") +
  ylab(expression(Profit~(umol~mol^{-1}~m^{-2}~s^{-1}))) +
  theme_classic()-> c

expanded_output %>%
    filter(abs(integrated_daily_profit - 89.16447) == min(abs(integrated_daily_profit - 89.16447))) %>%
    unnest(outputs) %>%
  ggplot(aes(x = instant_through_day_hr, y= profit)) +
  geom_line() +
  geom_polygon(data = . %>%
  filter(abs(instant_through_day_hr - 12) == min(abs(instant_through_day_hr - 12))) %>%
    sample_n(1) %>%
    tibble(x = c(sun_rise, sun_rise, 12, sun_down, sun_down), y = c(0, min(profit), min(profit), min(profit), 0)),
  aes(x = x, y = y), fill = NA, colour = "black", linetype = "dashed") +
  geom_point(data = . %>%
  filter(abs(instant_through_day_hr - 12) == min(abs(instant_through_day_hr - 12))) %>%
    sample_n(1),
  aes(x = instant_through_day_hr, y = profit), colour = "black") +
  xlab("Time (hrs)") +
  ylab(expression(Profit~(umol~mol^{-1}~m^{-2}~s^{-1}))) +
  theme_classic()-> d
```


```{r}
png("outputs.png", height = 850, width = 850)
cowplot::plot_grid(a,b,c,d, nrow = 2, ncol = 2, labels = "AUTO")
dev.off()
```

```{r}
expanded_output %>% 
  filter(integrated_daily_profit > 0 & one_point_profit > 0) %>% 
  ggplot(aes(x =integrated_daily_profit,  y= one_point_profit)) +
  geom_hex() +
  geom_abline(slope = 1, intercept =0) +
  theme_classic() +
  xlab("Integrated daily profit") +
  ylab("One-point esimate daily profit") -> p

png("outputs_hex.png", height = 850, width = 850)
p
dev.off()

```

```{r}
focal_parameters <- expand_grid(continuous_parameter = c("ca","VP","day","psi_soil", "temp_diff"), factorial_parameter = c("vcmax", "p_50","huber_value","h"))
```

```{r}

set_focal_parameters <- function(...){
  
  focal_parameters = tibble(...)
if(focal_parameters$continuous_parameter == "ca") {
  ca = seq(30, 45, length.out =10)
} else {
  ca = 40
}

if(focal_parameters$continuous_parameter == "VP") {
  VP = seq(0.5, 3, length.out = 10)
} else {
  VP = 1.5
}


if(focal_parameters$continuous_parameter == "day") {
  day = seq(0, 180, length.out = 10)
} else {
  day = 60
}


if(focal_parameters$continuous_parameter == "psi_soil") {
  psi_soil = seq(0, 5, length.out = 10)
} else {
  psi_soil = 1.5
}

if(focal_parameters$continuous_parameter == "temp_diff") {
  temp_diff = seq(0, 15, length.out = 10)
} else {
  temp_diff = 5
}  
  
if(focal_parameters$factorial_parameter == "vcmax") {
  vcmax = c(30,100)
} else {
  vcmax = 70
}

if(focal_parameters$factorial_parameter == "p_50") {
  p_50 = c(1, 6)
} else {
  p_50 = 3
}


if(focal_parameters$factorial_parameter == "huber_value") {
  huber_value = c(0.000157/2, 0.000157*2)
} else {
  huber_value = 0.000157
}


if(focal_parameters$factorial_parameter == "h") {
  h = c(10, 80)
} else {
  h = 10
}
  
expand_grid(mean_temp = 20, temp_diff = temp_diff, latitude = 27, VP = VP,
                   vcmax = vcmax, p_50 = p_50, huber_value = huber_value,
                   day = day, h = h, E = 0.7, psi_soil = psi_soil, ca = ca)  
}

focal_parameters %>%
  pmap(set_focal_parameters) %>%
  map(nest) %>%
  bind_rows() %>%
  mutate(run = 1:nrow(.)) %>%
  unnest(data) %>%
  pmap(expand_input_data) %>%
  bind_rows() %>%
  filter(PAR > 1e-10) %>%
  pmap(calc_profit_instant) -> object

object %>%
  bind_rows() %>%
  nest(outputs = c(instant_through_day_hr, instant_through_day_dec, dec_day_time, solar_angle, PAR, temp, VPD_hr, profit))%>%
  mutate(integrated_daily_profit = map_dbl(outputs, ~plant:::trapezium(.x$instant_through_day_hr, .x$profit))) %>%
  mutate(one_point_profit = map_dbl(outputs, calc_one_point)) %>%
  mutate(integrated_PAR = map_dbl(outputs, ~plant:::trapezium(.x$instant_through_day_hr, .x$PAR)))-> expanded_output

plotting_function <- function(...){
  
  if(focal_parameters$continuous_parameter[...] == "day"){
    expanded_output %>%
    filter(run == ...) %>%
    pivot_longer(cols = c(one_point_profit, integrated_daily_profit)) %>%
    ggplot(aes(x = .data[["integrated_PAR"]], y = value)) +
    geom_line(aes(group = interaction(name, .data[[set_names(focal_parameters$factorial_parameter[...])]]), colour = .data[[set_names(focal_parameters$factorial_parameter[...])]], linetype = name)) +
    theme_classic() +
    ylab("Profit") 
  } else{
  
  expanded_output %>%
    filter(run == ...) %>%
    pivot_longer(cols = c(one_point_profit, integrated_daily_profit)) %>%
    ggplot(aes(x = .data[[set_names(focal_parameters$continuous_parameter[...])]], y = value)) +
    geom_line(aes(group = interaction(name, .data[[set_names(focal_parameters$factorial_parameter[...])]]), colour = .data[[set_names(focal_parameters$factorial_parameter[...])]], linetype = name)) +
    theme_classic() +
    ylab("Profit")
  }
}


list_of_plots <- map(1:20, plotting_function)
cowplot::plot_grid(plotlist = list_of_plots)  
```



```{r}
params <- expand_grid(mean_temp = 20, temp_diff = 5, latitude = 27, VP = c(1,3),
                   vcmax = 70, p_50 = 1, huber_value = 0.000157,
                   day = 60, h = 10, E = 0.7, psi_soil = 1.5, ca = seq(30, 45, 1))
```

```{r}
PPFD_curve <- params %>%
  pmap(expand_input_data) %>%
  bind_rows() %>%
  filter(PAR > 1e-10) %>% 
  pmap(calc_profit_instant) %>%
  bind_rows() %>%
  nest(outputs = c(instant_through_day_hr, instant_through_day_dec, dec_day_time, solar_angle, PAR, temp, VPD_hr, profit))
```

```{r}
PPFD_curve %>%
  mutate(integrated_daily_profit = map_dbl(outputs, ~plant:::trapezium(.x$instant_through_day_hr, .x$profit))) %>%
  mutate(one_point_profit = map_dbl(outputs, calc_one_point)) -> expanded_output

expanded_output %>% 
  pivot_longer(cols = c(integrated_daily_profit, one_point_profit)) %>%
  ggplot(aes(x =ca,  y= value)) +
  geom_line(aes(group=interaction(name, VP), col = name, linetype = as.factor(VP))) +
  theme_classic() 

```


```{r}
exanded_input_data2 <- params %>%
  future_pmap(expand_input_data) %>%
  bind_rows() %>%
  filter(PAR > 1e-10) %>% 
  future_pmap(calc_profit_instant) %>%
  bind_rows() %>%
  nest(outputs = c(instant_through_day_hr, instant_through_day_dec, dec_day_time, solar_angle, PAR, temp, VPD_hr, profit))
```

```{r}
exanded_input_data2 %>%
  mutate(integrated_daily_profit = map_dbl(outputs, ~plant:::trapezium(.x$instant_through_day_hr, .x$profit))) %>%
  # slice(1:5000) %>%
  expand_grid(focal_points = seq(1,10,1)) %>%
  mutate(point_estimate_profit = pmap_dbl(., calc_mult_point)) -> expanded_output2
```


```{r}
calc_mult_point <- function(...){
  
  data = tibble(...) %>%
    nest(outputs = outputs)
  first_num = 1/data$focal_points/2 
  quantiles = seq(from = first_num, 1, by = first_num*2)

  data %>%
    unnest(outputs) -> data
  
  extract_time_slice <- function(quantile){
    data %>%
  filter(abs(outputs$instant_through_day_hr - quantile(outputs$instant_through_day_hr, quantile)) %in% min(abs(outputs$instant_through_day_hr - quantile(outputs$instant_through_day_hr, quantile)))) %>%
      sample_n(1)
  }
  
    map(quantiles, extract_time_slice) %>%
    bind_rows() %>%
    mutate(profit_integrated = outputs$profit * day_length/focal_points) %>%
    summarise(daily_profit_integrated = sum(profit_integrated)) -> out
  
    out$daily_profit_integrated

    }
```

```{r}
expanded_output2 %>%
  filter(integrated_daily_profit != 0 & point_estimate_profit != 0) %>%
  ggplot(aes(x = integrated_daily_profit, y = point_estimate_profit)) +
  geom_point() +
  theme_classic() +
  facet_wrap(~focal_points) +
  geom_abline(slope = 1, intercept = 0) -> a
```

```{r}
expanded_output2 %>%
  filter(integrated_daily_profit != 0 & point_estimate_profit != 0) %>%
  group_by(focal_points) %>%
  mutate(correlation = cor(integrated_daily_profit, point_estimate_profit)) %>%
  ggplot(aes(x = focal_points, y = correlation)) +
  geom_line() +
  theme_classic() -> b
```

```{r}
expanded_output2 %>%
  group_by(focal_points) %>%
  mutate(sample_size = n()) %>%
  mutate(diff = integrated_daily_profit - point_estimate_profit) %>%
  mutate(bias = sum(diff)/sample_size) %>%
  mutate(rmse = sqrt(sum(diff^2)/sample_size)) %>%
  ggplot(aes(x = focal_points, y = rmse)) +
  geom_line() +
  theme_classic() -> c
```

```{r}
expanded_output2 %>%
  group_by(focal_points) %>%
  mutate(sample_size = n()) %>%
  mutate(diff = integrated_daily_profit - point_estimate_profit) %>%
  mutate(bias = sum(diff)/sample_size) %>%
  mutate(rmse = sqrt(sum(diff^2)/sample_size)) %>%
  ggplot(aes(x = focal_points, y = bias)) +
  geom_line() +
  theme_classic() -> d
```

```{r}
png("outputs_multi_point.png", height = 850, width = 850)
cowplot::plot_grid(a,b,c,d, nrow = 2, ncol = 2, labels = "AUTO")
dev.off()
```

```{r}
expanded_output2 %>%
  group_by(focal_points) %>%
  nest() %>%
  ungroup() %>%
  mutate(asd = purrr::map(data, linear_model_func)) %>%
  pull(asd)
  
linear_model_func <- function(...){
  data <- tibble(...) 
  fit <- lm(point_estimate_profit ~ integrated_daily_profit + 0, data)
  summary(fit)
}  



```





























```{r}
example_params <- tibble(mean_temp = 20, temp_diff = 17.8, latitude = 15.3, VP = 1.52, vcmax = 35.7, p_50 = 2.35, huber_value = 0.000250, day = 68.7, h = 8.95, E = 0.0376, psi_soil = 0.643, ca = 42.6) %>%
  expand_input_data() %>%
  filter(PAR > 1e-10) %>%
  ungroup() %>%
  pmap(calc_profit_instant) %>%
  bind_rows() 
example_params %>% View()
example_params %>%
  mutate(VPD_hr_min = (VPD_hr == min(VPD_hr))) %>%
  pivot_longer(cols = c(profit,ci, VPD_hr, PAR)) %>%
  ggplot(aes(x= instant_through_day_hr, y=value)) +
  geom_line(aes(colour= VPD_hr_min)) +
  facet_wrap(~name, nrow= 2, scales= "free") 

example_params %>% slice(198) -> example_params


vcmax = example_params$vcmax #maximum carboxylation rate, defined by leaf nitrogen (umol m^-2 s^-1) 
p_50 = example_params$p_50 #stem water potential at 50% loss of conductivity
c = 2.04 #shape parameter for hydraulic vulnerability curve (unitless) estimated from trait data in Austraits from Choat et al. 2012
b = calc_vul_b(p_50, c) #shape parameter for vulnerability curve, point of 37% conductance (-MPa) 
psi_crit = calc_psi_crit(b, c) #stem water potential at which conductance is 95%
huber_value = example_params$huber_value #huber value (m^2 sapwood area m^-2 leaf area)
K_s = p_50_2_K_s(p_50) #stem-specific conductivity (kg h2o m^-1 stem s^-1 MPa^-1)
h = example_params$h #height or path length (m)
k_l_max = calc_k_l_max(K_s, huber_value, h)
l <- plant:::Leaf(vcmax = vcmax, p_50 = p_50, c = c, b = b, psi_crit = psi_crit, huber_value = huber_value, K_s = K_s, epsilon_leaf = 0.0001)
l$set_physiology(PPFD = example_params$PAR*example_params$E, psi_soil = example_params$psi_soil, k_l_max = k_l_max, atm_vpd = example_params$VPD_hr, example_params$ca)
l$get_leaf_states_rates_from_psi_stem(psi_crit)
tibble(psi = seq(l$psi_soil_, psi_crit, length.out = 100)) %>%
  rowwise() %>%
  mutate(profit = l$profit_psi_stem_Sperry(psi)) %>%
  mutate(cost = l$calc_hydraulic_cost_Sperry(psi)) %>%
  mutate(lambda = l$lambda_) %>%
  mutate(benefit = profit + cost*lambda) %>%
  mutate(cost_trans = cost*lambda) %>%
  pivot_longer(cols = c(profit, cost_trans, benefit)) %>%
  ggplot(aes(x=psi, y = value)) +
  geom_line(aes(group = name, colour = name))
  bind_rows(tibble(psi = l$opt_psi_stem, profit = l$profit)) %>%
  ggplot(aes(x = psi , y= profit)) +
  geom_line() +
  geom_point(aes(x = l$opt_psi_stem, y= l$profit))




l$optimise_psi_stem_Sperry()
l$opt_psi_stem
l$profit
l$get_leaf_states_rates_from_psi_stem(psi_crit)
max_ci <- l$ci

l$optimise_ci_Sperry(max_ci) 

l$opt_ci - max_ci
l$psi_stem

l$psi_soil_
l$g_c
```


```{r}

if (is.na(opt_psi_stem) | (opt_psi_stem > example_params$psi_crit)){

  max_psi = 1;

  opt_psi_stem = example_params$psi_crit - diff_value;
}

if((count > 15) | ((R_IsNA(opt_psi_stem) | (opt_psi_stem > psi_crit) | (opt_psi_stem < psi_soil)) && max_psi == 3)){
      method = 1;

      optimise_psi_stem_Sperry();
      return;
    }

  if ((R_IsNA(opt_psi_stem) | (opt_psi_stem > example_params$psi_crit)) && max_psi == 1){

std::cout << "using lower value" << std::endl;

    count = 0;
    max_psi = 2;

    opt_psi_stem = leaf$psi_soil_ + diff_value;

  }

    if ((R_IsNA(opt_psi_stem) | (opt_psi_stem > psi_crit)) && max_psi == 2){

std::cout << "using middle value" << std::endl;

    count = 0;

    max_psi = 3;

    opt_psi_stem = (psi_soil_ + psi_crit)/2;

  }

    psi_stem_initial = opt_psi_stem;

    x_1 = psi_stem_initial - diff_value;
    x_2 = psi_stem_initial + diff_value;
    
    y_2 = leaf[[1]]$profit_psi_stem_Sperry(x_2);

    y_1 = leaf[[1]]$profit_psi_stem_Sperry(x_1);

    y_0 = leaf[[1]]$profit_psi_stem_Sperry(psi_stem_initial);

    first_dev = (y_2 - y_1)/(2*diff_value);
    sec_dev = (y_2 - 2*y_0 + y_1)/diff_value^2;

    opt_psi_stem = psi_stem_initial -  first_dev/sec_dev;

    abs(opt_psi_stem - psi_stem_initial) < 0.0001
    leaf$profit_psi_stem_Sperry(opt_psi_stem)
    if(abs(opt_psi_stem - psi_stem_initial) < epsilon_leaf){
      finished = 0;
    }

    profit = y_0;
  }
    return;
  }
```




map(expanded_output$outputs, ~.x %>% ggplot(aes(x = instant_through_day_hr, y=profit)) + geom_line())

map(expanded_output$outputs, ~.x %>% ggplot(aes(x = instant_through_day_hr, y=transpiration)) + geom_line())

map(expanded_output$outputs, ~.x %>% ggplot(aes(x = instant_through_day_hr, y=psi_stem)) + geom_line())

map(expanded_output$outputs[24], ~.x %>% ggplot(aes(x = instant_through_day_hr, y=g_c)) + geom_line())
```

```{r}
expanded_output %>% slice(24) %>%unnest(outputs) %>%
mutate(profit_outputs = map2(leaf, psi_crit, profit_curves_from_leaf)) %>%
  mutate(profit_max = map_dbl(profit_outputs, ~max(.x$inst_profit))) %>%
  unnest(profit_outputs) %>%
  ggplot() +
  geom_line(aes(x = inst_psi_stem, y = inst_profit), col = "red") +
  geom_line(aes(x = inst_psi_stem, y = inst_cost), col = "blue") +
  geom_line(aes(x = inst_psi_stem, y = inst_ben), col = "green") +
  geom_vline(aes(xintercept = psi_stem)) +
  geom_hline(aes(yintercept = profit_max)) +
  facet_wrap(~instant_through_day_hr)
```


expanded_output_integration %>%
  ggplot(aes(x = integrated_daily_profit, y = point_estimate_integrated_daily_profit)) +
  geom_point() +
  facet_wrap(~focal_points)


expanded_output_integration %>%
  ggplot(aes(x = integrated_daily_transpiration, y = point_estimate_integrated_daily_transpiraiton)) +
  geom_point() +
  facet_wrap(~focal_points)
```
