---
title: "water_bucket_model"
author: "Isaac Towers"
date: "19/05/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE}
library(tidyverse)
library(deSolve)
```

Here I will develop the prototype for the water bucket model for the **plant** model tracking the rate of change in $\theta$


$$\frac{d\theta}{dt} = \frac{R *I(\theta) - E_T (\theta) - E_S (\theta) - q_D(\theta)}{z} $$
Variables

$\theta ( t)$: soil volumetric water content $(m^3 / m^3)$

Parameters

$R$: Rainfall $(mm / t)$
$I$: Infiltration rate (unitless)
$E_T$: Transpiration (water used by plants) $(mm / t)$
$E_S$: Soil surface evaporation $(mm / t)$
$q_D$: Drainage from soil profile $(mm /t)$
$z$: Soil depth (mm)

$$I(\theta) =  1-\bigg(\frac{\theta(t)}{\theta_{sat}}\bigg)^b$$
$b$: Unitless parameter determining the shape of water accumulation in the soil moisture bank
$\theta_sat$ Soil moisture capacity $(m^3 m^{-3})$

Ok, lets first show that rain will eventually fill up the soil without any water outputs if we include a saturated water content
$$\frac{d\theta}{dt} = R *\bigg(1-\bigg(\frac{\theta(t)}{\theta_{sat}}\bigg)^b\bigg)$$

```{r warning=FALSE}
# Timestep
t <- seq(0,365, by = 1)

# Variables
theta_init <- c(theta=0.210)

# Model parameters
params <- 
  expand_grid(theta_sat = 0.513, 
            R  = 1000/365,
            b = c(1, 0.5, 2, 4, 8),
            z = 1000) %>%
        split(., .$b)

# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    
    dtheta_dt = (R*(1-(theta/theta_sat)^b))/z

    return(list(c(dtheta_dt)))
  })
}
```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "b")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta, col=b)) +
  geom_line()

```

Now let's add evaporation. We will use the Penman-Monteith equation to model evaporation as a water flux (rather than an energy flux):

$$E=\frac{\Delta*\varphi_s+g_a*p_a*c_{pa}*D}{\Delta+\gamma(1+\frac{g_a}{g_s})}*\frac{1}{L},$$

Where E is the evaporation in m (i.e. $mm$). This equation is dependent on the air temperature, which manifests its effect on evaporation through determining the vapour pressure deficit $D$, and the derivative of the saturated vapour pressure curve at a given air temperature $\Delta$, the net radiation reaching the soil surface $\varphi_s$, and soil moisture which affects the soil conductance $g_s$.

The equation for the saturated vapour pressure is dependent on air temperature $(T_{air}~(^\circ C))$ only:

$$e_s=0.6108*exp\bigg(\frac{17.27*T_{air}}{T_{air} + 237.3}\bigg).$$
Thus, $$\Delta = \frac{\delta e_s}{\delta T_{air}}.$$

To estimate the daily average vapour pressure deficit, the maximum $T_{max}$ and minimum $T_{min}$ air temperature can be used $$\overline{D} = e_s(T_{max})-e_s(T_{min}),$$


and the daytime average vapour pressure deficit $D_sd$ can be calculated by assuming a fraction of daytime hours $\alpha$: $$\overline{D_{daytime}} =  \alpha * e_s(T_{max})- (1-\alpha) * e_s(T_{min}).$$


Radiation reaching the soil surface (i.e. passing through the tree canopy) is assumed to be entirely absorbed. Thus, net radiation at the soil surface is defined as the radiation incident on the tree canopy $\varphi_0$ minus the radiation absorbed the tree canopy $\varphi_na$

$$\varphi_s = \varphi_0 - \varphi_{na},$$
where $\varphi_{na}$ is dependent on the leaf area index (LAI) of the canopy and the light extinction coefficient (k):

$$\varphi_{na} = \varphi_0*(1-e^{-k*LAI})$$

Conductance of water vapour through the soil moisture boundary layer $g_c$ is dependent on soil moisture, because resistance to evaporation increases as the soil becomes increasingly dry according the following equation. Using an equation proposed by Soares and Almeida (2001):

$$g_{s} = g_{s_{max}}*(\theta/\theta_{sat})$$
Once through the soil moisture boundary layer, water vapour must then travel through the atmospheric boundary layer according to the rate $g_a$. For soil, a fixed value is currently assumed using value obtained from Soares and Almeida (2001) where $g_{a_s} = 0.01 m~s^{-1}$

```{r warning=FALSE}
# Timestep
t <- seq(0,3650, by = 1)

# Variables
theta_init <- c(theta=0.513)

# Model parameters
params <- 
  expand_grid(theta_sat = 0.513,
            R  = 1000/365,
            b = 8,
            gcmax=0.0025,
            gb=0.01,
            vpd=900,
            p = 1.204,
            cpa = 1004, 
            y=66.1, 
            s=145,
            air_temp=25,
            slope_sat_vap_curve = 4098*(0.6108*exp((17.27*air_temp)/(air_temp+237.3)))/(air_temp+237.3)^2 *1000,
            #radiation will be fed in from plant model based on its solar radiation submodel and plant lai
            rad = c(200,434),
            max_air_temp = 25,
            min_air_temp= 10,
            L = 2.45*10^6,
            z = 1500) %>%
            split(., .$rad)

# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    evap_numerator = slope_sat_vap_curve *rad + gb*p*cpa*vpd
    evap_denominator = slope_sat_vap_curve + y*(1+gb/(gcmax*(theta/theta_sat)))
    evap=evap_numerator/evap_denominator*(1/L)*12*60*60
    dtheta_dt = (R*(1-(theta/theta_sat)^b)- evap)/z

    return(list(c(dtheta_dt)))
  })
}
```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "rad")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta, col=rad)) +
  geom_line()

```


Cool, let's now include a drainage factor

$$\frac{d\theta}{dt} = R *\bigg(1-\bigg(\frac{\theta(t)}{\theta_{sat}}\bigg)^b\bigg)-J(\theta_z, \theta_{z+x})$$
Soil water movement through the soil at a given depth, $z$ is defined as:

$$J(\theta_z, \theta_{z+x}) = -D_z(\theta_z)\frac{\delta\theta_z}{\delta{x}},$$
Where the diffusivity, $D$ $(m^2 s^{-1})$ is defined as:

$$D_z(\theta_z) = \frac{K_z}{g*p_w}\frac{\delta\psi_d}{\delta\theta_d},$$
and the hydraulic conductivity $K_d$ $(m~yr^{-1})$ is defined as:

$$K_z=K_{sat}\bigg(\frac{\theta_z}{\theta_{sat}}\bigg)^{nk}.$$
$K_{sat}$ $(m~yr^{-1})$ and $nk$ (unitless) are empirical parameter which vary with soil type, $g$ is acceleration due to gravity $(9.8~m~s^{-2})$ and $p_w$ is the density of water $(1000~kg~m^{-3})$. Finally, $\psi$ is the soil water potential.

Because the soil water potential at a given depth is related to soil moisture content according to the following equation:

$$\psi_z=-a_\psi \theta_z^{-n_\psi},$$
where $a_\psi$ and $n_\psi$ are soil-dependent empirical parameters. It follows that diffusivity can be calculated in terms of soil moisture content at a given depth according to the following equation:

$$D_z(\theta_z) = \frac{K_z}{g*p_w}a_{\psi}n_{\psi}\theta_z^{-n_{\psi}-1},$$


```{r warning=FALSE}
# Timestep
t <- seq(0,100, by = 1)

# Variables
theta_init <- c(theta0=0.513)

# Model parameters
params <- 
  expand_grid(
    R = 1/365,
    b = c(2),
    theta_sat = 0.513, 
    ds = 0.86,
    nd = 8.5,
    ) %>%
        split(., .$b)

# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    
    dtheta0_dt =R-ds*((theta0/theta_sat)^nd)
    
    return(list(c(dtheta0_dt)))
  })
}

```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "b")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta0, col=b)) +
  geom_line()

```


```{r warning=FALSE}
# Timestep
t <- seq(0,100, by = 0.1)

# Variables
theta_init <- c(theta1=0.513, theta2=0, theta3=0)

# Model parameters
params <- 
  expand_grid(
    theta_sat = 0.513, 
    ds = 0.86,
    nd = 8.5,
    R = 0.00274
    ) %>%
        split(., .$ds)

# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    
    dtheta_dt1 = R -(ds*((theta1/theta_sat)^nd)) * (theta1-theta2)
    dtheta_dt2 = ((ds*((theta1/theta_sat)^nd)) * (theta1-theta2)) - ((ds*((theta2/theta_sat)^nd)) * (theta2-theta3))
    dtheta_dt3 = ((ds*((theta2/theta_sat)^nd)) * (theta2-theta3))

    return(list(c(dtheta_dt1,dtheta_dt2,dtheta_dt3)))
  })
}

```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "ks")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta1, col=ks)) +
  geom_line()

logistic_solution %>% 
  ggplot(aes(time, theta2, col=ks)) +
  geom_line()

logistic_solution %>% 
  ggplot(aes(time, theta3, col=ks)) +
  geom_line()


```



```{r warning=FALSE}
# Timestep
t <- seq(0,10, by = 0.1)

# Variables
theta_init <- c(theta=0.482)

# Model parameters
params <- 
  expand_grid(
    ks = c(0.00021388),
    theta_sat = 0.482, 
    nk = 0.007,
    g = 9.8,
    pw = 1000,
    apsi = 1.87,
    npsi = 11.3) %>%
        split(., .$ks)

# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    
    dtheta_dt = -(((ks*(theta/theta_sat)^nk)/(g*pw)) * apsi * npsi * theta^(-npsi-1))

    return(list(c(dtheta_dt)))
  })
}

```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "ks")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta, col=ks)) +
  geom_line()


```

Parameters

$K_s$: Saturated soil hydraulic conductivity $(m^3 m^{-2} yr^{-1})$
$c$: Unitless empirical soil parameter

```{r warning=FALSE}
# Timestep
t <- seq(0,10, by = 0.01)

# Variables
theta_init <- c(theta=0)

# Model parameters
params <- 
  expand_grid(theta_sat = 0.482, 
            R  = 1.0,
            ks = c(40.47, 146),
            c = 4,
            b = 4) %>%
        split(., .$ks)

# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    
    dtheta_dt = R*(1-(theta/theta_sat)^b)-ks*(theta/theta_sat)^c

    return(list(c(dtheta_dt)))
  })
}

```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "ks")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta, col=ks)) +
  geom_line()


```
Now with climatic variability of amplitude = the annual mean rainfall

```{r}
# Timestep
t <- seq(0,10, by = 0.01)

# Variables
theta_init <- c(theta=0)

# Model parameters
params <- 
  expand_grid(theta_sat = 0.482, 
            R  = 1.0,
            ks = c(40.47, 5550.37),
            c = 11,
            b = 4) %>%
        split(., .$ks)

# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    
    dtheta_dt = (R+R*sin(t/(2*pi)^-1))*(1-(theta/theta_sat)^b)-ks*(theta/theta_sat)^c

    return(list(c(dtheta_dt)))
  })
}

```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "ks")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta, col=ks)) +
  geom_line()


```
```{r}
# Timestep
t <- seq(0,10, by = 0.01)

# Variables
theta_init <- c(theta=0)

# Model parameters
params <- 
  expand_grid(theta_sat = 0.482, 
            R  = 1.0,
            ks = c(40.47),
            c = 11,
            b = 4,
            pe = c(0.5,1,1.4),
            k = 0.46,
            LAI = 0) %>%
        split(., .$pe)


# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    
    dtheta_dt = R*(1-(theta/theta_sat)^b)-ks*(theta/theta_sat)^c - 
    pe*(exp(-k * LAI)/(1 + exp (-12 * (theta /theta_sat - 0.5))))

    return(list(c(dtheta_dt)))
  })
}

```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "pe")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta, col=pe)) +
  geom_line()


```
```{r}
# Timestep
t <- seq(0,10, by = 0.01)

# Variables
theta_init <- c(theta=0)

# Model parameters
params <- 
  expand_grid(theta_sat = 0.482, 
            R  = 1.0,
            ks = c(40.47),
            c = 11,
            b = 4,
            pe = c(0.5, 1,1.4),
            k = 0.46,
            LAI = 0) %>%
        split(., .$pe)


# Rates of change
theta_rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {
    
    dtheta_dt = (R+R*sin(t/(2*pi)^-1))*(1-(theta/theta_sat)^b)-ks*(theta/theta_sat)^c - 
    pe*(exp(-k * LAI)/(1 + exp (-12 * (theta /theta_sat - 0.5))))

    return(list(c(dtheta_dt)))
  })
}

```

```{r}

logistic_solution <-
  map_dfr(
    params,
    ~ode(
    y = theta_init,
    times = t,
    func = theta_rates,
    parms = .x) %>%
  as.data.frame(), .id = "pe")
```


```{r}

logistic_solution %>% 
  ggplot(aes(time, theta, col=pe)) +
  geom_line()


```

Evaporation should occur only in the top 10 cm of soil https://pure.mpg.de/rest/items/item_1781494/component/file_1786328/content

Penman-montheith equation for transpiration

$$$$

```{r}
145*radiation+boundaryconductance * 1.204 *1004 * vapour_pressure_deficit / 145 +66.1(1)

pa = 101300*((293-0.0065*100)/293)^5.26
  
psych = 6.65 * 10^-4 * pa
delta <- (4098*es)/((298.16-35.86)^2)

es = 610.8 * exp((17.27*(298.16-273.16))/(298.16-35.86))

1.26*(delta/(delta + psych)) * 288

167*365
7000000 / (24*60*60)
```

