---
title: "Cohort spacing algorithm"
author: 
- Rich FitzJohn
- Daniel Falster
date: 2016
output:
  bookdown::html_document2:
    toc: yes
    number_sections: false
---

This vignette shows some details of cohort splitting.  It's
probably not very interesting to most people, only those
interested in knowing how the SCM technique works in detail.  It also uses a lot
of non-exported, non-documented functions from plant so you'll see
a lot of `plant:::` prefixes.

``` {r }
library(plant)
library(parallel)

p0 <- scm_base_parameters("FF16")
p <- expand_parameters(trait_matrix(0.0825, "lma"), p0, FALSE)
p$seed_rain <- 20 # close to equilibrium
```

The default cohort introduction times are designed to concentrate
cohort introductions onto earlier times, based on empirical
patterns of cohort refining:
``` {r }
t1 <- p$cohort_schedule_times[[1]]
plot(t1, cex=.2, pch=19, ylab="Time (years)", las=1)
```

The actual differences are stepped in order to increase the chance
that cohorts from different species will be introduced at the same
time and reduce the total amount of work being done.
``` {r }
plot(diff(t1), log="y", pch=19, cex=.2,
     ylab="Time difference (years)", las=1)
```

We can create more refined schedules by interleaving points between
these points:
``` {r }
interleave <- function(x) {
  n <- length(x)
  xp <- c(x[-n] + diff(x) / 2, NA)
  c(rbind(x, xp))[- 2 * n]
}

t2 <- interleave(t1)
t3 <- interleave(t2)
```

Consider running the SCM and computing seed rain at the end; this
is one of the key outputs from the model so a reasonable one to
look for differences in.
``` {r }
run_with_times <- function(p, t) {
  p$cohort_schedule_times[[1]] <- t
  run_scm(p)$seed_rains
}

sr_1 <- run_with_times(p, t1)
sr_2 <- run_with_times(p, t2)
sr_3 <- run_with_times(p, t3)
```

Seed rain increases as cohorts are introduced more finely,
though at a potentially saturating rate.  We're doing lots more
work at the more refined end!
``` {r }
sr <- c(sr_1, sr_2, sr_3)
sr
```

The differences in seed rain are not actually that striking
(perhaps `r signif(100 * diff(range(sr)) / mean(sr), 1)`%) but in
some runs can be more, and the *variation* creates instabilities.
``` {r }
plot(c(length(t1), length(t2), length(t3)), sr,
     las=1, xlab="Number of cohort introductions", ylab="Seed rain")
```

Where is the fitness difference coming from?

Consider adding a *single* additional cohort at one of the points
along the first vector of times `t1` and computing fitness:
``` {r }
insert_time <- function(i, x) {
  j <- seq_len(i)
  c(x[j], (x[i] + x[i+1])/2, x[-j])
}
run_with_insert <- function(i, p, t) {
  run_with_times(p, insert_time(i, t))
}
```

The internal function `plant:::run_scm_error` runs the SCM and
computes errors as the integration proceeds; this helps shed some
light.
``` {r }
i <- seq_len(length(t1) - 1)
res <- unlist(mclapply(i, run_with_insert, p, t1))
tm <- (t1[-1L] + t1[-length(t1)]) / 2
```

The biggest deviations in output seed rain come about half way
through the schedule:
``` {r }
plot(i, res - sr_1, xlab="Index", ylab="Seed rain", las=1)
abline(h=0, col="grey")
```

Though because of the compression of early times it's still fairly
early:
``` {r }
plot(tm, res - sr_1, xlab="Time (years)", ylab="Seed rain", las=1)
abline(h=0, col="grey")

ebt <- plant:::FF16_SCM(p)
ebt$run()
```

Now look at the contribution of different cohorts to see rain (x axis
log scaled for clarity).  In this case almost all the contribution
comes from early cohorts (this is essentially a single-age stand of
pioneers).  Overlaid on this are the five cohorts with the largest
change in total fitness (biggest difference in red).  The
difference is not coming from seed rain contributions from those
cohorts, which is basically zero, though it is higher than the
surrounding cohorts.
``` {r }
plot(t1[-1], ebt$seed_rain_cohort(1)[-1], log="xy",
     xlab="Time (years)", ylab="Seed rain")
j <- order(abs(res - sr_1), decreasing=TRUE)[1:5]
abline(v=tm[j], col=c("red", rep("grey", 4)))
```

Next up, need to work out what the fitness contribution of each
cohort is.
``` {r }
dat1 <- run_scm_collect(p)
p2 <- p
p2$cohort_schedule_times[[1]] <- insert_time(j[[1]], t1)
dat2 <- run_scm_collect(p2)
```

Then consider the light environment over time.  This reconstructs
the spline for the light environment for both runs, and computes
canopy openness in both and computes the difference in light
environments.  The resulting image plot is blue in regions where
the refined light environment is lighter (higher canopy openness)
and red in regions where the the light environment is darker in the
refined environment.
``` {r }
f <- function(e, h) {
  i <- plant:::Interpolator()
  i$init(e[, 1], e[, 2])
  y <- rep(1, length(h))
  j <- h < i$max
  y[j] <- i$eval(h[j])
  y
}
hmax <- max(dat1$light_env[[length(dat1$light_env)]][, "height"])
h <- seq(0, hmax, length.out=201)
y1 <- sapply(dat1$light_env, f, h)
y2 <- sapply(dat2$light_env, f, h)[, -(j[[1]] + 1L)]
dy <- t(y2 - y1)
dy[abs(dy) < 1e-10] <- NA
```
ColorBrewer's RdBu
``` {r }
cols <- c("#B2182B", "#D6604D", "#F4A582", "#FDDBC7",
          "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC")
pal <- colorRampPalette(cols)(20)
image(dat1$time, h, dy, xlab="Time (years)", ylab="Height (m)", las=1,
      col=pal)
```

Because the differences are mostly manifest in the leaf area, we
monitor error in both the leaf area and in fitness for all cohorts:
(black line indicates the cohort identified as problematic above)
``` {r }
dat <- plant:::run_scm_error(p)
image(dat1$time, dat1$time, dat$err$lai[[1]],
      xlab="Patch age (years)", ylab="Introduction time (years)", las=1)
abline(h=t1[j[[1]]])
```

We then run through rounds of refining cohorts until estimated
error has decreased down to an appropriate threshold:
``` {r }
p_refined <- build_schedule(p)
sr_refined <- attr(p_refined, "seed_rain_out")
```

Each round, the algorithm looks at the error in the leaf area
calculations and in the fitness calculations and refines the worst
cohorts, repeating as necessary.
``` {r }
t_refined <- p_refined$cohort_schedule_times[[1]]
i <- seq_len(length(t_refined) - 1)
res <- unlist(mclapply(i, run_with_insert, p_refined, t_refined))

tm <- (t1[-1L] + t1[-length(t1)]) / 2
```

The problem cohorts are still in about the same place, but are much
less pronounced (and less so if considering relative error)
``` {r }
plot(i, res - sr_refined, xlab="Index", ylab="Seed rain", las=1)
abline(h=0, col="grey")

plot(i, res / sr_refined - 1, xlab="Index", ylab="Seed rain", las=1)
```
