---
title: "Cohort spacing algorithm"
author: 
- Rich FitzJohn
- Daniel Falster
date: 2016
output:
  bookdown::html_document2:
    toc: yes
    number_sections: false
---

# Background

As described in the [demography vignette](https://traitecoevo.github.io/plant/articles/demography.html#approximation-of-size-density-distribution-via-the-scm),
the spacing of cohorts can affect the accuracy of integration over the
size-density distribution. `plant` uses an adaptive algorithm to build an 
appropriately spaced cohort schedule with the desired accuracy at every time 
point, using as few cohorts as possible.

The `build_schedule` function takes an
initial vector of introduction times and considers for each cohort
whether removing that cohort causes the error introduced when
integrating two specified functions over the size-density distribution to jump
over the allowable error threshold `schedule_eps`. This
calculation is repeated for every time step in the development of the
patch. A new cohort is introduced immediately prior to any cohort
failing these tests. The dynamics of the patch are
then simulated again and the process is repeated, until all integrations at
all time points have an error below the tolerable limit
`schedule_eps`. Decreasing `schedule_eps` demands higher
accuracy from the solver, and thus increases the number of cohorts being
introduced. Note that we are assessing whether removing an existing cohort
causes the error to jump above the threshold limit, and using this to decide
whether an extra cohort -- in addition to the one used in the test --
should be introduced. Thus, the actual error is likely to
be lower than, but at most equal to, `schedule_eps`.

This vignette shows some details of cohort splitting, most of which happens 
automatically. It's probably not very interesting to most people, only those
interested in knowing how the SCM technique works in detail.  It also uses a lot
of non-exported, non-documented functions from plant so you'll see
a lot of `plant:::` prefixes. 


# Cohort introduction times

<!-- TODO: elaborate on why concentrating concentrating early helps -->

The default cohort introduction times are designed to concentrate
cohort introductions onto earlier times, based on empirical
patterns of cohort refining:

```{r}
library(plant)
library(parallel)
n_cores = max(1, detectCores() - 1)

params <- scm_base_parameters("FF16")
patch <- expand_parameters(trait_matrix(0.0825, "lma"), params, mutant = FALSE)
patch$seed_rain <- 20

times_default <- patch$cohort_schedule_times[[1]]
plot(times_default, cex=.2, pch=19, ylab="Time (years)", las=1)
```
<!-- TODO: explain how introducing two species together reduces work -->

The actual differences are stepped in order to increase the chance
that cohorts from different species will be introduced at the same
time and reduce the total amount of work being done.

```{r}
plot(diff(times_default), pch=19, cex=.2, ylab="Time difference (years)", las=1)
```


# Seed rain

We can assess the impact of cohort spacing on integration error by examining
the cumulative propagule output (e.g. seed rain) of a patch.

Increasing the number of cohorts is expected to increase integration accuracy
at the expense of more computational effort. We can create more refined schedules
by interleaving points into the existing schedule:


```{r}
interleave <- function(x) {
  n <- length(x)
  xp <- c(x[-n] + diff(x) / 2, NA)
  c(rbind(x, xp))[- 2 * n]
}


times_2x <- interleave(times_default)
times_4x <- interleave(times_2x)
times_8x <- interleave(times_4x)

plot(times_default, xlab = "Cohort #", ylab="Time (years)", 
     xlim = c(1, length(times_8x)), las=1, pch = 19, cex=.2,)
points(times_2x, pch=19, cex=.2, las=1, col = "blue")
points(times_4x, pch=19, cex=.2, las=1, col = "red")
points(times_8x, pch=19, cex=.2, las=1, col = "black")
```
Each schedule runs for the same duration, but `times_high` has 4x as many cohorts.
Running a patch with each cohort schedule allows us to compare the propagule 
outputs:


```{r}
run_with_times <- function(patch, t) {
  patch$cohort_schedule_times[[1]] <- t
  
  start <- Sys.time()
  seed_rain <- run_scm(patch)$seed_rains
  print(paste("Time to solve:", round(Sys.time() - start, 2)))
  
  return(seed_rain)
}

seed_rain_default <- run_with_times(patch, times_default)
seed_rain_2x <- run_with_times(patch, times_2x)
seed_rain_4x <- run_with_times(patch, times_4x)
seed_rain_8x <- run_with_times(patch, times_8x)

cohort_range <- c(length(times_default), length(times_2x), 
                  length(times_4x), length(times_8x))
seed_rain_range <- c(seed_rain_default, seed_rain_2x, 
                     seed_rain_4x, seed_rain_8x)
```
Quadrupling the number of cohorts results in nearly a 6x increase in runtime, 
eight times the cohorts takes roughly 26x longer.

```{r}
plot(x = cohort_range, y = seed_rain_range,
     xlab="Number of cohort introductions", ylab="Seed rain", las=1)
```
Seed rain increases as cohorts are introduced more finely, 
though at a saturating rate, meaning there are diminishing returns in accuracy
for additional cohorts.


The differences in seed rain are not actually that striking
(perhaps 
`r signif(100 * diff(range(seed_rain_range)) / mean(seed_rain_range), 1)`%) but 
the introduced  *variation* creates instabilities of sufficient concern.

# Individual seed rain contributions

Where is the fitness difference driving change in seed rain coming from? 

Consider adding a *single* additional cohort at one of the points
along the first vector of times `times_default` and computing the 
output seed rain:

```{r}
insert_time <- function(i, x) {
  j <- seq_len(i)
  c(x[j], (x[i] + x[i+1])/2, x[-j])
}
run_with_insert <- function(i, patch, t) {
  run_with_times(patch, insert_time(i, t))
}

insert_positions <- seq_len(length(times_default) - 1)
seed_rains <- unlist(mclapply(insert_positions, run_with_insert, 
                              patch, times_default, mc.cores = n_cores))

seed_rain_differences <- seed_rains - seed_rain_default

plot(insert_positions, seed_rain_differences, 
     xlab="Cohort #", ylab="Difference from default seed rain", las=1)
abline(h=0, col="grey")
```

The biggest deviations in output seed rain come about half way
through the schedule, though because of the compression of early times 
this occurs fairly early in patch development:

```{r}
times_interpolated <- (times_default[-1L] + 
                         times_default[-length(times_default)]) / 2

plot(times_interpolated, seed_rain_differences,
     xlab="Time (years)", ylab="Difference from default seed rain", las=1)
abline(h=0, col="grey")
```
Now look at the contribution of different cohorts to the overall patch output
(log scaled for clarity):

```{r}
default <- run_scm(patch)

# remove first cohort to avoid log(0)
plot(times_default[-1], default$seed_rain_cohort(1)[-1], log = "y",
     xlab="Time (years)", ylab="Seed rain (log scale)")

top_5 <- order(abs(seed_rain_differences), decreasing=TRUE)[1:5]
abline(v=times_interpolated[top_5], col=c("red", rep("grey", 4)))
```
<!-- TODO: there might be a clearer way to summarise this point -->

In this case almost all the contribution comes from early cohorts 
(this is essentially a single-age stand of pioneers).  

Overlaid on this are the five cohorts where extra refinement 
(i.e. inserting an additional cohort) causes the largest
change in overall seed rain output (biggest difference in red).  

The difference is not coming from seed rain contributions of those
cohorts, which is basically zero, though it is higher than the
surrounding cohorts.


# Individual competitive impacts

If the inserted cohorts are not altering seed rain through individual 
contributions, they may be altering the output of neighbouring cohorts
through comepttive interactions.

Here we investigate the differences in patch light environment after
introducing the single cohort driving the greatest difference in seed output
identified above.

First we run patches with and without the extra cohort, collecting the complete
set of patch conditions at each timestep.

```{r}
collect_with_times <- function(patch, t) {
  patch$cohort_schedule_times[[1]] <- t
  result <- run_scm_collect(patch)
  return(result)
}

insert_position <- top_5[1]
times_insert <- insert_time(insert_position, times_default)

results_default <- collect_with_times(patch, times_default)
results_insert <- collect_with_times(patch, times_insert)
```
In order to make a direct comparison, we need to interpolate the light
environment over a common set of heights at each time step.

We reconstruct the light environment spline for both runs, and compute
canopy openness in both:

```{r}
interpolated_environment <- function(env, heights) {
  # intialise new spline with existing values
  spline <- plant:::Interpolator()
  spline$init(env[, 1], env[, 2])
  
  # subset to valid heights in patch
  patch_heights <- heights < spline$max
  
  # evaluate over list of heights and zero
  y <- spline$eval(heights)
  y[!patch_heights] = 0
  
  return(y)
} 

n = length(results_default$time)

max_height <- max(results_default$env[[n]][, 1],
                  results_insert$env[[n+1]][, 1])

heights <- seq(0, max_height, length.out=201)
canopy_default <- sapply(results_default$env, interpolated_environment, heights)
canopy_insert <- sapply(results_insert$env, interpolated_environment, heights)

canopy_difference <- t(canopy_insert[, -(insert_position + 1)] - canopy_default)
canopy_difference[abs(canopy_difference) < 1e-10] <- NA
```

Taking the difference between light environments at each height and timestep,
the resulting image plot is blue in regions where the refined light environment,
with the additional cohort is lighter (higher canopy openness)
and red in regions where the the light environment is darker with the additional
cohort.

```{r}
# ColorBrewer's RdBu palette
cols <- c("#B2182B", "#D6604D", "#F4A582", "#FDDBC7",
          "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC")
pal <- colorRampPalette(cols)(20)

image(results_default$time, heights, canopy_difference,
      xlab="Time (years)", ylab="Height (m)", las=1, col=pal)
```
Canopy openness is
`r round(100 * max(abs(canopy_difference), na.rm = T) / mean(canopy_default), 2)`%
lower for ~7 m tall cohorts at year 25. This increased competitive intensity
may be driving differences in seed rain.

# Monitoring integration accuracy

<!-- TODO: find more general terms for 'area_leaf_error' -->

The internal function `plant:::run_scm_error` runs the SCM and
computes errors as the integration proceeds. 

We integrate two different functions over the size-density distribution,
`area_leaf_error` and `seed_rain_error`, and assess how much removing the focal
cohort increases the error in these two integrations. 

`area_leaf_error`, determines how much the removal of the focal cohort 
increases the error in the estimate of the total leaf area in the patch. 

Similarly, `seed_rain_error`, determines how much the removal of the focal cohort 
increases the error in the estimate of the total seed
production from the patch. The relative error in each integration is then
calculated using the `local_error_integration` function.

We've shown that large differences  manifest in canopy openness.
Here we use `plant:::run_scm_error` to assess the integration error of
the leaf area for all cohorts. 

The black line indicates the cohort identified as problematic above.

```{r}
patch_error <- plant:::run_scm_error(patch)
image(times_default, times_default, patch_error$err$lai[[1]],
      xlab="Patch age (years)", ylab="Introduction time (years)", las=1)
abline(h=times_default[top_5[1]])
```

`plant:::run_scm_error` is used by the `build_schedule` function to find
an appropriately refined cohort spacing algorithm that minimises both 
integration errors. 

```{r}
patch_refined <- build_schedule(patch)
seed_rain_refined <- attr(patch_refined, "seed_rain_out")
```

We can see the impact of this refinement by repeating our earlier analysis
and inserting a new cohort at each timestep and extracting the seed rain:

```{r}
times_refined <- patch_refined$cohort_schedule_times[[1]]

insert_positions <- seq_len(length(times_refined) - 1)

seed_rains_refined <- unlist(mclapply(insert_positions, run_with_insert, 
                                      patch_refined, times_refined, 
                                      mc.cores = n_cores))

refined_differences = seed_rains_refined - seed_rain_refined

plot(insert_positions, refined_differences, 
     xlab="Index", ylab="Seed rain differences", las=1)
abline(h=0, col="grey")
```

Some cohorts still introduce variation, however errors in seed rain are reduced
by an order of magnitude, and are even smaller if considering relative error:

```{r}

plot(insert_positions, seed_rains_refined / seed_rain_refined - 1, 
     xlab="Index", ylab="Relative seed rain differences", las=1)
```
The refined cohort introduction schedule results in a seed rain near the 
saturating value identified in [# Seed rain] but with `r length(times_refined)`
cohorts rather than the `r length(times_8x)` cohorts used in naive interpolation.