---
title: "Plant-level properties"
author: 
- Rich FitzJohn
- Daniel Falster
date: 2016
output:
  bookdown::html_document2:
    toc: yes
    number_sections: false
---

This vignette illustrates the sort of analysis used to generate
the plant figure (Fig 2) in the manuscript, i.e. modelling the
dynamics of individual plants. It also shows some of the features of working
with plant components of the model.
```{r }
library(plant)
```

Plants are constructed with the `FF16_Plant` function.  That
function takes as its only argument a "strategy" object; the
default is `FF16_Strategy`, but alternative strategies can be
provided (see below). The "strategy" object contains all
the physiological underpinning the dynamics of individual plants
and entire metapopulations.
```{r }
pl <- FF16_Plant()
```

Plants are an [R6](https://github.com/wch/R6) class, and have a
number of elements and fields that can be accessed:
```{r }
pl
```

Things labelled 'active binding' are "fields" and can be read from
and (sometimes) set:
```{r }
pl$state("height")
pl$set_state("height", 10)
pl$state("height")

pl$state("fecundity")
pl$state("mortality")
```

Height, fecundity and mortality are the three key variables
propagated by the internal system of differential equations:
```{r }
pl$ode_state
```

To compute rates of change for these variables we need a light
environment.  The function `fixed_environment` creates an
environment that has the same canopy openness (here 100%) at all
heights.  The plant *does not affect* this light environment.
```{r }
env <- fixed_environment(1.0)
```

The `compute_vars_phys` method computes net mass production for
the plant, and from that demographic rates:
```{r }
pl$ode_rates
pl$compute_vars_phys(env)
pl$ode_rates
```

<!-- TODO: next bit is a bit vague -->
The names of the vector are available via
```{r}
pl$ode_names
```

There is also a function to directly access particular rates:

```{r }
pl$rate("height")
```

There's not actually that much that can be done with Plant
objects; they're designed to be small and light to work well with
the larger simulation code that does not particularly care about
most of the internal calculations.

<!-- TODO: PlantPlus 
Because of this, we have a "PlantPlus" object that exposes more of
a strategy, and in addition stem diameter growth:
```{r, eval=FALSE}
pp <- FF16_Plant(pl$strategy)
pp$compute_vars_phys(env)
names(pp$internals)
```

Some of the internals require `compute_vars_internals` to be run
(the zapsmall function rounds numbers close to zero to zero):
```{r }
pp$compute_vars_growth()
zapsmall(unlist(pp$internals))
```


This PlantPlus object also includes heartwood area and mass as two
more variables for the ODE system (this might move into Plant
soon -- see [this issue](https://github.com/traitecoevo/plant/issues/139)).
Tracking of these variables is needed to estimate stem diameter growth
```{r }
pp$ode_names
```
 -->

Plants are a type of *reference object*.  They are different to
almost every other R object you regularly interact with in that
they do not make copies when you rename them.  So changes to one
will be reflected in another.
```{r }
pl2 <- pl
pl2$set_state("height", 1)
pl2$state("height")
pl$state("height") # also 1!
```

# Growing plants

Rather than setting plant physical sizes to given values, it will
often be required to *grow* them to a size.  This is required to
compute seed output (integrated over the plant's lifetime)
stem diameter, survival, etc; basically everything except for
height.

It's possible to directly integrate the equations exposed by the
plant, using the `ode_state` field, `compute_vars_phys` method and
`ode_rates` field.  For example, we can use the R package `deSolve`:
```{r }
derivs <- function(t, y, plant, env) {
  plant$ode_state <- y
  plant$compute_vars_phys(env)
  list(plant$ode_rates)
}
pl <- FF16_Plant()
tt <- seq(0, 50, length.out=101)
y0 <- setNames(pl$ode_state, pl$ode_names)
yy <- deSolve::lsoda(y0, tt, derivs, pl, env=env)
plot(height ~ time, yy, type="l")
```

Alternatively, it might desirable to grow a plant to a particular
size.  We could interpolate from the previous results easily
enough.  E.g., to find a plant with height of 15 m:
```{r }
h <- 15.0
```

that happened approximately here:
```{r }
t <- spline(yy[, "height"], yy[, "time"], xout=h)$y
```

Interpolate to find the state:
```{r }
y <- apply(yy[, -1], 2, function(y) spline(yy[, "time"], y, xout=t)$y)

pl2 <- FF16_Plant()
pl2$ode_state <- y
pl2$compute_vars_phys(env)
```
Plant is expected height:
```{r }
pl2$state("height")
```
And at this height, here is the total seed production:
```{r }
pl2$state("fecundity")
```

To make this type of operation easier, we provide the function
`grow_plant_to_time`

```{r }
res <- grow_plant_to_time(FF16_Plant(FF16_Strategy()), tt, env)
```

Here is the result, plotted against the result obtained
from using `deSolve` above:
```{r }
plot(height ~ tt, res$state, type="l", las=1,
     xlab="Time (years)", ylab="Height (m)")
points(height ~ time, yy, col="grey", cex=.5)
```

Completing the set, `plant` also provides a function for growing
plants to a particular size; `grow_plant_to_size`.  This takes
*any* size measurement in the plant and can grow the plant to that
size.  So, for height:

```{r }
pl <- FF16_Plant(FF16_Strategy())
heights <- seq(pl$state("height"), pl$strategy$hmat, length.out=20)
res <- grow_plant_to_size(pl, heights, "height", env)
```

This returns a vector of times; this is when the heights were
achieved
```{r }
res$time
```

A matrix of state:
```{r }
head(res$state)
```

And a list of plants
```{r }
res$plant[[10]]
res$plant[[10]]$state("height")
heights[[10]]
```

Also included is `trajectory`; the points that the ODE stepper used
with the system state at those times.

<!-- 
TODO: Revise with new Plant Plus
There is a convenience function `run_plant_to_heights` that
achieves the same thing.  Alternatively, and variable within
`plant$internals` can be used, so long as it increases
monotonically with respect to time.
```{r }
pl <- FF16_Plant(FF16_Strategy())
mass <- seq_log(pl$internals$mass_above_ground + 1e-8, 1000, length.out=21)
res_mass <- grow_plant_to_size(pl, mass, "mass_above_ground", env,
                               time_max=100, warn=FALSE)
```
(this seems on the low side - see [this
issue](https://github.com/traitecoevo/plant/issues/142)).
```{r }
plot(res_mass$time, mass, type="o", pch=19, las=1, xlab="Time (years)")
```
 -->

With all these bits in place, let's look at growth trajectories of
two species that differ in their LMA values. This what is presented
In Fig. 2a of the paper.

```{r }
p <- scm_base_parameters("FF16")
```
Low LMA ("fast growth") species
```{r }
s1 <- strategy(trait_matrix(0.0825,  "lma"), p)
```
High LMA ("low growth") species
```{r }
s2 <- strategy(trait_matrix(0.2625, "lma"), p)
```

Note that we're using an alternative way of specifying strategies
here, to trigger our "hyper-parametrisation" approach.  This may be
simplified in future, but currently the "hyper-parametrisation" function
resides on the `p` object.

Then, generate a sequence of heights to collect information at
```{r }
pl1 <- FF16_Plant(s1)
pl2 <- FF16_Plant(s2)
```

(they are different for the two plants because they have different
starting heights, the lower LMA of s1 allows it to achieve a greater
initial height for given seed mass)
```{r }
heights1 <- seq(pl1$state("height"), s1$hmat, length.out=100L)
heights2 <- seq(pl2$state("height"), s2$hmat, length.out=100L)

dat1 <- grow_plant_to_height(pl1, heights1, env,
                             time_max=100, warn=FALSE, filter=TRUE)
dat2 <- grow_plant_to_height(pl2, heights2, env,
                             time_max=100, warn=FALSE, filter=TRUE)

plot(dat1$trajectory[, "time"], dat1$trajectory[, "height"],
     type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height (m)")
lines(dat2$trajectory[, "time"], dat2$trajectory[, "height"], lty=2)
legend("bottomright", c("Low LMA", "High LMA"), lty=1:2, bty="n")
```

Similarly, growing the plants under lower light:
```{r }
env_low <- fixed_environment(0.5)
dat1_low <- grow_plant_to_height(pl1, heights1, env_low,
                                 time_max=100, warn=FALSE, filter=TRUE)
dat2_low <- grow_plant_to_height(pl2, heights2, env_low,
                                 time_max=100, warn=FALSE, filter=TRUE)

cols <- c("black", "#e34a33")
plot(dat1$trajectory[, "time"], dat1$trajectory[, "height"],
     type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height (m)")
lines(dat2$trajectory[, "time"], dat2$trajectory[, "height"], lty=2)
lines(dat1_low$trajectory[, "time"], dat1_low$trajectory[, "height"],
      lty=1, col=cols[[2]])
lines(dat2_low$trajectory[, "time"], dat2_low$trajectory[, "height"],
      lty=2, col=cols[[2]])
legend("bottomright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

The *height growth rate* is the derivative of height with respect
to time - the slope of the plot above.  It is really the core
quantity in the model; the actual heights are computed by solving
the set of ODEs that includes height growth rate.

Growth rate with respect to height shows a hump-shaped pattern that
is affected by both traits and by light environment.  To extract
this information from the trajectories takes a little more work
though.

Here is a plant from part way through one run
```{r }
pl <- dat1$plant[[50]]
```
Here is the set of ODE state variables:
```{r }
setNames(pl$ode_state, pl$ode_names)
```
And the set of *rate* variables
```{r }
setNames(pl$ode_rates, pl$ode_names)
```
...however, the rates might not be correct.  They are whatever was
left by the ODE stepper when it was advancing the plant, so it's
best to update them:
```{r }
pl$compute_vars_phys(dat1$env)
setNames(pl$ode_rates, pl$ode_names)
```
(in this case they are the same because the light environment is
unchanging, but that not be the case generally)

Alternatively, we can access the height growth rate via the
internals, which is the same as accessing directly from the ODE
rates but more explicit:
```{r }
pl$rate("height")
pl$ode_rates[[1]]
```

Collecting height growth for all plants:
```{r }
f <- function(x) x$rate("height")
dhdt1 <- sapply(dat1$plant, f)
dhdt2 <- sapply(dat2$plant, f)
dhdt1_low <- sapply(dat1_low$plant, f)
dhdt2_low <- sapply(dat2_low$plant, f)

plot(dat1$time, dhdt1, type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height growth rate (m / yr)")
lines(dat2$time, dhdt2, lty=2)
lines(dat1_low$time, dhdt1_low, lty=1, col=cols[[2]])
lines(dat2_low$time, dhdt2_low, lty=2, col=cols[[2]])
legend("topright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

Alternatively, change in height plotted against height itself:
```{r }
ylim <- c(0, max(dhdt1))
plot(dat1$state[, "height"], dhdt1, type="l", lty=1,
     las=1, xlab="Height (m)", ylab="Height growth rate (m / yr)", ylim=ylim)
lines(dat2$state[, "height"], dhdt2, lty=2)
lines(dat1_low$state[, "height"], dhdt1_low, lty=1, col=cols[[2]])
lines(dat2_low$state[, "height"], dhdt2_low, lty=2, col=cols[[2]])
legend("topright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

<!-- 
TODO: Revise with new Plant Plus
Over a plant's life, allocation to different structures varies.
This is captured by a set of variables stored within the internals:
e.g., `mass_leaf` `mass_sapwood`.
```{r }
pl$internals$mass_leaf
pl$internals$mass_sapwood
```

(these numbers seem a bit off: one of the motivations here is to
develop and use better models of plant allometry.  The
parameterisation used at present are derived from adults and
perform poorly with small plants.  However, based on height / area
relationships [Falster 2011, supporting information], for an 8m
tall plant total leaf areas of 5-10 m are plausible and with an LMA
of 0.08 that implies a total *dry* weight of 400 - 800 g).

Total live dry mass fraction to leaf and stem can be computed as:
```{r }
f <- function(p) {
  p_ints <- p$internals
  c(leaf=p_ints$mass_leaf / p_ints$mass_live,
    sapwood=p_ints$mass_sapwood / p_ints$mass_live)
}

cols_part <- c("black", "#045a8d")
matplot(dat1$state[, "height"], t(sapply(dat1$plant, f)),
        type="l", col=cols_part, lty=1, ylim=c(0, 1),
        xlab="Height (m)", ylab="Fractional allocation", las=1)
matlines(dat2$state[, "height"], t(sapply(dat2$plant, f)),
         col=cols_part, lty=2)
legend("topright", c("Sapwood", "Leaf"), lty=1, col=rev(cols_part), bty="n")
```

Relative allocation to leaf mass drops steeply as a plant grows and
is replaced by allocation to sapwood mass.
-->

The growth rates vary with both size and light environment (see
above).

```{r }
pl <- FF16_Plant()
pl$set_state("height", 10)
pl$compute_vars_phys(fixed_environment(1.0))
pl$rate("height") # in full light

pl$compute_vars_phys(fixed_environment(0.5))
pl$rate("height") # in 1/2 light
```

At some point the plant cannot maintain positive carbon balance and
therefore cannot grow; for example at 25% canopy openness:
```{r }
pl$compute_vars_phys(fixed_environment(0.25))
pl$rate("height")
```

<!-- # TODO: PlantPlus revise once LCP available

The light level at which carbon gain becomes zero is the "whole
plant light compensation point".
```{r }
lcp_whole_plant(pl)
```

Consider a vector of canopy opennesses:
```{r }
openness <- seq(0, 1, length.out=51)

lcp <- lcp_whole_plant(pl)
```
 -->

<!-- TODO: Remove below once lap works -->

```{r}
openness <- seq(0, 1, length.out=51)
lcp <- 0.299
```

Height growth rate increases in a saturating fashion with
increasing canopy openness from the light compensation point.
```{r }
f <- function(x, pl) {
  env <- fixed_environment(x)
  pl$compute_vars_phys(env)
  pl$rate("height")
}
x <- c(lcp, openness[openness > lcp])
plot(x, sapply(x, f, pl), type="l", xlim=c(0, 1),
     las=1, xlab="Canopy openness", ylab="Height growth rate (m / yr)")
points(lcp, 0.0, pch=19)

g <- function(x, pl) {
  lcp <- lcp_whole_plant(pl)
  x <- c(lcp, openness[openness > lcp])
  cbind(x=x, y=sapply(x, f, pl))
}
```
<!-- # TODO: PlantPlus revise once LCP available

Now, consider this for a seedling and for a plant at half its
maximum size
size:
```{r }
pl_seed <- FF16_Plant(s1)
y_seed <- g(openness, pl_seed)
pl_adult <- FF16_Plant(s1)
pl_adult$state("height", pl_adult$strategy$hmat / 2)
y_adult <- g(openness, pl_adult)
cols_height <- c("#31a354", "black")

ymax <- max(y_seed[, 2], y_adult[, 2])
plot(y_seed, type="l", col=cols_height[[1]],
     xlim=c(0, 1), ylim=c(0, ymax), las=1,
     xlab="Canopy openness", ylab="Height growth year (m / yr)")
lines(y_adult, col=cols_height[[2]])
legend("bottomright", c("Seedling", "Adult"), lty=1, col=cols_height, bty="n")
```

The light compensation point and curve varies with traits, too:
```{r }
pl2_seed <- FF16_PlantPlus(s2)
pl2_adult <- FF16_PlantPlus(s2)
pl2_adult$height <- pl2_adult$strategy$hmat / 2
y2_seed <- g(openness, pl2_seed)
y2_adult <- g(openness, pl2_adult)

ymax <- max(ymax, y2_seed[, 2], y2_adult[, 2])
plot(y_seed, type="l", col=cols_height[[1]],
     xlim=c(0, 1), ylim=c(0, ymax), las=1,
     xlab="Canopy openness", ylab="Height growth year (m / yr)")
lines(y_adult, col=cols_height[[2]])
lines(y2_seed, col=cols_height[[1]], lty=2)
lines(y2_adult, col=cols_height[[2]], lty=2)
legend("bottomright", c("Seedling", "Adult"), lty=1, col=cols_height, bty="n")
```

Note that the traits with the lowest growth rate in most lights
while a seedling (dotted lines) has the highest growth rate in all
lights as an adult.
 -->
