---
title: "Individual-level properties"
author:
- Rich FitzJohn
- Daniel Falster
date: 2016
output:
  bookdown::html_document2:
    toc: yes
    number_sections: false
---

# Background

Any `plant` model defines a strategy that implements a system of equations
describing the change of one or more characteristics (state variables) of
individuals. These individuals are aggregated as cohorts of different ages
that compete within a patch.

This vignette illustrates the how to model the dynamics of individuals and
the analysis used to generate  figure (Fig 2) in the manuscript.
Throughout we use the `FF16` strategy, describing the Falster Fitzjohn model
of light competition in forest stands, however an equivalent `Individual` component
of the model will be present in all strategies.

```{r }
library(plant)
```

# Individuals

Individuals are constructed with the `FF16_Individual` function. That
function takes as its only argument a "strategy" object; the
default is `FF16_Strategy`, but alternative strategies can be
provided (see below). The "strategy" object contains all
the physiological underpinning the dynamics of individual plants
and entire meta-populations.
```{r }
indv <- FF16_Individual()
```

Individuals are an [R6](https://github.com/wch/R6) class, and have a
number of elements and fields that can be accessed:
```{r }
indv
```

Things labelled 'active binding' are "fields" and can be read from
and (sometimes) set:
```{r }
indv$state("height")
indv$set_state("height", 10)
indv$state("height")

indv$state("fecundity")
indv$state("mortality")
```

Height, fecundity and mortality are the three key variables of the FF16
strategy, each propagated by the internal system of differential equations:
```{r }
indv$ode_state
```

While we can get and set variable state for a single individual,
computing the rates of change requires information about the environment an
individual inhabits. In the FF16 model, individuals compete through a shared
light environment, described in terms of canopy openness.
The function `fixed_environment` creates an
environment that has the same canopy openness (here 100%) at all
heights. *Note: The `indv` individual does not affect this light environment*.
```{r }
env <- FF16_fixed_environment(1.0)
```

In the FF16 model, the `compute_rates` method computes net mass production for
the individual, and from that demographic rates:
```{r }
indv$ode_rates
indv$compute_rates(env)
indv$ode_rates
```

<!-- TODO: next bit is a bit vague -->
While we have focused on three key variables, height, mortality and fecundity,
the FF16 model describes variable for five characteristics.
The name vector of these variables is available via

```{r}
indv$ode_names
```

There is also a function to directly access particular rates:

```{r }
indv$rate("height")
```

Individuals are a type of *reference object*. They are different to
almost every other R object you regularly interact with in that
they do not make copies when you rename them, meaning changes to one
will be reflected in another.
```{r }
indv2 <- indv
indv2$set_state("height", 1)
indv2$state("height")
indv$state("height") # also 1!
```
There's not actually that much that can be done with Individual
objects; they're designed to be small and light to work well with
the larger simulation code that does not particularly care about
most of the internal calculations.

# Growing plants

Rather than setting the size of Individuals to given values,
it will often be required to *grow* them to a size. In the FF16 model,
this allows us to compute seed output (integrated over the plant's lifetime)
stem diameter, survival, etc; basically everything except for
height.

It's possible to directly integrate the equations exposed by the
Individual, using the `ode_state` field, `compute_rates` method and
`ode_rates` field.  For example, we can use the R package `deSolve`:
```{r }
# deSolve wrapper of Individual ODEs
derivs <- function(t, y, individual, env) {
  individual$ode_state <- y
  individual$compute_rates(env)
  list(individual$ode_rates)
}

# initialise indiidual, timesteps and initial conditions
indv <- FF16_Individual()
tt <- seq(0, 50, length.out=101)
y0 <- setNames(indv$ode_state, indv$ode_names)

# solve and plot
yy <- deSolve::lsoda(y0, tt, derivs, indv, env=env)
plot(height ~ time, yy, type="l")
```

Alternatively, it might desirable to grow a Individual to a particular
size.  We could interpolate from the previous results easily
enough.  E.g., to find a plant with height of 15 m:
```{r }
h <- 15.0
```

that happened approximately here:
```{r }
t <- spline(yy[, "height"], yy[, "time"], xout=h)$y
```

Interpolate to find the state:
```{r }
y <- apply(yy[, -1], 2, function(y) spline(yy[, "time"], y, xout=t)$y)

indv2 <- FF16_Individual()
indv2$ode_state <- y
indv2$compute_rates(env)
```
Plant is expected height:
```{r }
indv2$state("height")
```
And at this height, here is the total seed production:
```{r }
indv2$state("fecundity")
```

To make this type of operation easier, we provide the function
`grow_plant_to_time`

```{r }
res <- grow_plant_to_time(FF16_Individual(), tt, env)
```

Here is the result using the internal ODE solver (black line),
plotted against the result  obtained from using `deSolve` above (grey dots):
```{r }
plot(height ~ tt, res$state, type="l", las=1,
     xlab="Time (years)", ylab="Height (m)")

points(height ~ time, yy, col="grey", cex=.5)
```

Completing the set, `plant` also provides a function for growing
individuals to a particular size; `grow_plant_to_size`.  This takes
*any* size measurement in the individuals and can grow the individuals to that
size.  So, for height:

```{r }
# create a sequence from the initial size to height at maturity
indv <- FF16_Individual()
heights <- seq(indv$state("height"), indv$strategy$hmat, length.out=20)
res <- grow_plant_to_size(indv, heights, "height", env)
```

This returns a vector of times; this is when the heights were
achieved
```{r }
res$time
```

A matrix of states, showing the state of related characteristics:
```{r }
head(res$state)
```

And a list of individualss
```{r }
res$plant[[10]]
res$plant[[10]]$state("height")
heights[[10]]
```

# Comparing individuals

With all relevant components  in place, let's look at growth trajectories of
individuals of two species that differ in their trait values.
This presented in Fig. 2a of the paper.

Note that we're using an alternative way of specifying strategies
here, to trigger our "hyper-parameterisation" approach. This allows us to relate
the values of one trait, here leaf-mass area (LMA), to other traits that
form fundamental trade-offs in plant physiology. This may be
simplified in future, but currently the "hyper-parameterisation" function
resides on the `params` object.

First, we load the FF16 strategy pre-populated with a set of default parameters

```{r }
params <- scm_base_parameters("FF16")
```

Then we can use the `params` object to set a low LMA ("fast growth") species:
```{r }
s1 <- strategy(trait_matrix(0.0825,  "lma"), params)
```

and a high LMA ("low growth") species:
```{r }
s2 <- strategy(trait_matrix(0.2625, "lma"), params)
```

In the FF16 model, LMA is related to leaf turnover (`k_l`) and
the ratio of dark respiration to leaf mass (`r_l`):
```{r }
lapply(list(s1, s2), `[`, c("lma", "k_l", "r_l"))

```

We use these specific strategies to create Individuals:
```{r }
indv1 <- FF16_Individual(s1)
indv2 <- FF16_Individual(s2)
```


Then, collect information of their characteristic variables
over a sequence of heights. Note that range of heights are different
for the two Individuals because they have different
starting heights. The lower LMA of s1 allows it to achieve a greater
initial height for given seed mass.

```{r }
heights1 <- seq(indv1$state("height"), s1$hmat, length.out=100L)
heights2 <- seq(indv2$state("height"), s2$hmat, length.out=100L)

dat1 <- grow_plant_to_height(indv1, heights1, env,
                             time_max=100, warn=FALSE, filter=TRUE)
dat2 <- grow_plant_to_height(indv2, heights2, env,
                             time_max=100, warn=FALSE, filter=TRUE)

plot(dat1$trajectory[, "time"], dat1$trajectory[, "height"],
     type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height (m)")
lines(dat2$trajectory[, "time"], dat2$trajectory[, "height"], lty=2)
legend("bottomright", c("Low LMA", "High LMA"), lty=1:2, bty="n")
```

We cam see each species fares under lower light:

```{r }
env_low <- FF16_fixed_environment(0.5)
dat1_low <- grow_plant_to_height(indv1, heights1, env_low,
                                 time_max=100, warn=FALSE, filter=TRUE)
dat2_low <- grow_plant_to_height(indv2, heights2, env_low,
                                 time_max=100, warn=FALSE, filter=TRUE)

cols <- c("black", "#e34a33")
plot(dat1$trajectory[, "time"], dat1$trajectory[, "height"],
     type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height (m)")
lines(dat2$trajectory[, "time"], dat2$trajectory[, "height"], lty=2)
lines(dat1_low$trajectory[, "time"], dat1_low$trajectory[, "height"],
      lty=1, col=cols[[2]])
lines(dat2_low$trajectory[, "time"], dat2_low$trajectory[, "height"],
      lty=2, col=cols[[2]])
legend("bottomright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

The *height growth rate* is the derivative of height with respect
to time - the slope of the plot above.  It is really the core
quantity in the model; the actual heights are computed by solving
the set of ODEs that includes height growth rate.

Growth rate with respect to height shows a hump-shaped pattern that
is affected by both traits and by light environment.  To extract
this information from the trajectories takes a little more work
though.

Here is an inidivdual from part way through one run
```{r }
indv <- dat1$plant[[50]]
```

Here is the set of ODE state variables:
```{r }
setNames(indv$ode_state, indv$ode_names)
```
And the set of *rate* variables
```{r }
setNames(indv$ode_rates, indv$ode_names)
```
...however, the rates might not be correct.  They are whatever was
left by the ODE stepper when it was advancing the plant, so it's
best to update them:
```{r }
indv$compute_rates(dat1$env)
setNames(indv$ode_rates, indv$ode_names)
```
(in this case they are the same because the light environment is
unchanging, but that not be the case generally)

Alternatively, we can access the height growth rate via the
internals, which is the same as accessing directly from the ODE
rates but more explicit:
```{r }
indv$rate("height")
indv$ode_rates[[1]]
```

Collecting height growth for high and low LMA individauls
in high and low light environments:
```{r }
f <- function(x) x$rate("height")
dhdt1 <- sapply(dat1$plant, f)
dhdt2 <- sapply(dat2$plant, f)
dhdt1_low <- sapply(dat1_low$plant, f)
dhdt2_low <- sapply(dat2_low$plant, f)

plot(dat1$time, dhdt1, type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height growth rate (m / yr)")
lines(dat2$time, dhdt2, lty=2)
lines(dat1_low$time, dhdt1_low, lty=1, col=cols[[2]])
lines(dat2_low$time, dhdt2_low, lty=2, col=cols[[2]])
legend("topright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

Alternatively, change in height plotted against height itself:
```{r }
ylim <- c(0, max(dhdt1))
plot(dat1$state[, "height"], dhdt1, type="l", lty=1,
     las=1, xlab="Height (m)", ylab="Height growth rate (m / yr)", ylim=ylim)
lines(dat2$state[, "height"], dhdt2, lty=2)
lines(dat1_low$state[, "height"], dhdt1_low, lty=1, col=cols[[2]])
lines(dat2_low$state[, "height"], dhdt2_low, lty=2, col=cols[[2]])
legend("topright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

<!--
TODO: Remove Plant Plus
Over a plant's life, allocation to different structures varies.
This is captured by a set of variables stored within the internals:
e.g., `mass_leaf` `mass_sapwood`.
```{}
indv$internals$mass_leaf
indv$internals$mass_sapwood
```

(these numbers seem a bit off: one of the motivations here is to
develop and use better models of plant allometry.  The
parameterisation used at present are derived from adults and
perform poorly with small plants.  However, based on height / area
relationships [Falster 2011, supporting information], for an 8m
tall plant total leaf areas of 5-10 m are plausible and with an LMA
of 0.08 that implies a total *dry* weight of 400 - 800 g).

Total live dry mass fraction to leaf and stem can be computed as:
```{}
f <- function(p) {
  p_ints <- p$internals
  c(leaf=p_ints$mass_leaf / p_ints$mass_live,
    sapwood=p_ints$mass_sapwood / p_ints$mass_live)
}

cols_part <- c("black", "#045a8d")
matplot(dat1$state[, "height"], t(sapply(dat1$plant, f)),
        type="l", col=cols_part, lty=1, ylim=c(0, 1),
        xlab="Height (m)", ylab="Fractional allocation", las=1)
matlines(dat2$state[, "height"], t(sapply(dat2$plant, f)),
         col=cols_part, lty=2)
legend("topright", c("Sapwood", "Leaf"), lty=1, col=rev(cols_part), bty="n")
```

Relative allocation to leaf mass drops steeply as a plant grows and
is replaced by allocation to sapwood mass.
-->

The growth rates vary with both size and light environment (see
above).

```{r }
indv <- FF16_Individual()
indv$set_state("height", 10)
indv$compute_rates(FF16_fixed_environment(1.0))
indv$rate("height") # in full light

indv$compute_rates(FF16_fixed_environment(0.5))
indv$rate("height") # in 1/2 light
```

At some point the individual cannot maintain positive carbon balance and
therefore cannot grow; for example at 25% canopy openness
```{r }
indv$compute_rates(FF16_fixed_environment(0.25))
indv$rate("height")
```

The light level at which carbon gain becomes zero is the "whole
plant light compensation point".

<!-- TODO: Remove below once lcp works -->
```{r}
openness <- seq(0, 1, length.out=51)
lcp <- 0.299
```

Height growth rate increases in a saturating fashion with
increasing canopy openness from the light compensation point.
```{r }
f <- function(x, indv) {
  env <- FF16_fixed_environment(x)
  indv$compute_rates(env)
  indv$rate("height")
}
x <- c(lcp, openness[openness > lcp])
plot(x, sapply(x, f, indv), type="l", xlim=c(0, 1),
     las=1, xlab="Canopy openness", ylab="Height growth rate (m / yr)")
points(lcp, 0.0, pch=19)

```

<!-- TODO: refactor once lcp works

Now, consider this for a seedling and for a plant at half its
maximum size
size:
```{ }
indv_seed <- FF16_Individual(s1)
y_seed <- g(openness, indv_seed)
indv_adult <- FF16_Individual(s1)
indv_adult$state("height", indv_adult$strategy$hmat / 2)
y_adult <- g(openness, indv_adult)
cols_height <- c("#31a354", "black")

ymax <- max(y_seed[, 2], y_adult[, 2])
plot(y_seed, type="l", col=cols_height[[1]],
     xlim=c(0, 1), ylim=c(0, ymax), las=1,
     xlab="Canopy openness", ylab="Height growth year (m / yr)")
lines(y_adult, col=cols_height[[2]])
legend("bottomright", c("Seedling", "Adult"), lty=1, col=cols_height, bty="n")
```

The light compensation point and curve varies with traits, too:
```{ }
pl2_seed <- FF16_IndividualPlus(s2)
pl2_adult <- FF16_IndividualPlus(s2)
pl2_adult$height <- pl2_adult$strategy$hmat / 2
y2_seed <- g(openness, pl2_seed)
y2_adult <- g(openness, pl2_adult)

ymax <- max(ymax, y2_seed[, 2], y2_adult[, 2])
plot(y_seed, type="l", col=cols_height[[1]],
     xlim=c(0, 1), ylim=c(0, ymax), las=1,
     xlab="Canopy openness", ylab="Height growth year (m / yr)")
lines(y_adult, col=cols_height[[2]])
lines(y2_seed, col=cols_height[[1]], lty=2)
lines(y2_adult, col=cols_height[[2]], lty=2)
legend("bottomright", c("Seedling", "Adult"), lty=1, col=cols_height, bty="n")
```

Note that the traits with the lowest growth rate in most lights
while a seedling (dotted lines) has the highest growth rate in all
lights as an adult.

-->
