---
title: "Modifying parameters"
author: 
- Rich FitzJohn
- Daniel Falster
date: 2016
output:
  bookdown::html_document2:
    toc: yes
    number_sections: false
---

```{r}
library(plant)
```

There are a large number of parameters to the physiological model,
but all are changeable.  The default strategy is detailed in the
"physiology" vignette:

```{r}
s <- FF16_Strategy()
names(s)
```

The strategy object here is a special object of class `r class(s)`.
In contrast with most of the "reference" objects used by `plant`,
this is simply a list with a class attribute.  Some validation will
be done on the parameters every time it is passed into the C++ bits
of the model.

All parameters except for `control` are floating point (decimal)
numbers:

```{r}
unlist(s[names(s) != "control"])
```

All of these parameters can be directly changed.  For example, we
can double the leaf mass per unit leaf area (lma) value:

```{r}
s$lma <- s$lma * 2
```

and then from this construct a plant:

```{r}
pl <- FF16_Individual(s)
pl$strategy$lma
```

`lma` affects a few places in the model; see [the source
code](https://github.com/traitecoevo/plant/blob/master/src/ff16_strategy.cpp),
but only really for converting from leaf area to leaf mass (leaf
mass being leaf area multiplied by leaf mass per unit leaf area).
However, as a component of the leaf economic spectrum we imagine
LMA as affecting a number of other components of the model.

We capture this through what we call a "hyper-parameterisation";
additional parameters and functions that mean that changing one
parameter might affect a number of other lower-level parameters.
Our default hyper-parameterisation is via the
`FF16_hyperpar` function:

```{r}
FF16_hyperpar
```

You will rarely need to call this function directly (see below) but
note how setting of lma affects parameters `k_l`, `a_p1` and `r_l`

```{r}
FF16_hyperpar(trait_matrix(0.1, "lma"), s)
```

These are:
  * `k_l`: Turnover rate for leaves
  * `a_p1`: Leaf photosynthesis per area
  * `r_l`: Leaf respiration per mass

This means that it is possible to implement different trade-offs
between parameters relatively easily by modifying the
hyper-parameterisation function in R, rather than having to modify
the underlying physiological model in C++.

The `trait_matrix` function is a simple wrapper that just makes
sure the trait matrix has the right format:

```{r}
trait_matrix(0.1, "lma")
trait_matrix(c(0.1, 500), "rho")
```

To make use of the hyper-parameterisation, the preferred way of
setting parameters is through the utility functions `strategy` and
`strategy_list`.  These take a `Parameters` object:

```{r}
p <- FF16_Parameters()
```

The Parameters object mostly contains information about the patch:

```{r}
names(p)
```

and it comes pre-set with the hyper-parameterisation function:

```{r}
identical(p$hyperpar, FF16_hyperpar)
```

`k_I` is the light extinction coefficient,
`disturbance_mean_interval` is the mean disturbance interval.  The
`strategy_default` member is a Strategy:

```{r}
class(p$strategy_default)
```

This is the Strategy object that all others will be built from by
difference.  Running

```{r}
s <- strategy(trait_matrix(0.1, "lma"), p)
```

will create a strategy `s` where lma is set but also all the
parameters that *depend* on lma.

The function `strategy_list` can be used to create a list of
strategies:

```{r}
lma <- trait_matrix(seq(0.1, 0.5, length.out=5), "lma")
FF16_hyperpar(trait_matrix(lma, "lma"), s)

ss <- strategy_list(lma, p)
length(ss)
```

We can then use standard R commands to extract variable from this list

```{r}
sapply(ss, function(x) x$lma)
sapply(ss, function(x) x$k_l)
sapply(ss, function(x) x$a_p1)
sapply(ss, function(x) x$r_l)
```

There's a convenience function `plant_list` that returns a set of
*plants* based on a vector of traits:

```{r}
pp <- plant_list(lma, p)

sapply(pp, function(p) p$area_leaf_above(0))
```

In addition to the physiological parameters there are large number
of "control" parameters that affect the behaviour of the various
numerical algorithms used (note that in contrast to the
physiological parameters these have a variety of types)

```{r}
p$control
```

The defaults are rather too slow for many uses, so
`scm_base_parameters` provides a faster set by using `fast_control`
to set many of these to less accurate values.

```{r}
p2 <- scm_base_parameters("FF16")
p2$control[unlist(p$control) != unlist(p2$control)]
```
