---
title: "Patch-level dynamics"
author: 
- Rich FitzJohn
- Daniel Falster
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: yes
    number_sections: false
---

The aim here is to use plant to investigate dynamics of plants
within a single patch.

```{r}
library(plant)

p0 <- scm_base_parameters("FF16")
p0$control$equilibrium_nsteps <- 30
p0$control$equilibrium_solver_name <- "hybrid"
p0$disturbance_mean_interval <- 30.0
```

First, with a single species:

```{r}
p1 <- expand_parameters(trait_matrix(0.0825, "lma"), p0, FALSE)
```

Run everything out to equilibrium:

```{r}
p1_eq <- equilibrium_seed_rain(p1)
```

This collects information about the state of the system at every
ODE step:

```{r}
data1 <- run_scm_collect(p1_eq)
```

Entries are:
  * time: time of the step
  * species: a list with the state of each species (in this case
    there is only 1).  The contents is a 3d array of variable /
    time step / cohort
  * light_env: a list with the points of the light environment
    spline (height / canopy openness)
  * seed_rain: output seed rain
  * p: input parameters

```{r}
names(data1)

t <- data1$time
h <- data1$species[[1]]["height", , ]
```

h here is two of the three dimensions of the state array; each row
is a *time* and each column is a *cohort*.

```{r}
dim(h)
```

So the ith row is the height of all extant cohorts in the patch
(some are NA if they have not been introduced yet)

```{r}
h[150, ]
```

And the jth column is the heights of a particular cohort, NA if
that time is before it was introduced:

```{r}
h[, 60]
```

With a single species there is always one more time step than
cohort introduction, with the last two steps having the same number
of cohorts.  With multiple species there can be more time steps
than cohort introductions, as we'll record data every cohort
introduction for *either* species, which will be refined to
different schedules.

Individual plants increase in height with respect to time, but
because they are competing the growth rate depends on the amount of
shading above them, in addition to size dependent growth rate (see
vignette:plant)

```{r}
matplot(t, h, lty=1, col=make_transparent("black", 0.25), type="l",
        las=1, xlab="Time (years)", ylab="Height (m)")
```

The light environment is stored over each time step:

```{r}
xlim <- c(0, 1.1)
ylim <- range(data1$light_env[[length(data1$light_env)]][, "height"])
plot(NA, xlim=xlim, ylim=ylim, las=1,
     xlab="Canopy openness", ylab="Height (m)")
for (i in data1$light_env) {
  lines(i[, "canopy_openness"], i[, "height"], col="grey")
}

blues <- c("#DEEBF7", "#C6DBEF", "#9ECAE1", "#6BAED6",
           "#4292C6", "#2171B5", "#08519C", "#08306B")
times <- c(5, 10, 20, 40, data1$time[[length(data1$time)]])
cols <- colorRampPalette(blues[-(1:2)])(length(times))
for (i in seq_along(times)) {
  x <- data1$light_env[[which.min(abs(times[[i]] - data1$time))]]
  lines(x[, "canopy_openness"], x[, "height"], col=cols[[i]])
  y <- x[nrow(x), "height"]
  points(1, y, pch=19, col=cols[[i]])
  text(1 + strwidth("x"), y, paste(round(times[[i]]), "years"),
       adj=c(0, 0))
}
```

The amount of light at the ground level is perhaps the most
relevant metric:

```{r}
y <- sapply(data1$light_env, function(x) x[1, "canopy_openness"])
plot(data1$time, y, type="l", las=1,
     ylim=c(0, 1), xlab="Time (years)", ylab="Canopy openness")
```

The waves here are due to rounds of recruitment and self thinning.
Mortality is not instantaneous, so species recruit to a density
that generates a canopy that they cannot survive under.

Leaf area index is the driver that controls the canopy openness
(via the light extinction coefficient `p1$k_I`, following
exponential extinction).  This is not returned by
`run_scm_collect` so instead we need to rebuild patches using
`scm_patch`.

```{r}
patches1 <- lapply(seq_along(data1$time), scm_patch, data1)
```

Each element of the resulting list is a Patch object, the same as
was observed when running the model.

```{r}
lai1 <- sapply(patches1, function(x) x$area_leaf_above(0.0))
plot(data1$time, lai1, type="l", las=1, xlab="Time (years)",
     ylab="Leaf area index")
```

If multiple species are grown at once, they compete with one
another.  This adds a second species -- with a higher lma value
than the first species -- to the population and runs them until
they reach equilibrium seed rain at the metapopulation level.

```{r}
p2 <- expand_parameters(trait_matrix(0.2625, "lma"), p1, FALSE)
p2_eq <- equilibrium_seed_rain(p2)
```

Then collect the patch-level dynamics:

```{r}
data2 <- run_scm_collect(p2_eq)

t2 <- data2$time
h1 <- data2$species[[1]]["height", , ]
h2 <- data2$species[[2]]["height", , ]

cols <- c("#e34a33", "#045a8d")
matplot(t2, h1, lty=1, col=make_transparent(cols[[1]], .25), type="l",
        las=1, xlab="Time (years)", ylab="Height (m)")
matlines(t2, h2, lty=1, col=make_transparent(cols[[2]], .25))
```

Alternatively we can compare the growth of species 1 by itself or
with another species:

```{r}
matplot(t, h, lty=1, col=make_transparent("black", .25), type="l",
        las=1, xlab="Time (years)", ylab="Height (m)")
matlines(t2, h1, lty=1, col=make_transparent(cols[[1]], .25))
```

This shows that the additional species does not affect the growth
of the *initial* wave of cohorts (because the second species is
growing more slowly and is shorter than the first species), but
subsequent waves are slowed or eliminated.

The dynamics are easier to see when coded by cohort density (some
of the lines here represent cohorts at close to zero density).

Relativise the log densities onto (-4, max)

```{r}
d1 <- data2$species[[1]]["log_density", , ]
d2 <- data2$species[[2]]["log_density", , ]
rel <- function(x, xmin) {
  x[x < xmin] <- xmin
  xmax <- max(x, na.rm=TRUE)
  (x - xmin) / (xmax - xmin)
}
rd1 <- rel(d1, -4)
rd2 <- rel(d2, -4)
```

R doesn't seem to offer a way to plot lines that vary in colour, so
this is quite roundabout using `segments`, shaded by the density at
the first part of the line segment:

```{r}
n <- length(t2)
x <- matrix(rep(t2, ncol(h1)), nrow(h1))
col1 <- matrix(make_transparent(cols[[1]], rd1), nrow(d1))
col2 <- matrix(make_transparent(cols[[2]], rd2), nrow(d2))
plot(NA, xlim=range(t2), ylim=range(h1, na.rm=TRUE),
     las=1, xlab="Time (years)", ylab="Cohort height (m)")
segments(x[-1, ], h2[-1, ], x[-n, ], h2[-n, ], col=col2[-n, ], lend="butt")
segments(x[-1, ], h1[-1, ], x[-n, ], h1[-n, ], col=col1[-n, ], lend="butt")
```

Then, the total leaf area:

Each element of the resulting list is a Patch object, the same as
was observed when running the model.

```{r}
patches2 <- lapply(seq_along(data2$time), scm_patch, data2)
lai2 <- sapply(patches2, function(x) x$area_leaf_above(0.0))

lai2_1 <- sapply(patches2, function(x) x$species[[1]]$area_leaf_above(0.0))
lai2_2 <- sapply(patches2, function(x) x$species[[2]]$area_leaf_above(0.0))

plot(t2, lai2, type="l", las=1, lty=2,
     xlab="Time (years)", ylab="Leaf area index")
lines(t2, lai2_1, col=cols[[1]])
lines(t2, lai2_2, col=cols[[2]])
```

To find the average value over the metapopulation we weight by patch abundance:

```{r}
metapopulation <- function(x){
  plant:::trapezium(t2, x*data2$patch_density)
}

lai2_av <- metapopulation(lai2)
lai2_1_av <- metapopulation(lai2_1)
lai2_2_av <- metapopulation(lai2_2)

plot(t2, lai2, type="l", las=1, lty=2,
     xlab="Time (years)", ylab="Leaf area index")
lines(t2, lai2_1, col=cols[[1]])
lines(t2, lai2_2, col=cols[[2]])

axis(4, at=lai2_av,   tck=0.1, lty = 2, labels=NA)
axis(4, at=lai2_1_av, tck=0.1, col.ticks=cols[[1]], labels=NA)
axis(4, at=lai2_2_av, tck=0.1, col.ticks=cols[[1]], labels=NA)
axis(1, at=108, labels="Av")
```
