---
title: Report on FF16 stand
output:
  html_document:
    df_print: kable
    highlight: tango
    keep_md: no
    smart: no
    theme: yeti
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
params:
    results: provide
    target_ages: .na
editor_options:
  chunk_output_type: console
---

<!-- hack to get indentation on 3rd level of floating TOC; see
https://stackoverflow.com/questions/46201753/rmarkdown-indentation-of-toc-items-in-html-output
 -->

<script>
$(document).ready(function() {
  $items = $('div#TOC li');
  $items.each(function(idx) {
    num_ul = $(this).parentsUntil('#TOC').length;
    $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
  });

});
</script>


```{r setup, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
# knitr defaults
knitr::opts_chunk$set(echo=TRUE, cache=FALSE, message=FALSE, warning=FALSE)

options(
  # default for table format  
  knitr.table.format = "html", 
  # remove warnings from dplyr about "summarise"
  dplyr.summarise.inform = FALSE,
  readr.show_col_types = FALSE
  )

library(ggplot2)
library(dplyr)
library(purrr)
library(tidyr)
library(ggridges)
library(patchwork)

theme_set(theme_classic())

select_closest <- function(x, targets) {
  f <- function(target, x) {
    dist <- abs(x - target)
    x[which(dist == min(dist))]
  }

  x <- unique(x)

  targets %>%
    purrr:::map_dbl(f, x)
}

```

```{r, echo = FALSE, eval=FALSE}

# this is not run by default, just some code to generate some example data, if needed 
p0 <- scm_base_parameters("FF16", "FF16_Env")
p0$strategy_default$recruitment_decay <- 5
p0$strategy_default$a_dG2 <- 30

hyperpar <- make_FF16_hyperpar(B_dI1 = 0)

traits <- trait_matrix(c(0.7, 5), c("lma", "eta"))
p1 <- expand_parameters(traits, p0, hyperpar = hyperpar, mutant = FALSE,
                        birth_rate_list = list(
                          list(x = seq(0, 200, len = 20),
                               y = sin(seq(0, 200, len = 20)) + 2)
                          ))
                        
results <- build_schedule(p1)
results <- run_scm_collect(results)


```

This reports provides some standardised reporting for simulations of stands using the `FF16` strategy within the plant model. You can reuse code for your own reports, or take this report and modify it for your own use.

Each report documents a single simulation.

# Inputs

## Site properties


(This is a placeholder for now. We need further work to report on these, as they are not currently available within the parameters object.)

The site was simulated with following properties

- Latitude: XXXX (input to hyper par function, not accessible in output)
- Productivity: XXXX (input to hyper par function, not accessible in output)
- Solar input: XXXX (input to hyper par function, not accessible in output)

## Species 

The simulation was run with `r length(results$p$strategies)` species.

### Parameters

The following parameters were included in the model. Note, values for species are only shown when they differ to the default.

```{r}
p <- results$p

extract_strategy <- function(strat) {
  ex <- c("control" , "collect_all_auxiliary",   "birth_rate_x", "birth_rate_y", "is_variable_birth_rate")

  x <- strat[!(names(strat) %in% ex)]
  
  tibble(parameter = names(x), value = unlist(x))
}

pars <-  
  purrr::map_df(p$strategies, extract_strategy, .id = "species") %>% 
  mutate(species = paste("sp", species)) %>%
  bind_rows(
    extract_strategy(p$strategy_default) %>% mutate(species = "default"),
    .
  ) %>% 
  group_by(parameter, .drop = TRUE) %>%
  mutate(
    # format values
    value = format(value, scientific = TRUE, drop0trailing = TRUE, digits=4),
    # remove repated values
    value = ifelse(duplicated(value), NA, value)
  ) %>%
  pivot_wider(values_from = value, names_from = species)

pars_def <- 
 system.file("docs/FF16", "FF16-params-core.csv", package = "plant") %>% 
 read.csv(stringsAsFactors = FALSE, header = TRUE) %>%
 rename("parameter" = "Code")

names(pars_def) <- tolower(names(pars_def))

# detect rows for bold
i_bold <- grep("**", pars_def$description, fixed = TRUE)
pars_def$description <- gsub("**", "", pars_def$description, fixed = TRUE)

pars_def %>% 
  left_join(by = "parameter", pars) %>% 
  select(description, parameter, everything()) %>% 
  select(-symbol) %>% 
  mutate(across(everything(), ~replace_na(.x,""))) %>%
  knitr::kable(align = "l") %>% 
  kableExtra::kable_material(c("striped", "hover")) %>%
  kableExtra::row_spec(i_bold, bold = TRUE)
```

### Traits

- XXX Comparisons to data (Austraits)
- XXX Along trade-off curves(requires data from hyperpar?)

### Birth rates

XXX

```{r, eval = FALSE}

extract_birthrates <- function(strategy, times) {
  if (strategy$is_variable_birth_rate) {
    f <- splinefun(strategy$birth_rate_x, strategy$birth_rate_y)
  } else {
    f <- function(times) {strategy$birth_rate_y}
  }
  
  tibble(time = times,
        birth_rate = f(times))
  }

  df <- 
    p$strategies %>%
    map_df(extract_birthrates, times = results$time, .id = "species") 

  ggplot(df, aes(time, birth_rate, color = species)) +
    geom_line() +
    ylim(c(0, max(df$birth_rate)*1.2)) +
    labs(title = "Annual birth rate",
         x = "Time (years)",
         y = "Birth rate (# / m2 / yr)",
         color = "Species")

```

### Leaf photosynthesis

Species had the following potential leaf photosynthetic rates, calculated by integrating instantaneous photosynthesis across the solar path throughout the year.

```{r}
E <- seq(0,1, by=0.01)

leaf_rates <- 
  p$strategies %>% purrr::map_df(~tibble(E, A = .x$a_p1 * E / (E+ .x$a_p2)),.id = "species") %>% 
mutate(
  species = paste("sp", species)
  ) 
  
leaf_rates %>% 
ggplot(aes(E, A, group = species, col= species)) + 
geom_line() +
labs(
  title = "Annual leaf photosynthesis",
  x = "Canopy openness (0-1)",
  y = "Photosynthesis (mol/m2/yr)"
)

```

# Individual trajectories

XXX Compare to BAAD 

# Stand dynamics - size structured 

This section provides plots on size-structured population dynamics. For the plots below we're going to use the tidy outputs format of the results.

```{r}
results_tidy_expand <- 
  results %>% 
  tidy_patch() %>% 
  FF16_expand_state()

data_species <-
  results_tidy_expand$species %>%
  drop_na()

data_species_tot <-
  data_species %>%
  integrate_over_size_distribution()

canopy <-
  results_tidy_expand$env$canopy %>%
  group_by(step) %>% 
  mutate(
    # LAI at height
    LAI_above = -log(canopy_openness)/0.5,
    # Cumulative LAI from the rgound up
    LAI_below = max(LAI_above) - LAI_above
  ) %>% ungroup()


# unless specified in inputs, select ages to make plots
if(any(is.na(target_ages))) {
  target_ages <- c(0.1, 1, 2.5, 5, 10, 20, 40, 80, 160)
}

target_ages <- target_ages[target_ages <= max(results$time)]

```

## Size distributions

This plot shows the size-density distribution of the population. 

- XXX notes/links on interpreting density

```{r}
data_species %>%
  plot_size_distribution()
```


## Growth rates

These plots show the distribution of growth rates at different ages in the stand. 

```{r}
get_dt <- function(y,x) {
  c(diff(y)/diff(x), NA)
}

vars <- c("height", "diameter_stem", "mass_above_ground")

growth_rates <-
  data_species %>% 
  select(time, species, node, height, mortality, fecundity, any_of(vars)) %>%
  group_by(species, node) %>%
  mutate(
    dt = c(diff(time), NA),
    across(c("height", "mortality", "fecundity", any_of(vars)), ~ c(diff(.x), NA) / dt, .names = "{.col}_dt")
  ) %>% 
  ungroup() %>%
  filter(time %in% select_closest(time, target_ages)) %>% 
  mutate(
    species = paste("sp", species),
    time = round(time)) %>%
  select(-dt) %>%
  select(time, species, height, ends_with("dt"))
  
growth_rates_max <- 
  growth_rates %>%
  group_by(time, species) %>%
  slice(1) %>%
  ungroup() %>%
  pivot_longer(ends_with("dt"))

growth_rates %>% 
  select(-mortality_dt, -fecundity_dt) %>%
  pivot_longer(ends_with("dt")) %>%
  ggplot(aes(x = height, y = value, label = time, group = time, col = time)) +
  geom_line() +
  geom_text(data = growth_rates_max %>% filter(!name %in% c("mortality_dt", "fecundity_dt")), nudge_x = 0.5) +
  labs(
    title = "Growth rates vs size",
    x = "Individual height (m)",
    y = "Growth rate (unit / yr)"
  ) -> p
```
The first plot shows species with different x and y axis, enabling best visualisation for each species
```{r}
p + facet_wrap(vars(name, species), scales = "free")
```

The next plot shows all species with a common y axis and variable x, enabling comparison between species
```{r}
p + facet_grid(vars(name), vars(species), scales = "free")
```

## Mortality

These plots show the distribution of mortality rates at different ages in the stand. 

```{r}
growth_rates %>%
  select(species, height, time, mortality_dt) %>%
  pivot_longer(ends_with("dt")) %>%
  ggplot(aes(x = height, y = value, label = time, group = time, col = time)) +
  geom_line() +
  geom_text(data = growth_rates_max %>% filter(name %in% c("mortality_dt")), nudge_x = 0.5) +
  labs(
    title = "Mortality rates vs size",
    x = "Individual height (m)",
    y = "Mortality rate ( / yr)"
  ) -> p
```

The first plot shows species with different x and y axis, enabling best visualisation for each species
```{r}
p + facet_wrap(vars(species), scales = "free")
```

The next plot shows all species with a common y axis and variable x, enabling comparison between species
```{r}
p + facet_wrap(vars(species), scales = "free_x")
```

Todo:

- XXX Do we want to separate out growth vs independent?

## Fecundity

These plots show the distribution of fecundity rates at different ages in the stand. 

```{r}

growth_rates %>%
  select(species, height, time, fecundity_dt) %>%
  pivot_longer(ends_with("dt")) %>%
  ggplot(aes(x = height, y = value, label = time, group = time, col = time)) +
  geom_line() +
  geom_text(data = growth_rates_max %>% filter(name %in% c("fecundity_dt")), nudge_x = 0.5) +
  labs(
    title = "Fecundity rates vs size",
    x = "Individual height (m)",
    y = "Fecundity rate ( / yr)"
  ) -> p
```
The first plot shows species with different x and y axis, enabling best visualisation for each species
```{r}
p + facet_wrap(vars(species), scales = "free")
```

The next plot shows all species with a common y axis and variable x, enabling comparison between species
```{r}
p + facet_wrap(vars(species), scales = "free_x")
```

# Stand dynamics - aggregate outputs

## Leaf area

The following plot shows the total leaf area for each species in the stand over time.

```{r}
data_species_tot %>%
  ggplot(aes(time, area_leaf, colour = species, fill=species)) +
  geom_line() + 
  xlab("Patch age (yr") +
     ylab("Leaf area index (m2/m2)")
```

The next plots shows the amount of leaf area contributed by plants of different height in the stand. We have three different ways of visualising this.

Note, this is the leaf area contributed by whole individuals, against their height; so not the precise distribution of leaf area within the stand. The leaf area for each individual is distributed across a range of heights.

This first set shows the density of leaf area by individual height.

```{r}
d <- data_species %>%
  filter(time %in% select_closest(time, target_ages))

p <- 
  d %>%
  ggplot(aes(x = height, height = density * area_leaf, y = time, group = time, fill = species)) 
  
for(s in unique(d$species))
  p <- p + 
  geom_ridgeline(data = filter(d, species== s), stat = "identity", scale = 2)

  p +  
  labs(
    title = "Leaf area contributed by individual height",
    y = "Patch age (yr)",
    x = "Individual height (m)"
  ) +
  theme_ridges() +
  coord_flip()

```

This next set shows the CUMULATIVE leaf area by individual height.

```{r}
trapezium_segments <- function(x, y) c(0, diff(x) * 0.5 * (y[-length(y)] + y[-1]))

cumulative_trapezium <- function(x,y) {
  cumsum(trapezium_segments(x,y))
}

rev_cumulative_trapezium <- function(x,y) {
  rev(cumsum(trapezium_segments(rev(x), rev(y))))
}

d2 <- 
  data_species %>%
  stats::na.omit() %>%
  filter(time %in% select_closest(time, target_ages)) %>% 
  mutate(species = paste("sp ", species)) %>%
  group_by(time, species) %>%
  mutate(
    LAI_below = 
    rev_cumulative_trapezium(height, density * area_leaf)
  ) %>% ungroup()
  
d2_com <- 
  d2 %>%
  bind_rows(
    canopy %>% mutate(species= "whole stand") %>%
      filter(time %in% select_closest(time, target_ages))
   )

d2_max <- 
  d2_com %>%
    group_by(time, species) %>%
    filter(height == max(height)) %>%
    ungroup()

p <- 
  d2 %>%
  ggplot(aes(x = height, y=time, height = LAI_below, group = time, fill = species))

  for (s in unique(d2$species)) {
    p <- p +
      geom_ridgeline(data = filter(d2, species == s), stat = "identity", scale = 2)
  }
  
  p + 
  labs(
     title = "Cumulative leaf area by individual height",
     y = "Patch age (yr)",
     x = "Individual height (m)"
   ) +
   theme_ridges() +
   coord_flip()
```

Another way of looking at the same data:

```{r}
d2_com %>%
  ggplot(aes(y = height, x = LAI_below, label = round(time), group = time, col = time)) +
  geom_line() +
  geom_text(data = d2_max, nudge_x = 0.1) +
  labs(
    title = "Cumulative leaf area from ground, total (left) and by individual (right)",
    x = "Total leaf area from ground (m2/m2)",
    y = "Individual height (m)"
  ) + facet_wrap(~species, ncol =1)

```

## Basal area 

The following plot shows the total stem basal area for each species in the stand over time.

```{r}
data_species_tot %>%
  ggplot(aes(time, area_stem, colour = species)) +
  geom_line() + 
  xlab("Patch age (yr") +
  ylab("Total basal area (m2/m2)")
```

## Biomass distribution

The following plot shows the cumulative leaf area for each species in the stand.
Tissues 

```{r}
v <- c("mass_heartwood", "mass_sapwood", "mass_bark", "mass_leaf")

data_long <-
  data_species_tot %>%
  select(time, species, one_of(v)) %>%
  pivot_longer(cols = starts_with("mass"), names_to = "tissue")

data_long$tissue <- factor(data_long$tissue, levels = v)

species_names <- tibble(
  species = unique(data_long$species),
  species_name = paste("Species", species)
)

data_long %>%
  left_join(by = "species", species_names) %>%
  ggplot(aes(time, value, fill = tissue)) +
  geom_area() +
  labs(x = "Patch age (yr)", y = "Above ground mass (kg/m2)") +
  theme_classic() -> p
```

The first plot shows species with different y axis, enabling best visualisation for each species
```{r}
p + facet_wrap(vars(species), ncol = 1, scales = "free")
```

The next plot shows all species with a common y axis, enabling comparison between species
```{r}
p + facet_wrap(vars(species), , ncol = 1, scales = "free_x")
```

## Carbon budgets 

XXX

### NPP

XXX

### Turnover

XXX

## Environment

XXX

### Light availability

```{r}
d2 <-
    canopy %>% mutate(species = "canopy") %>%
    filter(time %in% select_closest(time, target_ages)) %>% 
    mutate(
      time = round(time),
      LAI = -log(canopy_openness)/ 0.5      
    )

d2_max <- d2 %>%
  group_by(time, species) %>%
  slice(n())

p1 <- 
  d2 %>%
  ggplot(aes(x = height, y = canopy_openness, label = time, group = time, col = time)) +
  geom_line() +
  geom_text(data = d2_max, nudge_y = 0.05) +
  labs(
    title = "Canopy openness against height, through time",
    x = "",
    y = "Canopy openness (0-1)"
  )


p2 <- d2 %>%
  ggplot(aes(x = height, y = LAI, label = time, group = time, col = time)) +
  geom_line() +
  geom_text(data = d2_max, nudge_y = 0.05) +
  labs(
    title = "Leaf area index against height, through time",
    x = "height from ground (m)",
    y = "LAI above (m2/m2)"
  )

p1 / p2 
```
