---
title: "TF24 hydraulic scenario scorecard"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
params:
  scorecard: NULL
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
scorecard <- params$scorecard
meta <- attr(scorecard, "metadata")
```

This scorecard evaluates how well the current TF24 build reproduces the
**expected** behaviours in `model_scenarios_hydraulic.csv` (issue
[#549](https://github.com/traitecoevo/plant/issues/549), epic
[#424](https://github.com/traitecoevo/plant/issues/424)). Each scenario's
qualitative descriptors are translated into concrete parameters via
`scenario_mapping.csv`, run through the SCM, and classified. A scenario
**matches** when the observed outcome equals the expected one — including an
*expected failure* that does fail.

## Provenance

```{r provenance}
if (!is.null(meta)) {
  knitr::kable(data.frame(
    field = names(meta),
    value = vapply(meta, function(x) paste(as.character(x), collapse = ", "),
                   character(1)),
    row.names = NULL))
}
```

## Summary

```{r summary}
s <- plant::scenario_summary(scorecard)
knitr::kable(s, digits = 3)
```

```{r headline, results = "asis"}
cat(sprintf("\n**Match rate: %d / %d (%.0f%%)**\n",
            s$n_match, s$n, 100 * s$match_rate))
```

## Scorecard

`observed` is *success* when the run completes with finite, positive offspring
production, else *failure*. `outcome` refines this into `persisted`
(finite, offspring > 0), `extinct` (finite run, no offspring) or `crashed`
(a numerical failure — thrown error or non-finite state). The distinction
matters: the hydraulic/NSC work (#549, #554) targets *crashes*, whereas
*extinct* is often correct model behaviour.

```{r scorecard}
cols <- c("scenario_id", "scenario", "expected", "observed", "match",
          "outcome", "offspring_production", "run_seconds", "error_message")
knitr::kable(as.data.frame(scorecard)[, cols], digits = 4)
```

## Known caveats in the scenario design

- **Confounded traits.** Within each named scenario the trait bundle moves
  together (acquisitive/mesic vs conservative/xeric), so a failure cannot be
  attributed to any single trait — these are paired archetypes, not a factorial.
- **Binary expectations.** The table records only pass/fail, with no graduated
  target or tolerance.
- **Derived hydraulic traits (#548).** `p_50`, `c`, `b` and `psi_crit` are
  derived from `K_s`, and `g1_TF24` from `rho`, inside `TF24_hyperpar`. The CSV's
  `Ks`/`Psi_crit,b,c` and `Wood density`/`g1_TF24` columns are consistent with
  those relationships, so the mapping sets `K_s` and `rho` and lets the hyperpar
  derive the rest.
```
