---
title: "Package Organisation and Management"
output: bookdown::html_document2
---

**authors**: DS Falster

```{r setup, include=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
```



# Development


## Directory structure

The code is organised as follows. This structure reflects structure generally used for R packages including C++ source.


```sh
├── inst
│   ├── RcppR6_classes.yml    # Used by RCppR6 to generate the interface for Rcpp (C++ classes)
│   ├── RcppR6_functions.yml  # Used by RCppR6 to generate the interface for Rcpp (C++ function)
│   ├── include
│   │   ├── plant.h # THE main header file for package - sources other headers
│   │   ├── plant 	# header (.h) files for different components, many including implementation
│   │   └── tk      # simple cubic spline interpolation library
│   ├── reference_plant_ff16 # Reference plant physiology ff16
│   ├── docs    		# Tables of variables for models
│   └── scripts 		# Tools used in package management
├── man 			  		# R Documentation files (*.Rd)
├── R 							# R functions
├── scripts 	  		# R scripts
├── src 			  		# C++ files with implementation where not included in headers
├── tests 		  		# Tests
└── vignettes   		# Vignettes
```

## Issues

We use 

- [github issues](https://github.com/traitecoevo/plant/issues/) for feature requests, bug reports and  developer coordination.
- [github projects](https://github.com/orgs/traitecoevo/projects) for planning of work, including roadmaps, milestones and sprint planning. 

## Gitflow workflow

We use the [Gitflow](https://www.atlassian.com/git/tutorials/comparing-workflows/gitflow-workflow) workflow.

This means we have at least two main branches always present:

1. a master branch (`master`) for stable releases
2. a development branch (`develop`) where all development takes place

We also use feature branches which branch from the development branch where we can then introduce new features and then pull them back into the development branch. (though we don't usually have a release branch).

![](https://wac-cdn.atlassian.com/dam/jcr:34c86360-8dea-4be4-92f7-6597d4d5bfae/02%20Feature%20branches.svg?cdnVersion=1744)

## Testing


## Continuous integration via GH Actions



## Compiling

There are two main ways to compile the package:

1. Via a makefile, which defines a set of targets that can be built via the `make` command in terminal/command line
2. Directly within R using `pkgbuild::compile_dll(compile_attributes = FALSE, debug=FALSE)`
3. Directly within R using `devtools::load_all()` (can be slwoer, use with cuation. See below.)

### Makefile

The makefile is defined in the root directory of the package and is called `Makefile`. It defines multiple targets in the package build process. Each rule defines a target and the commands to run to build that target. These are called by running 
```
make target_name
```
where `target_name` is the target to build. Targets are as follows:

- **`all`**: Compiles the package and builds the function documentation to `/man`
- **`compile`**: Builds the dll
- **`debug`**: Builds the dll with debug information
- **`install`**: Installs this package (` R CMD INSTALL .`)
- **`build`**: Builds plant (` R CMD build --no-build-vignettes`)
- **`RcppR6`**:  Update or install RcppR6 files
- **`attributes`**: Scans the source files for attributes and generate code as required. Generates the bindings required to call C++ functions from R for functions labelled with the `Rcpp::export` attribute
- **`roxygen`**: Compiles the documentation for R
- **`test`** Run the test suite
- **`check`**: [Checks](http://r-pkgs.had.co.nz/check.html) plant for common problems
- **`clean`**: Deletes `*.o` and `*.so` files that were produced when compiling the package
- **`vignettes`**: Builds the vignettes

We use R to execute the compiling via `pkgbuild::compile_dll`, evem within the makefile.

### devtools::load_all produces a slow binary

Which build method you use to compile can drastically change spped at runtime. In particular, different levels of compiler optimisation are triggered based on the method used to complile the pacakge.

An easy way to compile and load the binary is by running
```
devtools::load_all()
```

However, this compiles code without optimisation, cuasing the model to run much slower. Similarly, by default, `pkgbuild::compile_dll` has argument debug=TRUE which leads causes debug symbols to be included and slow runtime.

If you compile with optimisation, the code runs much faster. By default, `R CMD INSTALL`, `devtools::install()` or `make` will compile with optimisation. You will want to clean first. 

When developing, you can still get optimisation by compiling as follows. After changing some cpp code, run

```{r, eval=FALSE}
pkgbuild::compile_dll(debug=FALSE, compile_attributes = FALSE)
devtools::load_all()
```

The first line recompiles with optimisation, and the second line loads the package with the new code. If the code is already compiled, `devtools::load_all` won't recompile it. If you skip the first line, your code will be recompiled with debug symbols, and will be slow.

Also, including `compile_attributes = FALSE` in the first line avoids the need to recompile the Rcpp exports.

## Profiling



# Building/updating the website:

Website mainatined on spereate branch [`gh-pages`](https://github.com/traitecoevo/plant/tree/gh-pages)
Wesbite built with [pkgdown](http://pkgdown.r-lib.org), drawing on files in package such as
- Readme
- News
- folder `man`: 
- folder `vignettes`: 

To build and push the website, do as follows:

1. Open up package on develop or master
2. Clone branch containing the website into folder `gh-pages`, within root of pacakges
```
git clone --branch gh-pages --single-branch git@github.com:traitecoevo/plant.git gh-pages
```
3. Rebuild the website using `pkgdown::build_site()`
4. Go into the gh-apges folder, commit and push

If you just want to build the wsbite locally, you only need to complete step 3.



# Program Design

`plant` is build as an R package using C++ code, linked via [Rcpp](https://cran.r-project.org/package=Rcpp). Rcpp provides "provides R functions as well as C++ classes which offer a seamless integration of R and C++." 

Two common challenges with using Rcpp are 1) that there is a lot of interface code (or "boilerplate") needed, 2) Rcpp "modules" can be slow to load. To overcome these we use the package [RcppR6](https://github.com/richfitz/RcppR6) to generate the interface. The RcppR6 package aims to provide a simple way of generating boilerplate code for exposing C++ classes to R. It is similar in many ways to Rcpp "modules" but without the slow load time. RcppR6 creates an interface using Rcpp's ["attributes"](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-attributes.pdf) facilities to build the actual R/C++ glue. The [R6](https://github.com/wch/R6) R package is the reference class that we use for wrapping the generated class. To achieve the link, class definitions are written into a [YAML](http://en.wikipedia.org/wiki/YAML) file.

## Templating

We use templating to enable plant to work with multiple strategies and lets us write code that is strategy generic. The templated class definitions are found in `inst/include/plant` and their function implementations are found either in the `.h` files or in `/src`.

We then use the [RcppR6](https://github.com/richfitz/RcppR6) package combined with the `.h` files and class definitions (found in `inst/RcppR6_classes.yml`) into Rcpp which we can then be used in the `.R` files (in `/R`) or exported.

Templating Resources:

* [RccpR6 Templates](http://htmlpreview.github.io/?https://raw.githubusercontent.com/richfitz/RcppR6/master/inst/doc/templates.html)
* [Templating in C++](https://en.wikipedia.org/wiki/Template_(C%2B%2B))

## Classes and their functions

The main c++ header files (i.e. \*.h) are stored within the directory `inst/include/plant/`. In many cases, the implementations is written directly into the header files. Otherwise, the corresponding source file (i.e. \*.cpp) is stored within the folder `src`.

The role of these different files is as follows.

strategies: ff16_strategy.h

plant.h

node.h

species.h

patch.h

SCM.h

Helpers:

- parameters.h
- node_schedule.h
- control.h
- disturbance.h
- environment.h

plant_runner.h
individual_tools.h
scm_utils.h

stochastic_patch_runner.h
stochastic_patch.h
stochastic_species.h
stochastic_utils.h

### Utilities

adaptive_interpolator.h
gradient.h
interpolator.h
lorenz.h
ode_control.h
ode_interface.h
ode_r.h
ode_runner.h
ode_solver.h
ode_step.h
qag_internals.h
qag.h
qk_rules.h
qk.h

RcppR6_post.hpp
RcppR6_pre.hpp
RcppR6_support.hpp
uniroot.h
util_post_rcpp.h
util.h

## Strategies

The foundation of the `plant` package are sub-models for an individual species’ physiological strategy. The `FF16` strategy is the default strategy. However, the plant package has been written to enable new physiological strategies to be added, and then use the same machinery for modelling size-structured population dynamics, we do this by using templating in c++ (see [Templating](#templating)).

Adding a new strategy (and, optionally, environment) is a cross-cutting change: every templated C++ class and every R dispatch table must learn the new `<Strategy, Environment>` pair. Rather than making these edits by hand, use the **scaffolder** (`scripts/new_strategy_scaffolder.R`), which clones an existing model and wires the pair through the codebase. The full workflow — scaffolding, the own-environment vs reuse-environment modes, implementing the biology (worked Kohyama-1993 example), and the build/test loop — is documented in the **`new-strategy` skill** (`.claude/skills/new-strategy/`) and summarised in [agents.md](../../agents.md) §7.
