---
title: "recruitment_dynamics"
output: html_document
date: "2022-09-12"
editor_options: 
  chunk_output_type: console
---
 
```{r}
devtools::load_all(".")
```

```{r}
library(tidyverse)
```

```{r}
library(plant)
```

```{r}
library(furrr)
```

Set base parameters

```{r}
p0 <- scm_base_parameters("FF16w")
```

Define a number of hyper-parameterisations, including the relationship between K_s and WD

```{r}
p_50_2_K_s <- function(p50){
  1.638999*(p50/2)^(-1.38)
}

calc_vul_b <- function(p_50, c){
  num <- p_50
  den <- (-log(1-50/100))^(1/c)
  num/den
}


calc_psi_crit <- function(b,c) {
  b*(log(1/0.05))^(1/c)
}
```

Run the models, setting the max patch lifetime, the p50s, and hyperparametersising K_s, b, psi_ crit

```{r}
# set default values 
run_FF16w_model <- function(inputs){
  

  ctrl = scm_base_control()

  for(v in names(ctrl)){
    if(length(inputs[[v]]) > 0){
    if(!is.na(inputs[[v]])){
    ctrl[[v]] <- inputs[[v]]
    }
    }
  }
  
  p0 <- scm_base_parameters("FF16w")
  p0$max_patch_lifetime <- inputs$max_patch_lifetime

  p50_1 <- inputs$p50_1
  p50_2 <- inputs$p50_2

  if(is.na(inputs$epsilon_leaf)){
    epsilon_leaf = 0.00001
  } else{
    epsilon_leaf = inputs$epsilon_leaf
  }

  ## increase birth rates to 10 or 50
  
  p1 <- expand_parameters(trait_matrix(c(p50_1, p50_2), "p_50"), p0, mutant=FALSE, birth_rate_list = c(1,1))
  
  p1$strategies[[1]]$K_s <- p_50_2_K_s(p50_1)
  p1$strategies[[1]]$b <- calc_vul_b(p50_1, p1$strategies[[1]]$c)
  p1$strategies[[1]]$psi_crit <- calc_psi_crit(p1$strategies[[1]]$b, p1$strategies[[1]]$c)
  p1$strategies[[1]]$epsilon_leaf <- epsilon_leaf
  p1$strategies[[1]]$recruitment_decay <- inputs$recruitment_decay


  p1$strategies[[2]]$K_s <- p_50_2_K_s(p50_2)
  p1$strategies[[2]]$b <- calc_vul_b(p50_2, p1$strategies[[2]]$c)
  p1$strategies[[2]]$psi_crit <- calc_psi_crit(p1$strategies[[2]]$b, p1$strategies[[2]]$c)
  p1$strategies[[2]]$epsilon_leaf <- epsilon_leaf
  p1$strategies[[2]]$epsilon_leaf <- inputs$recruitment_decay

  env <- make_environment("FF16w", soil_initial_state = rep(inputs$initial_theta, 1), rainfall = inputs$rainfall)
  time1 <- system.time(result <- build_schedule(p1, env, ctrl))
  time2 <- system.time(result <- run_scm_collect(result, env, ctrl, collect_auxiliary_variables = TRUE))
  return(list(result, time1, time2))
}
```

Because I'm also looking to see where the `plant` model is breaking at the momemnt, I can use the `safely` function to run the function in a map or for loop, and where it breaks, I can ensure that it continues running, and saves the error for the defined parameters so that I can go back through and look at them later. 

```{r}
poss_run_plant = safely(.f = run_FF16w_model, otherwise = "Error")
```

Here, I set some random inputs for rainfall, p50_1, p50_2. I also defined rho as either NA (which will then be defined via hyperparametersiation depedning on p50, or set it at a constant vlaue of 0.6, which is the average value in `austraits`.)

```{r}
inputs <- expand_grid(rainfall = 2.93, p50_1 = 5.99, p50_2 = 8.46, initial_theta = c(0.1), max_patch_lifetime = 11, schedule_eps = NA, ode_tol_abs = NA, ode_tol_rel = NA, epsilon_leaf = NA, recruitment_decay = c(1)) %>%
  slice(sample(1:n()))
```

```{r}
plan(multisession, workers = 8)
```

```{r}
run_and_save_plant_water_model <- function(..., save = FALSE, dir = NULL){
  input_parameters <- tibble(...)

  if(save == TRUE){
  if(!file.exists(paste("examples_vignettes/outputs/", dir, sep = ""))){
  browser()

    dir.create(paste("examples_vignettes/outputs/", dir, sep = ""))
    
  }
  
  dir <- paste("examples_vignettes/outputs/", dir, "/", sep = "")
  
  filename <-paste("rainfall_", round(input_parameters$rainfall,2), 
                   "p501_", round(input_parameters$p50_1, 2), 
                   "p502_", round(input_parameters$p50_2, 2), 
                   "lifetime_", input_parameters$max_patch_lifetime,
                   "schedule_eps", round(input_parameters$schedule_eps, 6),
                   "ode_tol_abs", round(input_parameters$ode_tol_abs, 6),
                   "ode_tol_rel", round(input_parameters$ode_tol_rel, 6),
                   "epsilon_leaf", round(input_parameters$epsilon_leaf, 6),
                   "theta", input_parameters$initial_theta,
                   "recruitment_decay", input_parameters$recruitment_decay,
                   ".RDS", sep="")
                   

  path <- paste(dir, filename, sep="")
  
  if(!file.exists(path)){
  result <- poss_run_plant(input_parameters)
  result<-list(result, input_parameters)
  saveRDS(result, path)
  }
  else {
    print(paste("skipping", path, "already done", sep = " "))
  }
  } else {
  
  result <- poss_run_plant(input_parameters)
  result<-list(result, input_parameters)
  
  return(result)
  }
  
  
  return
}
```

```{r}
furrr::future_pmap(inputs, run_and_save_plant_water_model, save = TRUE, dir = "recruitment_dynamics")

result <- pmap(inputs, run_and_save_plant_water_model, save = FALSE, dir = "recruitment_dynamics")

```

```{r}
result <- readRDS(paste("examples_vignettes/outputs/recruitment_dynamics/",list.files("examples_vignettes/outputs/recruitment_dynamics")[2], sep = ""))

result[[1]][[1]]$result[[1]] %>%
  tidy_patch() %>%
  FF16_expand_state() %>%
  pluck("species") %>%
  plot_size_distribution()
  
result[[1]][[1]]$result[[1]] %>%
  tidy_patch() %>%
  FF16_expand_state() %>%
  pluck("env") %>%
  distinct(time, theta) %>%
  ggplot(aes(x = time, y= theta)) +
  geom_line()


result[[1]][[1]]$result[[1]] %>%
  tidy_patch() %>%
  FF16_expand_state() %>%
  pluck("species") %>%
  select(-opt_psi_stem) %>%
  ggplot(aes(x = time, y=  opt_ci)) +
  geom_line(aes(group = interaction(node, species), colour = species))
```

