---
title: "A canopy that thins"
subtitle: "Flexible allometry in TF24 — what it does, and why it does not yet buy survival"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5
---

```{r setup, message=FALSE, warning=FALSE}
# Staging demo: needs the in-tree dev build (#516), not necessarily the
# installed `plant`.
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

# Shared computation, kept out of this file so a test can exercise it: a chunk
# that runs is not a chunk that is checked.
source(file.path(pkg_root, "overstorey_staging", "allometry_demo_helpers.R"))

theme_set(theme_bw(base_size = 12))
```

Real plants under stress thin their canopies. TF24 could not: leaf area was pinned
to height, so a struggling plant had no way to economise and no option but to die.
It now has one.

This page shows the new mechanism working — a plant thinning through a drought and
rebuilding afterwards, and a suppressed plant doing the same in the shade — and
then shows the problem with it. **Thinning saves the plant almost nothing.** We
have now tested two different triggers for it, and both give a survival gain of
about one per cent while costing two metres of height. That rules the trigger out
and points somewhere else, which is the useful part.

The mechanism is off by default. Everything here runs with it switched on.

## Principles

These are the rules the build follows. Most of the design decisions below are
consequences of them rather than independent choices, and where a measurement
overturned one it is said so explicitly.

1. **The old model must be recoverable exactly.** Switching the mechanism off
   returns the fixed allometry bit for bit, not approximately. This is a
   diagnostic, not an end in itself: it is how we know a decomposition is
   algebraically right rather than merely plausible.
2. **Bounds should follow from the form, not from clamps.** Where a quantity must
   stay in a range, arrange the flow so it cannot leave — a rate that vanishes at
   the boundary. A clamp changes the answer silently, is a kink in every
   downstream derivative, and hides the overshoot it was added to catch.
3. **Everything smooth.** No `min`, `max` or branches in the carbon-to-growth map.
   Automatic differentiation is coming, and a kink propagates into every
   sensitivity that passes through it.
4. **Carbon and tissue balance as an identity.** Carbon a plant declines to spend
   and tissue it consequently loses are two halves of one decision, so they are
   computed from one number and asserted as an identity, never as a tolerance.
5. **Only gate a pool that has a state.** Withholding the cost of tissue that
   cannot shrink saves carbon no tissue paid for — a free lunch, and a silent one.
6. **Scale-free where possible.** Rates are relative, so the same parameter means
   the same thing for a seedling and a mature tree. This is what keeps the
   plasticity parameters from repeating `storage_prod_eps`'s mistake of being an
   absolute constant against fluxes spanning six orders of magnitude.
7. **Signals that drive slow structure must be slow.** Assimilation swings with
   the weather; leaf and stem take years to build. Anything reading the first to
   decide the second reads a running mean.
8. **Numerically safe always.** A state that runs somewhere its arithmetic stops
   meaning anything is a defect even when the biology it represents is right —
   a dying plant is still integrated alongside live ones.
9. **Follow the evidence, and record what it overturns.** Two of the decisions
   here reverse earlier ones after measurement, and both retractions are kept in
   place rather than quietly edited away.

## What changed

TF24 used to pin leaf area to height through a fixed allometry, `A = A*(h)`. A
plant could not carry less leaf than its height implied, and its height could only
ever increase, so *thinning was not expressible at all*.

Leaf area and sapwood area are now ODE states. But they are carried as **departures
from their preferred values** rather than as the areas themselves:

| state | in words | resting value |
|---|---|---|
| `log_area_leaf_departure` | how much less canopy the plant carries than its height implies | 0 |
| `log_area_sapwood_departure` | how much more conducting stem it carries than its canopy needs | 0 |

Leaf area is then derived rather than integrated. Writing `phi` for the first
departure, `A = A*(h) * exp(phi)`, so `phi = 0` means "exactly on the old
allometry" and `phi = -0.7` means "half a canopy".

::: {.callout-note}
## Why the two move in opposite directions
Both departures are defined the same way — the log of what a plant *has* over what
it *should* have. What differs is the reference, because the allometry is a chain:

$$h \;\longrightarrow\; A = A_\star(h) \;\longrightarrow\; A_s = \theta A$$

Each departure is measured against its own parent in that chain. `phi` asks *"how
much leaf do I carry, against what my height implies?"*; `psi` asks *"how much
conducting stem do I carry, against what my canopy needs?"*.

Under stress they then move oppositely, and that is the physical result rather than
an inconsistency. Leaves turn over faster than sapwood, so the canopy falls below
what its height implies (`phi` negative) while the stem is left carrying more than
the now-smaller canopy needs (`psi` positive).

Referencing both to height instead — which would be $\psi' = \psi + \phi$ — would
give them a shared sign. It would cost two things: `psi` would stop being the log
Huber value, which is the quantity the hydraulics actually reads, and the clean
`psi >= 0` invariant would become the messier $\psi' \ge \phi$.
:::

Carrying departures rather than levels is what makes the old model recoverable
*exactly*. Integrating leaf area alongside height would make `A = A*(h)` a
redundant relationship that the ODE solver holds only to its own error tolerance,
so height would end up reading a slightly drifted leaf area. On the departure
coordinate the allometric part stays exact and only the departure is integrated —
so with the mechanism off, its rate is exactly zero and the old model comes back
bit for bit.

**The switch is `a_pl0`**: the largest fraction of leaf and sapwood turnover a
plant may decline to replace. `a_pl0 = 0`, the default, is the old model.

```{r versions}
s <- TF24_Strategy()
data.frame(
  quantity = c("model", "number of states", "a_pl0 (default, = off)",
               "a_pl1 (reserve level where withholding is half its max)",
               "a_pl2 (gap scale)",
               "k_l (leaf turnover, /yr)", "k_s (sapwood turnover, /yr)"),
  value = c(model_id("TF24"), TF24_Individual(s)$ode_size,
            s$pars$a_pl0, s$pars$a_pl1, s$pars$a_pl2,
            round(s$pars$k_l, 4), s$pars$k_s)) |>
  knitr::kable()
```

### How it works, in words

1. **A plant sheds when its marginal leaf stops paying for itself.** Not when its
   reserves run out — when the last leaf's assimilation no longer covers its own
   upkeep and that of the fine root and bark that come with it. Below that point,
   the plant declines to replace some of the leaf and sapwood its turnover
   destroys, and *is not charged for what it does not replace*.
2. **It reads a running mean, not today's weather.** Assimilation swings daily; a
   canopy takes years to build. The gate reads a filtered balance with a memory of
   about four months — long against weather, short against a drought.
3. **Leaves go faster than stem.** Leaves turn over at `k_l = 0.46`/yr and sapwood
   at `k_s = 0.20`/yr, so withholding both thins the canopy faster than the stem.
   Conducting area per unit leaf area therefore *rises*. Nothing imposes this; it
   falls out of the two rate constants.
4. **When the balance recovers, the plant rebuilds**, redirecting growth from
   height into canopy, tapering off as the gap closes.
5. **Height never falls, and the canopy never empties.** A plant in deficit stops
   growing, and thinning tapers to nothing at a floor.

::: {.callout-important}
## Why the trigger is the carbon balance and not the reserve pool
The first version gated on the NSC reserve fraction: shed once reserves are nearly
gone. Two lines of evidence overturned it.

**Theory.** Manzoni et al. (2015) maximise a season's carbon gain over leaf area
and *derive* the trigger: `dG/dL = 0`, i.e. shed once a leaf stops covering its
upkeep. That fires while the plant is still carbon-rich — their Fig. 4D has
cumulative gain at its seasonal *maximum* when shedding starts. **Waiting for
reserves to run down is precisely the evergreen strategy in that paper, and it is
the one that loses a long drought**, running its season's gain from +150 to
−50 gC m⁻².

**Field data.** In Robinson's mulga, 85 per cent of trees shed — the healthiest
included — and the died and survived canopy distributions are indistinguishable in
the first year. Shedding is what healthy trees do, not what dying trees do last.

Measured here, the reserve gate bought a survival factor of 1.0–1.13 even when
moved as early as it would go. The carbon-balance gate is the better model on the
evidence; it does not, as it turns out, fix survival either — see below.
:::

## What it does

### The gate

The gate responds to soil water — through the marginal leaf's carbon balance —
while reserves are held *full* throughout. That is the demonstration that reserves
are no longer the axis it reads.

```{r gate, fig.height=4, fig.cap="A 5 m plant across soil water, with reserves at 0.8 of capacity throughout. The old gate could not fire anywhere on this axis."}
gate <- allometry_gate_map(a_pl0 = 1.0, height = 5)

p_margin <- ggplot(gate, aes(theta_soil, margin)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "soil water content", y = "marginal leaf return",
       title = "Is the last leaf paying for itself?",
       subtitle = "Zero is break-even: Manzoni's criterion")

p_rates <- gate |>
  select(theta_soil, leaf = dphi, sapwood = dpsi) |>
  pivot_longer(-theta_soil) |>
  ggplot(aes(theta_soil, value, colour = name)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "soil water content", y = "departure rate (per year)", colour = NULL,
       title = "What the plant does about it",
       subtitle = "Canopy thins; stem-per-leaf rises")

p_margin + p_rates
```

The two departure curves have **opposite signs** — that is the acclimation, and it
is `k_l > k_s` showing up. And the response is switch-like in soil water, because
the quantity it reads crosses zero sharply: between 0.15 and 0.13 the marginal
leaf goes from 88 per cent in profit to 106 per cent in loss.

### Through a drought and out the other side

A plant sits in wet soil, is dropped into a four-year drought, and is then returned
to wet soil. Soil water is prescribed directly here so the carbon balance can be
driven cleanly — this is *not* how the SCM drives soil water.

```{r trajectory, message=FALSE, warning=FALSE}
history <- demo_soil_history(wet = 0.30, dry = 0.11,
                             years_wet = 2, years_dry = 4)
traj <- bind_rows(
  allometry_trajectory(0.0, history, dt = 0.25),
  allometry_trajectory(1.0, history, dt = 0.25)) |>
  mutate(allometry = ifelse(a_pl0 == 0, "fixed", "flexible"))
```

```{r trajectory-plot, fig.height=7, fig.cap="Shaded band is the drought. Canopy fullness of 1 means the plant sits on the old allometry."}
dry_band <- annotate("rect", xmin = 2, xmax = 6, ymin = -Inf, ymax = Inf,
                     fill = "grey85", alpha = 0.6)

p1 <- ggplot(traj, aes(time, canopy_fraction, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = NULL, y = "canopy fullness", colour = NULL,
       title = "Canopy, relative to what its height implies")

p2 <- ggplot(traj, aes(time, huber_ratio, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = NULL, y = "stem per leaf, relative", colour = NULL,
       title = "Conducting area per unit leaf area")

p3 <- ggplot(traj, aes(time, height, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = "time (yr)", y = "height (m)", colour = NULL,
       title = "Height")

(p1 / p2 / p3) + plot_layout(guides = "collect")
```

The same trajectories seen in the plane the allometry lives in. The fixed plant is
confined to the curve; the flexible one leaves it and comes back.

```{r allometry-plane, fig.height=4.5, fig.cap="Leaf and sapwood area against height. Dashed lines are the fixed allometry. Note the two y-axes differ by four orders of magnitude."}
s_demo <- TF24_Strategy()
ref <- allometry_reference_curve(
  s_demo, seq(min(traj$height) * 0.95, max(traj$height) * 1.05,
              length.out = 200))

p_leaf <- ggplot(traj, aes(height, area_leaf, colour = allometry)) +
  geom_line(data = ref, aes(height, area_leaf), inherit.aes = FALSE,
            linetype = "dashed", colour = "grey40") +
  geom_path(linewidth = 0.9) +
  labs(x = "height (m)", y = "leaf area (m²)", colour = NULL,
       title = "Leaf area vs height",
       subtitle = "Dashed: the fixed allometry")

p_sap <- ggplot(traj, aes(height, area_sapwood, colour = allometry)) +
  geom_line(data = ref, aes(height, area_sapwood), inherit.aes = FALSE,
            linetype = "dashed", colour = "grey40") +
  geom_path(linewidth = 0.9) +
  labs(x = "height (m)", y = "sapwood area (m²)", colour = NULL,
       title = "Sapwood area vs height",
       subtitle = "Dashed: the pipe model, theta * A*(h)")

(p_leaf + p_sap) + plot_layout(guides = "collect")
```

These two panels are the clearest statement of what the change does. The fixed
plant cannot leave the dashed curve — that was the whole constraint. The flexible
one tracks it while conditions are good, drops **below** it during the drought as
the canopy thins at constant height — a vertical fall, since height is frozen —
and then climbs back onto it as the canopy rebuilds.

The sapwood panel shows the asymmetry — but compare the two *relative* to their
own curves rather than by eye, since the y-axes differ by four orders of magnitude.
At the worst of the drought, leaf area falls to **19 per cent** of what its height
implies while sapwood area falls only to **46 per cent**, because sapwood is lost
at `k_s` and leaf area at the faster `k_l`. That gap between the two is exactly the
rise in conducting area per unit leaf area.

```{r trajectory-table}
traj |>
  filter(allometry == "flexible", time %in% c(2, 3, 4, 5, 6, 6.5, 7.5, 8)) |>
  transmute(time, soil = theta_soil,
            `canopy fullness` = round(canopy_fraction, 3),
            `stem per leaf` = round(huber_ratio, 3),
            height = round(height, 2)) |>
  knitr::kable()
```

All of this emerges from the two rate constants and the reserve gate. None of it is
imposed:

- the canopy thins to about **19 per cent** of what its height implies, and can
  thin no faster than `k_l`, because the mechanism is *declining to replace* rather
  than actively shedding;
- conducting area per unit leaf area rises to about **2.4 times** its preferred
  value, purely because `k_l > k_s`;
- **height never falls** — it stops, and resumes when the drought breaks;
- on recovery the canopy rebuilds to about **98 per cent**, smoothly, with no jump
  as the gap closes.

So the mechanism does what it was designed to do.

### In the shade: self-thinning in a wet stand

Shade is as much the point as drought. A plant losing the light race has the same
problem as one losing water — its leaves stop paying for themselves — and the same
lever. Here a wet stand (rainfall 1.5 m/yr) is left to self-thin for 25 years.

```{r selfthin, message=FALSE, warning=FALSE}
thin_flex <- allometry_self_thinning(a_pl0 = 1.0)
thin_fix  <- allometry_self_thinning(a_pl0 = 0.0)
```

```{r selfthin-plot, fig.height=4.5, fig.cap="Every cohort alive at each census, against its height. Shading is a within-stand gradient, so the canopy response is a function of position in it."}
late <- thin_flex$cohorts |>
  filter(time %in% unique(time)[round(seq(0.3, 1, length.out = 4) *
                                      length(unique(time)))])

p_pos <- ggplot(late, aes(height, canopy_fraction, colour = factor(round(time)))) +
  geom_hline(yintercept = 1, linewidth = 0.3, colour = "grey50") +
  geom_point(size = 1.1) +
  labs(x = "height (m)", y = "canopy fullness", colour = "patch age",
       title = "Suppressed plants thin; dominants do not")

p_surv <- ggplot(late, aes(canopy_fraction, survivorship,
                           colour = factor(round(time)))) +
  geom_point(size = 1.1) + scale_y_log10() +
  labs(x = "canopy fullness", y = "survivorship", colour = "patch age",
       title = "And a thinned canopy goes with a dying plant")

(p_pos + p_surv) + plot_layout(guides = "collect")
```

```{r selfthin-table}
thin_flex$cohorts |>
  filter(time == max(time)) |>
  arrange(desc(height)) |>
  slice(unique(round(seq(1, n(), length.out = 7)))) |>
  transmute(height = round(height, 2),
            `canopy fullness` = signif(canopy_fraction, 3),
            `stem per leaf` = signif(huber_ratio, 3),
            survivorship = signif(survivorship, 2)) |>
  knitr::kable()
```

This is the behaviour asked for, and it emerges from competition alone: dominants
sit exactly on their allometry, suppressed plants thin progressively, and the
thinnest are the ones on their way out. Nothing in the model knows about canopy
position — the light gradient does all of it.

```{r selfthin-r0}
data.frame(allometry = c("fixed", "flexible"),
           R0 = c(thin_fix$R0, thin_flex$R0)) |>
  knitr::kable()
```

Lifetime offspring production is essentially unchanged, because the cohorts that
thin are the ones already contributing nothing.

## ⚠️ The problem: thinning does not buy survival

It does not, however, do what it was *for*. The point was to stop a drop in
productivity turning straight into death.

```{r survival, fig.height=4, fig.cap="Survivorship falls through ten orders of magnitude with or without the mechanism."}
surv <- traj |>
  select(time, theta_soil, allometry, survivorship) |>
  pivot_wider(names_from = allometry, values_from = survivorship) |>
  mutate(gain = flexible / fixed)

p_s <- traj |>
  ggplot(aes(time, survivorship, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) + scale_y_log10() +
  labs(x = "time (yr)", y = "survivorship", colour = NULL,
       title = "Survivorship is essentially unchanged")

p_g <- ggplot(surv, aes(time, gain)) +
  dry_band + geom_hline(yintercept = 1, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "time (yr)", y = "flexible / fixed",
       title = "What thinning buys")

p_s + p_g
```

```{r survival-numbers}
surv |>
  filter(time %in% c(2, 4, 6, 8)) |>
  transmute(time, soil = theta_soil,
            `survivorship, fixed` = signif(fixed, 3),
            `survivorship, flexible` = signif(flexible, 3),
            `ratio` = round(gain, 4)) |>
  knitr::kable()
```

**The gain is about one per cent**, against a survivorship that falls through ten
orders of magnitude either way. And thinning *costs* something real: the flexible
plant finishes around **two metres shorter**, having spent its recovery carbon on
canopy instead of height. That is why switching the mechanism on lowered lifetime
offspring production in every stand tested — by factors of 0.55 to 0.64 on seasonal
stands.

### Why — and it is not the trigger

Three measurements rule out the obvious explanations, and a fourth rules out the
trigger itself.

**It is not that the sheddable costs are too small.** Most of a plant's
maintenance *does* follow its canopy — leaf, fine root and bark all scale with
leaf area, and only sapwood is sticky:

| height | leaf | bark | sapwood | root | reachable by thinning |
|---|---|---|---|---|---|
| 10 m | 49.6% | 5.4% | 23.5% | 21.5% | **76.5%** |

(An earlier draft of this page claimed sapwood respiration dominated the drought
deficit. That was wrong, and this table is the retraction.)

**It is that thinning cuts income just as fast.** Assimilation scales with leaf
area too, so shrinking the canopy shrinks both sides of the budget together. Net
production per unit leaf area is flat as a plant thins, and slightly *worsens*
because the stem it still carries is spread over fewer leaves:

| canopy fullness | 1.00 | 0.50 | 0.20 | 0.10 | 0.05 | 0.018 |
|---|---|---|---|---|---|---|
| net production per m² leaf | −3.81 | −3.69 | −3.77 | −3.92 | −4.15 | −4.68 |
| stem per leaf, relative | 1.0 | 1.5 | 2.5 | 3.6 | 5.4 | 9.5 |

A plant whose leaves cost more than they earn cannot fix that by having fewer
leaves. Thinning slows the burn; it never reaches balance.

**There is no drought in which it changes the outcome.** Sweeping severity and
duration, the transition from fine to dead is a *cliff*:

| soil water | 0.20 | 0.17 | 0.15 | 0.13 | 0.11 |
|---|---|---|---|---|---|
| survivorship, 4-yr drought | 0.93 | 0.93 | 0.93 | 4.9e-10 | 3.4e-10 |

**And the trigger is not the problem.** This is the decisive one. We have now run
two different gates through the same drought:

| gate | fires when | survival gain | height cost |
|---|---|---|---|
| reserve fraction, centred at 0.05 | reserves nearly gone | 1.010 | ~2 m |
| reserve fraction, moved as early as it goes | reserves at 30% | 1.0–1.13 | more |
| **marginal leaf balance** (current) | **leaf stops paying, reserves still full** | **1.001** | ~2.5 m |

Firing *earlier*, on a *better-supported* signal, buys nothing. The carbon-balance
gate is still the right model — the evidence for it is independent of survival —
but survival is downstream of something else.

**And a grid of intensity against duration shows why.** The two thresholds are in
the wrong order in stress space:

| soil water | survivorship (fixed, 4 yr) | minimum canopy reached | survival gain |
|---|---|---|---|
| 0.145 | 0.911 | **1.000** — never thins | 1.00 |
| 0.140 | 1.5e-06 | **0.989** — barely thins | **1.06** |
| 0.135 | 1.0e-09 | 0.317 — thins hard | 1.01 |

At 0.145 the plant is healthy and its leaves are still profitable, so it has no
reason to shed. By 0.140 it is already effectively dead — and its canopy has moved
by one per cent. **Mortality saturates at a milder drought than shedding begins
at**, so there is no intensity at which a plant is both stressed enough to thin and
alive enough to profit from it. The largest gain in the whole grid, 1.06, sits in
that gap.

No gate can fix this, because the ordering is not the gate's doing: mortality
collapses as soon as net production turns slightly negative, since the pool has no
resting point and drains completely.

### The real cause is the reserve pool, and what mortality reads

The storage pool has **no resting point**. Its rate is
$\dot S = \text{charge}\,(1-r) - \text{drain}\,r$, so positive production fills it
to capacity and negative empties it, with nothing in between. The reserve fraction
is therefore close to a binary readout of the *sign* of net production. Mortality
then reads it through $a_{dG1}e^{-a_{dG2} r}$ with $a_{dG2}=20$, sharp enough to
turn a near-binary input into a near-binary output spanning eight orders.

So a mechanism that makes a plant modestly cheaper to run has nothing to act on,
whatever triggers it. The buffer does not buffer.

The literature says both halves of this are wrong (see the [NSC
appendix](#appendix-what-the-nsc-literature-says)): plants hold a **minimum NSC of
about 46 per cent** of their seasonal maximum and are found empty only when dead,
and they **sacrifice growth to defend storage** — the reverse of TF24, which gates
growth and leaves storage as the residual.

### Where the fix has to go

1. **Give the pool a resting point.** If the part of a deficit the pool cannot
   meet were paid by consuming tissue — which is what canopy thinning *is* — the
   pool would stabilise instead of emptying, and the reserve fraction would sit
   somewhere informative. NEWS already records that this unmet deficit currently
   leaves the budget, with "routing it into tissue loss" named as deferred work.
2. **Defend storage ahead of growth.** Beech under three years of drought
   *increased* NSC for two years before starch fell, and the authors conclude that
   maintaining storage at the expense of growth was key to survival. TF24 does the
   opposite.
3. **Change what mortality reads.** A reserve *fraction* cannot see a plant that
   has made itself cheaper to run — which is exactly what thinning does. Time to
   depletion at the current burn rate can, and predicted mortality better than any
   instantaneous level across 12 species.

## Also unresolved: nothing bounds the departures in permanent deficit

If reserves never recover, withheld leaf turnover (0.42/yr) outruns withheld
sapwood turnover (0.2/yr) whatever the plant does, so both departures grow without
limit — the canopy toward nothing, the stem-per-leaf ratio toward infinity.

This is arithmetic rather than a defect. A plant thinning at 46 per cent a year
while losing stem at 20 per cent really does have an ever-rising ratio. But what
removes such a plant is mortality, not the allometry, and a cohort model keeps
carrying it meanwhile.

Related, and worth watching: the rebuild rate grows as the canopy empties, which is
the same fast-boundary shape that #609 identified for the storage pool. It has not
caused a failure in any run so far.

## What is settled

```{r settled}
data.frame(
  claim = c("Switched off, the old model is recovered",
            "Carbon and tissue balance when replacement is withheld",
            "Stem per leaf never falls below its preferred value",
            "Height never decreases",
            "The canopy rebuilds after a drought"),
  basis = c("bit-identical, over 288 quantities and a full SCM run",
            "an exact identity: turnover not charged is mass lost",
            "follows from the form, given that k_l > k_s",
            "follows from the form: a product of non-negative factors",
            "measured: to ~98 per cent of preferred, smoothly")) |>
  knitr::kable()
```

The open questions are the two sections above: thinning does not yet buy survival,
and nothing bounds the departures for a plant that never recovers.

## Future refinements

Deliberately out of scope here, recorded so they are not rediscovered.

**A discrete re-flush drawing on stores.** Leaf construction *is* fully paid for
in the rebuild term — leaf, fine root, sapwood and bark at the preferred ratio —
so nothing is missing from the budget. What is missing is the *shape*: rebuilding
is continuous, financed out of current flux, whereas a real plant flushes a canopy
in a burst financed from reserves. Manzoni prices such a flush at 12–16 per cent of
a season's gain, and whether one happens changes his result qualitatively at long
drought duration. This is also the natural place for a reserve term to re-enter:
modulating the *cost of recovering*, not the *onset of shedding*.

**Acclimation of photosynthetic capacity.** `Amax` is fixed here, but it is
observed to scale back under sustained stress (Joshi et al. 2022). That is a
cheaper adjustment than shedding and a faster one, so it should sit between
stomatal closure and leaf loss in the ladder — and its absence may be part of why
the marginal balance crosses zero as abruptly as it does.

**Leaf turnover under an optimality argument.** `k_l` is currently a trait fixed by
the LMA trade-off. Leaf lifespan is itself an optimisation — hold a leaf while its
marginal return exceeds the cost of replacing it — so `k_l` could be derived rather
than imposed, in the same currency as the shedding gate. Attractive because it
would make shedding and leaf lifespan one mechanism instead of two.

**Optimality-derived allometric targets.** The preferred sapwood-to-leaf ratio, and
ultimately the preferred leaf area itself, could come from stationarity of the
plant's own growth rate rather than from fixed coefficients. Its own epic; the
sapwood target is written as a replaceable function so this can drop in.

**Drought-deciduous habit.** Setting the canopy floor near zero already permits
shedding to nothing and re-flushing, and that is supported behaviour for some
species. It is not the habit this model is aimed at, and the departure coordinate
degrades as the floor approaches zero, so it stays possible rather than
prioritised.

# Appendix: the mathematics {#appendix-the-mathematics}

::: {.callout-tip}
## These equations are checked, not just written down
`tests/testthat/test-allometry-demo.R` reimplements this appendix in R — working
from the prose rather than from the source — and compares it against what the C++
computes, across 81 combinations of height, both departures and reserve level.
They agree to 1e-10. So a divergence between this description and the model fails
a test, rather than quietly misleading a reader.
:::

### Symbols

| symbol | in the code | meaning |
|---|---|---|
| $h$ | `height` | height, a state (m) |
| $\phi$ | `log_area_leaf_departure` | leaf-area departure, a state |
| $\psi$ | `log_area_sapwood_departure` | sapwood departure, a state |
| $S$ | `storage` | NSC pool, a state (kg) |
| $A_\star(h)$ | `area_leaf(height)` | preferred leaf area (m²) |
| $A$, $A_s$, $A_b$ | — | actual leaf, sapwood, bark area (m²) |
| $\rho_\ell,\ \rho_s$ | `replacement`, `replacement_sapwood` | replaced share of leaf / sapwood turnover |
| $u_\ell,\ u_s$ | `withheld`, `withheld_sapwood` | withheld share, $1-\rho$ |
| $\sigma$ | `rebuild_share` | share of growth redirected to rebuilding |
| $b$ | `rebuild_rel` | relative rebuild rate (yr⁻¹) |
| $F$ | `growth_flux` | carbon for growth and reproduction (kg yr⁻¹) |

### Sizes

The allometry itself is unchanged, $A_\star(h) = (h/a_{l1})^{1/a_{l2}}$. What
changes is that a plant need not sit on it:

$$A = A_\star(h)\,e^{\phi}, \qquad A_s = \theta A\,e^{\psi} = \theta A_\star(h)\,e^{\phi+\psi}.$$

So $\phi$ is log canopy fullness, and $\psi$ is the log of the Huber value relative
to its preferred value $\theta$. Masses follow as before: $m_\ell = \lambda A$ (with
$\lambda$ the leaf mass per area), $m_r = a_{r1} A$, $m_s = A_s h \eta_c \rho$ and
$m_b = A_b h \eta_c \rho$.

::: {.callout-note}
Bark area is still pinned to *leaf* area, $A_b = a_{b1}\theta A$, rather than to
sapwood. Repinning it to sapwood is arguably more physical — bark sheathes the stem
— but it was left alone to keep this change bit-identical at rest.
:::

Hydraulic supply per unit leaf area reads the Huber value the plant actually has,

$$k_{\max,\text{leaf}} = \frac{K_s\,\theta e^{\psi}}{h\,\eta_c},$$

which is the channel through which thinning improves supply to the leaves that
remain.

### The replacement gate

With $r = S/S_{\max}$ and $S_{\max} = a_{st1} m_s$, the replaced share of leaf
turnover is a logistic on reserves. The sapwood share is that same quantity, scaled
down by how over-built the stem already is:

$$\rho_\ell = 1 - \frac{a_{pl0}}{1 + \exp\!\big((r - a_{pl1})/w_{pl}\big)},
\qquad
\rho_s = \rho_\ell\, e^{-\psi/a_{pl2}}.$$

$a_{pl0}$ caps how much may be withheld, so $a_{pl0}=0$ gives $\rho_\ell \equiv 1$.
Only the two pools that have states are gated; the turnover actually charged is

$$T = \rho_\ell k_\ell m_\ell + k_b m_b + \rho_s k_s m_s + k_r m_r.$$

Bark and root stay fully charged because neither has a state to lose. Withholding
their cost would save carbon that no tissue ever paid for.

### Growth, and where it goes

Net production and the existing NSC gate are untouched:

$$P = a_{bio}a_y(\tilde A - R) - T,\qquad
P_+ = \tfrac12\Big(P + \sqrt{P^2+\varepsilon^2}\Big),\qquad
F = P_+\,G(r).$$

$F$ then splits between **extension** along the allometry and **rebuilding** the
canopy, with the rebuild share gated on the same reserves as replacement:

$$\sigma = \rho_\ell\left(1 - e^{\phi/a_{pl2}}\right).$$

Extension moves the plant along the curve. Its marginal costs are evaluated at
$A_\star$ — the leaf area the plant's *height* implies — because they are properties
of the curve, and a plant's place on the curve is given by its height:

$$\dot A_\star = (1-\sigma)\,F f_g \left.\frac{\partial A}{\partial m_{\text{live}}}\right|_{A_\star},
\qquad
\frac{dh}{dt} = \left.\frac{\partial h}{\partial A}\right|_{A_\star}\dot A_\star
= a_{l1}a_{l2}A_\star^{\,a_{l2}-1}\,\dot A_\star .$$

Rebuilding buys leaf area at **unchanged height**, and buys the whole package —
leaf, fine root, and the sapwood and bark that supply it — at the ratio the plant
prefers rather than the one it currently has:

$$c_{\text{reb}}(h) = \lambda + a_{r1} + \theta h \eta_c \rho\,(1+a_{b1}),
\qquad
b = \frac{\sigma F f_g}{c_{\text{reb}}(h)\,A}.$$

### The two departure equations

$$\boxed{\;\frac{d\phi}{dt} = b - u_\ell k_\ell\;}
\qquad
\boxed{\;\frac{d\psi}{dt} = u_\ell k_\ell - u_s k_s + b\left(e^{-\psi}-1\right)\;}$$

In words: the canopy loses the turnover it declined to replace, and regains what
rebuilding buys. The Huber value moves on the **difference** between the two
losses. The final term is exact geometry rather than a parameter — rebuilding adds
leaf and sapwood in the ratio $\theta$, so it moves the two logs by $b/A$ and
$\theta b/A_s$, and their difference is $(b/A)(e^{-\psi}-1)$.

### Properties that follow from the form

**Thinning conserves carbon exactly.** Carbon not charged is
$u_\ell k_\ell m_\ell + u_s k_s m_s$. The withholding terms move the states by
$d(\ln A)/dt = -u_\ell k_\ell$ and, since
$d(\ln A_s)/dt = d(\ln A)/dt + d\psi/dt$, by $d(\ln A_s)/dt = -u_s k_s$. The mass
lost is therefore $\lambda u_\ell k_\ell A + h\eta_c\rho\,u_s k_s A_s$ — the same
two terms. An identity, not a tolerance.

**$\phi \le 0$ is invariant.** At $\phi=0$, $\sigma=0$ so $b=0$, leaving
$d\phi/dt = -u_\ell k_\ell \le 0$: the flow points inward. A plant never carries
*more* leaf than its height implies.

**$\psi \ge 0$ is invariant.** At $\psi=0$, $\rho_s = \rho_\ell$ so $u_s=u_\ell$,
and $e^{-\psi}-1 = 0$, leaving

$$\left.\frac{d\psi}{dt}\right|_{\psi=0} = u_\ell\,(k_\ell - k_s) \;\ge\; 0
\qquad\text{because } k_\ell > k_s .$$

So a stem never ends up with less conducting area than its canopy needs, the
scaling factor $e^{-\psi/a_{pl2}}$ stays within $(0,1]$, and no over-replacement
can arise to cost unbounded carbon. **⚠️ If $k_s$ were ever raised above $k_\ell$,
this argument reverses.**

**Height never falls.** $dh/dt$ is a product of $(1-\sigma) \in (0,1]$, $F \ge 0$,
$f_g \ge 0$, and two positive derivatives.

**$a_{pl0}=0$ is the old model exactly.** $\rho_\ell = 1$ gives $u_\ell = 0$;
$\phi=0$ gives $\sigma = 0$ and hence $b=0$; $\psi=0$ gives $\rho_s = 1$ and hence
$u_s = 0$. Both departure rates are exactly zero, $e^0 = 1$ leaves the sizes
bitwise unchanged, and odelia's **max**-norm error control ignores a zero-error
state, so the solver's step sizes are untouched as well.

### One approximation, stated

Extension is charged at $\partial m_{\text{live}}/\partial A$ evaluated at
$A_\star$, but a thinned plant adds only $e^{\phi}\,\dot A_\star$ of actual leaf
area for that money. A plant off its trajectory is therefore **over-charged** for
extension, by a factor $e^{-\phi}$.

The direction is conservative — a thinned plant grows more slowly than a strict
budget would allow, never faster — and the error vanishes as the canopy recovers.
It is the price of keeping extension exactly departure-preserving, which is what
makes the resting model exact and confines all of the departure's motion to
plasticity.

# Appendix: what the NSC literature says {#appendix-what-the-nsc-literature-says}

Gathered because the reserve pool turned out to be the binding constraint, not the
allometry. Every point here bears on a design decision above.

**Plants defend the pool rather than draining it.** A global synthesis of 121
studies and 177 species found **minimum NSC around 46 per cent of the seasonal
maximum**, with depletion of total NSC "rare"; the authors propose a dual function
in which starch is the reservoir while soluble sugars "are kept above some critical
threshold" (Martínez-Vilalta et al. 2016, *Ecol. Monogr.*,
[doi:10.1002/ecm.1231](https://doi.org/10.1002/ecm.1231)). TF24 lets the reserve
fraction run to ~1e-10.

**Storage outranks growth — the reverse of what we do.** Beech under three years of
drought *increased* NSC through years 1–2; starch fell only in year 3, and it was
the **dead** trees that were "virtually empty", with 87 per cent surviving. The
conclusion is explicit: *"maintaining an active C storage function at the expense
of growth was certainly key to survival"* (Chuste et al. 2019, *Trees*,
[doi:10.1007/s00468-019-01923-5](https://doi.org/10.1007/s00468-019-01923-5)).
After release from darkness, "NSC reformation occurs preferentially over growth"
(Weber et al. 2018, *New Phytol.*,
[doi:10.1111/nph.14987](https://doi.org/10.1111/nph.14987)).

**Under moderate stress the pool is actively regulated back up.** Across 47 studies
and 52 species, prolonged slight-to-moderate drought brought NSC back *toward*
control values — *"trees activated physiological regulation processes to increase
carbon storage and reduce the risks of carbon starvation"* — with net loss only
under long, severe drought (He et al. 2020, *Glob. Change Biol.*,
[doi:10.1111/gcb.15078](https://doi.org/10.1111/gcb.15078)).

**Depletion at death is partial.** Foliar NSC in *Pinus edulis* fell 30 per cent
through mortality, but "carbohydrates were not completely exhausted at mortality"
(Adams et al. 2013, *New Phytol.*,
[doi:10.1111/nph.12102](https://doi.org/10.1111/nph.12102)). Seedlings survive
weeks at starch of 0.4–0.8 per cent dry matter (Weber et al. 2018).

**What predicts death is a rate, not a level.** Across 12 species, total NSC and
soluble sugars did not decline at all; starch did, and *"the time leading to starch
depletion emerged as a strong mortality predictor"* (Trueba et al. 2024, *Ann. For.
Sci.*, [doi:10.1186/s13595-024-01246-7](https://doi.org/10.1186/s13595-024-01246-7)).
This is the direct argument for changing what mortality reads: a time-to-depletion
responds to a plant that has made itself cheaper to run, and an instantaneous
fraction cannot.

**Shade depletes reserves as drought does**, so the self-thinning case above is a
carbon problem and not only a light one (Maguire & Kobe 2015, *Ecol. Evol.*,
[doi:10.1002/ece3.1819](https://doi.org/10.1002/ece3.1819)); shade-induced depletion
also raises embolism vulnerability and blocks hydraulic recovery (Tomasella et al.
2021, *New Phytol.*, [doi:10.1111/nph.17384](https://doi.org/10.1111/nph.17384)).

**Life stage matters**, which our single size-scaled pool does not capture: mature
trees stayed homeostatic under throughfall exclusion while understorey seedlings
lost about 15 per cent of total NSC (Zhang et al. 2020, *For. Ecol. Manage.*,
[doi:10.1016/j.foreco.2020.118159](https://doi.org/10.1016/j.foreco.2020.118159)).

## And on leaf shedding itself

**The optimal trigger is a flux, and it fires early.** Manzoni et al. (2015,
*Adv. Water Resour.* 84:37–51) maximise a season's carbon gain over leaf area and
derive `dG/dL = 0` — shed when a leaf stops covering its upkeep — at about
−2 MPa soil water potential, *"a later and more drastic measure of water loss
regulation than stomatal closure"*. Gradual and instantaneous shedding give
*identical* carbon gain, so gradual is free in the optimality currency and is what
the field data show.

**Shedding is what healthy plants do.** In Robinson's mulga (573 trees, 49 sites,
2017–2024), 85 per cent shed, including the healthiest; the died and survived
canopy-density distributions are indistinguishable in the first year and only
separate later. Decline took 2–3 years; recovery took one season. Larger trees shed
*less* and survived better (canopy area the strongest mortality predictor), which
that work attributes to deep-water access rather than to carbon stores.

⚠️ **One correction to an earlier reading.** Severe de-leafing in that dataset went
with death: the trajectory groups that fell to near-zero canopy density all died.
Complete de-greening followed by re-leafing is documented for mulga but is rare,
and whether a plant can do it depends on its strategy. So a canopy floor is not
merely a numerical convenience for an evergreen — it is the observed behaviour.
