---
title: "A canopy that thins"
subtitle: "Flexible allometry in TF24 — what it does, and why it does not yet buy survival"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5
---

```{r setup, message=FALSE, warning=FALSE}
# Staging demo: needs the in-tree dev build (#516), not necessarily the
# installed `plant`.
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

# Shared computation, kept out of this file so a test can exercise it: a chunk
# that runs is not a chunk that is checked.
source(file.path(pkg_root, "overstorey_staging", "allometry_demo_helpers.R"))

theme_set(theme_bw(base_size = 12))
```

Real plants under stress thin their canopies. TF24 could not: leaf area was pinned
to height, so a struggling plant had no way to economise and no option but to die.
It now has one.

This page shows the new mechanism working — a plant thinning through a drought and
rebuilding afterwards — and then shows the problem with it. **Thinning currently
saves the plant almost nothing.** It survives about one per cent longer and ends up
two metres shorter. Finding out why is the useful part, and the answer turns out to
be about which costs a plant can actually shed.

The mechanism is off by default. Everything here runs with it switched on.

::: {.callout-note collapse="true"}
## Where things are
The maths is written out in full in the [appendix](#appendix-the-mathematics),
with each equation matched to the name it has in the code. The behaviour comes
first, on the grounds that it is easier to read equations once you know what they
do.
:::

## What changed

TF24 used to pin leaf area to height through a fixed allometry, `A = A*(h)`. A
plant could not carry less leaf than its height implied, and its height could only
ever increase, so *thinning was not expressible at all*.

Leaf area and sapwood area are now ODE states. But they are carried as **departures
from their preferred values** rather than as the areas themselves:

| state | in words | resting value |
|---|---|---|
| `log_area_leaf_departure` | how much less canopy the plant carries than its height implies | 0 |
| `log_area_sapwood_departure` | how much more conducting stem it carries than its canopy needs | 0 |

Leaf area is then derived rather than integrated. Writing `phi` for the first
departure, `A = A*(h) * exp(phi)`, so `phi = 0` means "exactly on the old
allometry" and `phi = -0.7` means "half a canopy".

::: {.callout-note}
## Why the two move in opposite directions
Both departures are defined the same way — the log of what a plant *has* over what
it *should* have. What differs is the reference, because the allometry is a chain:

$$h \;\longrightarrow\; A = A_\star(h) \;\longrightarrow\; A_s = \theta A$$

Each departure is measured against its own parent in that chain. `phi` asks *"how
much leaf do I carry, against what my height implies?"*; `psi` asks *"how much
conducting stem do I carry, against what my canopy needs?"*.

Under stress they then move oppositely, and that is the physical result rather than
an inconsistency. Leaves turn over faster than sapwood, so the canopy falls below
what its height implies (`phi` negative) while the stem is left carrying more than
the now-smaller canopy needs (`psi` positive).

Referencing both to height instead — which would be $\psi' = \psi + \phi$ — would
give them a shared sign. It would cost two things: `psi` would stop being the log
Huber value, which is the quantity the hydraulics actually reads, and the clean
`psi >= 0` invariant would become the messier $\psi' \ge \phi$.
:::

Carrying departures rather than levels is what makes the old model recoverable
*exactly*. Integrating leaf area alongside height would make `A = A*(h)` a
redundant relationship that the ODE solver holds only to its own error tolerance,
so height would end up reading a slightly drifted leaf area. On the departure
coordinate the allometric part stays exact and only the departure is integrated —
so with the mechanism off, its rate is exactly zero and the old model comes back
bit for bit.

**The switch is `a_pl0`**: the largest fraction of leaf and sapwood turnover a
plant may decline to replace. `a_pl0 = 0`, the default, is the old model.

```{r versions}
s <- TF24_Strategy()
data.frame(
  quantity = c("model", "number of states", "a_pl0 (default, = off)",
               "a_pl1 (reserve level where withholding is half its max)",
               "a_pl2 (gap scale)",
               "k_l (leaf turnover, /yr)", "k_s (sapwood turnover, /yr)"),
  value = c(model_id("TF24"), TF24_Individual(s)$ode_size,
            s$pars$a_pl0, s$pars$a_pl1, s$pars$a_pl2,
            round(s$pars$k_l, 4), s$pars$k_s)) |>
  knitr::kable()
```

### How it works, in words

A plant carries a pool of stored carbon (NSC). The fraction of that pool it still
holds, `r`, is the signal for everything below.

1. **When reserves run low, the plant stops replacing tissue it loses.** Leaves and
   sapwood turn over continuously; normally both are replaced automatically. Now a
   plant with low reserves declines to replace some of it, and *is not charged for
   what it does not replace*. Its canopy thins as a result.
2. **Leaves go faster than stem.** Leaves turn over at `k_l = 0.46`/yr and sapwood
   at `k_s = 0.20`/yr, so withholding both thins the canopy faster than the stem.
   Conducting area per unit leaf area therefore *rises* — the plant supplies each
   remaining leaf better. Nothing imposes this; it falls out of the two rates.
3. **When carbon returns, the plant rebuilds.** Some of its growth is redirected
   from getting taller into refilling the canopy, tapering off as the gap closes.
4. **Height never falls.** A plant in deficit simply stops growing.

The cuts are ordered, and the ordering emerges rather than being coded: growth and
reproduction are gated at $r = 0.10$, canopy replacement lower at $r = 0.05$. So a
struggling plant stops growing first, stops maintaining its canopy second, and only
then runs its reserves down.

## What it does

### The gate

Replacement is withheld only when reserves are genuinely low. Here is the response
at a fixed size, sweeping the reserve fraction:

```{r gate, fig.height=4, fig.cap="Departure rates against reserve fraction, for a 5 m plant with the mechanism fully on."}
gate <- allometry_gate_map(a_pl0 = 1.0, height = 5)

p_gate <- gate |>
  select(r, `leaf` = dphi, `sapwood` = dpsi) |>
  pivot_longer(-r) |>
  ggplot(aes(r, value, colour = name)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "reserve fraction  r", y = "departure rate (per year)",
       colour = NULL, title = "The replacement gate",
       subtitle = "Canopy thins (negative); stem-per-leaf rises (positive)")

p_h <- ggplot(gate, aes(r, dheight)) +
  geom_line(linewidth = 0.9) +
  labs(x = "reserve fraction  r", y = "height growth (m/yr)",
       title = "Height never falls",
       subtitle = "Gated separately, and at a higher reserve level")

p_gate + p_h
```

Two things to take from this. The leaf and sapwood curves have **opposite signs** —
that is the acclimation, and it is simply `k_l > k_s` showing up. And the gate is
**sharp**: by $r = 0.33$ both rates are down to ~1e-7. That is deliberate, since a
plant with reserves in hand should not be shedding leaves. But it is also why the
response is small in any stand where reserves rarely empty, which will matter
below.

### Through a drought and out the other side

A plant sits in wet soil, is dropped into a four-year drought, and is then returned
to wet soil. Soil water is prescribed directly here so the carbon balance can be
driven cleanly — this is *not* how the SCM drives soil water.

```{r trajectory, message=FALSE, warning=FALSE}
history <- demo_soil_history(wet = 0.30, dry = 0.11,
                             years_wet = 2, years_dry = 4)
traj <- bind_rows(
  allometry_trajectory(0.0, history, dt = 0.25),
  allometry_trajectory(1.0, history, dt = 0.25)) |>
  mutate(allometry = ifelse(a_pl0 == 0, "fixed", "flexible"))
```

```{r trajectory-plot, fig.height=7, fig.cap="Shaded band is the drought. Canopy fullness of 1 means the plant sits on the old allometry."}
dry_band <- annotate("rect", xmin = 2, xmax = 6, ymin = -Inf, ymax = Inf,
                     fill = "grey85", alpha = 0.6)

p1 <- ggplot(traj, aes(time, canopy_fraction, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = NULL, y = "canopy fullness", colour = NULL,
       title = "Canopy, relative to what its height implies")

p2 <- ggplot(traj, aes(time, huber_ratio, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = NULL, y = "stem per leaf, relative", colour = NULL,
       title = "Conducting area per unit leaf area")

p3 <- ggplot(traj, aes(time, height, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = "time (yr)", y = "height (m)", colour = NULL,
       title = "Height")

(p1 / p2 / p3) + plot_layout(guides = "collect")
```

The same trajectories seen in the plane the allometry lives in. The fixed plant is
confined to the curve; the flexible one leaves it and comes back.

```{r allometry-plane, fig.height=4.5, fig.cap="Leaf and sapwood area against height. Dashed lines are the fixed allometry. Note the two y-axes differ by four orders of magnitude."}
s_demo <- TF24_Strategy()
ref <- allometry_reference_curve(
  s_demo, seq(min(traj$height) * 0.95, max(traj$height) * 1.05,
              length.out = 200))

p_leaf <- ggplot(traj, aes(height, area_leaf, colour = allometry)) +
  geom_line(data = ref, aes(height, area_leaf), inherit.aes = FALSE,
            linetype = "dashed", colour = "grey40") +
  geom_path(linewidth = 0.9) +
  labs(x = "height (m)", y = "leaf area (m²)", colour = NULL,
       title = "Leaf area vs height",
       subtitle = "Dashed: the fixed allometry")

p_sap <- ggplot(traj, aes(height, area_sapwood, colour = allometry)) +
  geom_line(data = ref, aes(height, area_sapwood), inherit.aes = FALSE,
            linetype = "dashed", colour = "grey40") +
  geom_path(linewidth = 0.9) +
  labs(x = "height (m)", y = "sapwood area (m²)", colour = NULL,
       title = "Sapwood area vs height",
       subtitle = "Dashed: the pipe model, theta * A*(h)")

(p_leaf + p_sap) + plot_layout(guides = "collect")
```

These two panels are the clearest statement of what the change does. The fixed
plant cannot leave the dashed curve — that was the whole constraint. The flexible
one tracks it while conditions are good, drops **below** it during the drought as
the canopy thins at constant height — a vertical fall, since height is frozen —
and then climbs back onto it as the canopy rebuilds.

The sapwood panel shows the asymmetry — but compare the two *relative* to their
own curves rather than by eye, since the y-axes differ by four orders of magnitude.
At the worst of the drought, leaf area falls to **19 per cent** of what its height
implies while sapwood area falls only to **46 per cent**, because sapwood is lost
at `k_s` and leaf area at the faster `k_l`. That gap between the two is exactly the
rise in conducting area per unit leaf area.

```{r trajectory-table}
traj |>
  filter(allometry == "flexible", time %in% c(2, 3, 4, 5, 6, 6.5, 7.5, 8)) |>
  transmute(time, soil = theta_soil,
            `canopy fullness` = round(canopy_fraction, 3),
            `stem per leaf` = round(huber_ratio, 3),
            height = round(height, 2)) |>
  knitr::kable()
```

All of this emerges from the two rate constants and the reserve gate. None of it is
imposed:

- the canopy thins to about **19 per cent** of what its height implies, and can
  thin no faster than `k_l`, because the mechanism is *declining to replace* rather
  than actively shedding;
- conducting area per unit leaf area rises to about **2.4 times** its preferred
  value, purely because `k_l > k_s`;
- **height never falls** — it stops, and resumes when the drought breaks;
- on recovery the canopy rebuilds to about **98 per cent**, smoothly, with no jump
  as the gap closes.

So the mechanism does what it was designed to do.

## ⚠️ The problem: thinning does not buy survival

It does not, however, do what it was *for*. The point was to stop a drop in
productivity turning straight into death.

```{r survival, fig.height=4, fig.cap="Survivorship falls through ten orders of magnitude with or without the mechanism."}
surv <- traj |>
  select(time, theta_soil, allometry, survivorship) |>
  pivot_wider(names_from = allometry, values_from = survivorship) |>
  mutate(gain = flexible / fixed)

p_s <- traj |>
  ggplot(aes(time, survivorship, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) + scale_y_log10() +
  labs(x = "time (yr)", y = "survivorship", colour = NULL,
       title = "Survivorship is essentially unchanged")

p_g <- ggplot(surv, aes(time, gain)) +
  dry_band + geom_hline(yintercept = 1, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "time (yr)", y = "flexible / fixed",
       title = "What thinning buys")

p_s + p_g
```

```{r survival-numbers}
surv |>
  filter(time %in% c(2, 4, 6, 8)) |>
  transmute(time, soil = theta_soil,
            `survivorship, fixed` = signif(fixed, 3),
            `survivorship, flexible` = signif(flexible, 3),
            `ratio` = round(gain, 4)) |>
  knitr::kable()
```

**The gain is about one per cent**, against a survivorship that falls through ten
orders of magnitude either way. And thinning *costs* something real: the flexible
plant finishes around **two metres shorter**, having spent its recovery carbon on
canopy instead of height. That is why switching the mechanism on lowered lifetime
offspring production in every stand tested — by factors of 0.55 to 0.64 on seasonal
stands.

### Why

Mortality reads the reserve **fraction**. In a deep drought that fraction is pinned
near zero however thin the canopy gets, so mortality sits at its ceiling either
way. Thinning cannot help through that channel.

It ought to help through the **carbon balance** instead: less leaf to maintain
means a less negative carbon budget, so the pool drains more slowly. It does help —
production per unit leaf area climbs steadily as the canopy thins — but nowhere
near enough to matter.

```{r costs, fig.cap=NULL}
traj |>
  filter(allometry == "flexible", theta_soil < 0.2) |>
  transmute(time, `canopy fullness` = round(canopy_fraction, 3),
            `net production` = signif(P, 3),
            `per unit leaf area` = signif(P_per_area, 3)) |>
  head(9) |>
  knitr::kable()
```

The reason is which costs a plant can actually shed. Leaf and fine-root costs scale
with the canopy, so thinning cuts them. **Sapwood respiration scales with stem area
times height**, and thinning the canopy does not touch it — sapwood is lost only to
heartwood, at `k_s`, and deliberately so. For a tall plant in drought, that
unsheddable cost dominates the deficit, so cutting the sheddable part barely moves
the total.

There is a genuine tension here, and it has now been measured from both sides:

- if storage capacity followed **leaf area**, thinning would inflate the reserve
  fraction directly — mortality would fall for free, and a plant could shed its way
  to being immortal;
- because capacity follows **sapwood**, it cannot be inflated — but then thinning
  delivers no reserve benefit at all.

The way out has to be that thinning reduces the *deficit*. For that, the deficit
has to be dominated by costs that scale with the canopy. At TF24's defaults, for a
tall plant, it is not.

### Three candidates, none chosen

1. **Let the stem thin as well, faster than `k_s`.** Physically, heartwood
   conversion accelerating under stress, or sapwood being abandoned outright. It
   attacks the dominant cost — but it reopens the storage-capacity question and
   needs care not to restore the free lunch.
2. **Reconsider what mortality reads.** A reserve *fraction* is scale-free and
   bounded, which is why #517 chose it, but it is blind to a plant having made
   itself cheaper to run. Something like how long reserves would last at the
   current burn rate would respond to thinning.
3. **Check whether sapwood respiration is simply too high.** That `r_s` dominates a
   tall plant's drought deficit may be the real finding here — in which case this is
   a parameter question, not a structural one.

## Also unresolved: nothing bounds the departures in permanent deficit

If reserves never recover, withheld leaf turnover (0.42/yr) outruns withheld
sapwood turnover (0.2/yr) whatever the plant does, so both departures grow without
limit — the canopy toward nothing, the stem-per-leaf ratio toward infinity.

This is arithmetic rather than a defect. A plant thinning at 46 per cent a year
while losing stem at 20 per cent really does have an ever-rising ratio. But what
removes such a plant is mortality, not the allometry, and a cohort model keeps
carrying it meanwhile.

Related, and worth watching: the rebuild rate grows as the canopy empties, which is
the same fast-boundary shape that #609 identified for the storage pool. It has not
caused a failure in any run so far.

## What is settled

```{r settled}
data.frame(
  claim = c("Switched off, the old model is recovered",
            "Carbon and tissue balance when replacement is withheld",
            "Stem per leaf never falls below its preferred value",
            "Height never decreases",
            "The canopy rebuilds after a drought"),
  basis = c("bit-identical, over 288 quantities and a full SCM run",
            "an exact identity: turnover not charged is mass lost",
            "follows from the form, given that k_l > k_s",
            "follows from the form: a product of non-negative factors",
            "measured: to ~98 per cent of preferred, smoothly")) |>
  knitr::kable()
```

The open questions are the two sections above: thinning does not yet buy survival,
and nothing bounds the departures for a plant that never recovers.

# Appendix: the mathematics {#appendix-the-mathematics}

::: {.callout-tip}
## These equations are checked, not just written down
`tests/testthat/test-allometry-demo.R` reimplements this appendix in R — working
from the prose rather than from the source — and compares it against what the C++
computes, across 81 combinations of height, both departures and reserve level.
They agree to 1e-10. So a divergence between this description and the model fails
a test, rather than quietly misleading a reader.
:::

### Symbols

| symbol | in the code | meaning |
|---|---|---|
| $h$ | `height` | height, a state (m) |
| $\phi$ | `log_area_leaf_departure` | leaf-area departure, a state |
| $\psi$ | `log_area_sapwood_departure` | sapwood departure, a state |
| $S$ | `storage` | NSC pool, a state (kg) |
| $A_\star(h)$ | `area_leaf(height)` | preferred leaf area (m²) |
| $A$, $A_s$, $A_b$ | — | actual leaf, sapwood, bark area (m²) |
| $\rho_\ell,\ \rho_s$ | `replacement`, `replacement_sapwood` | replaced share of leaf / sapwood turnover |
| $u_\ell,\ u_s$ | `withheld`, `withheld_sapwood` | withheld share, $1-\rho$ |
| $\sigma$ | `rebuild_share` | share of growth redirected to rebuilding |
| $b$ | `rebuild_rel` | relative rebuild rate (yr⁻¹) |
| $F$ | `growth_flux` | carbon for growth and reproduction (kg yr⁻¹) |

### Sizes

The allometry itself is unchanged, $A_\star(h) = (h/a_{l1})^{1/a_{l2}}$. What
changes is that a plant need not sit on it:

$$A = A_\star(h)\,e^{\phi}, \qquad A_s = \theta A\,e^{\psi} = \theta A_\star(h)\,e^{\phi+\psi}.$$

So $\phi$ is log canopy fullness, and $\psi$ is the log of the Huber value relative
to its preferred value $\theta$. Masses follow as before: $m_\ell = \lambda A$ (with
$\lambda$ the leaf mass per area), $m_r = a_{r1} A$, $m_s = A_s h \eta_c \rho$ and
$m_b = A_b h \eta_c \rho$.

::: {.callout-note}
Bark area is still pinned to *leaf* area, $A_b = a_{b1}\theta A$, rather than to
sapwood. Repinning it to sapwood is arguably more physical — bark sheathes the stem
— but it was left alone to keep this change bit-identical at rest.
:::

Hydraulic supply per unit leaf area reads the Huber value the plant actually has,

$$k_{\max,\text{leaf}} = \frac{K_s\,\theta e^{\psi}}{h\,\eta_c},$$

which is the channel through which thinning improves supply to the leaves that
remain.

### The replacement gate

With $r = S/S_{\max}$ and $S_{\max} = a_{st1} m_s$, the replaced share of leaf
turnover is a logistic on reserves. The sapwood share is that same quantity, scaled
down by how over-built the stem already is:

$$\rho_\ell = 1 - \frac{a_{pl0}}{1 + \exp\!\big((r - a_{pl1})/w_{pl}\big)},
\qquad
\rho_s = \rho_\ell\, e^{-\psi/a_{pl2}}.$$

$a_{pl0}$ caps how much may be withheld, so $a_{pl0}=0$ gives $\rho_\ell \equiv 1$.
Only the two pools that have states are gated; the turnover actually charged is

$$T = \rho_\ell k_\ell m_\ell + k_b m_b + \rho_s k_s m_s + k_r m_r.$$

Bark and root stay fully charged because neither has a state to lose. Withholding
their cost would save carbon that no tissue ever paid for.

### Growth, and where it goes

Net production and the existing NSC gate are untouched:

$$P = a_{bio}a_y(\tilde A - R) - T,\qquad
P_+ = \tfrac12\Big(P + \sqrt{P^2+\varepsilon^2}\Big),\qquad
F = P_+\,G(r).$$

$F$ then splits between **extension** along the allometry and **rebuilding** the
canopy, with the rebuild share gated on the same reserves as replacement:

$$\sigma = \rho_\ell\left(1 - e^{\phi/a_{pl2}}\right).$$

Extension moves the plant along the curve. Its marginal costs are evaluated at
$A_\star$ — the leaf area the plant's *height* implies — because they are properties
of the curve, and a plant's place on the curve is given by its height:

$$\dot A_\star = (1-\sigma)\,F f_g \left.\frac{\partial A}{\partial m_{\text{live}}}\right|_{A_\star},
\qquad
\frac{dh}{dt} = \left.\frac{\partial h}{\partial A}\right|_{A_\star}\dot A_\star
= a_{l1}a_{l2}A_\star^{\,a_{l2}-1}\,\dot A_\star .$$

Rebuilding buys leaf area at **unchanged height**, and buys the whole package —
leaf, fine root, and the sapwood and bark that supply it — at the ratio the plant
prefers rather than the one it currently has:

$$c_{\text{reb}}(h) = \lambda + a_{r1} + \theta h \eta_c \rho\,(1+a_{b1}),
\qquad
b = \frac{\sigma F f_g}{c_{\text{reb}}(h)\,A}.$$

### The two departure equations

$$\boxed{\;\frac{d\phi}{dt} = b - u_\ell k_\ell\;}
\qquad
\boxed{\;\frac{d\psi}{dt} = u_\ell k_\ell - u_s k_s + b\left(e^{-\psi}-1\right)\;}$$

In words: the canopy loses the turnover it declined to replace, and regains what
rebuilding buys. The Huber value moves on the **difference** between the two
losses. The final term is exact geometry rather than a parameter — rebuilding adds
leaf and sapwood in the ratio $\theta$, so it moves the two logs by $b/A$ and
$\theta b/A_s$, and their difference is $(b/A)(e^{-\psi}-1)$.

### Properties that follow from the form

**Thinning conserves carbon exactly.** Carbon not charged is
$u_\ell k_\ell m_\ell + u_s k_s m_s$. The withholding terms move the states by
$d(\ln A)/dt = -u_\ell k_\ell$ and, since
$d(\ln A_s)/dt = d(\ln A)/dt + d\psi/dt$, by $d(\ln A_s)/dt = -u_s k_s$. The mass
lost is therefore $\lambda u_\ell k_\ell A + h\eta_c\rho\,u_s k_s A_s$ — the same
two terms. An identity, not a tolerance.

**$\phi \le 0$ is invariant.** At $\phi=0$, $\sigma=0$ so $b=0$, leaving
$d\phi/dt = -u_\ell k_\ell \le 0$: the flow points inward. A plant never carries
*more* leaf than its height implies.

**$\psi \ge 0$ is invariant.** At $\psi=0$, $\rho_s = \rho_\ell$ so $u_s=u_\ell$,
and $e^{-\psi}-1 = 0$, leaving

$$\left.\frac{d\psi}{dt}\right|_{\psi=0} = u_\ell\,(k_\ell - k_s) \;\ge\; 0
\qquad\text{because } k_\ell > k_s .$$

So a stem never ends up with less conducting area than its canopy needs, the
scaling factor $e^{-\psi/a_{pl2}}$ stays within $(0,1]$, and no over-replacement
can arise to cost unbounded carbon. **⚠️ If $k_s$ were ever raised above $k_\ell$,
this argument reverses.**

**Height never falls.** $dh/dt$ is a product of $(1-\sigma) \in (0,1]$, $F \ge 0$,
$f_g \ge 0$, and two positive derivatives.

**$a_{pl0}=0$ is the old model exactly.** $\rho_\ell = 1$ gives $u_\ell = 0$;
$\phi=0$ gives $\sigma = 0$ and hence $b=0$; $\psi=0$ gives $\rho_s = 1$ and hence
$u_s = 0$. Both departure rates are exactly zero, $e^0 = 1$ leaves the sizes
bitwise unchanged, and odelia's **max**-norm error control ignores a zero-error
state, so the solver's step sizes are untouched as well.

### One approximation, stated

Extension is charged at $\partial m_{\text{live}}/\partial A$ evaluated at
$A_\star$, but a thinned plant adds only $e^{\phi}\,\dot A_\star$ of actual leaf
area for that money. A plant off its trajectory is therefore **over-charged** for
extension, by a factor $e^{-\phi}$.

The direction is conservative — a thinned plant grows more slowly than a strict
budget would allow, never faster — and the error vanishes as the canopy recovers.
It is the price of keeping extension exactly departure-preserving, which is what
makes the resting model exact and confines all of the departure's motion to
plasticity.
