---
title: "TF24t leaf-level thermal response (ATLS): damage, and four strategies"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
---

```{r setup, message=FALSE, warning=FALSE}
# Staging demo: needs the in-tree dev build (ATLS, #566; builds on PM, #523), not
# necessarily the installed `plant`. Resolve the package root whether rendered
# from the repo root or from overstorey_staging/.
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

# Shared leaf-driving code (also exercised by tests/testthat/test-atls-leaf-demo.R,
# so this demo cannot silently rot).
source(file.path(pkg_root, "overstorey_staging", "atls_demo_helpers.R"))

theme_set(theme_bw(base_size = 12))
# One colour per strategy archetype, reused across the document.
strat_pal <- c("Generalist" = "#7570b3", "Tolerant" = "#d95f02",
               "Repairer" = "#1b9e77", "Acclimated" = "#e7298a")
```

The **ATLS** layer (leaf thermal damage / repair / acclimation, issue #566,
*merged revision*) gives a TF24 leaf a mechanistic response to heat. Its core is
the **deactivated fraction** `φ_d ∈ [0, 1]` — the transiently-unfolded share of
the machinery, read straight off the *same* Medlyn peaked-Arrhenius curve that
sets capacity. `φ_d` half-unfolds at an **emergent onset** `≈ H_d/ΔS ≈ 34.5 °C`
(no separate `T_crit` parameter). Unfolded protein leaks into **lasting damage**:
a recoverable pool `I_r` (resynthesised over ~4 days) and a permanent pool `I_p`
(cleared only as the plant grows new leaf). Both capacities carry the lasting
discount `(1 − I_r − I_p)`, so — unlike the memoryless as-built factor — a cooled
leaf does *not* rebound instantly.

A thermal *strategy* is a point in a four-axis trait space, all plugging into the
same physics:

| Axis | Knob | Moves |
|---|---|---|
| **Tolerance** (constitutive) | `topt_offset` | shifts the capacity optimum *and* the damage onset up together (one thermostability knob) |
| **Acclimation** (inducible) | `A` state | the same shift, induced when the regime is hot |
| **Repair** | preventive `k_i`, restorative `k_rec` | lower the leak into `I_r` / rebuild it faster (and cut the permanent floor) |
| **Avoidance** | transpirational cooling (PM) | lowers the operating leaf temperature, hence `φ_d` |

This demo drives a single well-watered leaf. It is a work-in-progress staging
document, not a built vignette. See `notes/atls/implementation-plan.md` for the
design and `notes/atls/gate-atls-vs-pm-evaluation.md` for the evaluation gate.

## 1. The deactivation curve `φ_d(Tleaf)` and the damage it drives

The generalist (default) leaf: `φ_d` sits near 0 while cold, passes through
**half-unfolding at the emergent onset ≈ 34.5 °C** (`H_d/ΔS`, marked), and
saturates toward 1 when hot — all from the *same* `K(T)` inside the peaked-
Arrhenius curve, with no separate `T_crit`. Leaf temperature is *prescribed*
(Fick path), so this is the pure mechanism with no optimiser in the loop.

The right-hand panel shows the standing recoverable damage this substrate would
drive at equilibrium, `I_r* = k_i·φ_d / (k_i·φ_d + k_rec_eff + k_mat)` (with
`I_p = 0`): the balance the `I_r` ODE relaxes toward at a held temperature. Fresh
leaves lose no capacity instantly — damage is a *memory* that fills this pool over
days, and each unit of `I_r`/`I_p` discounts **both** `vcmax` and `jmax`.

```{r damage-curve}
Tleaf <- seq(20, 50, by = 0.5)
phi_gen <- atls_phi_d_curve(Tleaf, atls_traits())
Ir_gen  <- atls_Ir_star_curve(Tleaf, atls_traits())
onset <- 200000 / 650 - 273.15   # H_d/dS in deg C ~ 34.5

p_phi <- ggplot(data.frame(Tleaf, phi = phi_gen), aes(Tleaf, phi)) +
  geom_line(colour = "#d95f02", linewidth = 0.9) +
  geom_vline(xintercept = onset, linetype = "dashed", colour = "grey50") +
  annotate("text", x = onset + 0.4, y = 0.05, label = "onset H_d/ΔS", hjust = 0,
           size = 3, colour = "grey40") +
  ylim(0, 1) +
  labs(x = "leaf temperature (°C)", y = "φ_d (unfolded fraction)",
       title = "Deactivation φ_d(Tleaf)")

p_Ir <- ggplot(data.frame(Tleaf, Ir = Ir_gen), aes(Tleaf, Ir)) +
  geom_line(colour = "#7570b3", linewidth = 0.9) +
  ylim(0, 1) +
  labs(x = "leaf temperature (°C)", y = "I_r* (quasi-steady)",
       title = "Standing recoverable damage")

p_phi | p_Ir
```

## 2. Four thermal-response strategies move the damage differently

Each archetype is a single-axis move off the generalist, shown as the standing
recoverable damage `I_r*(Tleaf)`. **Tolerance** and **acclimation** shift the
onset (and hence `I_r*`) to higher temperature; **repair** (higher `k_rec`, lower
`k_i`) lowers `I_r*` at a given temperature. (Avoidance is not a curve-shift — it
lowers the *operating* temperature, shown in §4.)

```{r strategy-curves}
strategies <- atls_strategies()
Ir_curves <- bind_rows(lapply(names(strategies), function(nm) {
  data.frame(strategy = nm, Tleaf = Tleaf,
             Ir = atls_Ir_star_curve(Tleaf, strategies[[nm]]))
}))

ggplot(Ir_curves, aes(Tleaf, Ir, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = strat_pal, name = NULL) +
  ylim(0, 1) +
  labs(x = "leaf temperature (°C)", y = "I_r* (quasi-steady recoverable damage)",
       title = "Standing damage I_r*(Tleaf) by thermal strategy",
       subtitle = "Tolerant/Acclimated shift the onset right; Repairer lowers the whole curve") +
  theme(legend.position = "top")
```

## 3. Who keeps carbon where? Assimilation across a heat gradient

The payoff, on fresh leaves (`I_r = I_p = 0`, so the assimilation panel isolates
the reversible-capacity effect). Tolerance and acclimation raise the capacity
optimum (`topt_offset`/`A`), buying carbon at hot leaf temperatures; the standing-
damage panel shows where lasting damage *would* accrue over days — repair (lower
`k_i`, higher `k_rec`) and tolerance both suppress `I_r*`, while the generalist
accrues the most.

```{r strategy-assim}
grad <- atls_solve_strategies(seq(24, 48, by = 2), strategies,
                              PAR = 1500, VPD = 2)

p_grad_A <- ggplot(grad, aes(Tenv, A, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = strat_pal, name = NULL) +
  labs(x = "leaf temperature (°C)", y = "assimilation (µmol C m⁻² s⁻¹)",
       title = "Carbon gain by strategy")

p_grad_Ir <- ggplot(grad, aes(Tenv, Ir_star, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = strat_pal, name = NULL) +
  labs(x = "leaf temperature (°C)", y = "I_r* at the operating point",
       title = "Standing damage by strategy")

(p_grad_Ir | p_grad_A) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

## 4. Avoidance: transpirational cooling as the fourth axis

Avoidance is not a damage trait — it is the leaf keeping itself cooler than the
air by transpiring, through the Penman–Monteith energy balance (#523). A
better-coupled leaf (smaller dimension `d`, higher conductance) runs closer to
air temperature, so on a hot day it sits at a lower `φ_d` and drives less lasting
damage — and because `φ_d` (and capacity) feed the carbon the psi-stem optimiser
sees, cooling is *self-selecting*. Here both leaves are the generalist; only the
coupling differs.

```{r avoidance}
avoid_cfgs <- list(
  "poorly coupled (large leaf, low g)" =
    atls_leaf_config(d = 0.10, leaf_specific_conductance_max = 2e-3),
  "well coupled (small leaf, high g)" =
    atls_leaf_config(d = 0.02, leaf_specific_conductance_max = 1e-2))

avoid <- bind_rows(lapply(names(avoid_cfgs), function(nm) {
  bind_rows(lapply(seq(24, 40, by = 0.5), function(Tair) {
    r <- atls_solve_cell(1500, Tair, 2, atls_traits(), pm = TRUE,
                         cfg = avoid_cfgs[[nm]])
    data.frame(coupling = nm, Tair = Tair, Tleaf = r$Tleaf,
               phi_d = r$phi_d, A = r$A)
  }))
}))

avoid_pal <- c("poorly coupled (large leaf, low g)" = "#b2182b",
               "well coupled (small leaf, high g)" = "#2166ac")

p_av_T <- ggplot(avoid, aes(Tair, Tleaf, colour = coupling)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey60") +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = avoid_pal, name = NULL) +
  labs(x = "air temperature (°C)", y = "operating leaf temperature (°C)",
       title = "Cooling by coupling", subtitle = "dotted = Tleaf = Tair")

p_av_phi <- ggplot(avoid, aes(Tair, phi_d, colour = coupling)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = avoid_pal, name = NULL) +
  labs(x = "air temperature (°C)", y = "φ_d at the operating point",
       title = "... lowers the deactivated fraction")

(p_av_T | p_av_phi) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

> Note the minimal-cut PM overheats an isolated leaf well above air (up to ~20 °C
> under high radiation, #523 caveats), so the *absolute* temperatures here are an
> upper bound; the **contrast** between couplings is the robust signal.

## 5. From one leaf to a community: persistence under warming (SCM)

Everything above is a single isolated leaf. The real question is demographic:
does the damage feedback change whether a *population* persists? Here we run the
full size-structured community model (SCM) — a single-species patch to
equilibrium under a constant climate (the `air_temp` driver is the midday air
temperature) — and read the net reproduction ratio **R0** (a resident persists
iff R0 ≥ 1) across a warming gradient, TF24t (PM + ATLS) vs a PM-only TF24.

Each SCM run takes ~30 s, so the grid is deliberately small and the chunk is
cached.

```{r scm-climate, cache=TRUE}
climate <- atls_scm_climate(seq(24, 30, by = 1))
lab_type <- c(TF24 = "PM only", TF24t = "PM + ATLS")
climate$model <- lab_type[climate$type]

ggplot(climate, aes(air_temp, R0, colour = model)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.9) + geom_point(size = 2) +
  annotate("text", x = 24, y = 1.04, label = "persistence (R0 = 1)",
           hjust = 0, size = 3, colour = "grey40") +
  scale_colour_manual(values = c("PM only" = "#1b9e77",
                                 "PM + ATLS" = "#d95f02"), name = NULL) +
  labs(x = "midday climate temperature (°C)", y = "net reproduction ratio R0",
       title = "Population persistence collapses as the climate warms",
       subtitle = "single-species SCM to equilibrium; R0 ≥ 1 required to persist") +
  theme(legend.position = "top")
```

Two things stand out. First, persistence collapses steeply between ~26 and 28 °C
for **both** models — the constant-climate community is fragile, and (as at the
leaf level) that collapse is driven by the minimal-cut PM overheating the canopy,
not by ATLS. Second, within the persistence band ATLS is **not** a pure penalty:
TF24t actually clears R0 = 1 at a slightly *hotter* climate than PM-only, because
its acclimation raises the photosynthetic optimum (a capacity *gain*) while
leaves are not yet hot enough for the damage term to bite hard. Damage and
acclimation are two sides of the same layer; which wins depends on the climate.

## 6. Which thermal strategy persists hottest? (SCM tournament)

Now hold the climate at the warm edge and vary the *strategy* — each a heritable
single-axis move off the generalist (tolerance = constitutive `topt_offset`;
repair = higher restorative `k_rec` + lower preventive-leak `k_i`; acclimator =
faster acclimation kinetics). Higher R0 = more able to persist.

```{r scm-tournament, cache=TRUE}
tourney <- atls_scm_tournament(c(26, 28))
tourney$strategy <- factor(tourney$strategy,
  levels = c("Generalist", "Tolerant", "Repairer", "Acclimator"))

ggplot(tourney, aes(strategy, R0, fill = strategy)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ sprintf("climate = %d °C", air_temp)) +
  scale_fill_manual(values = strat_pal, guide = "none") +
  labs(x = NULL, y = "net reproduction ratio R0",
       title = "Tolerance persists hottest; repair does not pay here",
       subtitle = "TF24t residents at the warm edge; dashed = persistence R0 = 1") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

Constitutive **tolerance** persists hottest — it raises the capacity optimum and
the damage onset together and clears R0 = 1 at a warmer climate than the
generalist. Dynamic **acclimation** buys a similar shift inducibly. **Repair**
must earn back its standing maintenance (paid for resynthesis capacity `k_rec` and
protection `k_i`, whether or not damage is present), so whether it pays depends on
how hot the climate runs. That cost/benefit tension is exactly what the merged
cost structure was built to expose. (Runs are cached; exact rankings depend on the
calibration-target rate/cost defaults.)

## 7. Memory: the running-mean leaf temperature

The sections above all read the *instantaneous* operating leaf temperature. TF24t
also carries a diagnostic ODE state, **`mean_leaf_temp`**, an exponential moving
average of that operating-point `Tleaf`:

$$
\frac{\mathrm{d}\,\overline{T_\ell}}{\mathrm{d}t}
  = D\,k_{\overline{T}}\,\bigl(T_\mathrm{op} - \overline{T_\ell}\bigr),
$$

where `T_op` is the PM operating-point leaf temperature (published as the
`operating_leaf_temp` aux), `k_mean_leaf_temp` is the memory decay rate (day⁻¹,
`D = 365.25` puts it on the yearly ODE clock), and the time constant is
`1/k_mean_leaf_temp`. It is **purely diagnostic** — it does not feed back into
damage or acclimation — so it is a clean read-out of the temperature history a
plant has actually experienced. Below, a single TF24t individual is driven with a
sub-seasonally oscillating `air_temp`; the running mean lags and smooths the
operating temperature (memory ≈ 20 d here).

```{r running-mean, cache=TRUE, message=FALSE, warning=FALSE}
e <- "TF24_Env"
k_mem  <- 0.05                 # day^-1 -> ~20-day memory (1 / (k*365.25) yr)
period <- 0.12                 # yr (~44 d) forcing, comparable to the memory
t_end  <- 0.6                  # yr
t_grid <- seq(0, t_end, length.out = 200)
Tair   <- 27 + 6 * sin(2 * pi * t_grid / period)

s <- TF24t_Strategy(); s$k_mean_leaf_temp <- k_mem
p <- Individual("TF24t", e)(s)
env <- Environment("TF24t")
env$extrinsic_drivers_set_variable("air_temp", t_grid, Tair)

runner <- OdeRunner("TF24t")(IndividualRunner("TF24t", e)(p, env))
rec <- list()
while (runner$time < t_end) {
  runner$step()
  ind <- runner$object$individual
  rec[[length(rec) + 1L]] <- data.frame(
    time      = runner$time,
    air_temp  = env$extrinsic_drivers_evaluate("air_temp", min(runner$time, t_end)),
    operating = ind$aux("operating_leaf_temp"),
    running   = ind$state("mean_leaf_temp"))
}
traj <- do.call(rbind, rec) |>
  pivot_longer(c(air_temp, operating, running), names_to = "series",
               values_to = "temp")

series_lab <- c(air_temp = "air temperature (driver)",
                operating = "operating leaf temp (instantaneous)",
                running = "mean_leaf_temp (running mean)")
series_pal <- c(air_temp = "grey60", operating = "#d95f02", running = "#1b9e77")

ggplot(traj, aes(time, temp, colour = series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = series_pal, labels = series_lab, name = NULL) +
  labs(x = "time (years)", y = "temperature (°C)",
       title = "mean_leaf_temp lags and smooths the operating leaf temperature",
       subtitle = "single TF24t individual, oscillating air_temp; memory ≈ 20 d") +
  theme(legend.position = "top")
```

The running mean (green) tracks the operating leaf temperature (orange) with a
lag set by `k_mean_leaf_temp` and a reduced amplitude — the "average growing
temperature" the plant carries. The operating temperature sits well above air
because the minimal-cut PM overheats an isolated leaf (§4 caveat); `mean_leaf_temp`
averages that operating signal, not the air-temperature driver.

## 8. Summary

The deactivation curve `φ_d(Tleaf)` is the single substrate: strategies win by
moving *where it turns on* (tolerance, acclimation shift the emergent onset),
*how much unfolds leaks into lasting damage and how fast it is rebuilt* (preventive
`k_i`, restorative `k_rec`), or *where the leaf actually operates* (avoidance).
Damage is now a **memory** — the pools `I_r`/`I_p` discount both capacities and
persist after cooling — so the same insult that a memoryless factor would forget is
carried forward, and permanent damage clears only as the plant grows new leaf. All
plug into the one Medlyn deactivation curve and the Penman–Monteith energy balance.
At the community scale the same layer sets *where a population can persist* under
warming, and *which strategy* persists hottest is a genuine trade-off. See
`notes/atls/gate-atls-vs-pm-evaluation.md` for the whole-plant materiality gate
(verdict: keep) and the caveats (notably the minimal-cut PM overheating, which
makes the constant-climate collapse temperatures an upper bound).
