---
title: "TF24t leaf-level thermal response (ATLS): damage, and four strategies"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
---

```{r setup, message=FALSE, warning=FALSE}
# Staging demo: needs the in-tree dev build (ATLS, #566; builds on PM, #523), not
# necessarily the installed `plant`. Resolve the package root whether rendered
# from the repo root or from overstorey_staging/.
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

# Shared leaf-driving code (also exercised by tests/testthat/test-atls-leaf-demo.R,
# so this demo cannot silently rot).
source(file.path(pkg_root, "overstorey_staging", "atls_demo_helpers.R"))

theme_set(theme_bw(base_size = 12))
# One colour per strategy archetype, reused across the document.
strat_pal <- c("Generalist" = "#7570b3", "Tolerant" = "#d95f02",
               "Repairer" = "#1b9e77", "Acclimated" = "#e7298a")
```

The **ATLS** layer (Lumry–Eyring leaf thermal damage / repair / acclimation,
issue #566) gives a TF24 leaf a mechanistic response to heat. At its core is a
single quasi-steady **functional fraction** `N ∈ (0, 1]` — the share of the
electron-transport machinery still working — which downscales `jmax → jmax·N`.
`N` is a smooth turn-off centred just above the critical temperature `T_crit`:
cold leaves are undamaged (`N ≈ 1`); hot leaves lose capacity (`N → 0`).

A thermal *strategy* is a point in a four-axis trait space, all plugging into the
same physics:

| Axis | Knob | Moves |
|---|---|---|
| **Tolerance** (constitutive) | `tcrit_0`, `topt_offset` | shifts the damage curve and the capacity optimum to higher T |
| **Acclimation** (inducible) | `A_crit`, `A_opt` states | *raises* `T_crit`/`T_opt` when the regime is hot |
| **Repair** | `k_r1_0` (refold capacity) | lifts `N` at a given hot T (starves the damage pool) |
| **Avoidance** | transpirational cooling (PM) | lowers the operating leaf temperature itself |

This demo drives a single well-watered leaf. It is a work-in-progress staging
document, not a built vignette. See `notes/atls/implementation-plan.md` for the
design and `notes/atls/gate-atls-vs-pm-evaluation.md` for the evaluation gate.

## 1. The damage curve `N(Tleaf)` and its carbon cost

The generalist (ATLS-default) leaf: `N` holds at 1 until ~34 °C, then falls
sharply through `T_crit = 38 °C`. Because `jmax` is scaled by `N`, assimilation
follows it down. Here the leaf temperature is *prescribed* (Fick path), so this
is the pure mechanism with no optimiser in the loop.

In the right-hand panel, **damage off** vs **on** is the *same* leaf at the same
temperature, differing only in whether the damage kinetics run: "off" zeroes the
unfolding rate (`k_d1_0 = 0`), forcing `N = k_r/(k_r + k_d) = 1` at every
temperature (no `jmax` penalty); "on" uses the ATLS defaults, so `N < 1` once hot.
The gap between the curves is the carbon lost to thermal damage.

```{r damage-curve}
Tleaf <- seq(20, 50, by = 0.5)
N_gen <- atls_N_curve(Tleaf, atls_traits())

# Assimilation with the damage layer OFF vs ON, prescribed Tleaf, solved optimum.
gradA <- bind_rows(lapply(seq(20, 50, by = 2), function(T) {
  off <- atls_solve_cell(1500, T, 2, atls_traits(k_d1_0 = 0))  # k_d=0 -> N==1
  on  <- atls_solve_cell(1500, T, 2, atls_traits())
  data.frame(Tleaf = T, `damage off` = off$A, `damage on` = on$A,
             check.names = FALSE)
})) |>
  pivot_longer(-Tleaf, names_to = "layer", values_to = "A")

p_N <- ggplot(data.frame(Tleaf, N = N_gen), aes(Tleaf, N)) +
  geom_line(colour = "#d95f02", linewidth = 0.9) +
  geom_vline(xintercept = 38, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 38.4, y = 0.05, label = "T_crit", hjust = 0,
           size = 3, colour = "grey40") +
  ylim(0, 1) +
  labs(x = "leaf temperature (°C)", y = "N (functional fraction)",
       title = "Damage factor N(Tleaf)")

p_A <- ggplot(gradA, aes(Tleaf, A, colour = layer)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("damage off" = "grey55",
                                 "damage on" = "#d95f02"), name = NULL) +
  labs(x = "leaf temperature (°C)", y = "assimilation (µmol C m⁻² s⁻¹)",
       title = "Carbon cost of damage") +
  theme(legend.position = c(0.75, 0.85))

p_N | p_A
```

## 2. Four thermal-response strategies move the curve differently

Each archetype is a single-axis move off the generalist. **Tolerance** and
**acclimation** slide the whole turn-off to higher temperature; **repair** lifts
the floor without moving the threshold. (Avoidance is not a curve-shift — it
lowers the *operating* temperature, shown in §4.)

```{r strategy-curves}
strategies <- atls_strategies()
Ncurves <- bind_rows(lapply(names(strategies), function(nm) {
  data.frame(strategy = nm, Tleaf = Tleaf,
             N = atls_N_curve(Tleaf, strategies[[nm]]))
}))

ggplot(Ncurves, aes(Tleaf, N, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 38, linetype = "dashed", colour = "grey70") +
  scale_colour_manual(values = strat_pal, name = NULL) +
  ylim(0, 1) +
  labs(x = "leaf temperature (°C)", y = "N (functional fraction)",
       title = "N(Tleaf) by thermal strategy",
       subtitle = "Tolerant/Acclimated shift T_crit right; Repairer lifts the hot-end floor") +
  theme(legend.position = "top")
```

## 3. Who keeps carbon where? Assimilation across a heat gradient

The payoff. Solving the optimum at each prescribed leaf temperature, tolerance
and acclimation buy the most carbon in the 36–44 °C band (they also lift the cool
end, because `topt_offset`/`A_opt` raise the capacity optimum); repair buffers the
extreme hot end; the generalist collapses first.

```{r strategy-assim}
grad <- atls_solve_strategies(seq(24, 48, by = 2), strategies,
                              PAR = 1500, VPD = 2)

p_grad_A <- ggplot(grad, aes(Tenv, A, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = strat_pal, name = NULL) +
  labs(x = "leaf temperature (°C)", y = "assimilation (µmol C m⁻² s⁻¹)",
       title = "Carbon gain by strategy")

p_grad_N <- ggplot(grad, aes(Tenv, N, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = strat_pal, name = NULL) +
  labs(x = "leaf temperature (°C)", y = "N at the operating point",
       title = "Retained capacity by strategy")

(p_grad_N | p_grad_A) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

## 4. Avoidance: transpirational cooling as the fourth axis

Avoidance is not a damage trait — it is the leaf keeping itself cooler than the
air by transpiring, through the Penman–Monteith energy balance (#523). A
better-coupled leaf (smaller dimension `d`, higher conductance) runs closer to
air temperature, so on a hot day it sits at a higher `N` and keeps assimilating —
and because `N` feeds the carbon the psi-stem optimiser sees, cooling is
*self-selecting*. Here both leaves are the generalist; only the coupling differs.

```{r avoidance}
avoid_cfgs <- list(
  "poorly coupled (large leaf, low g)" =
    atls_leaf_config(d = 0.10, leaf_specific_conductance_max = 2e-3),
  "well coupled (small leaf, high g)" =
    atls_leaf_config(d = 0.02, leaf_specific_conductance_max = 1e-2))

avoid <- bind_rows(lapply(names(avoid_cfgs), function(nm) {
  bind_rows(lapply(seq(24, 40, by = 0.5), function(Tair) {
    r <- atls_solve_cell(1500, Tair, 2, atls_traits(), pm = TRUE,
                         cfg = avoid_cfgs[[nm]])
    data.frame(coupling = nm, Tair = Tair, Tleaf = r$Tleaf, N = r$N, A = r$A)
  }))
}))

avoid_pal <- c("poorly coupled (large leaf, low g)" = "#b2182b",
               "well coupled (small leaf, high g)" = "#2166ac")

p_av_T <- ggplot(avoid, aes(Tair, Tleaf, colour = coupling)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey60") +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = avoid_pal, name = NULL) +
  labs(x = "air temperature (°C)", y = "operating leaf temperature (°C)",
       title = "Cooling by coupling", subtitle = "dotted = Tleaf = Tair")

p_av_N <- ggplot(avoid, aes(Tair, N, colour = coupling)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = avoid_pal, name = NULL) +
  labs(x = "air temperature (°C)", y = "N at the operating point",
       title = "... buys retained capacity")

(p_av_T | p_av_N) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

> Note the minimal-cut PM overheats an isolated leaf well above air (up to ~20 °C
> under high radiation, #523 caveats), so the *absolute* temperatures here are an
> upper bound; the **contrast** between couplings is the robust signal.

## 5. From one leaf to a community: persistence under warming (SCM)

Everything above is a single isolated leaf. The real question is demographic:
does the damage feedback change whether a *population* persists? Here we run the
full size-structured community model (SCM) — a single-species patch to
equilibrium under a constant climate (the `leaf_temp` driver is the midday air
temperature) — and read the net reproduction ratio **R0** (a resident persists
iff R0 ≥ 1) across a warming gradient, TF24t (PM + ATLS) vs a PM-only TF24.

Each SCM run takes ~30 s, so the grid is deliberately small and the chunk is
cached.

```{r scm-climate, cache=TRUE}
climate <- atls_scm_climate(seq(24, 30, by = 1))
lab_type <- c(TF24 = "PM only", TF24t = "PM + ATLS")
climate$model <- lab_type[climate$type]

ggplot(climate, aes(leaf_temp, R0, colour = model)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.9) + geom_point(size = 2) +
  annotate("text", x = 24, y = 1.04, label = "persistence (R0 = 1)",
           hjust = 0, size = 3, colour = "grey40") +
  scale_colour_manual(values = c("PM only" = "#1b9e77",
                                 "PM + ATLS" = "#d95f02"), name = NULL) +
  labs(x = "midday climate temperature (°C)", y = "net reproduction ratio R0",
       title = "Population persistence collapses as the climate warms",
       subtitle = "single-species SCM to equilibrium; R0 ≥ 1 required to persist") +
  theme(legend.position = "top")
```

Two things stand out. First, persistence collapses steeply between ~26 and 28 °C
for **both** models — the constant-climate community is fragile, and (as at the
leaf level) that collapse is driven by the minimal-cut PM overheating the canopy,
not by ATLS. Second, within the persistence band ATLS is **not** a pure penalty:
TF24t actually clears R0 = 1 at a slightly *hotter* climate than PM-only, because
its acclimation raises the photosynthetic optimum (a capacity *gain*) while
leaves are not yet hot enough for the damage term to bite hard. Damage and
acclimation are two sides of the same layer; which wins depends on the climate.

## 6. Which thermal strategy persists hottest? (SCM tournament)

Now hold the climate at the warm edge and vary the *strategy* — each a heritable
single-axis move off the generalist (tolerance = constitutive `tcrit_0` +
`topt_offset`; repair = higher `k_r1_0`; acclimator = faster acclimation
kinetics). Higher R0 = more able to persist.

```{r scm-tournament, cache=TRUE}
tourney <- atls_scm_tournament(c(26, 28))
tourney$strategy <- factor(tourney$strategy,
  levels = c("Generalist", "Tolerant", "Repairer", "Acclimator"))

ggplot(tourney, aes(strategy, R0, fill = strategy)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ sprintf("climate = %d °C", leaf_temp)) +
  scale_fill_manual(values = strat_pal, guide = "none") +
  labs(x = NULL, y = "net reproduction ratio R0",
       title = "Tolerance persists hottest; repair does not pay here",
       subtitle = "TF24t residents at the warm edge; dashed = persistence R0 = 1") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

The ranking is consistent across the warm edge: **Tolerant > Acclimator >
Generalist > Repairer**. Constitutive tolerance wins decisively (at 26 °C it lifts
R0 from ~1.0 to ~1.8 and is the only strategy still viable at 28 °C); dynamic
acclimation is a solid second; and **repair actually underperforms the
generalist** here — its standing maintenance cost (`∝ k_r1_0`, paid even when
cool) is not repaid at these moderate temperatures. That cost/benefit reversal is
exactly the kind of trade-off the cost structure (Phase 3) was built to expose.

## 7. Summary

`N(Tleaf)` is the single lever: strategies win carbon by moving *where* it turns
off (tolerance, acclimation), *how high the floor sits* when hot (repair), or
*where the leaf actually operates* (avoidance). All four plug into the same
Lumry–Eyring kinetics and Penman–Monteith energy balance. At the community scale
the same layer sets *where a population can persist* under warming, and — because
tolerance and acclimation carry benefits while repair and construction carry
costs — *which strategy* persists hottest is a genuine trade-off, not a given.
See `notes/atls/gate-atls-vs-pm-evaluation.md` for the whole-plant materiality
gate (verdict: keep) and the caveats (notably the minimal-cut PM overheating,
which makes the constant-climate collapse temperatures an upper bound).
