---
title: "TF24 leaf-level Penman-Monteith energy balance: Fick vs PM"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
---

```{r setup, message=FALSE, warning=FALSE}
# Staging demo: needs the in-tree dev build (PM, #523), not necessarily the
# installed `plant`. Resolve the package root whether rendered from the repo root
# or from overstorey_staging/.
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

# Shared leaf-driving code (also exercised by tests/testthat/test-pm-leaf-demo.R,
# so this demo cannot silently rot).
source(file.path(pkg_root, "overstorey_staging", "pm_demo_helpers.R"))

theme_set(theme_bw(base_size = 12))
pal <- c("Fick (Tleaf = Tair)" = "#1b9e77", "Penman-Monteith" = "#d95f02")
lab_pm <- function(pm) ifelse(pm, "Penman-Monteith", "Fick (Tleaf = Tair)")
```

This document demonstrates the TF24 leaf submodel with the Penman-Monteith leaf
energy balance (`use_energy_balance_`, issue #523) turned **on** vs the default
**off** (Fick's law, `Tleaf = Tair`). It drives a single well-watered leaf across
an environment grid; it is a work-in-progress staging document, not a built
vignette. See `notes/penman-monteith/` for the plan and the Step 5 evaluation.

## 1. Profit anatomy at a single hot environment

The leaf chooses its **root-collar** water potential to maximise carbon
**profit** = assimilation benefit − hydraulic cost; `find_root_collar_psi()`
searches over this variable directly, not ψ_stem — ψ_stem is *derived* from
each candidate collar potential via the stem's hydraulic transport
(`find_psi_stem_from_psi_root`). This scan drives the leaf the same way, via
`evaluate_root_collar_psi()`, so the dashed real-solver optimum lands on the
visible curve. Under PM, the assimilation curve bends because a hotter leaf
shifts the Farquhar temperature response. Here at a hot, high-radiation,
high-VPD environment (Tair = 38 °C, PAR = 1800, VPD = 2.5 kPa).

```{r profit-anatomy}
env1 <- list(PAR = 1800, Tair = 38, VPD = 2.5)
root_psi_seq <- seq(0.31, 5.5, length.out = 120)

collar_curve <- bind_rows(
  pm_collar_curve(env1$PAR, env1$Tair, env1$VPD, FALSE, root_psi_seq),
  pm_collar_curve(env1$PAR, env1$Tair, env1$VPD, TRUE,  root_psi_seq)
) |>
  mutate(model = lab_pm(pm)) |>
  filter(is.finite(profit))

# Optima from the actual solver, on the same (positive-magnitude) root-collar
# potential axis as root_psi_seq.
opt <- bind_rows(
  pm_solve_cell(env1$PAR, env1$Tair, env1$VPD, FALSE),
  pm_solve_cell(env1$PAR, env1$Tair, env1$VPD, TRUE)
) |>
  # opt_root_psi is already the positive magnitude the plot's axis wants
  # (phylloptim #25), so there is no negation here any more.
  mutate(model = lab_pm(pm), root_psi = opt_root_psi)

collar_curve_long <- collar_curve |>
  pivot_longer(c(profit, assim, hydraulic_cost, leaf_temp),
               names_to = "component", values_to = "value") |>
  mutate(component = recode(component,
    profit = "net profit", assim = "assimilation benefit",
    hydraulic_cost = "hydraulic cost"))

ggplot(collar_curve_long, aes(root_psi, value, colour = model)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = opt, aes(xintercept = root_psi, colour = model),
             linetype = "dashed") +
  facet_wrap(~component, scales = "free_y") +
  scale_colour_manual(values = pal, name = NULL) +
  labs(x = expression(psi[rc]~"(-MPa, root collar)"), y = "µmol C m⁻² s⁻¹",
       title = "Profit anatomy: Fick vs Penman-Monteith",
       subtitle = sprintf("Tair = %g °C, PAR = %g, VPD = %g kPa; dashed = optimum",
                          env1$Tair, env1$PAR, env1$VPD)) +
  theme(legend.position = "top")
```

## 2. Optimal outcomes across environments

Sweep Tair × PAR at a fixed VPD, solving the optimum with Fick and with PM, and
compare leaf temperature, the optimum stem potential, stomatal conductance,
transpiration, assimilation, and profit.

```{r sweep}
grid <- expand.grid(PAR = c(400, 900, 1500, 2000),
                    Tair = c(15, 25, 33, 40),
                    VPD = 2.0)
sweep <- pm_solve_grid(grid) |> mutate(model = lab_pm(pm))

sweep_w <- sweep |>
  select(PAR, Tair, model, Tleaf, opt_psi_stem, gs, E, A, profit) |>
  mutate(dTleaf = Tleaf - Tair)

p_tleaf <- ggplot(subset(sweep_w, model == "Penman-Monteith"),
                  aes(PAR, dTleaf, colour = factor(Tair))) +
  geom_line() + geom_point() +
  labs(x = "PAR (µmol m⁻² s⁻¹)", y = "Tleaf − Tair (°C)",
       colour = "Tair (°C)", title = "PM leaf warming")

metric_plot <- function(var, ylab) {
  ggplot(sweep_w, aes(PAR, .data[[var]], colour = model,
                      linetype = factor(Tair))) +
    geom_line() +
    scale_colour_manual(values = pal, name = NULL) +
    labs(x = "PAR", y = ylab, linetype = "Tair (°C)")
}

(p_tleaf | metric_plot("A", "assimilation (µmol C m⁻² s⁻¹)")) /
  (metric_plot("gs", "gs (mol CO₂ m⁻² s⁻¹)") | metric_plot("profit", "profit")) +
  plot_annotation(title = "Optimal leaf outcomes across environments (VPD = 2 kPa)")
```

## 3. Where PM matters

The correction is negligible in cool/low-light conditions and large in the hot,
high-radiation corner — where the overheated leaf loses most of its carbon gain.

```{r where}
grid2 <- expand.grid(PAR = c(200, 500, 1000, 2000),
                     Tair = c(15, 25, 35, 40), VPD = 2.0)
both <- pm_solve_grid(grid2)
wide <- both |>
  select(PAR, Tair, pm, A, Tleaf) |>
  tidyr::pivot_wider(names_from = pm, values_from = c(A, Tleaf),
                     names_glue = "{.value}_{ifelse(pm, 'pm', 'fick')}") |>
  mutate(dTleaf = Tleaf_pm - Tair,
         relA = ifelse(A_fick > 1, 100 * (A_pm - A_fick) / A_fick, NA_real_))

ggplot(wide, aes(factor(PAR), factor(Tair), fill = relA)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = ifelse(is.na(relA), "·", sprintf("%.0f", relA))), size = 3) +
  scale_fill_gradient2(low = "#b2182b", mid = "#f7f7f7", high = "#2166ac",
                       midpoint = 0, name = "ΔA %") +
  labs(x = "PAR (µmol m⁻² s⁻¹)", y = "Tair (°C)",
       title = "Relative change in assimilation, PM vs Fick",
       subtitle = "VPD = 2 kPa; blue = PM raises A, red = PM lowers A, · = A≈0")
```

See `notes/penman-monteith/step5-fick-vs-pm-evaluation.md` for the full factorial
gate and verdict.
