---
title: "One sweep, forty-eight trade-offs: reading the iso-biomass surface"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5.5
---

```{r setup, message=FALSE, warning=FALSE}
# Staging demo: needs the in-tree dev build, not necessarily the installed
# `plant`. Resolves whether rendered from the repo root or overstorey_staging/.
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(odelia)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

# Shared driving and interpretation code. Nothing exercises it: the smoke test
# that did was dropped from this branch, so a plant interface change rots these
# helpers and this document is where it surfaces.
source(file.path(pkg_root, "overstorey_staging", "gradient_demo_helpers.R"))

theme_set(theme_bw(base_size = 12))
LIFETIME <- 30
METRIC <- "mass_above_ground"
```

## What one reverse sweep buys

A stand gradient is one number per parameter: how much the stand's above-ground
mass moves when that parameter moves. One reverse sweep produces every one of
them, and what it costs is set by the length of the forward run rather than by
how many parameters there are. Finite differences would want another forward run
per parameter.

```{r one-point}
tr0 <- unlist(gd_traits())
pt <- gd_point(tr0, lifetime = LIFETIME)   # refine, run, sweep

tibble::tibble(quantity = names(pt$value), value = as.numeric(pt$value)) |>
  knitr::kable(digits = 4, caption = "The census at this stand.")
```

```{r shape}
# Every count in the prose below reads from P rather than repeating a literal.
# A literal disagrees with the sweep silently: nothing here would fail if TF24
# gained a parameter.
P <- ncol(pt$gradient)
cat(sprintf("gradient: %d metrics x %d parameters, %d refused, %d ODE steps\n",
            nrow(pt$gradient), P, sum(pt$refused), pt$steps))
```

The point of this document is what that vector *is*, geometrically. It is the
normal to the iso-biomass surface. So one sweep does not just rank the
parameters — it hands over the entire `r P - 1`-dimensional space of parameter
changes that leave the stand's biomass unchanged, and any interpretable pair can
be read straight out of it.

## Raw partials are not comparable, and the reason is units

The first thing to do with a gradient is *not* to rank it. A partial derivative
carries the units of its parameter, so the largest entry is usually the one whose
parameter has the smallest units.

```{r units, fig.height=4.6}
el  <- gd_elasticity(pt$gradient, pt$value, pt$x)
raw <- sort(abs(pt$gradient[METRIC, ]), decreasing = TRUE)[1:12]
ela <- sort(abs(el[METRIC, ]), decreasing = TRUE)[1:12]

# Two plots rather than two facets: each panel has to be ordered by its OWN
# values, and reorder() inside a facet orders both by the same ones. The raw
# panel is log-scaled because it spans four decades; a bar on a log axis starts
# at the axis floor rather than at zero, so the elasticity panel, which spans
# less than two decades, stays linear where a bar means what it looks like.
bars <- function(v, title, sub, logx) {
  d <- tibble::tibble(name = sub("^1\\.", "", names(v)), v = as.numeric(v))
  p <- ggplot(d, aes(stats::reorder(name, v), v)) +
    geom_col(fill = "#3b6ea5") + coord_flip() +
    labs(x = NULL, y = NULL, title = title, subtitle = sub)
  if (logx) p + scale_y_continuous(trans = "log10", labels = scales::comma) else p
}

bars(raw, "Raw partial  |dy/dx|", "ranked by the size of the units", TRUE) +
  bars(ela, "Elasticity  |(x/y) dy/dx|", "ranked by influence", FALSE) +
  patchwork::plot_annotation(
    title = "The same gradient, ranked two ways",
    subtitle = "The left ordering is an artefact; only the right one is about the stand.")
```

The two orderings disagree, and the left one is an artefact. `omega` and `theta`
lead it by three orders of magnitude because they are small numbers, not because
the stand cares about them. Everything below is in elasticities — percentage
change in the metric per percentage change in the parameter — which is
dimensionless and therefore comparable across columns, and which is also the form
a trait-database correlation is quoted in.

## The trade-offs

Holding the metric fixed, the elasticity ratio gives the percentage change in one
parameter that offsets a one-percent rise in another. This is `-e_i / e_j`, and
it is available for every one of the `r P` × `r P` ordered pairs from the single
sweep above.

```{r tradeoffs}
tro <- gd_tradeoff(el, METRIC)
interp <- c("lma", "rho", "hmat", "a_l1", "k_I", "vcmax_25", "jmax_25", "K_s",
            "a_bio", "r_l")
have <- intersect(paste0("1.", interp), colnames(tro))
sub <- tro[have, have]
dimnames(sub) <- list(sub("^1\\.", "", have), sub("^1\\.", "", have))

as.data.frame(as.table(sub)) |>
  setNames(c("rises", "offset_by", "pct")) |>
  filter(is.finite(pct)) |>
  ggplot(aes(offset_by, rises, fill = pct)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", pct)), size = 3) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac",
                       midpoint = 0, limits = c(-3, 3), oob = scales::squish) +
  labs(title = "Compensating changes that hold above-ground mass fixed",
       subtitle = paste("Cell = % change in the row trait offsetting a 1% rise",
                        "in the column trait; colour saturates beyond +-3, so",
                        "read the numbers"),
       x = "a 1% rise in…", y = "…is offset by this % change in", fill = "%") +
  theme(panel.grid = element_blank())
```

A cell reads: *a one percent rise in the column parameter is offset by this many
percent in the row parameter.* A negative entry means the two push the same way,
so compensating means moving one back.

## The hyperpar wedge, which changes the answer

The columns above are **partial** derivatives: each holds all other parameters
fixed. That is not what an ecologist means by "the LMA derivative", because
`TF24_hyperpar` derives further parameters from the traits — `k_l`, `r_l` and
`nmass_l` from `lma`; `d_I`, `g1_TF24`, `k_s`, `r_s`, `r_b` from `rho`; `a_f3`
from `omega`; and `c`, `p_50`, `b`, `psi_crit` from `K_s`. Raising LMA in the
field raises those with it.

The total derivative chains the gradient through that map. The Jacobian is taken
by differencing `TF24_hyperpar` itself rather than restating its algebra, so
there is one source of truth for the derived forms.

```{r chain}
J <- gd_hyperpar_jacobian(tr0)
gt <- gd_trait_gradient(pt$gradient, J)
te <- gd_trait_elasticity(gt, pt$value, tr0)

moved <- rownames(J)[apply(J, 1, function(r) any(r != 0))]
cat("parameters that move when a trait moves:", paste(moved, collapse = ", "), "\n")

partial <- el[METRIC, paste0("1.", names(tr0))]
tibble::tibble(trait = names(tr0),
               partial = as.numeric(partial),
               total = as.numeric(te[METRIC, names(tr0)])) |>
  mutate(ratio = total / partial) |>
  knitr::kable(digits = 4,
               caption = "Elasticity of above-ground mass: holding the derived parameters fixed, against carrying them along.")
```

Only `lma` has derived children in this trait set, so it is the only row where
the two columns differ — which is the control: the other four traits chain
through the identity and must come out unchanged.

## What this costs, and why reverse mode is the reason it is possible

```{r cost, fig.height=4}
# Counted, not timed. The axis is forward stand solves -- what each method has
# to run -- because a seconds axis is a figure from whenever it was last taken.
# Reverse mode replaces the extra solves with one reverse pass over the steps
# the first solve already recorded.
sp <- 1:3
tibble::tibble(species = rep(sp, 2),
               solves = c(rep(1L, length(sp)), 1L + P * sp),
               method = rep(c("one reverse sweep", "finite differences"),
                            each = length(sp))) |>
  ggplot(aes(species, solves, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  scale_y_continuous(trans = "log10", labels = scales::comma) +
  scale_x_continuous(breaks = sp) +
  scale_colour_manual(values = c("finite differences" = "#b2182b",
                                 "one reverse sweep" = "#2166ac")) +
  labs(title = "The whole gradient, against differencing for it",
       subtitle = sprintf("%d parameters per species, on a %d-year stand",
                          P, LIFETIME),
       x = "species in the stand", y = "forward stand solves (log scale)",
       colour = NULL)
```

The count does not move with the number of parameters: three species means
`r 3 * P` of them and still one forward solve, while differencing wants a solve
per parameter on top of it. That is the whole argument, and it gets better as the
community grows.

## What to be careful of

**A near-zero denominator is not a trade-off.** If a parameter barely moves the
metric, the change in it that offsets anything is unbounded; `gd_tradeoff()`
returns `NA` there rather than a large number that looks like an answer.

**The tangent plane is local.** These ratios hold for infinitesimal changes. A
10% move is outside where a linearisation is trustworthy, and the honest way to
check one is to take a second gradient at the moved point and see whether it
still says the same thing.

**Two limits are structurally zero here.** `psi_crit` and `root_psi_crit` carry
no row while the leaf's operating point is away from the bound they set, and come
alive the moment it sits on one. A zero in this table can mean "no effect" or
"not currently binding", and those are different statements.
