---
title: "TF24 flexible allometry: a canopy that thins, and why it does not yet buy survival"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
---

```{r setup, message=FALSE, warning=FALSE}
# Staging demo: needs the in-tree dev build (#516), not necessarily the
# installed `plant`.
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

# Shared computation, kept out of this file so a test can exercise it: a chunk
# that runs is not a chunk that is checked.
source(file.path(pkg_root, "overstorey_staging", "allometry_demo_helpers.R"))

theme_set(theme_bw(base_size = 12))
```

## What changed

TF24 used to pin leaf area to height, `A = A*(h) = (h/a_l1)^(1/a_l2)`. A plant
could not economise by carrying less leaf, and `dh/dt >= 0` was structural, so
thinning was not expressible at all.

Leaf area and sapwood area are now ODE states — but carried as **log departures
from their preferred values** rather than as levels:

| state | meaning | resting value |
|---|---|---|
| `log_area_leaf_departure` (`phi`) | `ln(A / A*(h))` | 0 |
| `log_area_sapwood_departure` (`psi`) | `ln(A_s / (theta * A))` | 0 |

Leaf area is derived, `A = A*(h) * exp(phi)`. The reason for the departure
coordinate is exactness: integrating `A` alongside `h` would make `A = A*(h)` a
redundant invariant that Runge–Kutta preserves only to truncation error, so
height's rate would read a drifted leaf area. On the departure coordinate the
allometric motion stays analytic and only the departure is integrated, so with
plasticity off `dphi/dt` is exactly zero and the fixed-allometry model is
recovered **bit-for-bit**.

The switch is `a_pl0`, the maximum fraction of leaf and sapwood turnover a plant
may decline to replace. `a_pl0 = 0` (the default) is the old model exactly.

```{r versions}
s <- TF24_Strategy()
data.frame(
  quantity = c("model", "state_size", "a_pl0 (default)", "a_pl1", "a_pl2",
               "k_l", "k_s"),
  value = c(model_id("TF24"), TF24_Individual(s)$ode_size,
            s$pars$a_pl0, s$pars$a_pl1, s$pars$a_pl2,
            round(s$pars$k_l, 4), s$pars$k_s))
```

## The gate: replacement is withheld only at very low reserves

Replacement is gated on the NSC reserve fraction $r = S/S_{max}$, on a logistic
centred at `a_pl1 = 0.05` — below the growth gate `a_st2 = 0.10`, so the
allocation ladder emerges without any branching: reproduction and growth are cut
first, canopy replacement next, and only then do reserves run out.

```{r gate, fig.height=4}
gate <- allometry_gate_map(a_pl0 = 1.0, height = 5)

p_gate <- gate |>
  select(r, `leaf  d(phi)/dt` = dphi, `sapwood  d(psi)/dt` = dpsi) |>
  pivot_longer(-r) |>
  ggplot(aes(r, value, colour = name)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "reserve fraction  r = S / S_max", y = "departure rate (1/yr)",
       colour = NULL, title = "The replacement gate",
       subtitle = "Leaf area thins; sapwood per leaf area rises")

p_h <- ggplot(gate, aes(r, dheight)) +
  geom_line(linewidth = 0.9) +
  labs(x = "reserve fraction  r", y = "d(height)/dt  (m/yr)",
       title = "Height never falls",
       subtitle = "Growth is gated separately, and earlier")

p_gate + p_h
```

Two things to read off this. The leaf and sapwood rates have **opposite signs**,
which is the acclimation: leaf area thins at up to `k_l` while conducting area
is only lost to heartwood at `k_s`, so sapwood per leaf area rises at up to
`k_l - k_s`. And the gate is **sharp** — by $r = 0.33$ the rates are already
~1e-7. That is deliberate (a plant should not thin its canopy while it still has
reserves) but it is also why the response is small in any stand where reserves
rarely empty.

## The mechanism works: thin through a drought, rebuild after it

A plant is held at wet soil, dropped into a four-year drought, then returned.
Soil water is prescribed piecewise-constant here so the carbon balance can be
driven directly — this is *not* how the SCM drives soil water.

```{r trajectory, message=FALSE, warning=FALSE}
history <- demo_soil_history(wet = 0.30, dry = 0.11,
                             years_wet = 2, years_dry = 4)
traj <- bind_rows(
  allometry_trajectory(0.0, history, dt = 0.25),
  allometry_trajectory(1.0, history, dt = 0.25)) |>
  mutate(allometry = ifelse(a_pl0 == 0, "fixed", "flexible"))
```

```{r trajectory-plot, fig.height=7}
dry_band <- annotate("rect", xmin = 2, xmax = 6, ymin = -Inf, ymax = Inf,
                     fill = "grey85", alpha = 0.6)

p1 <- ggplot(traj, aes(time, canopy_fraction, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = NULL, y = "A / A*(h)", colour = NULL,
       title = "Canopy relative to what its height prefers")

p2 <- ggplot(traj, aes(time, huber_ratio, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = NULL, y = "A_s / (theta A)", colour = NULL,
       title = "Huber value relative to preferred")

p3 <- ggplot(traj, aes(time, height, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) +
  labs(x = "time (yr)", y = "height (m)", colour = NULL,
       title = "Height")

(p1 / p2 / p3) + plot_layout(guides = "collect")
```

```{r trajectory-table}
traj |>
  filter(allometry == "flexible", time %in% c(2, 3, 4, 5, 6, 6.5, 7.5, 8)) |>
  transmute(time, soil = theta_soil,
            `A / A*` = round(canopy_fraction, 3),
            `Huber ratio` = round(huber_ratio, 3),
            height = round(height, 2)) |>
  knitr::kable()
```

This is the intended behaviour, and all of it emerges rather than being imposed:

- the canopy thins to about **19 per cent** of what its height prefers, at a rate
  bounded by `k_l` because the mechanism is declining to replace what died
  rather than actively shedding;
- sapwood per leaf area rises to about **2.4 times** its preferred value, purely
  because `k_l > k_s` — no asymmetry is imposed anywhere;
- **height never falls**, and simply stops while the plant is in deficit;
- on recovery the canopy rebuilds to ~98 per cent of preferred and the Huber
  value returns toward 1, asymptotically and with no jump.

## ⚠️ The issue to solve: thinning does not buy survival

Here is what the mechanism was *for* — buffering mortality, so a productivity
drop does not translate straight into death. It does not yet do that.

```{r survival, fig.height=4}
surv <- traj |>
  select(time, theta_soil, allometry, survivorship) |>
  pivot_wider(names_from = allometry, values_from = survivorship) |>
  mutate(gain = flexible / fixed)

p_s <- traj |>
  ggplot(aes(time, survivorship, colour = allometry)) +
  dry_band + geom_line(linewidth = 0.9) + scale_y_log10() +
  labs(x = "time (yr)", y = "survivorship  exp(-mortality)", colour = NULL,
       title = "Survivorship is essentially unchanged")

p_g <- ggplot(surv, aes(time, gain)) +
  dry_band + geom_hline(yintercept = 1, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "time (yr)", y = "flexible / fixed",
       title = "Survival gain from thinning")

p_s + p_g
```

```{r survival-numbers}
surv |>
  filter(time %in% c(2, 4, 6, 8)) |>
  transmute(time, soil = theta_soil,
            `surv. fixed` = signif(fixed, 3),
            `surv. flexible` = signif(flexible, 3),
            `gain` = round(gain, 4)) |>
  knitr::kable()
```

**The survival gain is about 1 per cent**, against a survivorship that falls
through ten orders of magnitude either way. And flexibility *costs* height: the
flexible plant finishes around 2 m shorter, having spent its recovery carbon
rebuilding canopy rather than growing. That is why turning plasticity on lowered
R0 in every stand tested (factors of 0.55 to 0.64 on seasonal stands).

### Why

Mortality reads the reserve **fraction** $r = S/S_{max}$, and during a deep drought
`r` is pinned near zero however thin the canopy gets, so mortality sits at its
bounded maximum `a_dG1` regardless.

Thinning ought to help through the **carbon balance** instead — less leaf to
maintain means a less negative `P`, so the pool drains more slowly. It does, but
not nearly enough, and the reason is a scaling mismatch:

```{r costs}
traj |>
  filter(allometry == "flexible", theta_soil < 0.2) |>
  transmute(time, `A / A*` = round(canopy_fraction, 3),
            `net production P` = signif(P, 3),
            `P per leaf area` = signif(P_per_area, 3)) |>
  head(9) |>
  knitr::kable()
```

Leaf and fine-root costs scale with the canopy, so thinning reduces them. But
**sapwood respiration scales with `A_s * h`**, which thinning the canopy does not
touch — sapwood is only lost to heartwood at `k_s`, and deliberately so, since
that is what closed the free lunch where shedding leaves would have shrunk the
denominator of `r` and made a plant immortal for free. For a tall plant in
drought, that untouchable maintenance cost dominates the deficit.

So there is a real tension, now measured from both sides:

- if storage capacity followed **leaf area**, thinning would inflate `r` directly
  — a free lunch, and an immortal skeleton;
- because it follows **sapwood** it cannot be inflated — but then thinning gives
  no reserve benefit either.

The resolution has to be that thinning helps by reducing the *deficit*, which
means the deficit has to be dominated by costs that scale with the canopy. At
TF24's defaults, for a tall plant, it is not.

### Candidates, none chosen

1. **Let the stem thin too, faster than `k_s`.** Physically this is heartwood
   conversion accelerating under stress, or sapwood being abandoned. It reduces
   the dominant cost, but it re-opens the `S_max` question and needs care not to
   restore the free lunch.
2. **Reconsider what mortality reads.** A reserve *fraction* is scale-free and
   bounded, which is why #517 chose it, but it is insensitive to a plant having
   made itself cheaper to run. Some measure of how long reserves would last at
   the current burn rate would respond to thinning.
3. **Check whether `r_s` is too high.** Sapwood respiration per unit stem mass
   dominating a drought deficit for a tall plant may be the real finding here,
   and it is a parameter question rather than a structural one.

## Also unresolved: neither departure is bounded in permanent deficit

With reserves empty, `withheld * k_l` (0.42/yr) outruns `withheld_sapwood * k_s`
(0.2/yr) whatever `psi` does, so both departures diverge linearly — `phi` toward
`-inf` and `psi` toward `+inf`. That is arithmetic rather than a defect: a plant
thinning at 46 per cent a year while losing conducting area at 20 per cent really
does raise its Huber value without limit. What removes such a plant is mortality,
not the allometry, and in a cohort model that plant is still carried.

Related: `rebuild_rel` scales as `exp(-phi)`, so it grows without bound as a
canopy empties — structurally the fast-attracting-boundary shape that #609's
analysis identified for `log S`. It has not caused a failure in any run so far.

## What is settled

```{r settled}
data.frame(
  claim = c("a_pl0 = 0 reproduces the fixed allometry",
            "carbon and tissue balance under withholding",
            "psi >= 0 is invariant",
            "height never decreases",
            "the canopy rebuilds after a drought"),
  status = c("bit-identical, verified over 288 quantities and an SCM run",
             "exact by construction: turnover not charged is mass lost",
             "by the form: at psi = 0 the rate is (1-repl)(k_l - k_s) >= 0",
             "by the form: its rate is a product of non-negative factors",
             "measured: to ~98 per cent of preferred, asymptotically")) |>
  knitr::kable()
```
