---
title: "TF24t leaf-level thermal response (ATLS): damage, and four strategies"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
---

```{r setup, message=FALSE, warning=FALSE}
# Staging demo: needs the in-tree dev build (ATLS, #566; builds on PM, #523), not
# necessarily the installed `plant`. Resolve the package root whether rendered
# from the repo root or from overstorey_staging/.
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

# Shared leaf-driving code (also exercised by tests/testthat/test-atls-leaf-demo.R,
# so this demo cannot silently rot).
source(file.path(pkg_root, "overstorey_staging", "atls_demo_helpers.R"))

theme_set(theme_bw(base_size = 12))
# One colour per strategy archetype, reused across the document.
strat_pal <- c("Generalist" = "#7570b3", "Tolerant" = "#d95f02",
               "Repairer" = "#1b9e77", "Acclimated" = "#e7298a")

# File-based cache for the expensive (~30 s each) SCM runs: compute once, reuse
# the .rds on re-render. Preferred over Quarto's `cache=TRUE` so the cached
# *data* is a plain, inspectable artefact independent of the freezer/render
# machinery. Keyed on the cache name AND the deparsed call, so editing a chunk's
# call (e.g. widening the temperature grid) auto-invalidates that entry; the
# expression is evaluated only on a miss. A model recompile is NOT tracked --
# delete overstorey_staging/cache/ to force a full rerun after changing the model.

cache_dir <- file.path(pkg_root, "overstorey_staging", "cache")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
cached_rds <- function(name, expr) {
  call_txt <- paste(deparse(substitute(expr)), collapse = " ")
  path <- file.path(cache_dir, paste0(name, ".rds"))
  if (file.exists(path)) {
    hit <- readRDS(path)
    if (identical(hit$call, call_txt)) return(hit$value)
  }
  value <- expr                       # evaluated only here (cache miss)
  saveRDS(list(call = call_txt, value = value), path)
  value
}
```

The **ATLS** layer (leaf thermal damage / repair / acclimation, issue #566,
*merged revision*) gives a TF24 leaf a mechanistic response to heat. Its core is
the **deactivated fraction** `φ_d ∈ [0, 1]` — the transiently-unfolded share of
the machinery, read straight off the *same* Medlyn peaked-Arrhenius curve that
sets capacity. `φ_d` half-unfolds at an **emergent onset** `≈ H_d/ΔS ≈ 34.5 °C`
(no separate `T_crit` parameter). Unfolded protein leaks into **lasting damage**:
a recoverable pool `I_r` (resynthesised over ~4 days) and a permanent pool `I_p`
(cleared only as the plant grows new leaf). Both capacities carry the lasting
discount `(1 − I_r − I_p)`, so — unlike the memoryless as-built factor — a cooled
leaf does *not* rebound instantly.

A thermal *strategy* is a point in a four-axis trait space, all plugging into the
same physics:

| Axis | Knob | Moves |
|---|---|---|
| **Tolerance** (constitutive) | `topt_offset` | shifts the capacity optimum *and* the damage onset up together (one thermostability knob) |
| **Acclimation** (inducible) | `A` state | the same shift, induced when the regime is hot |
| **Repair** | preventive `k_i`, restorative `k_rec` | lower the leak into `I_r` / rebuild it faster (and cut the permanent floor) |
| **Avoidance** | transpirational cooling (PM) | lowers the operating leaf temperature, hence `φ_d` |

This demo drives a single well-watered leaf. It is a work-in-progress staging
document, not a built vignette. See `notes/atls/implementation-plan.md` for the
design and `notes/atls/gate-atls-vs-pm-evaluation.md` for the evaluation gate.

## 1. The deactivation curve `φ_d(Tleaf)` and the damage it drives

The generalist (default) leaf: `φ_d` sits near 0 while cold, passes through
**half-unfolding at the emergent onset ≈ 34.5 °C** (`H_d/ΔS`, marked), and
saturates toward 1 when hot — all from the *same* `K(T)` inside the peaked-
Arrhenius curve, with no separate `T_crit`. Leaf temperature is *prescribed*
(Fick path), so this is the pure mechanism with no optimiser in the loop.

The right-hand panel shows the standing recoverable damage this substrate would
drive at equilibrium, `I_r* = k_i·φ_d / (k_i·φ_d + k_rec_eff + k_mat)` (with
`I_p = 0`): the balance the `I_r` ODE relaxes toward at a held temperature. Fresh
leaves lose no capacity instantly — damage is a *memory* that fills this pool over
days, and each unit of `I_r`/`I_p` discounts **both** `vcmax` and `jmax`.

```{r damage-curve}
Tleaf <- seq(20, 50, by = 0.5)
phi_gen <- atls_phi_d_curve(Tleaf, atls_traits())
Ir_gen  <- atls_Ir_star_curve(Tleaf, atls_traits())
onset <- 200000 / 650 - 273.15   # H_d/dS in deg C ~ 34.5

p_phi <- ggplot(data.frame(Tleaf, phi = phi_gen), aes(Tleaf, phi)) +
  geom_line(colour = "#d95f02", linewidth = 0.9) +
  geom_vline(xintercept = onset, linetype = "dashed", colour = "grey50") +
  annotate("text", x = onset + 0.4, y = 0.05, label = "onset H_d/ΔS", hjust = 0,
           size = 3, colour = "grey40") +
  ylim(0, 1) +
  labs(x = "leaf temperature (°C)", y = "φ_d (unfolded fraction)",
       title = "Deactivation φ_d(Tleaf)")

p_Ir <- ggplot(data.frame(Tleaf, Ir = Ir_gen), aes(Tleaf, Ir)) +
  geom_line(colour = "#7570b3", linewidth = 0.9) +
  ylim(0, 1) +
  labs(x = "leaf temperature (°C)", y = "I_r* (quasi-steady)",
       title = "Standing recoverable damage")

p_phi | p_Ir
```

## 2. Four thermal-response strategies move the damage differently

Each archetype is a single-axis move off the generalist, shown as the standing
recoverable damage `I_r*(Tleaf)`. **Tolerance** and **acclimation** shift the
onset (and hence `I_r*`) to higher temperature; **repair** (higher `k_rec`, lower
`k_i`) lowers `I_r*` at a given temperature. (Avoidance is not a curve-shift — it
lowers the *operating* temperature, shown in §4.)

```{r strategy-curves}
strategies <- atls_strategies()
Ir_curves <- bind_rows(lapply(names(strategies), function(nm) {
  data.frame(strategy = nm, Tleaf = Tleaf,
             Ir = atls_Ir_star_curve(Tleaf, strategies[[nm]]))
}))

ggplot(Ir_curves, aes(Tleaf, Ir, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = strat_pal, name = NULL) +
  ylim(0, 1) +
  labs(x = "leaf temperature (°C)", y = "I_r* (quasi-steady recoverable damage)",
       title = "Standing damage I_r*(Tleaf) by thermal strategy",
       subtitle = "Tolerant/Acclimated shift the onset right; Repairer lowers the whole curve") +
  theme(legend.position = "top")
```

## 3. Who keeps carbon where? Assimilation across a heat gradient

The payoff, on fresh leaves (`I_r = I_p = 0`, so the assimilation panel isolates
the reversible-capacity effect). Tolerance and acclimation raise the capacity
optimum (`topt_offset`/`A`), buying carbon at hot leaf temperatures; the standing-
damage panel shows where lasting damage *would* accrue over days — repair (lower
`k_i`, higher `k_rec`) and tolerance both suppress `I_r*`, while the generalist
accrues the most.

```{r strategy-assim}
grad <- atls_solve_strategies(seq(24, 48, by = 2), strategies,
                              PAR = 1500, VPD = 2)

p_grad_A <- ggplot(grad, aes(Tenv, A, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = strat_pal, name = NULL) +
  labs(x = "leaf temperature (°C)", y = "assimilation (µmol C m⁻² s⁻¹)",
       title = "Carbon gain by strategy")

p_grad_Ir <- ggplot(grad, aes(Tenv, Ir_star, colour = strategy)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = strat_pal, name = NULL) +
  labs(x = "leaf temperature (°C)", y = "I_r* at the operating point",
       title = "Standing damage by strategy")

(p_grad_Ir | p_grad_A) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

## 4. Avoidance: transpirational cooling as the fourth axis

Avoidance is not a damage trait — it is the leaf keeping itself cooler than the
air by transpiring, through the Penman–Monteith energy balance (#523). A
better-coupled leaf (smaller dimension `d`, higher conductance) runs closer to
air temperature, so on a hot day it sits at a lower `φ_d` and drives less lasting
damage — and because `φ_d` (and capacity) feed the carbon the psi-stem optimiser
sees, cooling is *self-selecting*. Here both leaves are the generalist; only the
coupling differs.

```{r avoidance}
avoid_cfgs <- list(
  "poorly coupled (large leaf, low g)" =
    atls_leaf_config(d = 0.10, leaf_specific_conductance_max = 2e-3),
  "well coupled (small leaf, high g)" =
    atls_leaf_config(d = 0.02, leaf_specific_conductance_max = 1e-2))

avoid <- bind_rows(lapply(names(avoid_cfgs), function(nm) {
  bind_rows(lapply(seq(24, 40, by = 0.5), function(Tair) {
    r <- atls_solve_cell(1500, Tair, 2, atls_traits(), pm = TRUE,
                         cfg = avoid_cfgs[[nm]])
    data.frame(coupling = nm, Tair = Tair, Tleaf = r$Tleaf,
               phi_d = r$phi_d, A = r$A)
  }))
}))

avoid_pal <- c("poorly coupled (large leaf, low g)" = "#b2182b",
               "well coupled (small leaf, high g)" = "#2166ac")

p_av_T <- ggplot(avoid, aes(Tair, Tleaf, colour = coupling)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey60") +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = avoid_pal, name = NULL) +
  labs(x = "air temperature (°C)", y = "operating leaf temperature (°C)",
       title = "Cooling by coupling", subtitle = "dotted = Tleaf = Tair")

p_av_phi <- ggplot(avoid, aes(Tair, phi_d, colour = coupling)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = avoid_pal, name = NULL) +
  labs(x = "air temperature (°C)", y = "φ_d at the operating point",
       title = "... lowers the deactivated fraction")

(p_av_T | p_av_phi) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

> Note the minimal-cut PM overheats an isolated leaf well above air (up to ~20 °C
> under high radiation, #523 caveats), so the *absolute* temperatures here are an
> upper bound; the **contrast** between couplings is the robust signal.

## 5. From one leaf to a community: persistence under warming (SCM)

Everything above is a single isolated leaf. The real question is demographic:
does the damage feedback change whether a *population* persists? We run the full
size-structured community model (SCM) — a single-species patch under a constant
climate (the `air_temp` driver = midday air temperature) — across a warming
gradient, TF24t (PM + ATLS) vs a PM-only TF24.

**How we measure fitness matters.** The net reproduction ratio R0 at a *fixed*
seed rain is **not** a valid cross-strategy comparison: it conflates a plant's
per-capita productivity with density-dependent recruitment. A less-fecund
resident shades the patch less, so at a shared seed rain each of its seeds does
*better* — inflating its R0 even though it is the weaker plant. (R0 falls as seed
rain rises for every strategy; the ranking at one shared rate is a density
artifact, not fitness.)

The density-free measure is **viability**: when *rare*, does a resident more than
replace itself? We evaluate R0 in the low-density limit (a small `birth_rate`, so
competition is negligible where it matters). R0 ≥ 1 ⇒ the population grows from
rare and persists; R0 < 1 ⇒ it dies out. The warmest climate with R0 ≥ 1 is the
**persistence boundary**, and *that* is comparable across strategies — fecundity
and density effects vanish as density → 0. We read the boundary, never the R0
magnitude away from it.

Each SCM run takes ~30 s; the grid is cached to an `.rds` under
`overstorey_staging/cache/` (delete that folder to force a rerun) and the runs
are parallelised across cores.

```{r scm-viability}
# The lasting-damage maturation rate k_mat is a CALIBRATION TARGET. Its default
# (0.02/day) makes permanent damage I_p ratchet to ~0.9 (maturation influx
# DAYS_PER_YEAR*k_mat ~ 7.3/yr vs the only sink, leaf turnover k_l ~ 0.46/yr),
# which renders TF24t inviable at EVERY climate. Here we use the balanced value
# k_l/DAYS_PER_YEAR ~ 0.00125/day so the section is illustrative. PROVISIONAL,
# pending the k_mat calibration decision -- set this to 0.02 to see the collapse.
atls_scm_kmat <- 0.00125

scen5 <- list(
  "PM only (TF24)"    = list(type = "TF24"),
  "PM + ATLS (TF24t)" = list(type = "TF24t",
                             mutate = function(s) { s$k_mat <- atls_scm_kmat; s }))
# Cache name carries k_mat: cached_rds keys on the call *text*, and the scenario
# list (hence k_mat) hides behind the `scen5` symbol -- so encode k_mat in the
# name to force a recompute when the knob changes.
viab5  <- cached_rds(sprintf("scm-viability-k%.5g", atls_scm_kmat),
                     atls_scm_viability(seq(23, 31, by = 1), scen5))
bound5 <- atls_persistence_boundary(viab5)
knitr::kable(bound5, digits = 1, col.names = c("scenario", "persistence °C"))

ggplot(viab5, aes(air_temp, pmax(log10(R0), -8), colour = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(data = bound5, aes(xintercept = boundary, colour = scenario),
             linetype = "dotted", show.legend = FALSE) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.8) +
  annotate("text", x = 23, y = 0.35, label = "persistence (R0 = 1)",
           hjust = 0, size = 3, colour = "grey40") +
  scale_colour_manual(values = c("PM only (TF24)"    = "#1b9e77",
                                 "PM + ATLS (TF24t)" = "#d95f02"), name = NULL) +
  labs(x = "midday climate temperature (°C)",
       y = expression(log[10] * " R0 (low density)"),
       title = "Persistence boundary: warmest climate a rare population survives",
       subtitle = "low-density R0; the R0 = 1 crossing (dotted) is the valid, density-free boundary") +
  theme(legend.position = "top")
```

The valid signal is the **persistence boundary** — the dotted crossings, not the
curve heights (those still carry the density artifact and are shown only on a log
scale to expose the crossing). Two honest readings, under the provisional `k_mat`:

- **ATLS shifts persistence *cooler*, not hotter.** TF24t clears R0 = 1 only up to
  a *lower* climate temperature than PM-only TF24 — adding lasting thermal damage
  is a net persistence penalty here. (This corrects the earlier draft's claim that
  TF24t persisted *hotter*; that was an artifact of comparing R0 at a fixed seed
  rain.)
- **The collapse is a near-cliff**, driven by the minimal-cut PM overheating the
  canopy leaf 9–20 °C above air (the §4/#523 caveat) feeding the lasting-damage
  loop. So the absolute boundary temperatures are an upper-bound artifact of the
  minimal-cut PM, not a calibrated prediction.

> Caveat stack: this section rests on (1) the PM-overheating artifact above and
> (2) the `k_mat` permanent-damage ratchet — at the **default** `k_mat = 0.02`
> TF24t is inviable at *every* climate; the balanced `0.00125` used here is
> provisional. Both are open calibration items, not settled model behaviour.

## 6. Which thermal strategy persists hottest? (SCM tournament)

Same valid measure, now varying the *strategy* — each a heritable single-axis
move off the generalist (tolerance = constitutive `topt_offset`; repair = higher
restorative `k_rec` + lower preventive-leak `k_i`; acclimator = faster
acclimation kinetics). "Persists hottest" = the highest **persistence boundary**
(warmest climate at which a rare resident still replaces itself), *not* the R0
height at a shared seed rain.

```{r scm-tournament}
# Each strategy carries the provisional balanced k_mat from §5, so the ranking is
# not swamped by the default permanent-damage ratchet (see the §5 caveat).
scm_strats <- atls_scm_strategies()
scen6 <- setNames(lapply(names(scm_strats), function(nm) {
  base <- scm_strats[[nm]]
  list(type = "TF24t",
       mutate = function(s) { s <- base(s); s$k_mat <- atls_scm_kmat; s })
}), names(scm_strats))

viab6  <- cached_rds(sprintf("scm-viability-strat-k%.5g", atls_scm_kmat),
                     atls_scm_viability(seq(23, 33, by = 1), scen6))
bound6 <- atls_persistence_boundary(viab6)
bound6 <- bound6[order(-bound6$boundary), ]
knitr::kable(bound6, digits = 1, row.names = FALSE,
             col.names = c("strategy", "persistence °C"))

strat_pal6 <- c("Generalist" = "#7570b3", "Tolerant" = "#d95f02",
                "Repairer" = "#1b9e77", "Acclimator" = "#e7298a")
ggplot(bound6, aes(reorder(scenario, boundary), boundary, fill = scenario)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f °C", boundary)), hjust = -0.15, size = 3) +
  coord_flip() +
  scale_fill_manual(values = strat_pal6, guide = "none") +
  expand_limits(y = max(bound6$boundary, na.rm = TRUE) + 1.5) +
  labs(x = NULL,
       y = "persistence boundary (warmest climate with R0 ≥ 1, °C)",
       title = "Which thermal strategy persists hottest?",
       subtitle = "warmest climate at which a rare TF24t resident still replaces itself") +
  theme(axis.text.y = element_text(size = 10))
```

Reading the ranking off the persistence boundaries (the bars), the order is
**tolerance > repair > acclimation > generalist**. Constitutive **tolerance**
(`topt_offset`) extends persistence furthest — it shifts the capacity optimum
*and* the damage onset up together, so a rare resident still replaces itself at
the warmest climate. **Repair** (higher `k_rec`, lower `k_i`) is a close second:
in this damage-dominated regime, directly cutting the leak into `I_r`/`I_p` and
rebuilding faster pays for its standing maintenance cost. Inducible
**acclimation** helps less than constitutive tolerance (its onset shift is
smaller and lags), and the **generalist** collapses coolest.

Note the boundary — not R0 height — is doing the work: tolerance and repair are
*less* fecund, so their low-density R0 at cool climates is actually **lower** than
the generalist's (the density artifact §5 warned about), yet they persist to
hotter climates. That inversion is exactly why we rank by the R0 = 1 crossing.
The ordering is robust; the exact boundary temperatures are provisional (they
inherit both §5 caveats — the PM-overheating artifact and the `k_mat`
calibration).

## 7. Memory: the running-mean leaf temperature

The sections above all read the *instantaneous* operating leaf temperature. TF24t
also carries a diagnostic ODE state, **`mean_leaf_temp`**, an exponential moving
average of that operating-point `Tleaf`:

$$
\frac{\mathrm{d}\,\overline{T_\ell}}{\mathrm{d}t}
  = D\,k_{\overline{T}}\,\bigl(T_\mathrm{op} - \overline{T_\ell}\bigr),
$$

where `T_op` is the PM operating-point leaf temperature (published as the
`operating_leaf_temp` aux), `k_mean_leaf_temp` is the memory decay rate (day⁻¹,
`D = 365.25` puts it on the yearly ODE clock), and the time constant is
`1/k_mean_leaf_temp`. It is **purely diagnostic** — it does not feed back into
damage or acclimation — so it is a clean read-out of the temperature history a
plant has actually experienced. Below, a single TF24t individual is driven with a
sub-seasonally oscillating `air_temp`; the running mean lags and smooths the
operating temperature (memory ≈ 20 d here).

```{r running-mean, message=FALSE, warning=FALSE}
traj_wide <- cached_rds("running-mean", {
  e <- "TF24_Env"
  k_mem  <- 0.05               # day^-1 -> ~20-day memory (1 / (k*365.25) yr)
  period <- 0.12               # yr (~44 d) forcing, comparable to the memory
  t_end  <- 0.6                # yr

  s <- TF24t_Strategy(); s$k_mean_leaf_temp <- k_mem
  p <- Individual("TF24t", e)(s)
  env <- Environment("TF24t")
  # The adaptive ODE stepper can probe the driver just past t_end, so set air_temp
  # on a grid extending beyond the loop horizon (values on [0, t_end] unchanged) --
  # otherwise the spline extrapolates and run halts.
  t_drive <- seq(0, t_end * 1.15, length.out = 230)
  env$extrinsic_drivers_set_variable("air_temp", t_drive,
                                     27 + 6 * sin(2 * pi * t_drive / period))

  runner <- OdeRunner("TF24t")(IndividualRunner("TF24t", e)(p, env))
  rec <- list()
  while (runner$time < t_end) {
    runner$step()
    ind <- runner$object$individual
    rec[[length(rec) + 1L]] <- data.frame(
      time      = runner$time,
      air_temp  = env$extrinsic_drivers_evaluate("air_temp", min(runner$time, t_end)),
      operating = ind$aux("operating_leaf_temp"),
      running   = ind$state("mean_leaf_temp"))
  }
  do.call(rbind, rec)
})
traj <- traj_wide |>
  pivot_longer(c(air_temp, operating, running), names_to = "series",
               values_to = "temp")

series_lab <- c(air_temp = "air temperature (driver)",
                operating = "operating leaf temp (instantaneous)",
                running = "mean_leaf_temp (running mean)")
series_pal <- c(air_temp = "grey60", operating = "#d95f02", running = "#1b9e77")

ggplot(traj, aes(time, temp, colour = series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = series_pal, labels = series_lab, name = NULL) +
  labs(x = "time (years)", y = "temperature (°C)",
       title = "mean_leaf_temp lags and smooths the operating leaf temperature",
       subtitle = "single TF24t individual, oscillating air_temp; memory ≈ 20 d") +
  theme(legend.position = "top")
```

The running mean (green) tracks the operating leaf temperature (orange) with a
lag set by `k_mean_leaf_temp` and a reduced amplitude — the "average growing
temperature" the plant carries. The operating temperature sits well above air
because the minimal-cut PM overheats an isolated leaf (§4 caveat); `mean_leaf_temp`
averages that operating signal, not the air-temperature driver.

## 8. Summary

The deactivation curve `φ_d(Tleaf)` is the single substrate: strategies win by
moving *where it turns on* (tolerance, acclimation shift the emergent onset),
*how much unfolds leaks into lasting damage and how fast it is rebuilt* (preventive
`k_i`, restorative `k_rec`), or *where the leaf actually operates* (avoidance).
Damage is now a **memory** — the pools `I_r`/`I_p` discount both capacities and
persist after cooling — so the same insult that a memoryless factor would forget is
carried forward, and permanent damage clears only as the plant grows new leaf. All
plug into the one Medlyn deactivation curve and the Penman–Monteith energy balance.
At the community scale the same layer sets a **persistence boundary** — the
warmest climate at which a rare population still replaces itself (the density-free
viability measure, §5). Under the current calibration ATLS shifts that boundary
*cooler* (a net damage penalty), and among strategies, onset-shifting tolerance
and acclimation extend it furthest. Both community results are provisional: they
rest on the minimal-cut PM overheating (which makes the absolute boundary
temperatures an upper-bound artifact) and on the `k_mat` permanent-damage rate
(whose default makes TF24t inviable everywhere — an open calibration item). See
`notes/atls/gate-atls-vs-pm-evaluation.md` for the whole-plant materiality gate
(verdict: keep) and the caveats.
